---
title: "Syllabus Economía I (Introducción a la Economía) Module B AE"
author:
  - name: Luis Francisco Gómez López
    email: luis.gomezl@unimilitar.edu.co
    affiliations:
      - name: Universidad Militar Nueva Granada
date: 2025-01-13
format:
  html: 
    number-sections: true
    toc: true
    embed-resources: true
execute: 
  echo: false
  warning: false
  message: false
params:
  meeting_info:
    value:
      phone: +57 3124541536
      days: [Tuesday, Wednesday, Friday]
      start_hour: 17:00
      end_hour: 18:00
  block:
    value:
      start_date: 2025-01-20
      end_date: 2025-03-22
  schedule:
    value:
      meeting_dates: [3, 5]
      dates_recess: null
      start_presentation_hour: 20:00
      end_presentation_hour: 20:45  
      start_virtual_meeting_hour: 19:00
      end_virtual_meeting_hour: 20:30
      start_activity_hour: 00:00
      end_activity_hour: 22:00
      start_exam_hour: 18:30
      end_exam_hour: 20:30
engine: knitr
bibliography: intro_econ.bib
editor: visual
---

```{r}
#| label: libraries

library(tidyverse)
library(glue)
library(knitr)
```

```{=html}
<style>
  p { 
    text-align: justify;
  }
</style>
```

# Professor and meeting hours with students

```{r}
#| label: meeting-info

# Days ----
if(length(params$meeting_info$days) > 1) {
  days_str <- glue_collapse(params$meeting_info$days, 
                            sep = ", ", last = " or ") 
} else {
  days_str <- params$meeting_info$days[1]
}
```

-   **Name**

    -   `r rmarkdown::metadata$author[[1]]$name`

-   **Communication through the virtual classroom**

    -   For doubts and concerns of the topics of the course, please use the *Forum: Questions or concerns* that can be accessed at the top of the platform

        -   In this way, doubts and concerns can be shared with all students and all course participants can contribute to the learning process

    -   Complaints regarding grades, please send an email to [**`{r} rmarkdown::metadata$author[[1]]$email`**](mailto:%60r%20rmarkdown::metadata$author%5B%5B1%5D%5D$email%60)

        -   Please wait for grades to be published on the News Bulletin Board before submitting any complaints. Complaints should be based on the feedback provided and must take into account the evaluation parameters.

    -   For urgent issues, you can call me to `{r} params$meeting_info$phone` between `{r} params$meeting_info$start_hour` hours and `{r} params$meeting_info$end_hour` hours on `{r} days_str`. But, please do not send me whatsapp messages or call me on weekends.

# Presentation of the subject

## Description

```{r}
#| label: block

# Dates ----
start_date <- ymd(params$block$start_date) 
end_date   <- ymd(params$block$end_date)

str_start_end_date <- paste(format(start_date, 
                                   format = "%B %d"),
                            " - ",
                            format(end_date, 
                                   format = "%B %d"), 
                            sep = "",
                            ", ",
                            year(start_date))
```

-   Credits: 4
-   Type: Theoretical-Practice
-   Area of Knowledge: Professional
-   Semester: 1
-   Block: 1 (`{r} str_start_end_date`)

## Justification

To succeed in today's globalized business environment, business administrators require a comprehensive understanding of various disciplines. Among these, economics is particularly relevant. A strong grasp of economic principles enables business administrators to effectively analyze the economic landscape in which their organizations operate.

Effective economic training equips prospective business administrators with the theoretical knowledge and practical skills necessary for sound decision-making. This ensures the optimal utilization of an organization's resources.

Therefore, understanding economic ideas and concepts is crucial for business administrators. It empowers them to make informed decisions based on economic criteria, comprehend market dynamics, and anticipate the economic behavior of stakeholders such as customers, intermediaries, suppliers, and competitors.

## General objective

Familiarize students with the main economic concepts and their relevance in the current Colombian economic situation, through the analysis of data and simple technical tools.

## Learning outcomes

-   Course learning outcome:

    -   Diagnose the economic environment of the organization by utilizing economic theory to contribute to the improvement of the company's competitiveness.

-   Course specific learning outcomes:

    -   Interpret economic data based on information from the organizational environment for decision-making.

    -   Understand the proximate and fundamental determinants of growth through economic theory to comprehend how it affects the competitiveness of organizations.

    -   Identify the economic environment of the organization through the fields of public, monetary, and financial economics to justify investment and financing decisions.

    -   Understand the economic concepts of the labor market from the field of labor economics to improve the organizational environment and promote personal and professional development

-   Competencies SABE curriculum

    -   Skills in handling a second language
    -   Ability to work in teams
    -   Ability to use the scientific method and logical thinking

# Schedule and contents

## Reading schedule

```{r}
#| label: tbl-reading-schedule
#| tbl-cap: Reading schedule

tibble(Week = c(1:9),
       Material       = c(rep("[@cardenas_introduccion_2020]", 
                              times = 9)),
       Chapter        = c("1", "2", "3", "4", "5", "6", "7", "8", "9"),
       `Initial page` = c(  1,  19,  53,  81, 115, 167, 225, 263, 303),
       `Final page`   = c( 18,  52,  80, 114, 166, 224, 262, 302, 348)) |> 
    mutate(`Number of pages` = (`Final page` - `Initial page`) + 1) |> 
    kable()
```

## Detailed schedule

```{r}
#| label: detailed-schedule

source(file = "000_scripts/000_tbl_schedule.R")

tbl_schedule <- tbl_schedule(start_date   = params$block$start_date, 
                             end_date     = params$block$end_date,
                             dates_recess = params$schedule$dates_recess,
                             meeting_days = params$schedule$meeting_dates)

virtual_meeting <- tbl_schedule |> 
  filter(virtual_meeting == 1) |> 
  pull(index)

activity <-  tbl_schedule |> 
  filter(activity == 1) |> 
  pull(index)

exam <-  tbl_schedule |> 
  filter(exam == 1) |> 
  pull(index)
```

### Primer corte 30%

-   `r format(x = activity[[1]], format = "%B %e")`

    -   **Questionnaire 1 Opening**
        -   **Opening time**: `r paste(params$schedule$start_activity_hour)` hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapters 1-2]

-   `r format(x = ymd(params$block$start_date), format = "%B %e")`

    -   **Virtual meeting Presentation Syllabus**

        -   **Meeting time**: `r paste(params$schedule$start_presentation_hour, params$schedule$end_presentation_hour, sep = " - ")` hours
        -   **Prerequisites**:
            -   Review the course program in advance, which can be downloaded by following the route *Primer corte 30% \> Learning Resources \> Support material \> Presentations \> 000_syllabus_intro_econ\_...*
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   `r format(x = virtual_meeting[[1]], format = "%B %e")`

    -   **Virtual meeting Introduction I**
        -   **Meeting time**: `r paste(params$schedule$start_presentation_hour, params$schedule$end_presentation_hour, sep = " - ")` hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 1]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   `r format(x = virtual_meeting[[2]], format = "%B %e")`

    -   **Virtual meeting Introduction II**
        -   **Meeting time**: `r paste(params$schedule$start_virtual_meeting_hour, params$schedule$end_virtual_meeting_hour, sep = " - ")` hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 1]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   `r format(x = virtual_meeting[[3]], format = "%B %e")`

    -   **Virtual meeting Production and Income I**
        -   **Meeting time**: `r paste(params$schedule$start_virtual_meeting_hour, params$schedule$end_virtual_meeting_hour, sep = " - ")` hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 2]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   `r format(x = virtual_meeting[[4]], format = "%B %e")`

    -   **Virtual meeting Production and Income II**
        -   **Meeting time**: `r paste(params$schedule$start_virtual_meeting_hour, params$schedule$end_virtual_meeting_hour, sep = " - ")` hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 2]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   `r format(x = activity[[2]], format = "%B %e")`

    -   **Questionnaire 1 Closing**
        -   **Closing time**: `r paste(params$schedule$end_activity_hour)` hours
        -   **Prerequisites**:
            -   Have answered the Questionnaire 1 before `r paste(params$schedule$end_activity_hour)` hours

-   `r format(x = virtual_meeting[[5]], format = "%B %e")`

    -   **Virtual meeting Economic Growth I**
        -   **Meeting time**: `r paste(params$schedule$start_virtual_meeting_hour, params$schedule$end_virtual_meeting_hour, sep = " - ")` hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 3]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   `r format(x = virtual_meeting[[6]], format = "%B %e")`

    -   **Virtual meeting Economic Growth II**
        -   **Meeting time**: `r paste(params$schedule$start_virtual_meeting_hour, params$schedule$end_virtual_meeting_hour, sep = " - ")` hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 3]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   `r format(x = exam[[1]], format = "%B %e")`

    -   **Oral presentation Midterm Exam 1 Opening**
        -   **Opening time**: `r paste(params$schedule$start_exam_hour)` hours
        -   **Closing time**: `r paste(params$schedule$end_exam_hour)` hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 3]
            -   Have a headset with a built-in microphone to isolate noise from the environment

### Segundo corte 30%

-   `r format(x = activity[[3]], format = "%B %e")`

    -   **Questionnaire 2 Opening**
        -   **Opening time**: `r paste(params$schedule$start_activity_hour)` hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapters 4-5]

-   `r format(x = virtual_meeting[[7]], format = "%B %e")`

    -   **Virtual meeting Economy and institutions I**
        -   **Meeting time**: `r paste(params$schedule$start_virtual_meeting_hour, params$schedule$end_virtual_meeting_hour, sep = " - ")` hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 4]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   `r format(x = virtual_meeting[[8]], format = "%B %e")`

    -   **Virtual meeting Economy and institutions II**
        -   **Meeting time**: `r paste(params$schedule$start_virtual_meeting_hour, params$schedule$end_virtual_meeting_hour, sep = " - ")` hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 4]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   `r format(x = virtual_meeting[[9]], format = "%B %e")`

    -   **Virtual meeting External sector I**
        -   **Meeting time**: `r paste(params$schedule$start_virtual_meeting_hour, params$schedule$end_virtual_meeting_hour, sep = " - ")` hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 5]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   `r format(x = virtual_meeting[[10]], format = "%B %e")`

    -   **Virtual meeting External sector II**
        -   **Meeting time**: `r paste(params$schedule$start_virtual_meeting_hour, params$schedule$end_virtual_meeting_hour, sep = " - ")` hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 5]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   `r format(x = activity[[4]], format = "%B %e")`

    -   **Questionnaire 2 Closing**
        -   **Closing time**: `r paste(params$schedule$end_activity_hour)` hours
        -   **Prerequisites**:
            -   Have answered the Questionnaire 2 before `r paste(params$schedule$end_activity_hour)` hours

-   `r format(x = virtual_meeting[[11]], format = "%B %e")`

    -   **Virtual meeting Public Finances I**
        -   **Meeting time**: `r paste(params$schedule$start_virtual_meeting_hour, params$schedule$end_virtual_meeting_hour, sep = " - ")` hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 6]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   `r format(x = virtual_meeting[[12]], format = "%B %e")`

    -   **Virtual meeting Public Finances II**
        -   **Meeting time**: `r paste(params$schedule$start_virtual_meeting_hour, params$schedule$end_virtual_meeting_hour, sep = " - ")` hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 6]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   `r format(x = exam[[2]], format = "%B %e")`

    -   **Oral presentation Midterm Exam 2 Opening**
        -   **Opening time**: `r paste(params$schedule$start_exam_hour)` hours
        -   **Closing time**: `r paste(params$schedule$end_exam_hour)` hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 6]
            -   Have a headset with a built-in microphone to isolate noise from the environment

### Tercer corte 40%

-   `r format(x = activity[[5]], format = "%B %e")`

    -   **Questionnaire 3 Opening**
        -   **Opening time**: `r paste(params$schedule$start_activity_hour)` hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapters 7-8]

-   `r format(x = virtual_meeting[[13]], format = "%B %e")`

    -   **Virtual meeting Money, prices and the exchange rate I**
        -   **Meeting time**: `r paste(params$schedule$start_virtual_meeting_hour, params$schedule$end_virtual_meeting_hour, sep = " - ")` hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 7]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   `r format(x = virtual_meeting[[14]], format = "%B %e")`

    -   **Virtual meeting Money, prices and the exchange rate II**
        -   **Meeting time**: `r paste(params$schedule$start_virtual_meeting_hour, params$schedule$end_virtual_meeting_hour, sep = " - ")` hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 7]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   `r format(x = virtual_meeting[[15]], format = "%B %e")`

    -   **Virtual meeting Financial Market I**
        -   **Meeting time**: `r paste(params$schedule$start_virtual_meeting_hour, params$schedule$end_virtual_meeting_hour, sep = " - ")` hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 8]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   `r format(x = virtual_meeting[[16]], format = "%B %e")`

    -   **Virtual meeting Financial Market II**
        -   **Meeting time**: `r paste(params$schedule$start_virtual_meeting_hour, params$schedule$end_virtual_meeting_hour, sep = " - ")` hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 8]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   `r format(x = activity[[6]], format = "%B %e")`

    -   **Questionnaire 3 Closing**
        -   **Closing time**: `r paste(params$schedule$end_activity_hour)` hours
        -   **Prerequisites**:
            -   Have answered the Questionnaire 2 before `r paste(params$schedule$end_activity_hour)` hours

-   `r format(x = virtual_meeting[[17]], format = "%B %e")`

    -   **Virtual meeting Labor Market I**
        -   **Meeting time**: `r paste(params$schedule$start_virtual_meeting_hour, params$schedule$end_virtual_meeting_hour, sep = " - ")` hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 9]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   `r format(x = virtual_meeting[[18]], format = "%B %e")`

    -   **Virtual meeting Labor Market II**
        -   **Meeting time**: `r paste(params$schedule$start_virtual_meeting_hour, params$schedule$end_virtual_meeting_hour, sep = " - ")` hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 9]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   `r format(x = exam[[3]], format = "%B %e")`

    -   **Oral presentation Final Exam Opening**
        -   **Opening time**: `r paste(params$schedule$start_exam_hour)` hours
        -   **Closing time**: `r paste(params$schedule$end_exam_hour)` hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 9]
            -   Have a headset with a built-in microphone to isolate noise from the environment

# Qualifiable activities, weights and process in case of no presentation

## Activities and weights

-   **Primer corte**: 30%

    -   **Questionnaire 1**: 15%
    -   **Oral presentation Midterm Exam 1**: 15%

-   **Segundo corte**: 30%

    -   **Questionnaire 2**: 15%
    -   **Oral presentation Midterm Exam 2**: 15%

-   **Tercer corte**: 40%

    -   **Questionnaire 3**: 15%
    -   **Oral presentation Final Exam**: 25 %

## Process in case of no presentation

-   Students who do not present a **Questionare** due to force majeure or illness, must send the respective excuse scanned through the [**`r rmarkdown::metadata$author[[1]]$email`**](mailto:%60%7Br%7D%20rmarkdown::metadata$author%5B%5B1%5D%5D$email%60). If it is not sent, the note will be zero. Once the excuse has been satisfactorily received, the student should **not** present the activity and the corresponding grade will be equal to the immediately subsequent grade obtained in the **Midterm Exam 1** or **Midterm Exam 2** or **Final Exam** according to the case.

-   Students who do not present a **Midterm Exam 1** or **Midterm Exam 2** or the **Final Exam**, must send and email to [**admon.distancia\@unimilitar.edu.co**](mailto:admon.distancia@unimilitar.edu.co){.email} two days before or two days after the closing date of these activities. However, this request can be made only once during the course. If it is not requested under these conditions, the grade will be zero.

# Use of Chat GPT-4, Gemini 1.5, Copilot and its friends

In this course, the use of generative artificial intelligence tools is allowed as long as you follow the following good practices [@equipo_proyecto_ia-uniandes_lineamientos_2024, pp. 20-21]:

-   **Informed use**: learn beforehand how the tool works, its limitations, and risks
-   **Transparent use**: report which tool you used and how it was used it
-   **Ethical use**: do not present material generated by the tool as your own and follow the corresponding citation rules (see [@mcadoo_how_2024]).
-   **Critical evaluation**: always contrast and compare any response obtained through the tool with reliable or primary sources of information
-   **Responsible use**: do not include personal or confidential information (own or others) when entering instructions (prompts)

For more detailed guidance on good practices and recommended uses, check out [@equipo_proyecto_ia-uniandes_lineamientos_2024, Section 3.2]

# Study material

-   [@cardenas_introduccion_2020, Chapters 1-9]

    -   <a href="https://www.umng.edu.co/inicio" target="_blank" rel="noopener noreferrer">https://www.umng.edu.co/</a> \> Biblioteca \> Ingrese acá a Bases virtuales y libros electrónicos \> AlfaOmega \> Ingresar usuario y contraseña [^1] \> Buscar "Introducción a la economía colombiana 4ed" \> Leer

[^1]: Remember that the user and password to access and read the e-book are the same that you use to check your grades

# About plagiarism

-   **Acuerdo 02 de 2015 (7 de Mayo): Artículo 68. Escala de Calificaciones Numeral f**

*"La comprobación de todo fraude o intento de fraude en cualquier tipo de prueba, acarreará al estudiante y a su(s) colaborador(es), una calificación de CERO PUNTO CERO (0.0) en la prueba respectiva. Como quiera que el fraude o intento de cometerlo es una falta grave, el docente debe informar de este hecho al Director del programa respectivo, quien enviará el informe a la Decanatura de la Facultad, con el fin de iniciar los trámites correspondientes para aplicar la sanción disciplinaria prevista en éste reglamento.”*

-   **Acuerdo 05 de 2018 (6 de Junio): Artículo 20 que modifica el Artículo 107 Faltas Disciplinarias Numeral c. Número 6 del Acuerdo 02 de 2015 (7 de Mayo)**

*"Hacer plagio en cualquier tipo de documento o desplegar alguna conducta dolosa que viole el régimen de propiedad intelectual de la Universidad."*

# Study material references
