---
title: "Syllabus Economía I (Introducción a la Economía) Module B AE"
author:
  - name: Luis Francisco Gómez López
    email: luis.gomezl@unimilitar.edu.co
    affiliations:
      - name: Universidad Militar Nueva Granada
date: 2024-07-09
date-modified: last-modified
format:
  html: 
    number-sections: true
    toc: true
    embed-resources: true
bibliography: intro_econ.bib
editor: visual
---

```{r setup}
#| include: false
knitr::opts_chunk$set(echo    = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```

```{r libraries}
library(tidyverse)
library(knitr)
```

```{css}
p {
  text-align: justify;
}
```

# Professor and meeting hours with students

-   **Name**

    -   Luis Francisco Gomez Lopez

-   **Communication through the virtual classroom**

    -   Every Saturday between 09:00 hours and 09:45 hours:

        -   *Communication channels \> Link Virtual meetings*

    -   Doubts and concerns of the topics of the course, please use the *Forum: Questions or concerns* that can be accessed at the top of the platform

        -   In this way, doubts can be shared with all students and all course participants can contribute to the learning process

    -   Complaints regarding grades, please send an email to [**luis.gomezl\@unimilitar.edu.co**](mailto:luis.gomezl@unimilitar.edu.co){.email}

        -   To make any complaints, you are asked to wait first for the grades to be published on the News Bulletin Board and to be made based on the feedbacks that will be made taking into account the evaluation parameters

        -   Urgent issues, you can call me to +57 3124541536 Tuesday, Wednesday and Friday between 17:00 hours and 18:00 hours. Please do not send me whatsapp messages or call me on weekends.

# Presentation and objectives of the subject

## Description

-   Credits: 4
-   Type: Theoretical-Practice
-   Area of Knowledge: Professional
-   Semester: 1
-   Block: 1 (July 22 - September 21, 2024)

## Justification

Comprehensive training of business managers involves knowledge of aspects related to multiple disciplines. One of the disciplines that have greater importance for such instruction is economics, where it allows you to recognize the economic environment in which the different organizations participate.

In this context, business managers need to understand the economic environment of a country in which a company operates by using economic indicators in the decision-making processes of a business.

## General objective

-   Familiarize students with the main economic concepts and their relevance in the current Colombian economic situation, through the analysis of data and the use of simple technical tools.

# Competencies

-   Global competency:

    -   Diagnose the economic environment of the organization through the use of economic theory to contribute to the competitive improvement of the company.

-   Specific competencies:

    -   Interpret economic data based on information from the organizational environment for decision-making.

    -   Understand the proximate and fundamental determinants of growth through economic theory in order to comprehend how it affects the competitiveness of organizations.
    
    -   Identify the economic environment of the organization through the fields of public, monetary, and financial economics in order to justify investment and financing decisions.
    
    -   Understand the economic concepts of the labor market from the field of labor economics in order to improve the organizational environment and promote personal and professional development.

-   Competencies SABE curriculum

    -   Managing data for decision making
    -   Skills in handling a second language
    -   Ability to work in teams

# Schedule and contents

## Reading schedule

```{r}

tibble(Week = c(1:9),
       Material       = c(rep("[@cardenas_introduccion_2020]", 
                              times = 9)),
       Chapter        = c("1", "2", "3", "4", "5", "6", "7", "8", "9"),
       `Initial page` = c(  1,  19,  53,  81, 115, 167, 225, 263, 303),
       `Final page`   = c( 18,  52,  80, 114, 166, 224, 262, 302, 348)) %>%
    mutate(`Number of pages` = (`Final page` - `Initial page`) + 1) %>% 
    kable()
```

## Detailed schedule of virtual meetings and activities

### Primer corte 30%

-   July 22

    -   Start Block 1

    -   **Virtual meeting Presentation Syllabus**

        -   **Meeting time**: 20:00 - 20:45 hours
        -   **Prerequisites**:
            -   Review the course program in advance, which can be downloaded by following the route *Primer corte 30% \> Learning Resources \> Support material \> Presentations \> 000_syllabus_intro_econ_2024_2*
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   July 24

    -   **Virtual meeting Introduction**
        -   **Meeting time**: 20:00 - 20:45 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 1]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   July 29

    -   **Questionnaire 1 Opening**
        -   **Opening time**: 00:00 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, , chap. 1-2]

-   July 31

    -   **Virtual meeting Production and Income I**
        -   **Meeting time**: 20:00 - 20:45 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 2]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   August 2

    -   **Virtual meeting Production and Income II**
        -   **Meeting time**: 20:00 - 20:45 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 2]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   August 4

    -   **Questionnaire 1 Closing**
        -   **Closing time**: 22:00 hours
        -   **Prerequisites**:
            -   Have answered the Questionnaire 1 before 22:00 hours

-   August 7

    -   **Virtual meeting Economic Growth**
        -   **Meeting time**: 20:00 - 20:45 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 3]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   August 10

    -   **Oral presentation Midterm Exam 2 Opening**
        -   **Opening time**: 16:00 hours
        -   **Closing time**: 18:00 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 3]
            -   Have a headset with a built-in microphone to isolate noise from the environment

### Segundo corte 30%

-   August 12

    -   **Questionnaire 2 Opening**
        -   **Opening time**: 00:00 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapters 4-5]

-   August 14

    -   **Virtual meeting Economy and institutions**
        -   **Meeting time**: 20:00 - 20:45 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 4]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   August 21

    -   **Virtual meeting External sector I**
        -   **Meeting time**: 20:00 - 20:45 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 5]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   August 23

    -   **Virtual meeting External sector II**
        -   **Meeting time**: 20:00 - 20:45 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 5]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   August 25

    -   **Questionnaire 2 Closing**
        -   **Closing time**: 22:00 hours
        -   **Prerequisites**:
            -   Have answered the Questionnaire 2 before 22:00 hours

-   August 28

    -   **Virtual meeting Public Finances**
        -   **Meeting time**: 20:00 - 20:45 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 6]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   August 31

    -   **Oral presentation Midterm Exam 2 Opening**
        -   **Opening time**: 18:30 hours
        -   **Closing time**: 20:30 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 6]
            -   Have a headset with a built-in microphone to isolate noise from the environment

### Tercer corte 40%

-   September 2

    -   **Questionnaire 3 Opening**
        -   **Opening time**: 00:00 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapters 7-8]

-   September 4

    -   **Virtual meeting Money, prices and the exchange rate**
        -   **Meeting time**: 20:00 - 20:45 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 7]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   September 11

    -   **Virtual meeting Financial Market**
        -   **Meeting time**: 20:00 - 20:45 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 8]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   September 15

    -   **Questionnaire 3 Closing**
        -   **Closing time**: 22:00 hours
        -   **Prerequisites**:
            -   Have answered the Questionnaire 2 before 22:00 hours

-   September 18

    -   **Virtual meeting Labor Market**
        -   **Meeting time**: 20:00 - 20:45 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 9]
            -   Have a headset with a built-in microphone to isolate noise from the environment

-   September 21

    -   **Oral presentation Final Exam Opening**
        -   **Opening time**: 18:30 hours
        -   **Closing time**: 20:30 hours
        -   **Prerequisites**:
            -   Read and review [@cardenas_introduccion_2020, Chapter 9]
            -   Have a headset with a built-in microphone to isolate noise from the environment

# Qualifiable activities, weights and process in case of no presentation

## Activities and weights

-   **Primer corte**: 30%

    -   **Questionnaire 1**: 15%
    -   **Oral presentation Midterm Exam 1**: 15%

-   **Segundo corte**: 30%

    -   **Questionnaire 2**: 15%
    -   **Oral presentation Midterm Exam 2**: 15%

-   **Tercer corte**: 40%

    -   **Questionnaire 3**: 15%
    -   **Oral presentation Final Exam**: 25 %

## Process in case of no presentation

-   Students who do not present a **Questionare** due to force majeure or illness, must send the respective excuse scanned through the [**luis.gomezl\@unimilitar.edu.co**](mailto:luis.gomezl@unimilitar.edu.co){.email}. If it is not sent, the note will be zero. Once the excuse has been satisfactorily received, the student should **not** present the activity and the corresponding grade will be equal to the immediately subsequent grade obtained in the **Midterm Exam 1** or **Midterm Exam 2** or **Final Exam** according to the case.

-   Students who do not present a **Midterm Exam 1** or **Midterm Exam 2** or the **Final Exam**, must send and email to [**admon.distancia\@unimilitar.edu.co**](mailto:admon.distancia@unimilitar.edu.co){.email} two days before or two days after the closing date of these activities. However, this request can be made only once during the course. If it is not requested under these conditions, the grade will be zero.

# Use of Chat GPT-3, Gemini 1.5 and its friends

You can use any artificial intelligence tool and in particular any Large Language Model but please use them in a responsible way by taking into account its capabilities and limitations and always pointing out to the professor and your colleges that you have use them.

# Study material

-   [@cardenas_introduccion_2020, Chapters 1-9]

    -   [https://www.umng.edu.co/inicio](https://www.umng.edu.co/inicio) \> Biblioteca \> Inicio Bases virtuales y libros electrónicos \> Ingresar usuario y contraseña [^1] \> Buscar "Introducción a la economía colombiana 4ed" \> Leer

[^1]: Remember that the user and password to access and read the e-book are the same that you use to check your grades

# About plagiarism

-   **Acuerdo 02 de 2015 (7 de Mayo): Artículo 68. Escala de Calificaciones Numeral f**

*"La comprobación de todo fraude o intento de fraude en cualquier tipo de prueba, acarreará al estudiante y a su(s) colaborador(es), una calificación de CERO PUNTO CERO (0.0) en la prueba respectiva. Como quiera que el fraude o intento de cometerlo es una falta grave, el docente debe informar de este hecho al Director del programa respectivo, quien enviará el informe a la Decanatura de la Facultad, con el fin de iniciar los trámites correspondientes para aplicar la sanción disciplinaria prevista en éste reglamento.”*

-   **Acuerdo 05 de 2018 (6 de Junio): Artículo 20 que modifica el Artículo 107 Faltas Disciplinarias Numeral c. Número 6 del Acuerdo 02 de 2015 (7 de Mayo)**

*"Hacer plagio en cualquier tipo de documento o desplegar alguna conducta dolosa que viole el régimen de propiedad intelectual de la Universidad."*

# Study material references
