---
title: "Syllabus Economía I (Introducción a la Economía) Module B AE"
author: "Distance Learning Faculty (FAEDIS)"
date: "2022-1"
output:
  pdf_document:
    number_sections: yes
  html_document:
    df_print: paged
    highlight: tango
    number_sections: yes
    theme: flatly
    toc: yes
    toc_depth: 4
bibliography: intro_econ.bib
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo    = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```

```{r}

library(tidyverse)
library(knitr)
```

# Professor and meeting hours with students

- __Name__

    + Luis Francisco Gomez Lopez

- __Communication through the virtual classroom__

    + Every Saturday between 08:00 hours and 08:45 hours: 
    
        + _Primer Corte 30% > Virtual Meetings > Meeting hours with students_
    
    + Doubts and concerns of the topics of the course, please use the _Forum: Questions or concerns_ that can be accessed at the top of the platform
    
        + In this way, doubts can be shared with all students and all course participants can contribute to the learning process
    
    + Complaints regarding grades, please send an email to __luis.gomezl@unimilitar.edu.co__
    
        + To make any complaints, you are asked to wait first for the grades to be published on the News Bulletin Board and to be made based on the feedbacks that will be made taking into account the evaluation parameters
        
        + Urgent issues, you can call me to +57 3124541536 Tuesday, Wednesday and Friday between 17:00 hours and 18:00 hours. Please do not send me whatsapp messages or call me on weekends.

# Presentation and objectives of the subject

## Description

- Credits: 4
- Type: Theoretical-Practice
- Area of Knowledge: Professional
- Semester: 1
- Block: 2 (March 28 – June 03, 2021) 

## Justification

Comprehensive training of business managers involves knowledge of aspects related to multiple disciplines. One of the disciplines that have greater importance for such instruction is economics, where it allows you to recognize the economic environment in which the different organizations participate.

In this context, business managers need to understand the economic environment of a country in which a company operates by using economic indicators in the decision-making processes of a business.

## General objective

Introduce and familiarize students with the economic environment of the Colombian economy to understand how it affects an organization that operates in our country

# Competencies

- Global competencies:

    + Formulate strategies through planning and prospective techniques to contribute to the competitive improvement of the organization
    
    + Make investment decisions based on the financial information of the organization and its external environment to generate value for the business.

- Specific competencies:

    + Identify the external forces that affect an organization using tools from economic theory to develop and formulate strategies that create value for the business.
    
    + Interpret the economic information that affects the organization to make investment and financing decisions.

- Competencies SABE curriculum

    + Skills in handling a second language
    + Ability to work in teams
    + Ability to use the scientific method and logical thinking
    
# Schedule and contents

## Reading schedule

```{r}

tibble(Week = c(1:9),
       Material       = c(rep("[@cardenas_santamaria_introduccion_2013]", times = 9)),
       Chapter        = c("1", "2", "3", "4", "5", "6", "7", "8", "9"),
       `Initial page` = c(17, 41, 75, 111, 149, 211, 267, 319, 379),
       `Final page`   = c(39, 74, 109, 147, 209, 265, 318, 377, 431)) %>%
    mutate(`Number of pages` = (`Final page` - `Initial page`) + 1) %>% 
    kable()
```

## Detailed schedule of virtual meetings and activities

### Primer corte 30%

- March 28

    + Start Block 2
    
    + __Virtual meeting Presentation Syllabus__
        + __Meeting time__: 20:00 - 20:45 hours
        + __Prerequisites__:
            + Review the course program in advance, which can be downloaded by following the route _Primer corte 30% > Learning Resources > Support material > Presentations > 000_syllabus_intro_econ_2021_2_
            + Have a headset with a built-in microphone to isolate noise from the environment
            
- March 30

    + __Virtual meeting Introduction__
        + __Meeting time__: 20:00 - 20:45 hours 
        + __Prerequisites__: 
            + Read and review [@cardenas_santamaria_introduccion_2013, Chapter 1]
            + Have a headset with a built-in microphone to isolate noise from the environment

- April 4

    + __Questionnaire 1 Opening__
        + __Opening time__: 00:00 hours
        + __Prerequisites__:
            + Read and review [@cardenas_santamaria_introduccion_2013, Chapters 1 & 2]
            
- April 6

    + __Virtual meeting Production and Income I__
        + __Meeting time__: 20:00 - 20:45 hours
        + __Prerequisites__:  
            + Read and review [@cardenas_santamaria_introduccion_2013, Chapter 2]
            + Have a headset with a built-in microphone to isolate noise from the environment

- April 8

    + __Virtual meeting Production and Income II__
        + __Meeting time__: 20:00 - 20:45 hours
        + __Prerequisites__:  
            + Read and review [@cardenas_santamaria_introduccion_2013, Chapter 2]
            + Have a headset with a built-in microphone to isolate noise from the environment
            
- April 10

    + __Questionnaire 1 Closing__
        + __Closing time__: 22:00 hours
        + __Prerequisites__: 
            + Have answered the Questionnaire 1 before 22:00 hours

- April 11 - 17 

    + Holy Week
            
- April 20

    + __Virtual meeting Economic Growth__
        + __Meeting time__: 20:00 - 20:45 hours
        + __Prerequisites__:
            + Read and review [@cardenas_santamaria_introduccion_2013, Chapter 3]
            + Have a headset with a built-in microphone to isolate noise from the environment
            
- April 23

    + __Oral presentation Midterm Exam 1 Opening__
        + __Opening time__: 20:30 hours
        + __Closing time__: 22:30 hours
        + __Prerequisites__:
            + Read and review [@cardenas_santamaria_introduccion_2013, Chapter 3]
            + Have a headset with a built-in microphone to isolate noise from the environment

### Segundo corte 30%

- April 25

    + __Questionnaire 2 Opening__
        + __Opening time__: 00:00 hours
        + __Prerequisites__:
            + Read and review [@cardenas_santamaria_introduccion_2013, Chapters 4 & 5]

- April 27

    + __Virtual meeting Economy and institutions__
        + __Meeting time__: 20:00 - 20:45 hours
        + __Prerequisites__:  
            + Read and review [@cardenas_santamaria_introduccion_2013, Chapter 4]
            + Have a headset with a built-in microphone to isolate noise from the environment
            
- May 4

    + __Virtual meeting External sector I__
        + __Meeting time__: 20:00 - 20:45 hours
        + __Prerequisites__:  
            + Read and review [@cardenas_santamaria_introduccion_2013, Chapter 5]
            + Have a headset with a built-in microphone to isolate noise from the environment

- May 6

    + __Virtual meeting External sector II__
        + __Meeting time__: 20:00 - 20:45 hours
        + __Prerequisites__:  
            + Read and review [@cardenas_santamaria_introduccion_2013, Chapter 5]
            + Have a headset with a built-in microphone to isolate noise from the environment
            
- May 8

    + __Questionnaire 2 Closing__
        + __Closing time__: 22:00 hours
        + __Prerequisites__: 
            + Have answered the Questionnaire 2 before 22:00 hours
            
- May 11

    + __Virtual meeting Public Finances__
        + __Meeting time__: 20:00 - 20:45 hours
        + __Prerequisites__:
            + Read and review [@cardenas_santamaria_introduccion_2013, Chapter 6]
            + Have a headset with a built-in microphone to isolate noise from the environment
            
- May 14

    + __Oral presentation Midterm Exam 2 Opening__
        + __Opening time__: 20:30 hours
        + __Closing time__: 22:30 hours
        + __Prerequisites__:
            + Read and review [@cardenas_santamaria_introduccion_2013, Chapter 6]
            + Have a headset with a built-in microphone to isolate noise from the environment

### Tercer corte 40%

- May 16

    + __Questionnaire 3 Opening__
        + __Opening time__: 00:00 hours
        + __Prerequisites__:
            + Read and review [@cardenas_santamaria_introduccion_2013, Chapters 7 & 8]
            
- May 18           

    + __Virtual meeting Money, prices and the exchange rate__
        + __Meeting time__: 20:00 - 20:45 hours
        + __Prerequisites__:  
            + Read and review [@cardenas_santamaria_introduccion_2013, Chapter 7]
            + Have a headset with a built-in microphone to isolate noise from the environment
            
- May 20

    + __Virtual meeting Financial Market__
        + __Meeting time__: 20:00 - 20:45 hours
        + __Prerequisites__:  
            + Read and review [@cardenas_santamaria_introduccion_2013, Chapter 8]
            + Have a headset with a built-in microphone to isolate noise from the environment
            
- May 22

    + __Questionnaire 3 Closing__
        + __Closing time__: 22:00 hours
        + __Prerequisites__: 
            + Have answered the Questionnaire 2 before 22:00 hours

- May 25

    + __Virtual meeting Labor Market__
        + __Meeting time__: 20:00 - 20:45 hours
        + __Prerequisites__:
            + Read and review [@cardenas_santamaria_introduccion_2013, Chapter 9]
            + Have a headset with a built-in microphone to isolate noise from the environment
            
- May 28

    + __Oral presentation Final Exam Opening__
        + __Opening time__: 20:30 hours
        + __Closing time__: 22:30 hours
        + __Prerequisites__:
            + Read and review [@cardenas_santamaria_introduccion_2013, Chapter 9]
            + Have a headset with a built-in microphone to isolate noise from the environment
            
# Qualifiable activities, weights and process in case of no presentation

## Activities and weights

- __Primer corte__: 30%

    + __Questionnaire 1__: 15% 
    + __Oral presentation Midterm Exam 1__: 15% 

- __Segundo corte__: 30%

    + __Questionnaire 2__: 15% 
    + __Oral presentation Midterm Exam 2__: 15%

- __Tercer corte__: 40%

    + __Questionnaire 3__: 15%
    + __Oral presentation Final Exam__: 25 %
    
## Process in case of no presentation

- Students who do not present a __Questionare__ due to force majeure or illness, must send the respective excuse scanned through the __luis.gomezl@unimilitar.edu.co__. If it is not sent, the note will be zero. Once the excuse has been satisfactorily received, the student should __not__ present the activity and the corresponding grade will be equal to the immediately subsequent grade obtained in the __Midterm Exam 1__ or __Midterm Exam 2__ or __Final Exam__ according to the case.

- Students who do not present a __Midterm Exam 1__ or __Midterm Exam 2__ or the __Final Exam__, must send and email to _admon.distancia@unimilitar.edu.co_ two days before or two days after the closing date of these activities. However, this request can be made only once during the course. If it is not requested under these conditions, the grade will be zero.

# Study material

- [@cardenas_santamaria_introduccion_2013, Chapters 1-9]

    + The book can be consulted through the route _https://www-alphaeditorialcloud-com.ezproxy.umng.edu.co/reader/introduccion-a-la-economia-colombiana-3ed > Login with username and password_ [^1]

[^1]: Remember that the user and password to access and read the e-book are the same that you use to check your grades

# About plagiarism

- __Acuerdo 02 de 2015 (7 de Mayo): Artículo 68. Escala de Calificaciones Numeral f__ 

_"La comprobación de todo fraude o intento de fraude en cualquier tipo de prueba, acarreará al estudiante y a su(s) colaborador(es), una calificación de CERO PUNTO CERO (0.0) en la prueba respectiva. Como quiera que el fraude o intento de cometerlo es una falta grave, el docente debe informar de este hecho al Director del programa respectivo, quien enviará el informe a la Decanatura de la Facultad, con el fin de iniciar los trámites correspondientes para aplicar la sanción disciplinaria prevista en éste reglamento.”_

- __Acuerdo 05 de 2018 (6 de Junio): Artículo 20 que modifica el Artículo 107 Faltas Disciplinarias Numeral c. Número 6 del Acuerdo 02 de 2015 (7 de Mayo)__

_"Hacer plagio en cualquier tipo de documento o desplegar alguna conducta dolosa que viole el régimen de propiedad intelectual de la Universidad."_

# Study material references
