---
title: Introduction
subtitle: Session 2
author:
  - name: 
      given: Luis F.
      family: Gómez L.
    orcid: 0000-0002-2877-9015
    email: luis.gomezl@unimilitar.edu.co
    affiliations:
      - name: Distance Learning Faculty
toc: true
toc-depth: 1
slide-level: 2
format:
  presentation-umng-revealjs:
    logo-vertical-strip: 000_logos/logo-white-vertical.png 
    embed-resources: true
    reference-location: document
date: last-modified
date-format: "D MMMM, YYYY"
knitr: 
  opts_chunk: 
    echo: false
    warning: false
    fig-align: center
    out-width: 85%
lang: en
bibliography: intro_econ.bib
csl: apa-7.csl
editor: source
---

```{r}
#| label: libraries

library(tidyverse)
library(gt)
```

# Please Read Me

##

-   Check the message **Welcome greeting** published in the News Bulletin Board.

-   Dear student please edit your profile uploading a photo where your face is clearly visible.

-   If you want to participate, please fill out the following survey: **Primer corte 30% > Learning Activities > Tu opinión sobre la economía colombiana**

-   This presentation is based on [@cardenas_introduccion_2020, Chapter 1]

## 

- **Purpose**

    - Identify the main characteristics of the Colombian economy, especially those that differentiate it from other economies
    
# Firms or enterprises

##

```{r}
# Complete table
tbl_isic_col <- tibble(Section = LETTERS[1:21],
       Division = c("01-03", "05-09", "10-33",
                    "35", "36-39", "41-43",
                    "45-47", "49-53", "55-56",
                    "58-63", "64-66",
                    "68", "69-75",
                    "77-82", "84", "85",
                    "86-88", "90-93", "94-96",
                    "97-98", "99"),
       Description = c("Agricultura, ganadería, caza, silvicultura y pesca",
                       "Explotación de minas y canteras",
                       "Industrias manufactureras",
                       "Suministro de electricidad, gas, vapor, y aire acondicionado",
                       "Distribución de agua; evacuación y tratamiento de aguas residuales, gestión de desechos y
actividades de saneamiento ambiental",
                       "Construcción",
                       "Comercio al por mayor y al por menor; reparación de vehículos automotores y motocicletas",
                       "Transporte y almacenamiento",
                       "Alojamiento y servicios de comida",
                       "Información y comunicaciones", 
                       "Actividades financieras y de seguros",
                       "Actividades inmobiliarias",
                       "Actividades profesionales, científicas y técnicas",
                       "Actividades de servicios administrativos y de apoyo",
                       "Administración pública y defensa; planes de seguridad social de afiliación obligatoria",
                       "Educación",
                       "Actividades de atención de la salud humana y de asistencia social",
                       "Actividades artísticas, de entretenimiento y recreación",
                       "Otras actividades de servicios",
                       "Actividades de los hogares en calidad de empleadores; actividades no diferenciadas de los
hogares individuales como productores de bienes y servicios para uso propio",
                       "Actividades de organizaciones y entidades extraterritoriales"))

# Helpers
n_row <- nrow(tbl_isic_col)
middle <- ceiling(n_row/2)
first <- 1:middle
second <- (middle + 1):n_row
```


```{r}
#| label: tbl-isic-col-1
#| tbl-cap: ISIC adapted for Colombia [@dane_clasificacion_2022, pp. 134-677]

tbl_isic_col_1 <- tbl_isic_col[first, ]  |> 
  gt() |> 
  tab_style(
    style = cell_text(weight = "bold"), 
    locations = cells_column_labels() 
      ) |> 
  tab_options(
    row.striping.include_table_body = TRUE,
    table.border.top.width = px(3),
    table.border.top.color = "black",
    column_labels.border.bottom.width = px(2),
    column_labels.border.bottom.color = "black",
    table.border.bottom.width = px(3),
    table.border.bottom.color = "black",
    table.font.size = px(20)
    )

tbl_isic_col_1
```

##

```{r}
#| label: tbl-isic-col-2
#| tbl-cap: ISIC adapted for Colombia [@dane_clasificacion_2022, pp. 134-677]

tbl_isic_col_2 <- tbl_isic_col[second, ]  |> 
  gt() |> 
  tab_style(
    style = cell_text(weight = "bold"), 
    locations = cells_column_labels() 
      ) |> 
  tab_options(
    row.striping.include_table_body = TRUE, 
    table.border.top.width = px(3),
    table.border.top.color = "black",
    column_labels.border.bottom.width = px(2),
    column_labels.border.bottom.color = "black",
    table.border.bottom.width = px(3),
    table.border.bottom.color = "black",
    table.font.size = px(20)
    )

tbl_isic_col_2
```

## 

::: {.incremental}

- Some key definitions:

    - **Enterprise**: *"An economic unit that produces goods and services, made up of one or more establishments, which has the capacity in its own right to own assets, incur liabilities and carry out economic activities and transactions with other units."* [@dane_metodologigeneral_2024, p. 92]

    - **Establishment**: *"An economic unit with facilities defined by constructions that are permanently located in the same place, in which factors and resources are combined for the development of one or more economic activities."* [@dane_metodologigeneral_2024, p. 92]

:::

## {.smaller} 

::: {.incremental}

- Some key definitions *in the context of manufacturing, assuming the unit of observation is the establishment and the data is reported yearly*:

    - **Aggregate value (Valor Agregado)**: Gross Production (Producción Bruta) - Intermediate Consumption (Consumo Intermedio) [@dane_metodologigeneral_2024, p. 32]
    
    - **Gross Production (Producción Bruta)** [@dane_metodologigeneral_2024, p. 28]: 
    
        - Value of all products, byproducts and work manufactured by the establishment in the year (+)
        - Value of electrical energy sold (+)
        - Value of subsidies accrued (causados) in the year (+)
        - Value of inventories of products in the process of manufacture at year end (+)
        - Value of products in the process of manufacture at the beginning of the year (-) 

:::

##

::: {.incremental}

- Some key definitions *in the context of manufacturing, assuming the unit of observation is the establishment and the data is reported yearly*:

    - **Intermediate Consumption**: Value of non-durable goods and services used as inputs in the production process to produce other goods and services (Consumo Intermedio) [@dane_metodologigeneral_2024, p. 29]
    
        - For a detailed list about the variables included when calculating intermediate consumption check out [@dane_metodologigeneral_2024, pp. 29-31] 

:::

# Flows and stocks

## {.smaller}

-   **Stock**: a variable that is measured at a particular point in time

-   **Flow**: a variable that is measured over a period of time

-   Example *Declaración de renta Gustavo Francisco Petro Urrego Año Gravable 2023*:

    -   [https://www.funcionpublica.gov.co/fdci/consultaCiudadana](https://www.funcionpublica.gov.co/fdci/consultaCiudadana)

        -   Tipo de persona: NATURAL
        -   Primer nombre: Gustavo
        -   Segundo nombre: Francisco
        -   Primer apellido: Petro
        -   Segundo apellido: Urrego

## 

```{r}
#| label: tbl-net-worth
#| tbl-cap: Net worth (stock)

date_net_worth <- ymd("2023-12-31")

income_tax_net_worth <- tibble(cat = c("Total patrimonio bruto",
                               "Deudas",
                               "Total patrimonio líquido"),
                               val = c(1573515000,
                                       280563000,
                                       1292952000))

tbl_income_tax_net_worth <- income_tax_net_worth |> 
  gt() |> 
  fmt_currency(
    columns = val, 
    decimals = 0  
      ) |> 
  tab_style(
    style = cell_text(weight = "bold"), 
    locations = cells_column_labels() 
      ) |> 
  tab_header(
    title = md(str_glue("**Patrimonio ({date_net_worth})**"))
  ) |> 
  tab_options(
    row.striping.include_table_body = TRUE,
    table.border.top.width = px(3),
    table.border.top.color = "black",
    column_labels.border.bottom.width = px(2),
    column_labels.border.bottom.color = "black",
    table.border.bottom.width = px(3),
    table.border.bottom.color = "black",
    column_labels.hidden = TRUE,
    table.font.size = px(24)
    )

tbl_income_tax_net_worth
```

```{r}
#| label: tbl-labor-income
#| tbl-cap: Labor income (flow)

interval_labor_income <- c(floor_date(date_net_worth,
                                      unit = "year"),
                           date_net_worth)

income_tax_labor_income <- tibble(cat = c("Ingresos brutos",
                                          "Ingresos no constitutivos de renta",
                                          "Renta líquida"),
                               val = c(718441000,
                                       13920000,
                                       704521000)) |> 
  gt() |> 
  fmt_currency(
    columns = val, 
    decimals = 0  
  ) |> 
  tab_style(
    style = cell_text(weight = "bold"), 
    locations = cells_column_labels() 
  ) |> 
  tab_header(
    title = md(str_glue("**Rentas de trabajo ({interval_labor_income[1]} - {interval_labor_income[2]})**"))
  ) |> 
  tab_options(
    row.striping.include_table_body = TRUE,
    table.border.top.width = px(3),
    table.border.top.color = "black",
    column_labels.border.bottom.width = px(2),
    column_labels.border.bottom.color = "black",
    table.border.bottom.width = px(3),
    table.border.bottom.color = "black",
    column_labels.hidden = TRUE,
    table.font.size = px(24)
  )

income_tax_labor_income
```

# References {.allowframebreaks}