---
title: Introduction
subtitle: Session 1
author:
  - name: 
      given: Luis F.
      family: Gómez L.
    orcid: 0000-0002-2877-9015
    email: luis.gomezl@unimilitar.edu.co
    affiliations:
      - name: Distance Learning Faculty
toc: true
toc-depth: 1
slide-level: 2
format:
  presentation-umng-revealjs:
    logo-vertical-strip: 000_logos/logo-white-vertical.png 
    embed-resources: true
    scrollable: true
    mermaid-format: png
    reference-location: document
date: last-modified
date-format: "D MMMM, YYYY"
knitr: 
  opts_chunk: 
    echo: false
    warning: false
    fig-align: center
    out-width: 85%
lang: en
bibliography: intro_econ.bib
csl: apa-7.csl
editor: source
---

```{r}
#| label: libraries

library(readxl)
library(tidyverse)
library(janitor)
library(plotly)
library(scales)
```

# Please Read Me

##

-   Check the message **Welcome greeting** published in the News Bulletin Board.

-   Dear student please edit your profile uploading a photo where your face is clearly visible.

-   If you want to participate, please fill out the following survey: **Primer corte 30% > Learning Activities > Tu opinión sobre la economía colombiana**

-   This presentation is based on [@cardenas_introduccion_2020, Chapter 1]

## 

- Purpose

    - Identify the main characteristics of the Colombian economy, especially those that differentiate it from other economies


# The big picture and connecting ideas

##

![Macro, micro and organizational level business [@leal_filho_business_2020]](_000_images/001_business_environment.png){.nostretch #fig-business-environment width=85%}

## 

![A model of the economy: flows of resources [@the_core_econ_team_economy_2023, chap. 1, sec. 1.13]](_000_images/001_model_of_the_economy.png){.nostretch #fig-model-of-the-economy width=75%}

##

```{mermaid}
%%| label: fig-management-functions
%%| fig-cap: POLC framework [@kinicki_management_2022, pp. 8-10]

flowchart TB
    A("`Management
    functions`")  --> A1("`**P**lanning`") --> A11("`Setting goals 
    and deciding 
    how to achieve 
    them`")
    A --> A2("`**O**rganizing`") --> A21("`Arranging tasks, 
    people and 
    other resources 
    to accomplish 
    the work`")
    A --> A3("`**L**eading`") --> A31("`Motivating, 
    directing and 
    influencing 
    people to 
    achieve the 
    organization’s
    goals`")
    A --> A4("`**C**ontrolling`") --> A41("`Monitoring 
    performance, 
    comparing it 
    with goals and 
    taking corrective 
    action as needed`")

    classDef UmngNode fill:#fdc600,stroke:#043074,stroke-width:2px;
    class A,A1,A11,A2,A21,A3,A31,A4,A41 UmngNode;
```

##

```{mermaid}
%%| label: fig-manager-roles
%%| fig-cap: The roles managers play [@openstax_featured_principles_2019, chap. 2, sec. 1.2]

flowchart TB
    A("`Manager
    roles`") --> A1(Informational)
    A1 --> A11("`Monitor
    Disseminator
    Spokesperson`")
    A --> A2(Interpersonal)
    A2 --> A21("`Figurehead
    Leader
    Liason`")
    A --> A3(Decisional)
    A3 --> A31("`Entrepreneur
    Disturbance handler
    Resource allocator
    Negotiator`")

    classDef UmngNode fill:#fdc600,stroke:#043074,stroke-width:2px;
    class A,A1,A11,A2,A21,A3,A31 UmngNode;
```

# People

##

```{r}
# Data
# https://www.dane.gov.co/ > Estadísticas por tema >
# Demografía y población > Proyecciones de población >
# Proyecciones y retroproyecciones de población nacional para el periodo 1950-2019 y 2020-2070 con base en el CNPV 2018 >
# Serie nacional de población por área, sexo y edad para el periodo 2020-2070

# Import 
pop_pyramid_tbl <- readxl::read_excel(path = "000_data/001_serie_nacional_de_poblacion_por_area_sexo_y_edad_para_el_periodo_2020_2070.xlsx", 
                                      sheet = 1, 
                                      range = "C12:KX95") 

# Tidy
pop_pyramid_tbl_tidy <- pop_pyramid_tbl |>  
  clean_names() |>  
  filter(area_geografica == "Total") |> 
  select(-c(area_geografica, 
            total_hombres, 
            total_mujeres)) |> 
  select(ano:mujeres_100_y_mas, total) |> 
  rename(mujeres_100 = mujeres_100_y_mas, 
         hombres_100 = hombres_100_y_mas) |> 
  pivot_longer(cols = hombres_0:mujeres_100, 
               names_to = "genero_edad", 
               values_to = "pop") |> 
  separate(col = genero_edad, 
           into = c("genero", "edad"), 
           sep = "_", 
           convert = TRUE) |>  
  mutate(pct_pop = pop / total) |> 
  group_by(ano, genero) |>  
  mutate(rango_edad = cut(edad, 
                          breaks = c(seq.int(0, 85, 5), "Inf"),
                          labels = c(str_c(seq(0,80,5), 
                                           seq(4,85,5), 
                                           sep = "-"),
                                     "85+"), 
                          right = FALSE)) |> 
  ungroup() |> 
  group_by(ano, genero, rango_edad) |> 
  summarize(pop = sum(pop), 
            pct_pop = sum(pct_pop)) |>  
  ungroup() |> 
  mutate(pct_pop_label = number(x = pct_pop, 
                                scale = 100, 
                                accuracy = 0.01, 
                                suffix = "%")) |> 
  # https://github.com/plotly/plotly.R/issues/1930
  mutate(genero_en = ifelse(test = genero == "hombres", 
                            yes = "Men", 
                            no = "Women"), 
         .after = genero) |> 
  mutate(pct_pop_gen = ifelse(test = genero_en == "Men", 
                              yes = pct_pop, 
                              no = -pct_pop), 
         .after = pct_pop) |> 
  mutate(label = str_glue("Population: {pop}
                          Percentage: {pct_pop_label}"))
```

::: {.r-stack}

```{r}
#| label: fig-population-col
#| fig-cap: Population pyramids in Colombia
#| fig-height: 6

# Visualization
ticks <- seq(-0.04, 0.04, 0.02)
padding <- 0.005
  
pop_pyramid_tbl_tidy |>
  plot_ly(x = ~pct_pop_gen, 
          y = ~rango_edad,
          color = ~genero_en,
          frame = ~ano) |> 
  add_bars(orientation = "h",
           text = ~label,
           textposition = "none", 
           hoverinfo = "text",
           colors = c("Men" = "#043074", 
                      "Women" = "#fdc600"),
           marker = list(
             line = list(color = "black", 
                         width = 1))) |> 
  layout(barmode = "overlay",
         bargap = 0.1,
         xaxis = list(
           title = "",
           tickvals = ticks,
           ticktext = number(x = abs(ticks),
                             suffix = "%",
                             scale = 100,
                             accuracy = 1),
           range = c(ticks[1] - padding, 
                     ticks[length(ticks)] + padding)),
         yaxis = list(
           title = ""),
         width = 850,
         height = 550) |> 
  animation_slider(
    currentvalue = list(prefix = "Year: "))
```

:::

##

- Why the behavior of population is important?

    - *"Demographic change must be taken into account to redesign human development and comprehensive well-being policies within the framework of longer lifespans."* [@dnp_visioncolombia_2022, p. 36]

##

```{r}
# Import
height_adults_col <- read_csv(file = "000_data/001_average_height_by_year_of_birth/average_height_by_year_of_birth.csv") 

# Tidy
height_adults_col_tidy <- height_adults_col |> 
  filter(Code == "COL") |> 
  set_names(nm = c("entity", "code", "year", 
                   "male", "female")) |> 
  pivot_longer(cols = c(male, female), 
               names_to = "sex",
               values_to = "mean_height")
```


```{r}
#| label: fig-avg-adult-height-by-year-of-birth
#| fig-cap: Average adult height by year of birth in Colombia [@roser_human_2024]

# Visualize
height_adults_col_tidy |> 
  ggplot(aes(x = year, y = mean_height)) + 
  geom_line(aes(group = sex, color = sex)) + 
  scale_y_continuous(breaks = seq(round(min(height_adults_col_tidy$mean_height)), 
                                  round(max(height_adults_col_tidy$mean_height)), 
                                  by = 2)) + 
  scale_color_manual(labels = c("Men", "Women"),
                     values = c("#043074", "#fdc600")) + 
  labs(x = NULL,
       y = "Centimeters",
       color = NULL,
       subtitle = str_glue("Period: {min(height_adults_col_tidy$year)} - {max(height_adults_col_tidy$year)}"), 
       caption = str_glue("Source: NCD RisC and Our Word In Data - Human Height
                          Last update: 2019-05-14")) + 
  theme(panel.border      = element_rect(fill = NA, 
                                         color = "black"),
        plot.background   = element_rect(fill = "white"),
        panel.background  = element_rect(fill = "white"),
        legend.background = element_rect(fill = "white"),
        axis.text         = element_text(size = 10))
```

##

- Why height is important?

    - *"Height provides a measure of biological quality of life, one of the dimensions of overall quality of life."* [@meisel_estatura_2007, p. 37]

    - *"Poor nutrition and illness in childhood limit human growth. As a consequence, the average height of a population is strongly correlated with living standards in a population."* [@roser_human_2024]
    
##

```{r}
# Data
# https://www.dane.gov.co/ > Datos Abiertos Microdatos y Metadatos >
# Sociedad > Gobierno >
# Encuesta de Cultura Política - ECP - 2023  > 
## Tabla de viviendas 
## Elecciones y partidos

# Import
viviendas <- read_csv(file = "000_data/001_tabla_de_viviendas_2023.csv")
### Personas de 18 años y más
elecciones_partidos <- read_csv(file = "000_data/001_elecciones_y_partidos_2023.csv")

# Tidy
## Viviendas
viviendas_clean <- viviendas |> 
  clean_names() |> 
  select(
    # Llave vivienda
    directorio, 
    # Región a la que corresponde 
    # la vivienda
    region, 
    # Total de hogares en la 
    # vivienda?
    p70,
    # Factor de Expansión
    fex_p)

## Elecciones partidos
### Personas de 18 años y más
posicion_ideologica <- elecciones_partidos |> 
  clean_names() |> 
  select(
    # Llave vivienda 
    directorio,
    # Tabla en que se cargo la informacion
    ## 252. Tabla de Personas
    nro_encuesta,
    # Llave hogar
    hogar_numero,
    # Llave persona
    persona_numero,
    # 14. Las personas cuando piensan en política 
    # utilizan los términos izquierda y derecha. En 
    # una escala de 1 a 10 donde 1 significa izquierda 
    # y 10 significa derecha ¿dónde se ubicaria usted?
    ## 1- Izquierda
    ## 2- Izquierda
    ## 3- Izquierda
    ## 4- Centro izquierda
    ## 5- Centro
    ## 6- Centro
    ## 7- Centro derecha
    ## 8- Derecha
    ## 9- Derecha
    ## 10-Derecha
    ## 98- No sabe
    ## 99- No informa
    p5328)

## Posición idelógica
### Personas de 18 años y más
posicion_ideologica_clean <- posicion_ideologica |> 
  left_join(y = viviendas_clean, 
            by = join_by(directorio)) |> 
  mutate(p5328_label = case_when(
    p5328 %in% 1:3  ~ "Izquierda",
    p5328 %in% 4    ~ "Centro izquierda",
    p5328 %in% 5:6  ~ "Centro",
    p5328 %in% 7    ~ "Centro derecha",
    p5328 %in% 8:10 ~ "Derecha",
    p5328 %in% 98   ~ "No sabe",
    p5328 %in% 99   ~ "No informa")) |> 
  group_by(p5328_label) |> 
  summarize(n = sum(fex_p)) |> 
  ungroup() |> 
  mutate(total = sum(n)) |> 
  mutate(pct_n = n/total) |> 
  mutate(p5328_label = fct_relevel(.f = p5328_label,
                                   "Izquierda",
                                   "Centro izquierda",
                                   "Centro",
                                   "Centro derecha",
                                   "Derecha",
                                   "No sabe",
                                   "No informa")) |> 
  mutate(pct_n_label = number(x = pct_n, 
                              accuracy = 0.01,  
                              scale = 100, 
                              suffix = "%"))
```

```{r}
#| label: fig-ideological-position
#| fig-cap: Ideological position in Colombia

# Visualization
posicion_ideologica_clean |>  
  ggplot(aes(pct_n, 
             p5328_label)) +
  geom_segment(aes(xend = 0, yend = p5328_label), 
               color = "#043074") + 
  geom_point(shape = 21, 
             fill = "#fdc600",
             size = 3) +
  geom_text(aes(label = pct_n_label),
            hjust = -0.5, size = 5) + 
  scale_x_continuous(limits = c(0, 
                                max(posicion_ideologica_clean$pct_n) + 0.05)) +
  labs(x = NULL,
       y = NULL,
       subtitle = str_glue("Year: 2023
                           Data only apply for people 18 years of age and older 
                           Total target population: {number(unique(posicion_ideologica_clean$total), accuracy = 1, big.mark = ',')}"),
       caption = str_glue('Source: DANE - Encuesta de Cultura Política - ECP - 2023
                          Last update date: 2024-04-15')) + 
  theme(panel.border      = element_rect(fill = NA, 
                                         color = "black"),
        plot.background   = element_rect(fill = "white"),
        panel.background  = element_rect(fill = "white"),
        legend.background = element_rect(fill = "white"),
        axis.text         = element_text(size = 10),
        axis.text.x       = element_blank(),
        axis.ticks.x      = element_blank())
```

## 

- Why political preferences matter in an economic context?

    - In a democracy political preferences matter because the representatives that we choose determine some economic outcomes. For example:
    
        - Level of taxation and spending
        - Level of regulation like in the case of the environment or the labor market
        - Trade policy in relation to more or less free trade or proteccionism
        
## 

- If you want to learn more about people in Colombia check out:

    - Primer corte 30% \> Learning Resources \> Links of interest \> Así se ve Colombia en 100 personas - La Silla Vacía

# Households

## {.smaller} 

-   **ingtotug**: *"Ingreso total de la unidad de gasto **antes** de imputación de arriendo a propietarios y usufructuarios"* [@dnp_pobreza_2012, pp. 16-17]

    -   *Ingreso monetario primera actividad (IMPA)*
    -   *Ingreso en especie (IE)*
    -   *Ingreso segunda actividad (ISA)*
    -   *Ingreso monetario de desocupados e inactivos (IMDI)*
    -   *Ingresos por otras fuentes (IOF)*
    
-   **ingtotugarr**: "Ingreso total de la unidad de gasto con imputación de arriendo a propietarios y usufructuarios" [@dnp_pobreza_2012, pp. 24-26]

    -   Modulo B Datos de la Vivienda: 11. Si tuviera que pagar arriendo por esta vivienda, ¿cuánto estima que tendría que pagar mensualmente? [@dane_cuestionario_2019, p. 2]

##

```{r}
# Data
# https://www.dane.gov.co/ > ANDA | ARCHIVO NACIONAL DE DATOS >
# Sociedad > Pobreza y condiciones de vida >
# Medición de Pobreza Monetaria y Desigualdad 2023 > Hogares

# Import
household_tbl <- read_delim(file = "000_data/001_hogares_pobreza_monetaria_desigualdad_2023.csv", 
                            delim = ",", 
                            locale = locale(decimal_mark = ".")) 

# Tidy
household_tbl_tidy <- household_tbl |> 
  clean_names() |>  
  select(
    # Llave vivienda 
    directorio,
    # Llave hogar
    secuencia_p,
    # La vivienda ocupada por este hogar es:
    # a. Propia, totalmente pagada      (1) 
    # b. Propia, la están pagando       (2) 
    # c. En arriendo o subarriendo      (3)
    # d. En usufructo                   (4)
    # e. Posesión sin titulo (ocupante) (5)
    # f. Propiedad colectiva            (6)
    # g. Otra                           (7)
    p5090,
    # Personas en el hogar
    nper, 
    # Ingreso total de la unidad de
    # gasto antes de imputación de
    # arriendo a propietarios y usufructuarios
    ingtotug,
    # Factor de expansión anualizado 
    fex_c)

## Total househods zero income
ingtotug_zero <- filter(household_tbl_tidy, 
                        ingtotug == 0) |>  
  _$fex_c |>  
  sum() |>  
  floor()

## Total households
total_fex_c <- summarize(household_tbl_tidy, 
                         sum_fex_c = sum(fex_c))$sum_fex_c

## Income distribution of households
household_income_nper_tbl <- household_tbl_tidy |> 
  mutate(pct_fex_c = fex_c / total_fex_c) |> 
  mutate(range_ingtotug = cut(x = ingtotug,
                              breaks = c(seq.int(from = min(household_tbl_tidy$ingtotug),
                                                 by = 500000,
                                                 length.out = 20), 
                                         max(household_tbl_tidy$ingtotug) + 1),
                              right = FALSE,
                              # Check this if there is problem 
                              # with the last range in relation
                              # to the upper limit
                              dig.lab = 9)) |>
  group_by(range_ingtotug) |>  
  summarize(pct_fex_c = sum(pct_fex_c),
            mean_nper = mean(nper),
            .groups = "drop") |>  
  mutate(pct_fex_c_text = number(pct_fex_c,
                                 suffix = "%",
                                 scale = 100,
                                 accuracy = 0.01),
         mean_nper_text = number(mean_nper, 
                                 accuracy = 0.01))
```

```{r}
#| label: fig-household-income-col
#| fig-cap: Household income distribution in Colombia

# Visualization
household_income_nper_tbl |>  
  ggplot(aes(pct_fex_c, fct_rev(range_ingtotug))) +
  geom_segment(aes(xend = 0, yend = fct_rev(range_ingtotug)), 
               color = "#043074") + 
  geom_point(shape = 21, 
             fill = "#fdc600",
             size = 3) +
  geom_text(aes(label = pct_fex_c_text),
            hjust = -0.5, 
            size = 4) + 
  scale_x_continuous(limits = c(0, 
                                max(household_income_nper_tbl$pct_fex_c) + 0.025)) +
  labs(x = NULL,
       y = "Monthly mean income range (2023 COP)",
       subtitle = str_glue("Year: 2023
                           Total households: {number(total_fex_c, accuracy = 1, big.mark = ',')}
                           Total households with an income of zero: {number(ingtotug_zero, accuracy = 1, big.mark = ',')}"),
       caption = str_glue('Source: DANE - Medición de Pobreza Monetaria y Desigualdad 2023
                          Last update date: 2024-09-02')) + 
  theme(panel.border      = element_rect(fill = NA, 
                                         color = "black"),
        plot.background   = element_rect(fill = "white"),
        panel.background  = element_rect(fill = "white"),
        legend.background = element_rect(fill = "white"),
        axis.text         = element_text(size = 10),
        axis.text.x       = element_blank(),
        axis.ticks.x      = element_blank())
```

## {.smaller} 

-   If you want to explore more about this topic using data from the year 2019 check out[^1]:

    -   <https://www.ofiscal.org/> \> Interactúa \> Calcule dónde se ubica su hogar según su ingreso →

        -   <https://www.ofiscal.org/ingresosxhogares>

-   If you want to explore more about this topic from around the world in a visual way check out[^2]:

    -   [See how the rest of the world lives, organized by income](https://youtu.be/u4L130DkdOw) (Configure spanish subtitles in the setting options)

    -   <https://www.gapminder.org> \> Resources \> Tools \> Dollar Street

        -   <https://www.gapminder.org/dollar-street>

[^1]: Observatorio Fiscal de la Pontifica Universidad Javeriana

[^2]: Dollar Street

## {.smaller} 

-   Housing (vivienda) ownership status[^3]

    -   *Propia, totalmente pagada*
    -   *Propia, la están pagando*
    -   *En arriendo o subarriendo*
    -   *En usufructo*
    -   *Posesión sin título*
    -   *Propiedad colectiva*
    -   *Otra*

-   The concept of housing (vivienda) is different from a household (hogar)

    -   Zero or more households can live in a housing (vivienda)
    -   For a detail definition of a househould check out [@united_nations_principles_2017, p. 38, 2.33]

[^3]: For a detail definition of these categories checkout [@dane_manual_2022, p. 35]

##

```{r}
# Tidy
household_housing_tbl <- household_tbl_tidy |>  
  mutate(pct_fex_c = fex_c / total_fex_c) |>  
  group_by(p5090) |>  
  summarize(pct_fex_c = sum(pct_fex_c), .groups = "drop") |> 
  mutate(pct_fex_c_text = number(pct_fex_c,
                                 suffix = "%",
                                 scale = 100,
                                 accuracy = 0.01),
         p5090 = factor(p5090,
                        labels = c(str_glue("Propia, 
                                            totalmente pagada"),
                                   str_glue("Propia, 
                                            la están pagando"),
                                   str_glue("En arriendo o 
                                            subarriendo"),
                                   "En usufructo",
                                   str_glue("Posesión 
                                           sin título"),
                                   str_glue("Propiedad 
                                            colectiva"),
                                   "Otra")))
```

```{r}
#| label: fig-housing-ownership-col
#| fig-cap: Housing ownership status of households in Colombia

# Visualization
household_housing_tbl |>  
  ggplot(aes(pct_fex_c, 
             fct_reorder(.f = p5090, .x = pct_fex_c))) +
  geom_segment(aes(xend = 0, yend = p5090), 
               color = "#043074") +
  geom_point(shape = 21, 
             fill = "#fdc600",
             size = 3) +
  geom_text(aes(label = pct_fex_c_text),
            hjust = -0.5, size = 5) + 
  scale_x_continuous(limits = c(0, 
                                max(household_housing_tbl$pct_fex_c) + 0.05)) +
  labs(x = NULL,
       y = NULL,
       subtitle = str_glue("Year: 2023
                           Total households: {number(total_fex_c, accuracy = 1, big.mark = ',')}"),
       caption = str_glue('Source: DANE - Medición de Pobreza Monetaria y Desigualdad 2023
                          Last update date: 2024-09-02')) + 
  theme(panel.border      = element_rect(fill = NA, 
                                         color = "black"),
        plot.background   = element_rect(fill = "white"),
        panel.background  = element_rect(fill = "white"),
        legend.background = element_rect(fill = "white"),
        axis.text         = element_text(size = 10),
        axis.text.x       = element_blank(),
        axis.ticks.x      = element_blank())
```

# School education system

## 

```{r}
# https://www.dane.gov.co/ > Datos Abiertos Microdatos y Metadatos >
# Sociedad > Educación > Educación Formal 2022 > Obtener Microdatos >
# Caratula única sede educativa

# Import
schools_tbl <- read_csv(file = '000_data/001_caratula_unica_de_la_sede_educativa_2023.csv')

# Tidy
schools_bilin_tbl <- schools_tbl |> 
  count(
    # Sector 01 Oficial 02 No Oficial
    SECTOR_NOMBRE, 
    # ¿La sede educativa es bilingüe? 0 NO 1 SÍ
    ES_BILINGUE, 
    #  ¿La sede educativa es bilingüe? 
    # No\Si 1 Programa académico 100% bilingüe 
    # 2 Bilingüismo como profundización y énfasis en el plan de estudios
    TIPOBIL_NOMBRE) |> 
  mutate(pct_n = n / nrow(schools_tbl),
         pct_n_text = number(pct_n,
                             suffix = "%",
                             scale = 100,
                             accuracy = 0.01),
         TIPOBIL_NOMBRE = case_when(
           ES_BILINGUE == "NO" ~ "No bilingüe",
           is.na(ES_BILINGUE) == TRUE ~ "Sin información",
           TIPOBIL_NOMBRE == "Bilingüismo como profundización y énfasis en el plan de estudios" ~ "Profundización/énfasis",
           TIPOBIL_NOMBRE == "Programa académico 100% bilingüe" ~ "100% bilingüe",
           .default = TIPOBIL_NOMBRE),
         TIPOBIL_NOMBRE = fct_reorder2(TIPOBIL_NOMBRE, SECTOR_NOMBRE, pct_n))
```

```{r}
#| label: fig-bilingualism-school-col
#| fig-cap: Bilingualism and educational establishments (preschool, elementary, middle and high school) in Colombia

# Visualization
schools_bilin_tbl |>
  ggplot(aes(TIPOBIL_NOMBRE, pct_n)) + 
  geom_col(color = "black",
           fill = "#043074") +
  geom_text(aes(label = pct_n_text), 
            vjust = -0.8,
            size = 5) + 
  scale_y_continuous(limits = c(0, 
                                max(schools_bilin_tbl$pct_n) + 0.1)) +
  facet_wrap(vars(SECTOR_NOMBRE)) + 
  labs(x = NULL,
       y = NULL,
       subtitle = str_glue("Year: 2023
                           Total educational establishments: {scales::number(sum(schools_bilin_tbl$n), big.mark = ',')} 
                           Total non official educational establishments: {scales::number(sum(schools_bilin_tbl$n[1:4]), big.mark = ',')}
                           Total official educational establishments: {scales::number(sum(schools_bilin_tbl$n[5:8]), big.mark = ',')}"),
       caption = str_glue('Fuente: DANE - Educación Formal - 2023 - Carátula única sede educativa
                          Última fecha de actualización: 2024-06-28')) + 
  theme(panel.border      = element_rect(fill = NA, 
                                         color = "black"),
        plot.background   = element_rect(fill = "white"),
        panel.background  = element_rect(fill = "white"),
        legend.background = element_rect(fill = "white"),
        axis.text         = element_text(size = 10),
        axis.text.x = element_text(angle = 90, 
                                   hjust = 1, 
                                   vjust = 0.5),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.background.x = element_rect(color = "black", 
                                          fill = "#fdc600"),
        strip.text.x = element_text(margin = margin(t = 5,
                                                    r = 5,
                                                    b = 5,
                                                    l = 5, 
                                                    unit = "pt"), 
                                    face = "bold", 
                                    size = 12))
```

## {.smaller}

- Why learn a foreign language?

    - The benefits of foreign language learning can be divided into 3 categories: intercultural understanding, economic benefits and cognitive [@oecd_pisa_2021, p. 10]

        - Intercultural understanding: by learning a foreign language you can understand other cultures and visions of the world [@oecd_pisa_2021, p. 11].
        
        - Economic: people who know more than one language are more likely to be employed and have better career opportunities [@oecd_pisa_2021, pp. 10-11].
        
        - Cognitive: students how learn a foreign language or more creative and better and solving complex problems [@oecd_pisa_2021, p. 11].

# References {.allowframebreaks}