---
title: Introduction
subtitle: Session 2
author:
  - name: 
      given: Luis F.
      family: Gómez L.
    orcid: 0000-0002-2877-9015
    email: luis.gomezl@unimilitar.edu.co
    affiliations:
      - name: Distance Learning Faculty
toc: true
toc-depth: 1
slide-level: 2
format:
  presentation-umng-revealjs:
    logo-vertical-strip: 000_logos/logo-white-vertical.png 
    embed-resources: true
    reference-location: document
date: last-modified
date-format: "D MMMM, YYYY"
knitr: 
  opts_chunk: 
    echo: false
    warning: false
    fig.align: center
    out.width: 85%
lang: en
bibliography: intro_econ.bib
csl: apa-7.csl
editor: source
---

```{r}
#| label: libraries

library(tidyverse)
library(gt)
library(janitor)
library(scales)
library(readxl)
library(ggrepel)
```

```{r}
#| label: palette

umng_palette <- c("#043074",
                  "#fdc600",
                  "#0d5e30",
                  "#ee2a24",
                  "#fc6700",
                  "#00b3f0",
                  "#6e3c1a",
                  "#f8941c",
                  "#8e44ad",
                  "#2c3e50",
                  "#16a085",
                  "#c0392b")
```

# Please Read Me

##

-   Check the message **Welcome greeting** published in the News Bulletin Board.

-   Dear student please edit your profile uploading a photo where your face is clearly visible.

-   If you want to participate, please fill out the following survey: **Primer corte 30% > Learning Activities > Tu opinión sobre la economía colombiana**

-   This presentation is based on [@cardenas_introduccion_2020, Chapter 1]

## 

- **Purpose**

    - Identify the main characteristics of the Colombian economy, especially those that differentiate it from other economies
    
# Firms or enterprises

##

```{r}
# Create table
tbl_isic_col <- tibble(Section = LETTERS[1:21],
       Division = c("01-03", "05-09", "10-33",
                    "35", "36-39", "41-43",
                    "45-47", "49-53", "55-56",
                    "58-63", "64-66",
                    "68", "69-75",
                    "77-82", "84", "85",
                    "86-88", "90-93", "94-96",
                    "97-98", "99"),
       Description = c("Agricultura, ganadería, caza, silvicultura y pesca",
                       "Explotación de minas y canteras",
                       "Industrias manufactureras",
                       "Suministro de electricidad, gas, vapor, y aire acondicionado",
                       "Distribución de agua; evacuación y tratamiento de aguas residuales, gestión de desechos y
actividades de saneamiento ambiental",
                       "Construcción",
                       "Comercio al por mayor y al por menor; reparación de vehículos automotores y motocicletas",
                       "Transporte y almacenamiento",
                       "Alojamiento y servicios de comida",
                       "Información y comunicaciones", 
                       "Actividades financieras y de seguros",
                       "Actividades inmobiliarias",
                       "Actividades profesionales, científicas y técnicas",
                       "Actividades de servicios administrativos y de apoyo",
                       "Administración pública y defensa; planes de seguridad social de afiliación obligatoria",
                       "Educación",
                       "Actividades de atención de la salud humana y de asistencia social",
                       "Actividades artísticas, de entretenimiento y recreación",
                       "Otras actividades de servicios",
                       "Actividades de los hogares en calidad de empleadores; actividades no diferenciadas de los
hogares individuales como productores de bienes y servicios para uso propio",
                       "Actividades de organizaciones y entidades extraterritoriales"))

# Helpers
n_row <- nrow(tbl_isic_col)
middle <- ceiling(n_row/2)
first <- 1:middle
second <- (middle + 1):n_row
```


```{r}
#| label: tbl-isic-col-1
#| tbl-cap: ISIC adapted for Colombia [@dane_clasificacion_2022, pp. 134-677]

# Table
tbl_isic_col_1 <- tbl_isic_col[first, ]  |> 
  gt() |> 
  tab_style(
    style = cell_text(weight = "bold"), 
    locations = cells_column_labels() 
      ) |> 
  tab_options(
    row.striping.include_table_body = TRUE,
    table.border.top.width = px(3),
    table.border.top.color = "black",
    column_labels.border.bottom.width = px(2),
    column_labels.border.bottom.color = "black",
    table.border.bottom.width = px(3),
    table.border.bottom.color = "black",
    table.font.size = px(20)
    )

tbl_isic_col_1
```

##

```{r}
#| label: tbl-isic-col-2
#| tbl-cap: ISIC adapted for Colombia [@dane_clasificacion_2022, pp. 134-677]

# Table
tbl_isic_col_2 <- tbl_isic_col[second, ]  |> 
  gt() |> 
  tab_style(
    style = cell_text(weight = "bold"), 
    locations = cells_column_labels() 
      ) |> 
  tab_options(
    row.striping.include_table_body = TRUE, 
    table.border.top.width = px(3),
    table.border.top.color = "black",
    column_labels.border.bottom.width = px(2),
    column_labels.border.bottom.color = "black",
    table.border.bottom.width = px(3),
    table.border.bottom.color = "black",
    table.font.size = px(20)
    )

tbl_isic_col_2
```

## 

::: {.incremental}

- Some key definitions:

    - **Enterprise**: *"An economic unit that produces goods and services, made up of one or more establishments, which has the capacity in its own right to own assets, incur liabilities and carry out economic activities and transactions with other units."* [@dane_metodologigeneral_2024, p. 92]

    - **Establishment**: *"An economic unit with facilities defined by constructions that are permanently located in the same place, in which factors and resources are combined for the development of one or more economic activities."* [@dane_metodologigeneral_2024, p. 92]

:::

## {.smaller} 

::: {.incremental}

- Some key definitions *in the context of manufacturing, assuming the unit of observation is the establishment and the data is reported yearly*:

    - **Aggregate value (Valor Agregado)**: Gross Production (Producción Bruta) - Intermediate Consumption (Consumo Intermedio) [@dane_metodologigeneral_2024, p. 32]
    
    - **Gross Production (Producción Bruta)** [@dane_metodologigeneral_2024, p. 28]: 
    
        - Value of all products, byproducts and work manufactured by the establishment in the year (+)
        - Value of electrical energy sold (+)
        - Value of subsidies accrued (causados) in the year (+)
        - Value of inventories of products in the process of manufacture at year end (+)
        - Value of products in the process of manufacture at the beginning of the year (-) 

:::

##

::: {.incremental}

- Some key definitions *in the context of manufacturing, assuming the unit of observation is the establishment and the data is reported yearly*:

    - **Intermediate Consumption**: Value of non-durable goods and services used as inputs in the production process to produce other goods and services (Consumo Intermedio) [@dane_metodologigeneral_2024, p. 29]
    
        - For a detailed list about the variables included when calculating intermediate consumption check out [@dane_metodologigeneral_2024, pp. 29-31] 

:::

##

```{r}
# https://www.dane.gov.co/ > 
# ANDA | ARCHIVO NACIONAL DE DATOS > 
# Microdatos > 
# Economía > 
# Industria > 
# Encuesta Anual Manufacturera – EAM - 2022 > 
# Obtener Microdatos

# Import
manufac_tbl <- read_csv(file = "000_data/001_encuesta_anual_manufacturera_2022.csv") |>  
  clean_names()

# Tidy
manufac_tbl_clean <- manufac_tbl |> 
  select(
    # Identificador empresa
    nordemp, 
    # Identificador establecimiento
    nordest,
    # Clasificación Industrial Internacional 
    # Uniforme Rev. 4. A.C.
    ciiu4, 
    # Año EAM
    periodo,
    # Total personal promedio ocupado en el año - mujeres
    c4r4c9t,
    # Total personal promedio ocupado en el año - hombres
    c4r4c10t,
    # Total en cantidades de energía eléctrica consumida (Kwh)
    c5r1c4,
    # Producción bruta
    prodbr2,
    # Consumo intermedio
    consin2,
    # Valor Agregado: Producción bruta - consumo intermedio 
    valagri)

## Auxiliary variables
total_nordest <- nrow(manufac_tbl_clean)

total_valagri <- manufac_tbl_clean |>  
  summarise(sum_valagri = sum(valagri)) |>  
  _$sum_valagri
```

```{r}
#| label: fig-value-added-manufacturing-sector-col
#| fig-cap: Distribution of aggregate value by establishment in the manufacturing Colombian sector

# Visualize
manufac_tbl_clean |>
  ggplot(aes(0, valagri)) +
  geom_point(shape = 21,
             fill = "#fdc600",
             size = 3,
             position = position_jitter(width = 0.2, 
                                        seed = 1234)) +
  scale_y_continuous(labels = scales::label_number(scale = 1e-9,
                                                   accuracy = 0.01,
                                                   suffix = "B", 
                                                   big.mark = ",")) + 
  labs(x = NULL, 
       y = "COP 2022 (B = Billions (10^12))",
       subtitle = str_glue("Period: 2022
                           Total aggregate value: {number(total_valagri, scale = 1e-9, accuracy = 0.01, big.mark = ',', suffix = 'B')} COP
                           Number of establisments: {total_nordest}
                           Observation: Every point represents an establishment"),
       caption = 'Source: DANE - Encuesta Anual Manufacturera – EAM - 2022
                          Last update date: 2024-11-06') +
  theme(panel.border      = element_rect(fill = NA, 
                                         color = "black"),
        plot.background   = element_rect(fill = "white"),
        panel.background  = element_rect(fill = "white"),
        legend.background = element_rect(fill = "white"),
        axis.text.x       = element_blank(),
        axis.ticks.x      = element_blank(),
        axis.text.y       = element_text(size = 10))
```

##

```{r}
#| label: fig-total-personnel-employed-manufacturing-sector-col
#| fig-cap: Distribution of average yearly total personnel employed in the manufacturing Colombian sector

# Auxiliar variables
total_personnel_employed <- manufac_tbl_clean |>
  mutate(total_personnel_employed = c4r4c9t + c4r4c10t)

less_than_10 <- total_personnel_employed$total_personnel_employed[total_personnel_employed$total_personnel_employed <= 10] |> 
  length()

# Visualization
total_personnel_employed |>   
  ggplot(aes(total_personnel_employed)) + 
  geom_histogram(boundary = 0, 
                 bins = 100,
                 color = "black",
                 fill = "#fdc600") + 
  labs(x = "Average total personnel employed", 
       y = "Frequency",
       subtitle = str_glue("Period: 2022
                           Min value: {min(total_personnel_employed$total_personnel_employed)}
                           Max value: {max(total_personnel_employed$total_personnel_employed)}
                           Number of establisments: {total_nordest}
                           Establishments with 10 or fewer employees: {less_than_10}"),
       caption = 'Source: DANE - Encuesta Anual Manufacturera – EAM - 2022
                          Last update date: 2024-11-06') +
  theme(panel.border      = element_rect(fill = NA, 
                                         color = "black"),
        plot.background   = element_rect(fill = "white"),
        panel.background  = element_rect(fill = "white"),
        legend.background = element_rect(fill = "white"),
        axis.text         = element_text(size = 10))
```

##

```{r}
#| label: fig-total-electrical-energy-consumed-manufacturing-sector-col
#| fig-cap: Distribution of electrical energy consumed in the manufacturing Colombian sector

# Visualization
manufac_tbl_clean |>   
  ggplot(aes(x = c5r1c4)) + 
  geom_histogram(boundary = 0, 
                 bins = 50,
                 color = "black",
                 fill = "#fdc600") + 
  scale_x_continuous(labels = label_number(big.mark = ",")) +
  labs(x = "Electrical energy consumed (kilowatt-hours kWh)", 
       y = "Frequency",
       subtitle = str_glue("Period: 2022
                           Min value: {number(min(manufac_tbl_clean$c5r1c4), big.mark = ',')} kWh
                           Max value: {number(max(manufac_tbl_clean$c5r1c4), big.mark = ',')} kWh
                           Number of establisments: {total_nordest}"),
       caption = 'Source: DANE - Encuesta Anual Manufacturera – EAM - 2022
                          Last update date: 2024-11-06') +
  theme(panel.border      = element_rect(fill = NA, 
                                         color = "black"),
        plot.background   = element_rect(fill = "white"),
        panel.background  = element_rect(fill = "white"),
        legend.background = element_rect(fill = "white"),
        axis.text         = element_text(size = 10))
```

# Natural environment

##

```{r}
# https://www.xm.com.co/ > 
# Generación >
# Tipos
## See types of sources that are used to 
## generate electric energy

# https://acolgen.org.co/ >
# Capacidad instalada en Colombia
## Some information about
## "matriz de generación eléctrica colombiana"

# https://ember-energy.org/ > 
# Data >
# Our Data -->
# Electricity Data Explorer
## Parameters
### Select dataset: Electricity generation
### Metric: Absolute
### Overview
#### Source: All electricity sources
#### Geography: Countries or economies
##### Colombia
#### View
##### Trend over time
##### Display: stacked
##### Reported by: year
##### From: 2000
##### to: 2024

# Import
generation_yearly <- read_delim(file = "000_data/001_yearly_electricity_generation_col.csv", 
                                delim = ",")

# Tidy
generation_yearly_clean <- generation_yearly |> 
  select(entity, 
         date,
         series,
         generation_twh) |> 
  filter(series %in% c("Bioenergy",
                       "Coal",
                       "Gas",
                       "Hydro",
                       "Nuclear",
                       "Other fossil",
                       "Other renewables",
                       "Solar",
                       "Wind")) |> 
  group_by(date) |> 
  mutate(pct_generation_twh = (generation_twh / sum(generation_twh))*100) |> 
  ungroup()
```

```{r}
#| label: fig-electricity-generation-source-col
#| fig-cap: Percentage share of Colombia electricity generation by source

# Visualization
generation_yearly_clean |> 
  ggplot(aes(x = date, pct_generation_twh, group = series)) +
  geom_area(aes(fill = series)) +
  scale_y_continuous(labels = label_number(suffix = "%")) +
  scale_fill_manual(values = umng_palette[1:9]) +
  labs(x = NULL, 
       y = "Percentage",
       fill = NULL,
       subtitle = str_glue("Period: {min(generation_yearly_clean$date)} - {max(generation_yearly_clean$date)}"),
       caption = 'Source: Ember Electricity Data Explorer, ember-energy.org
                          Last update date: 2025-07-04') +
  theme(panel.border      = element_rect(fill = NA, 
                                         color = "black"),
        plot.background   = element_rect(fill = "white"),
        panel.background  = element_rect(fill = "white"),
        legend.background = element_rect(fill = "white"),
        axis.text         = element_text(size = 10))
```

# Products

##

```{r}
# https://www.dane.gov.co/ > 
# Estadísticas por tema > 
# Precios y costos > 
# Precios > 
# Sistema de Información de Precios Agropecuarios (SIPSA) >
# Resultados por componente > 
# Precios mayoristas >
# Boletín mensual > 
# Series históricas

# Import
egg_jan_2021_tbl <- read_excel(path = "000_data/001_series_historicas_precios_mayoristas_2021_jan_july.xlsx", 
                               sheet = 2, 
                               range = "A7:E4608") |> 
  clean_names() |> 
  select(fecha, producto:precio_promedio_por_kilogramo) |>  
  filter(str_detect(string = producto, pattern = "^[Hh]uevo"))

egg_may_2025_tbl <- read_excel(path = "000_data/001_series_historicas_precios_mayoristas_2025_jan_may.xlsx", 
                                     sheet = 1, 
                                     range = "A6:F22500") |>  
  clean_names() |> 
  select(fecha, producto:precio_promedio_por_kilogramo) |> 
  filter(str_detect(string = producto, pattern = "^[Hh]uevo")) |>
  # The format of the file has change
  ## Now information is presented in only on spreadsheet
  filter(month(fecha) == 5)

# Tidy
## Auxiliar variables
market_list <- c("Cali, Siloé",
                  "Pereira, La 41",
                  "Valledupar, Mercado Nuevo")

egg_jan_may_tbl <- egg_jan_2021_tbl |>  
  bind_rows(egg_may_2025_tbl) |> 
  rename(precio = precio_promedio_por_kilogramo) |>  
  mutate(fecha = ymd(fecha),
         label_mercado = ifelse(mercado %in% market_list, mercado, ""))

n_market <- egg_jan_may_tbl |>  
  count(mercado) |>  
  nrow()
```

```{r}
#| label: fig-egg-wholesale-prices-col
#| fig-cap: Egg mean wholesale prices in Colombian market places

# Visualize
egg_jan_may_tbl |> 
  ggplot(aes(precio, producto)) + 
  geom_point(aes(fill = producto),
             shape = 21,
             color = "black",
             size = 3,
             show.legend = FALSE) + 
  geom_label_repel(aes(label = label_mercado),
                   seed = 1234,
                   size = 3, 
                   min.segment.length = 0,
                   box.padding = 0.5,
                   force = 5,
                   max.overlaps = Inf) + 
  scale_y_discrete(expand = expansion(mult = 0.05)) +
  scale_fill_manual(values = umng_palette[1:8]) +
  facet_wrap(vars(fecha)) + 
  labs(x = "COP/1 egg",
       y = NULL,
       subtitle = str_glue("Number of market places: {n_market}"),
       caption = 'Source: DANE - Sistema de información de precios SIPSA
                          Last update date: 2025-06-10')  + 
  theme(panel.border      = element_rect(fill = NA, 
                                         color = "black"),
        plot.background   = element_rect(fill = "white"),
        panel.background  = element_rect(fill = "white"),
        legend.background = element_rect(fill = "white"),
        axis.text         = element_text(size = 10),
        axis.text.x = element_text(angle = 90, 
                                   hjust = 1, 
                                   vjust = 0.5, 
                                   size = 12),
        strip.background.x = element_rect(color = "black", 
                                          fill = "#fdc600"),
        strip.text.x = element_text(margin = margin(t = 5,
                                                    r = 5,
                                                    b = 5,
                                                    l = 5, 
                                                    unit = "pt"), 
                                    face = "bold",
                                    size = 12))
```

# Flows and stocks

## {.smaller}

-   **Stock**: a variable that is measured at a particular point in time

-   **Flow**: a variable that is measured over a period of time

-   Example *Declaración de renta Gustavo Francisco Petro Urrego Año Gravable 2023*:

    -   [https://www.funcionpublica.gov.co/fdci/consultaCiudadana](https://www.funcionpublica.gov.co/fdci/consultaCiudadana)

        -   Tipo de persona: NATURAL
        -   Primer nombre: Gustavo
        -   Segundo nombre: Francisco
        -   Primer apellido: Petro
        -   Segundo apellido: Urrego

## 

```{r}
#| label: tbl-net-worth
#| tbl-cap: Net worth (stock)

date_net_worth <- ymd("2023-12-31")

income_tax_net_worth <- tibble(cat = c("Total patrimonio bruto",
                               "Deudas",
                               "Total patrimonio líquido"),
                               val = c(1573515000,
                                       280563000,
                                       1292952000))

tbl_income_tax_net_worth <- income_tax_net_worth |> 
  gt() |> 
  fmt_currency(
    columns = val, 
    decimals = 0  
      ) |> 
  tab_style(
    style = cell_text(weight = "bold"), 
    locations = cells_column_labels() 
      ) |> 
  tab_header(
    title = md(str_glue("**Patrimonio ({date_net_worth})**"))
  ) |> 
  tab_options(
    row.striping.include_table_body = TRUE,
    table.border.top.width = px(3),
    table.border.top.color = "black",
    column_labels.border.bottom.width = px(2),
    column_labels.border.bottom.color = "black",
    table.border.bottom.width = px(3),
    table.border.bottom.color = "black",
    column_labels.hidden = TRUE,
    table.font.size = px(24)
    )

tbl_income_tax_net_worth
```

```{r}
#| label: tbl-labor-income
#| tbl-cap: Labor income (flow)

interval_labor_income <- c(floor_date(date_net_worth,
                                      unit = "year"),
                           date_net_worth)

income_tax_labor_income <- tibble(cat = c("Ingresos brutos",
                                          "Ingresos no constitutivos de renta",
                                          "Renta líquida"),
                               val = c(718441000,
                                       13920000,
                                       704521000)) |> 
  gt() |> 
  fmt_currency(
    columns = val, 
    decimals = 0  
  ) |> 
  tab_style(
    style = cell_text(weight = "bold"), 
    locations = cells_column_labels() 
  ) |> 
  tab_header(
    title = md(str_glue("**Rentas de trabajo ({interval_labor_income[1]} - {interval_labor_income[2]})**"))
  ) |> 
  tab_options(
    row.striping.include_table_body = TRUE,
    table.border.top.width = px(3),
    table.border.top.color = "black",
    column_labels.border.bottom.width = px(2),
    column_labels.border.bottom.color = "black",
    table.border.bottom.width = px(3),
    table.border.bottom.color = "black",
    column_labels.hidden = TRUE,
    table.font.size = px(24)
  )

income_tax_labor_income
```

# Acknowledgments

##

-   To my family that supports me

-   To the taxpayers of Colombia and the [**UMNG students**](https://www.umng.edu.co/estudiante) who pay my salary

-   To the [**Business Science**](https://www.business-science.io/) and [**R4DS Online Learning**](https://www.rfordatasci.com/) communities where I learn [**R**](https://www.r-project.org/about.html) and [**$\pi$-thon**](https://www.python.org/about/)

-   To the [**R Core Team**](https://www.r-project.org/contributors.html), the creators of [**RStudio IDE**](https://posit.co/download/rstudio-desktop/), [**Positron**](https://positron.posit.co/), [**Quarto**](https://quarto.org/) and the authors and maintainers of the packages [**tidyverse**](https://CRAN.R-project.org/package=tidyverse),
[**gt**](https://CRAN.R-project.org/package=gt),
[**janitor**](https://CRAN.R-project.org/package=janitor), [**scales**](https://CRAN.R-project.org/package=scales), 
[**readxl**](https://CRAN.R-project.org/package=readxl),
[**ggrepel**](https://CRAN.R-project.org/package=ggrepel) and [**tinytex**](https://CRAN.R-project.org/package=tinytex) for allowing me to access these tools without paying for a license

# References {.allowframebreaks}