---
title: Introduction
subtitle: Session 1
author:
  - name: 
      given: Luis F.
      family: Gómez L.
    orcid: 0000-0002-2877-9015
    email: luis.gomezl@unimilitar.edu.co
    affiliations:
      - name: Distance Learning Faculty
toc: true
toc-depth: 1
slide-level: 2
format:
  presentation-umng-revealjs:
    logo-vertical-strip: 000_logos/logo-white-vertical.png 
    embed-resources: true
    scrollable: true
    mermaid-format: png
date: last-modified
date-format: "D MMMM, YYYY"
knitr: 
  opts_chunk: 
    echo: false
    warning: false
    fig-align: center
    out-width: 85%
lang: en
bibliography: intro_econ.bib
csl: apa-7.csl
editor: source
---

```{r}
#| label: libraries

library(readxl)
library(tidyverse)
library(janitor)
library(plotly)
library(scales)
```

# Please Read Me

##

-   Check the message **Welcome greeting** published in the News Bulletin Board.

-   Dear student please edit your profile uploading a photo where your face is clearly visible.

-   If you want to participate, please fill out the following survey: **Primer corte 30% > Learning Activities > Tu opinión sobre la economía colombiana**

-   This presentation is based on [@cardenas_introduccion_2020, Chapter 1]

## 

- Purpose

    - Identify the main characteristics of the Colombian economy, especially those that differentiate it from other economies


# The big picture and connecting ideas

##

![Macro, micro and organizational level business [@leal_filho_business_2020]](_000_images/001_business_environment.png){.nostretch #fig-business-environment width=85%}

## 

![A model of the economy: flows of resources [@the_core_econ_team_economy_2023, chap. 1, sec. 1.13]](_000_images/001_model_of_the_economy.png){.nostretch #fig-model-of-the-economy width=75%}

##

```{mermaid}
%%| label: fig-management-functions
%%| fig-cap: POLC framework [@kinicki_management_2022, pp. 8-10]

flowchart TB
    A("`Management
    functions`")  --> A1("`**P**lanning`") --> A11("`Setting goals 
    and deciding 
    how to achieve 
    them`")
    A --> A2("`**O**rganizing`") --> A21("`Arranging tasks, 
    people and 
    other resources 
    to accomplish 
    the work`")
    A --> A3("`**L**eading`") --> A31("`Motivating, 
    directing and 
    influencing 
    people to 
    achieve the 
    organization’s
    goals`")
    A --> A4("`**C**ontrolling`") --> A41("`Monitoring 
    performance, 
    comparing it 
    with goals and 
    taking corrective 
    action as needed`")

    classDef UmngNode fill:#fdc600,stroke:#043074,stroke-width:2px;
    class A,A1,A11,A2,A21,A3,A31,A4,A41 UmngNode;
```

##

```{mermaid}
%%| label: fig-manager-roles
%%| fig-cap: The roles managers play [@openstax_featured_principles_2019, chap. 2, sec. 1.2]

flowchart TB
    A("`Manager
    roles`") --> A1(Informational)
    A1 --> A11("`Monitor
    Disseminator
    Spokesperson`")
    A --> A2(Interpersonal)
    A2 --> A21("`Figurehead
    Leader
    Liason`")
    A --> A3(Decisional)
    A3 --> A31("`Entrepreneur
    Disturbance handler
    Resource allocator
    Negotiator`")

    classDef UmngNode fill:#fdc600,stroke:#043074,stroke-width:2px;
    class A,A1,A11,A2,A21,A3,A31 UmngNode;
```

# People

##

::: {.r-stack}

```{r}
#| label: fig-population-col
#| fig-cap: Population pyramids in Colombia
#| fig-height: 6

# Data
# https://www.dane.gov.co/ > Estadísticas por tema >
# Demografía y población > Proyecciones de población >
# Proyecciones y retroproyecciones de población nacional para el periodo 1950-2019 y 2020-2070 con base en el CNPV 2018 >
# Serie nacional de población por área, sexo y edad para el periodo 2020-2070

# Import 
pop_pyramid_tbl <- readxl::read_excel(path = "000_data/001_serie_nacional_de_poblacion_por_area_sexo_y_edad_para_el_periodo_2020_2070.xlsx", 
                                      sheet = 1, 
                                      range = "C12:KX95") 

# Tidy
pop_pyramid_tbl_tidy <- pop_pyramid_tbl |>  
  clean_names() |>  
  filter(area_geografica == "Total") |> 
  select(-c(area_geografica, 
            total_hombres, 
            total_mujeres)) |> 
  select(ano:mujeres_100_y_mas, total) |> 
  rename(mujeres_100 = mujeres_100_y_mas, 
         hombres_100 = hombres_100_y_mas) |> 
  pivot_longer(cols = hombres_0:mujeres_100, 
               names_to = "genero_edad", 
               values_to = "pop") |> 
  separate(col = genero_edad, 
           into = c("genero", "edad"), 
           sep = "_", 
           convert = TRUE) |>  
  mutate(pct_pop = pop / total) |> 
  group_by(ano, genero) |>  
  mutate(rango_edad = cut(edad, 
                          breaks = c(seq.int(0, 85, 5), "Inf"),
                          labels = c(str_c(seq(0,80,5), 
                                           seq(4,85,5), 
                                           sep = "-"),
                                     "85+"), 
                          right = FALSE)) |> 
  ungroup() |> 
  group_by(ano, genero, rango_edad) |> 
  summarize(pop = sum(pop), 
            pct_pop = sum(pct_pop)) |>  
  ungroup() |> 
  mutate(pct_pop_label = scales::number(x = pct_pop, 
                                        scale = 100, 
                                        accuracy = 0.01, 
                                        suffix = "%")) |> 
  # https://github.com/plotly/plotly.R/issues/1930
  mutate(genero_en = ifelse(test = genero == "hombres", 
                            yes = "Men", 
                            no = "Women"), 
         .after = genero) |> 
  mutate(pct_pop_gen = ifelse(test = genero_en == "Men", 
                              yes = pct_pop, 
                              no = -pct_pop), 
         .after = pct_pop) |> 
  mutate(label = str_glue("Population: {pop}
                          Percentage: {pct_pop_label}"))

# Visualization
ticks <- seq(-0.04, 0.04, 0.02)
padding <- 0.005
  
pop_pyramid_tbl_tidy |>
  plot_ly(x = ~pct_pop_gen, 
          y = ~rango_edad,
          color = ~genero_en,
          frame = ~ano) |> 
  add_bars(orientation = "h",
           text = ~label,
           textposition = "none", 
           hoverinfo = "text",
           colors = c("Men" = "#043074", 
                      "Women" = "#fdc600"),
           marker = list(
             line = list(color = "black", 
                         width = 1))) |> 
  layout(barmode = "overlay",
         bargap = 0.1,
         xaxis = list(
           title = "",
           tickvals = ticks,
           ticktext = number(x = abs(ticks),
                             suffix = "%",
                             scale = 100,
                             accuracy = 1),
           range = c(ticks[1] - padding, 
                     ticks[length(ticks)] + padding)),
         yaxis = list(
           title = ""),
         width = 850,
         height = 550) |> 
  animation_slider(
    currentvalue = list(prefix = "Year: "))
```

:::

# References {.allowframebreaks}