---
title: Introduction I
author: 
  - Luis Francisco Gómez López
institute: 
  - FAEDIS
date: last-modified
format: 
  beamer:
    colortheme: dolphin
    fonttheme: structurebold
    theme: AnnArbor
    link-citations: true
    colorlinks: true
    toc: true
    slide-level: 2
    section-titles: false
    include-in-header:
      file: 000_tex_files/preamble.tex
fig-cap-location: bottom
tbl-cap-location: top
knitr: 
  opts_chunk: 
    echo: false
    warning: false
    fig-align: center
    out-width: 85%
lang: en
bibliography: intro_econ.bib
---

```{r}
#| label: libraries

library(tidyverse)
library(readxl)
library(janitor)
library(scales)
library(knitr)
library(kableExtra)
library(lubridate)
library(ggrepel)
```

# Please Read Me

## 

-   Check the message **Welcome greeting** published in the News Bulletin Board.

-   Dear student please edit your profile uploading a photo where your face is clearly visible.

-   The purpose of the virtual meetings is to answer questions and not to make a summary of the study material.

-   If you want to participate, please fill out the following survey:  Primer corte 30% > Learning Activities > Tu opinión sobre la economía colombiana

-   This presentation is based on [@cardenas_introduccion_2020, Chapter 1]

# Purpose

## 

Identify the main characteristics of the Colombian economy, especially those that differentiate it from other economies

# The economy as a system

##

![A model of the economy: flows of resources [@the_core_econ_team_economy_2023, fig. 1.21]](_000_images/001_model_economy.png){#fig-model-of-the-economy fig-alt="A model of the economy: flows of resources" width="400px" height="250px"}

## 

```{r}
#| label: fig-econ-env-macro
#| fig-cap: Set of economic factors and forces that influence the development of an organization

#Plot
ggplot(data = tibble(x = 0, y = 0)) + 
    ## Company
    annotate(geom = "text", 
             x        = 0, 
             y        = 0,
             label    = "Company",
             size     = 3,
             fontface = "bold") +
    geom_rect(aes(xmin = -0.4,
                  xmax =  0.4,
                  ymin = -0.4,
                  ymax =  0.4),
              alpha = 0.3,
              fill  = "#E31A1C",
              color = "#E31A1C") +
    ## Micro environment
    geom_rect(aes(xmin = -2,
                  xmax =  2,
                  ymin = -2,
                  ymax =  2),
              alpha = 0.25,
              color = "#2C3E50") +
    annotate(geom = "text", 
             x        = 0, 
             y        = 1,
             label    = str_glue("Micro environment"),
             size     = 6,
             fontface = "bold") +
    annotate(geom = "text", 
             x     = 0, 
             y     = -1.2,
             label = str_glue("Clients - Suppliers - Intermediaries - Competitors - General Public"),
             size  = 3.5) +
    ## Macro environment
    geom_rect(aes(xmin = -3,
                  xmax =  3,
                  ymin = -3,
                  ymax =  3),
              alpha = 0.1,
              color = "#6A3D9A") +
    annotate(geom     = "text", 
             x        = 0, 
             y        = 2.5,
             label    = str_glue("Macro environment"),
             size     = 8,
             fontface = "bold") +
    annotate(geom = "text", 
             x     =  0, 
             y     = -2.5,
             label = str_glue("Economic - Political - Social - Technological - Ecological - Legal"),
             size  = 5) +
    theme_void() +
    labs(caption = "Based on https://foda-dafo.com/ ")
```

# People

## 

```{r}
#| label: fig-population-col
#| fig-cap: Population pyramid in Colombia

# Data
# https://www.dane.gov.co/ > Estadísticas por tema >
# Demografía y población > Proyecciones de población >
# Proyecciones y retroproyecciones de población nacional para el periodo 1950-2019 y 2020-2070 con base en el CNPV 2018 >
# Serie nacional de población por área, sexo y edad para el periodo 2020-2070

# Choose a year
ano_choose <- 2025

pop_pyramid_tbl <- readxl::read_excel(path = "000_data/001_serie_nacional_de_poblacion_por_area_sexo_y_edad_para_el_periodo_2020_2070.xlsx", 
                                      sheet = 1, 
                                      range = "C12:KX95") |>  
  janitor::clean_names() |>  
  filter(area_geografica == "Total") |> 
  select(-c(area_geografica, total_hombres, total_mujeres)) |> 
  select(ano:mujeres_100_y_mas, total) |> 
  rename(mujeres_100 = mujeres_100_y_mas, hombres_100 = hombres_100_y_mas) |> 
  pivot_longer(cols = hombres_0:mujeres_100, 
               names_to = "genero_edad", 
               values_to = "pop") |>  
  filter(ano %in% c(ano_choose, ano_choose + 30)) |>  
  separate(col = genero_edad, 
           into = c("genero", "edad"), 
           sep = "_", 
           convert = TRUE) |>  
  mutate(pct_pop = pop / total) |> 
  group_by(ano, genero) |>  
  mutate(rango_edad = cut(edad, 
                          breaks = c(seq.int(0, 85, 5), "Inf"),
                          labels = c(str_c(seq(0,80,5), seq(4,85,5), sep = "-"),
                                     "85+"), 
                          right = FALSE)) |> 
  ungroup() |> 
  group_by(ano, genero, rango_edad) |> 
  summarize(pop = sum(pop), pct_pop = sum(pct_pop)) |>  
  ungroup()

# Labels facets: Year, Total population
pop_years <- pop_pyramid_tbl |> 
  group_by(ano) |>
  summarize(pop = sum(pop)) |> 
  mutate(label = str_glue("{ano} (Total population: {pop})")) |> 
  select(ano, label) |> 
  deframe()

# Plot
pop_pyramid_tbl |>  
  ggplot(aes(x = ifelse(test = genero == "hombres", 
                        yes = -pct_pop, no = pct_pop) , 
             y = rango_edad,
             fill = genero)) + 
  geom_col(color = "black") + 
  scale_x_continuous(breaks = seq(-0.06, 0.06, 0.02),
                     labels = \(x) scales::number(x = abs(x),
                                                  suffix = "%",
                                                  scale = 100,
                                                  accuracy = 1), 
                     limits = c(-0.05, 0.05)) +
  scale_fill_manual(labels = c("Men", "Women"),
                    values = c("#2C3E50", "#E31A1C")) + 
  facet_wrap(facets = vars(ano),
             labeller = labeller(ano = pop_years)) + 
  labs(x = NULL,
       y = NULL,
       fill = NULL,
       caption = str_glue('Source: DANE - Proyecciones y retroproyecciones de población nacional 2020-2070 con base en el CNPV 2018 (área, sexo y edad)
                          Last update date: 2023-03-22')) + 
  theme(legend.position = "bottom",
        panel.border      = element_rect(fill = NA, 
                                         color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"),
        legend.text       = element_text(size = 12), 
        axis.text.x       = element_text(hjust = 0.2, size = 12),
        axis.text.y       = element_text(size = 12),
        strip.background.x = element_rect(color = "black", 
                                          fill = "#fefff0"),
        strip.text.x = element_text(margin = margin(t = 5,
                                                    r = 5,
                                                    b = 5,
                                                    l = 5, 
                                                    unit = "pt"), 
                                    face = "bold",
                                    size = 12))
```

##

- Why the behavior of population is important?

    - *"Demographic change must be taken into account to redesign human development and comprehensive well-being policies within the framework of longer lifespans."* [@dnp_visioncolombia_2022, p. 36]

## 

```{r}
#| label: fig-avg-height-19-year-old-col
#| fig-cap: Average height by gender for 19 year olds in Colombia

# Data
# https://ncdrisc.org/index.html > DATA DOWNLOADS > 
# NATIONAL > HEIGHT > 
# Country-specific data for all countries
height_tbl <- read_csv(file = "000_data/001_NCD_RisC_Lancet_2020_height_child_adolescent_country.csv") |> 
  janitor::clean_names() |> 
  filter(country == "Colombia", age_group == 19)

height_tbl |> 
  ggplot(aes(x = year, y = mean_height)) + 
  geom_line(aes(group = sex, color = sex)) + 
  scale_x_continuous(breaks = seq(min(height_tbl$year), 
                                  round(max(height_tbl$year), digits = -1), 
                                  by = 5)) +
  scale_y_continuous(breaks = seq(round(min(height_tbl$mean_height)), 
                                  round(max(height_tbl$mean_height)), 
                                  by = 2)) + 
  scale_color_manual(labels = c("Men", "Women"),
                     values = c("#2C3E50", "#E31A1C")) + 
  labs(x = NULL,
       y = "Centimeters",
       color = NULL,
       subtitle = str_glue("Period: {min(height_tbl$year)} - {max(height_tbl$year)}"), 
       caption = str_glue("Source: NCD Risk Factor Collaboration (NCD-RisC)
                          Last update: 2021-08-13")) +
  theme(legend.position = "bottom",
        panel.border      = element_rect(fill = NA, 
                                         color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        legend.text       = element_text(size = 12),
        axis.text         = element_text(face = "bold"),
        axis.text.x       = element_text(angle = 90, 
                                         vjust = 0.5,
                                         size = 12),
        axis.text.y       = element_text(size = 12))
```

##

- Why height is important?

    - *"Height provides a measure of biological quality of life, one of the dimensions of overall quality of life."* [@meisel_estatura_2007, p. 37]

    - *"Poor nutrition and illness in childhood limit human growth. As a consequence, the average height of a population is strongly correlated with living standards in a population."* [@roser_human_2024]
    
##

```{r}
#| label: fig-ideological-position
#| fig-cap: Ideological position in Colombia

# Data
# https://www.dane.gov.co/ > Datos Abiertos Microdatos y Metadatos >
# Sociedad > Gobierno >
# Encuesta de Cultura Política - ECP - 2023  > 
## Tabla de viviendas 
## Elecciones y partidos 

viviendas <- read_csv(file = "000_data/001_tabla_de_viviendas_2023.csv")
# Personas de 18 años y más
elecciones_partidos <- read_csv(file = "000_data/001_elecciones_y_partidos_2023.csv")

viviendas_clean <- viviendas |> 
  janitor::clean_names() |> 
  select(
    # Llave vivienda
    directorio, 
    # Región a la que corresponde 
    # la vivienda
    region, 
    # Total de hogares en la 
    # vivienda?
    p70,
    # Factor de Expansión
    fex_p)

posicion_ideologica <- elecciones_partidos |> 
  janitor::clean_names() |> 
  select(
    # Llave vivienda 
    directorio,
    # Tabla en que se cargo la informacion
    ## 252. Tabla de Personas
    nro_encuesta,
    # Llave hogar
    hogar_numero,
    # Llave persona
    persona_numero,
    # 14. Las personas cuando piensan en política 
    # utilizan los términos izquierda y derecha. En 
    # una escala de 1 a 10 donde 1 significa izquierda 
    # y 10 significa derecha ¿dónde se ubicaria usted?
    ## 1- Izquierda
    ## 2- Izquierda
    ## 3- Izquierda
    ## 4- Centro izquierda
    ## 5- Centro
    ## 6- Centro
    ## 7- Centro derecha
    ## 8- Derecha
    ## 9- Derecha
    ## 10-Derecha
    ## 98- No sabe
    ## 99- No informa
    p5328)

posicion_ideologica_clean <- posicion_ideologica |> 
  left_join(y = viviendas_clean, 
            by = join_by(directorio)) |> 
  mutate(p5328_label = case_when(
    p5328 %in% 1:3  ~ "Izquierda",
    p5328 %in% 4    ~ "Centro izquierda",
    p5328 %in% 5:6  ~ "Centro",
    p5328 %in% 7    ~ "Centro derecha",
    p5328 %in% 8:10 ~ "Derecha",
    p5328 %in% 98   ~ "No sabe",
    p5328 %in% 99   ~ "No informa")) |> 
  group_by(p5328_label) |> 
  summarize(n = sum(fex_p)) |> 
  ungroup() |> 
  mutate(total = sum(n)) |> 
  mutate(pct_n = n/total) |> 
  mutate(p5328_label = fct_relevel(.f = p5328_label,
                                   "Izquierda",
                                   "Centro izquierda",
                                   "Centro",
                                   "Centro derecha",
                                   "Derecha",
                                   "No sabe",
                                   "No informa")) |> 
  mutate(pct_n_label = scales::number(x = pct_n, 
                                      accuracy = 0.01,  
                                      scale = 100, 
                                      suffix = "%"))

posicion_ideologica_clean |>  
  ggplot(aes(pct_n, 
             p5328_label)) +
  geom_point(shape = 21, 
             fill = "#E31A1C",
             size = 3) +
  geom_segment(aes(xend = 0, yend = p5328_label), 
               color = "#2C3E50") + 
  geom_text(aes(label = pct_n_label),
            hjust = -0.5, size = 5) + 
  scale_x_continuous(limits = c(0, 0.7)) +
  labs(x = NULL,
       y = NULL,
       subtitle = str_glue("Year: 2023
                           Data only apply for people 18 years of age and older 
                           Total target population: {scales::number(unique(posicion_ideologica_clean$total), accuracy = 1, big.mark = ',')}"),
       caption = str_glue('Source: DANE - Encuesta de Cultura Política - ECP - 2023
                          Last update date: 2024-04-15')) + 
  theme(legend.position = "bottom",
        panel.border      = element_rect(fill = NA, 
                                         color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y       = element_text(size = 12))
```

## 

- Why political preferences matter in an economic context?

    - In a democracy political preferences matter because the representatives that we choose determine some economic outcomes. For example:
    
        - Level of taxation and spending
        - Level of regulation like in the case of the environment or the labor market
        - Trade policy in relation to more or less free trade or proteccionism
        
## 

- If you want to learn more about people in Colombia check out:

    - Primer corte 30% \> Learning Resources \> Links of interest \> Así se ve Colombia en 100 personas - La Silla Vacía 

# Households

## 

-   **ingtotug**: *"Ingreso total de la unidad de gasto **antes** de imputación de arriendo a propietarios y usufructuarios"* [@dnp_pobreza_2012, pp. 16-17]

    -   *Ingreso monetario primera actividad (IMPA)*
    -   *Ingreso en especie (IE)*
    -   *Ingreso segunda actividad (ISA)*
    -   *Ingreso monetario de desocupados e inactivos (IMDI)*
    -   *Ingresos por otras fuentes (IOF)*

-   **ingtotugarr**: "Ingreso total de la unidad de gasto con imputación de arriendo a propietarios y usufructuarios" [@dnp_pobreza_2012, pp. 24-26]

    -   Modulo B Datos de la Vivienda: 11. Si tuviera que pagar arriendo por esta vivienda, ¿cuánto estima que tendría que pagar mensualmente? [@dane_cuestionario_2019, p 2]

## 

```{r }
#| label: fig-household-income-col
#| fig-cap: Household income distribution in Colombia

# Data
# https://www.dane.gov.co/ > Datos Abiertos Microdatos y Metadatos >
# Sociedad > Pobreza y condiciones de vida >
# Medición de Pobreza Monetaria y Desigualdad 2023 > Hogares
household_tbl <- read_delim(file = "000_data/001_hogares_pobreza_monetaria_desigualdad_2023.csv", 
                            delim = ",", 
                            locale = locale(decimal_mark = ".")) |> 
  janitor::clean_names() |>  
  select(
    # Llave vivienda 
    directorio,
    # Llave hogar
    secuencia_p,
    # La vivienda ocupada por este hogar es:
    # a. Propia, totalmente pagada      (1) 
    # b. Propia, la están pagando       (2) 
    # c. En arriendo o subarriendo      (3)
    # d. En usufructo                   (4)
    # e. Posesión sin titulo (ocupante) (5)
    # f. Propiedad colectiva            (6)
    # g. Otra                           (7)
    p5090,
    # Personas en el hogar
    nper, 
    # Ingreso total de la unidad de
    # gasto antes de imputación de
    # arriendo a propietarios y usufructuarios
    ingtotug,
    # Factor de expansión anualizado 
    fex_c)

ingtotug_zero <- filter(household_tbl, 
                        ingtotug == 0) |>  
  _$fex_c |>  
  sum() |>  
  floor()

total_fex_c <- summarize(household_tbl, 
                         sum_fex_c = sum(fex_c))$sum_fex_c

household_income_nper_tbl <- household_tbl |> 
  mutate(pct_fex_c = fex_c / total_fex_c) |> 
  mutate(range_ingtotug = cut(x = ingtotug,
                              breaks = c(seq.int(from = min(household_tbl$ingtotug),
                                                 by = 500000,
                                                 length.out = 20), 
                                         max(household_tbl$ingtotug) + 1),
                              right = FALSE,
                              # Check this if there is problem 
                              # with the last range in relation
                              # to the upper limit
                              dig.lab = 9)) |>
  group_by(range_ingtotug) |>  
  summarize(pct_fex_c = sum(pct_fex_c),
            mean_nper = mean(nper),
            .groups = "drop") |>  
  mutate(pct_fex_c_text = scales::number(pct_fex_c,
                                         suffix = "%",
                                         scale = 100,
                                         accuracy = 0.01),
         mean_nper_text = scales::number(mean_nper, 
                                         accuracy = 0.01))

household_income_nper_tbl |>  
  ggplot(aes(pct_fex_c, fct_rev(range_ingtotug))) +
  geom_point(shape = 21, 
             fill = "#E31A1C",
             size = 3) +
  geom_segment(aes(xend = 0, yend = fct_rev(range_ingtotug)), 
               color = "#2C3E50") + 
  geom_text(aes(label = pct_fex_c_text),
            hjust = -0.5, 
            size = 4) + 
  scale_x_continuous(limits = c(0, 0.3)) +
  labs(x = NULL,
       y = "Monthly mean income range (2023 COP)",
       subtitle = str_glue("Year: 2023
                           Total households: {scales::number(total_fex_c, accuracy = 1, big.mark = ',')}
                           Total households with an income of zero: {scales::number(ingtotug_zero, accuracy = 1, big.mark = ',')}"),
       caption = str_glue('Source: DANE - Medición de Pobreza Monetaria y Desigualdad 2023
                          Last update date: 2024-09-02')) + 
  theme(legend.position = "bottom",
        panel.border      = element_rect(fill = NA, 
                                         color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y       = element_text(size = 12))
```

## 

-   If you want to explore more about this topic using data from the year 2019 check out[^1]:

    -   <https://www.ofiscal.org/> \> Interactúa \> Calcule dónde se ubica su hogar según su ingreso →

        -   <https://www.ofiscal.org/ingresosxhogares>

-   If you want to explore more about this topic from around the world in a visual way check out[^2]:

    -   [See how the rest of the world lives, organized by income](https://youtu.be/u4L130DkdOw) (Configure spanish subtitles in the setting options)

    -   <https://www.gapminder.org> \> Resources \> Tools \> Dollar Street

        -   <https://www.gapminder.org/dollar-street>

[^1]: Observatorio Fiscal de la Pontifica Universidad Javeriana

[^2]: Dollar Street

## 

```{r}
#| label: fig-household-size-col
#| fig-cap: Mean individuals per household by income range in Colombia

household_income_nper_tbl |> 
  ggplot(aes(mean_nper, fct_rev(range_ingtotug))) +
  geom_point(shape = 21, 
             fill = "#E31A1C",
             size = 3) +
  geom_segment(aes(xend = 0, yend = fct_rev(range_ingtotug)), 
               color = "#2C3E50") + 
  geom_text(aes(label = mean_nper_text),
            hjust = -0.5, 
            size = 4) + 
  scale_x_continuous(limits = c(0, 5)) +
  labs(x = NULL,
       y = "Monthly mean income range (2023 COP)",
       subtitle = str_glue("Year: 2023
                           Total households: {scales::number(total_fex_c, accuracy = 1, big.mark = ',')}"),
       caption = str_glue('Source: DANE - Medición de Pobreza Monetaria y Desigualdad 2022
                          Last update date: 2024-09-02')) + 
  theme(legend.position = "bottom",
        panel.border      = element_rect(fill = NA, 
                                         color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y       = element_text(size = 12))
```

## 

-   Housing (vivienda) ownership status[^3]

    -   *Propia, totalmente pagada*
    -   *Propia, la están pagando*
    -   *En arriendo o subarriendo*
    -   *En usufructo*
    -   *Posesión sin título*
    -   *Propiedad colectiva*
    -   *Otra*

-   The concept of housing (vivienda) is different from a household (hogar)

    -   Zero or more households can live in a housing (vivienda)
    -   For a detail definition of a househould check out [@united_nations_principles_2017, p. 38, 2.33]

[^3]: For a detail definition of these categories checkout [@dane_manual_2022, p. 35]

## 

```{r}
#| label: fig-housing-ownership-col
#| fig-cap: Housing ownership status of households in Colombia

# Data
household_housing_tbl <- household_tbl |>  
  mutate(pct_fex_c = fex_c / total_fex_c) |>  
  group_by(p5090) |>  
  summarize(pct_fex_c = sum(pct_fex_c), .groups = "drop") |> 
  mutate(pct_fex_c_text = scales::number(pct_fex_c,
                                         suffix = "%",
                                         scale = 100,
                                         accuracy = 0.01),
         p5090 = factor(p5090,
                        labels = c(str_glue("Propia, 
                                            totalmente pagada"),
                                   str_glue("Propia, 
                                            la están pagando"),
                                   str_glue("En arriendo o 
                                            subarriendo"),
                                   "En usufructo",
                                   str_glue("Posesión 
                                           sin título"),
                                   str_glue("Propiedad 
                                            colectiva"),
                                   "Otra")))

household_housing_tbl |>  
  ggplot(aes(pct_fex_c, 
             fct_reorder(.f = p5090, .x = pct_fex_c))) +
  geom_point(shape = 21, 
             fill = "#E31A1C",
             size = 3) +
  geom_segment(aes(xend = 0, yend = p5090), 
               color = "#2C3E50") + 
  geom_text(aes(label = pct_fex_c_text),
            hjust = -0.5, size = 5) + 
  scale_x_continuous(limits = c(0, 0.5)) +
  labs(x = NULL,
       y = NULL,
       subtitle = str_glue("Year: 2023
                           Total households: {scales::number(total_fex_c, accuracy = 1, big.mark = ',')}"),
       caption = str_glue('Source: DANE - Medición de Pobreza Monetaria y Desigualdad 2023
                          Last update date: 2024-09-02')) + 
  theme(legend.position = "bottom",
        panel.border      = element_rect(fill = NA, 
                                         color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y       = element_text(size = 12))
```

# School education system

## 

```{r}
#| label: fig-bilingualism-school-col
#| fig-cap: Bilingualism and educational establishments (preschool, elementary, middle and high school) in Colombia

# https://www.dane.gov.co/ > Datos Abiertos Microdatos y Metadatos >
# Sociedad > Educación > Educación Formal 2022 > Obtener Microdatos >
# Caratula única sede educativa
schools_tbl <- read_csv(file = '000_data/001_caratula_unica_de_la_sede_educativa_2023.csv')

schools_bilin_tbl <- schools_tbl |> 
  count(# Sector 01 Oficial 02 No Oficial
    SECTOR_NOMBRE, 
    # ¿La sede educativa es bilingüe? 0 NO 1 SÍ
    ES_BILINGUE, 
    #  ¿La sede educativa es bilingüe? 
    # No\Si 1 Programa académico 100% bilingüe 
    # 2 Bilingüismo como profundización y énfasis en el plan de estudios
    TIPOBIL_NOMBRE) |> 
  mutate(pct_n = n / nrow(schools_tbl),
         pct_n_text = scales::number(pct_n,
                                     suffix = "%",
                                     scale = 100,
                                     accuracy = 0.01),
         TIPOBIL_NOMBRE = case_when(
           ES_BILINGUE == "NO" ~ "No bilingüe",
           is.na(ES_BILINGUE) == TRUE ~ "Sin información",
           TIPOBIL_NOMBRE == "Bilingüismo como profundización y énfasis en el plan de estudios" ~ "Profundización/énfasis",
           TIPOBIL_NOMBRE == "Programa académico 100% bilingüe" ~ "100% bilingüe",
           .default = TIPOBIL_NOMBRE),
         TIPOBIL_NOMBRE = fct_reorder2(TIPOBIL_NOMBRE, SECTOR_NOMBRE, pct_n))

schools_bilin_tbl |>
  ggplot(aes(TIPOBIL_NOMBRE, pct_n)) + 
  geom_col(color = "black",
           fill = "#2C3E50") +
  geom_text(aes(label = pct_n_text), 
            vjust = -0.8,
            size = 5) + 
  scale_y_continuous(limits = c(0, 0.8)) +
  facet_wrap(vars(SECTOR_NOMBRE)) + 
  labs(x = NULL,
       y = NULL,
       subtitle = str_glue("Year: 2023
                           Total educational establishments: {scales::number(sum(schools_bilin_tbl$n), big.mark = ',')} 
                           Total non official educational establishments: {scales::number(sum(schools_bilin_tbl$n[1:4]), big.mark = ',')}
                           Total official educational establishments: {scales::number(sum(schools_bilin_tbl$n[5:8]), big.mark = ',')}"),
       caption = str_glue('Fuente: DANE - Educación Formal - 2023 - Carátula única sede educativa
                          Última fecha de actualización: 2024-06-28')) + 
  theme(legend.position = "bottom",
        panel.border      = element_rect(fill = NA, 
                                         color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"),
        axis.text.x = element_text(angle = 90, 
                                   hjust = 1, 
                                   vjust = 0.5, size = 12),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.background.x = element_rect(color = "black", 
                                          fill = "#fefff0"),
        strip.text.x = element_text(margin = margin(t = 5,
                                                    r = 5,
                                                    b = 5,
                                                    l = 5, 
                                                    unit = "pt"), 
                                    face = "bold",
                                    size = 12))
```

## 

- Why learn a foreign language?

    - The benefits of foreign language learning can be divided into 3 categories: intercultural understanding, economic benefits and cognitive [@oecd_pisa_2021, p. 10]

        - Intercultural understanding: by learning a foreign language you can understand other cultures and visions of the world [@oecd_pisa_2021, p. 11].
        
        - Economic: people who know more than one language are more likely to be employed and have better career opportunities [@oecd_pisa_2021, pp. 10-11].
        
        - Cognitive: students how learn a foreign language or more creative and better and solving complex problems [@oecd_pisa_2021, p. 11]. 

# Acknowledgments

## 

-   To my family that supports me

-   To the taxpayers of Colombia and the [**UMNG students**](https://www.umng.edu.co/estudiante) who pay my salary

-   To the [**Business Science**](https://www.business-science.io/) and [**R4DS Online Learning**](https://www.rfordatasci.com/) communities where I learn [**R**](https://www.r-project.org/about.html) and [**$\pi$-thon**](https://www.python.org/about/)

-   To the [**R Core Team**](https://www.r-project.org/contributors.html), the creators of [**RStudio IDE**](https://rstudio.com/products/rstudio/), [**Quarto**](https://quarto.org/) and the authors and maintainers of the packages [**tidyverse**](https://CRAN.R-project.org/package=tidyverse), [**readxl**](https://CRAN.R-project.org/package=readxl), [**janitor**](https://CRAN.R-project.org/package=janitor), [**scales**](https://CRAN.R-project.org/package=scales), [**knitr**](https://CRAN.R-project.org/package=knitr), [**kableExtra**](https://CRAN.R-project.org/package=kableExtra), [**lubridate**](https://CRAN.R-project.org/package=lubridate), [**ggrepel**](https://CRAN.R-project.org/package=ggrepel), and [**tinytex**](https://CRAN.R-project.org/package=tinytex) for allowing me to access these tools without paying for a license

-   To the [**Linux kernel community**](https://www.kernel.org/category/about.html) for allowing me the possibility to use some [**Linux distributions**](https://static.lwn.net/Distributions/) as my main [**OS**](https://en.wikipedia.org/wiki/Operating_system) without paying for a license

# References {.allowframebreaks}
