---
title: "Production and Income II"
author: "Luis Francisco Gomez Lopez"
date: 2023-07-17
format:
  beamer:
    colortheme: dolphin
    fonttheme: structurebold
    theme: AnnArbor
    link-citations: true
    linkcolor: blue
    include-in-header:
      - text: |
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{multirow}
          \usepackage{wrapfig}
          \usepackage{float}
          \usepackage{colortbl}
          \usepackage{pdflscape}
          \usepackage{tabu}
          \usepackage{threeparttable}
          \usepackage{threeparttablex}
          \usepackage[normalem]{ulem}
          \usepackage{makecell}
          \usepackage{xcolor}
bibliography: intro_econ.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo      = FALSE, 
                      warning   = FALSE, 
                      message   = FALSE,
                      fig.align = "center")
```

```{r}
library(tidyverse)
library(readxl)
library(knitr)
library(kableExtra)
library(janitor)
library(treemapify)
library(tidyquant)
library(wbstats)
```

# Contents

- Please Read Me
- Purpose
- Inflation adjustments in GPD
- Economics sectors and GDP
- Macroeconomic identity
- Resource of interest 
- Acknowledgments
- References

# Please Read Me

- Check the message __Welcome greeting__ published in the News Bulletin Board.

- Dear student please edit your profile uploading a photo where your face is clearly visible.

- The purpose of the virtual meetings is to answer questions and not to make a summary of the study material.

- This presentation is based on [@cardenas_introduccion_2020, Chapter 2]

# Purpose

Understand how production is measured using the concept of Gross Domestic Product (GDP)

# Inflation adjustments in GPD

- Initially the Gross Domestic Product is expressed in current Local Currency Units (LCU) which is the sum of monetary values

- A particular monetary value is the product of a quantity and a unit price

- A change in the level of Gross Domestic Product, measure in current LCU, is a combination of changes in quantities and prices

    + Inflation adjustments, applied to GDP, **try** to eliminate the changes of prices
    
    + There are 2 approaches to make inflation adjustments [@abs_demystifying_2003]:
    
        + **Constant price estimates**
        + **Chain volume measures**

# Inflation adjustments in GPD

- **Chain volume measures** is the alternative used by the Departamento Administrativo Nacional de Estadística (DANE) to make inflation adjustments

    + A gentle introduction to this approach is explained in [@abs_demystifying_2003]
    
    + The monetary values adjusted using **chain volume measures** are not additive. This means that the sum of the components of an aggregate is not necessarily equal to the aggregate 
    
        + In addition to the values adjusted using **chain volume measures**, DANE reports a value known as statistical discrepancy because of non additivity

# Inflation adjustments in GPD

- Expenditure/Final demand approach using chain volume measures and expressed in thousands of millions (Colombia)

```{r}
# Data
# Estadísticas por tema > Cuentas Nacionales >  Cuentas nacionales anuales >
# Cuentas nacionales anuales > Principales agregados > Principales agregados macroeconómicos
gdp_expenditure_annual <- readxl::read_excel("000_data/002_agregados_macroeconomicos_cuentas_nal_anuales_2005_2020p_2021pr.xlsx", 
                                             sheet = 3, 
                                             range = "E129:V147", 
                                             col_names = FALSE, 
                                             trim_ws = TRUE)

gdp_expenditure_annual_table <- gdp_expenditure_annual %>% 
  select(1, 12, ncol(gdp_expenditure_annual) - 1, ncol(gdp_expenditure_annual)) %>% 
  slice(1:2, 5, 7, 16:19) %>% 
  rename(concept = `...1`, value_2015 = `...12`, value_2020 = `...17`, value_2021 = `...18`) %>% 
  mutate(concept = case_when(
    concept == "Gasto de consumo final individual de los hogares; gasto de consumo final de las ISFLH1" ~ "Gasto de consumo final individual de los hogares y las ISFLH",
    concept == "Gasto de consumo final del gobierno general" ~ "Gasto de consumo final del gobierno general",
    TRUE ~ concept),
    value_2015 = round(as.numeric(value_2015), digits = 0),
    value_2020 = round(as.numeric(value_2020), digits = 0),
    value_2021 = round(as.numeric(value_2021), digits = 0))

gdp_expenditure_annual_table_footnote <- gdp_expenditure_annual_table

names(gdp_expenditure_annual_table_footnote)[1] <- "Concepto"
names(gdp_expenditure_annual_table_footnote)[2] <- "2015"
names(gdp_expenditure_annual_table_footnote)[3] <- "2020"
names(gdp_expenditure_annual_table_footnote)[4] <- "2021"

names(gdp_expenditure_annual_table_footnote)[3] <- paste0(names(gdp_expenditure_annual_table_footnote)[3],                                                 footnote_marker_alphabet(1, "latex"))
names(gdp_expenditure_annual_table_footnote)[4] <- paste0(names(gdp_expenditure_annual_table_footnote)[4],                                                 footnote_marker_alphabet(2, "latex"))
gdp_expenditure_annual_table_footnote[[2,1]] <- paste0(gdp_expenditure_annual_table_footnote[[2,1]], footnote_marker_alphabet(3, "latex"))

gdp_expenditure_annual_table_footnote %>% 
  kable(format = "latex", booktabs = TRUE, escape = FALSE) %>%
  kable_styling(font_size = 7) %>%
  # Indent some rows
  add_indent(positions = c(2, 3), level_of_indent = 1) %>% 
  row_spec(row = 0, bold = TRUE) %>%
  row_spec(row = c(1, 4:5), background = '#18BC9C') %>% 
  row_spec(row = 6, background = '#FF7F00') %>% 
  row_spec(row = 7, background = '#e31a1c') %>% 
  row_spec(row = 8, background = '#CCBE93') %>% 
  footnote(general       = c('Source: DANE - Cuentas Nacionales Anuales - Producto Interno Bruto (PIB) - Series encadenadas de volumen con año de referencia 2015', 'Last update: 2022-06-14'),
           general_title = "",
           alphabet = c('Provisional data',
                        'Preliminary data', 
                        'Instituciones sin fines de lucro que sirven a los hogares'), 
           threeparttable = TRUE)
```

# Inflation adjustments in GPD

- Expenditure/Final demand approach using current prices and expressed in thousands of millions (Colombia)

```{r}
# Data
# Estadísticad por tema > Cuentas Nacionales >  Cuentas nacionales anuales >
# Cuentas nacionales anuales > Principales agregados > Principales agregados macroeconómicos
gdp_expenditure_annual_current <- readxl::read_excel("000_data/002_agregados_macroeconomicos_cuentas_nal_anuales_2005_2020p_2021pr.xlsx",
                                                     sheet = 2,
                                                     range = "E119:V136",
                                                     col_names = FALSE,
                                                     trim_ws = TRUE)

gdp_expenditure_annual_current_table <- gdp_expenditure_annual_current %>%
  select(1, 12, ncol(gdp_expenditure_annual_current) - 1, ncol(gdp_expenditure_annual_current)) %>%
  slice(1:2, 5, 7, 16:18) %>%
  rename(concept = `...1`, value_2015 = `...12`, value_2020 = `...17`, value_2021 = `...18`) %>%
  mutate(concept = case_when(
    concept == "Gasto de consumo final individual de los hogares; gasto de consumo final de las ISFLH1" ~ "Gasto de consumo final individual de los hogares y las ISFLH",
    concept == "Gasto de consumo final del gobierno general" ~ "Gasto de consumo final del gobierno general",
    TRUE ~ concept),
    value_2015 = round(as.numeric(value_2015), digits = 0),
    value_2020 = round(as.numeric(value_2020), digits = 0),
    value_2021 = round(as.numeric(value_2021), digits = 0))

gdp_expenditure_annual_current_table_footnote <- gdp_expenditure_annual_current_table

names(gdp_expenditure_annual_current_table_footnote)[1] <- "Concepto"
names(gdp_expenditure_annual_current_table_footnote)[2] <- "2015"
names(gdp_expenditure_annual_current_table_footnote)[3] <- "2020"
names(gdp_expenditure_annual_current_table_footnote)[4] <- "2021"

names(gdp_expenditure_annual_current_table_footnote)[3] <- paste0(names(gdp_expenditure_annual_current_table_footnote)[3],                                                 footnote_marker_alphabet(1, "latex"))
names(gdp_expenditure_annual_current_table_footnote)[4] <- paste0(names(gdp_expenditure_annual_current_table_footnote)[4],                                                 footnote_marker_alphabet(2, "latex"))
gdp_expenditure_annual_current_table_footnote[[2,1]] <- paste0(gdp_expenditure_annual_current_table_footnote[[2,1]], footnote_marker_alphabet(3, "latex"))

gdp_expenditure_annual_current_table_footnote %>%
  kable(format = "latex", booktabs = TRUE, escape = FALSE) %>%
  kable_styling(font_size = 7) %>%
  # Indent some rows
  add_indent(positions = c(2, 3), level_of_indent = 1) %>%
  row_spec(row = 0, bold = TRUE) %>%
  row_spec(row = c(1, 4:5), background = '#18BC9C') %>%
  row_spec(row = 6, background = '#FF7F00') %>%
  row_spec(row = 7, background = '#e31a1c') %>%
  footnote(general       = c('Source: DANE - Cuentas Nacionales Anuales - Producto Interno Bruto (PIB) - Precios corrientes', 'Last update: 2022-06-14'),
           general_title = "",
           alphabet = c('Provisional data',
                        'Preliminary data',
                        'Instituciones sin fines de lucro que sirven a los hogares'),
           threeparttable = TRUE)
```

# Economics sectors and GDP

-  ISIC adapted for Colombia [@dane_clasicacion_2020, p 85]

```{r}
tibble(Section = LETTERS[1:11],
       Division = c("01-03", "05-09", "10-33",
                    "35", "36-39", "41-43",
                    "45-47", "49-53", "55-56",
                    "58-63", "64-66"),
       Description = c("Agricultura, ganadería, caza, silvicultura y pesca",
                       "Explotación de minas y canteras",
                       "Industrias manufactureras",
                       "Suministro de electricidad, gas, vapor, y aire acondicionado",
                       "Distribución de agua; evacuación y tratamiento de aguas residuales, gestión de desechos y
actividades de saneamiento ambiental",
"Construcción",
"Comercio al por mayor y al por menor; reparación de vehículos automotores y motocicletas",
"Transporte y almacenamiento",
"Alojamiento y servicios de comida",
"Información y comunicaciones", 
"Actividades financieras y de seguros")) %>% 
  kable(format = "latex", booktabs = TRUE) %>% 
  kable_styling(font_size = 9, latex_options = "striped") %>%
  row_spec(row = 0, bold = TRUE) %>%
  column_spec(column = 1, width = "0.3in") %>%
  column_spec(column = 2, width = "0.3in") %>%
  column_spec(column = 3, width = "3.4in")
```

# Economics sectors and GDP

-  ISIC adapted for Colombia [@dane_clasicacion_2020, p 85]

```{r}
isic_L_U_table <-  tibble(Section = LETTERS[12:21],
       Division = c("68", "69-75",
                    "77-82", "84", "85",
                    "86-88", "90-93", "94-96",
                    "97-98", "99"),
       Description = c("Actividades inmobiliarias",
                       "Actividades profesionales, científicas y técnicas",
                       "Actividades de servicios administrativos y de poyo",
                       "Administración pública y defensa; planes de seguridad social de afiliación obligatoria",
"Educación",
"Actividades de atención de la salud humana y de asistencia social",
"Actividades artísticas, de entretenimiento y recreación",
"Otras actividades de servicios",
"Actividades de los hogares en calidad de empleadores; actividades no diferenciadas de los
hogares individuales como productores de bienes y servicios para uso propio",
"Actividades de organizaciones y entidades extraterritoriales"))

isic_L_U_table_footnote <- isic_L_U_table

isic_L_U_table_footnote[[10,3]] <- paste0(isic_L_U_table_footnote[[10,3]], footnote_marker_alphabet(1, "latex"))

isic_L_U_table_footnote %>%
  kable(format = "latex", booktabs = TRUE, escape = FALSE) %>% 
  kable_styling(font_size = 9, latex_options = "striped") %>%
  row_spec(row = 0, bold = TRUE) %>%
  column_spec(column = 1, width = "0.3in") %>% 
  column_spec(column = 2, width = "0.3in") %>%
  column_spec(column = 3, width = "3.4in") %>% 
  row_spec(row = 10, background = '#e31a1c') %>% 
  footnote(alphabet = c('Because of extraterritoriality, in the context of international law, the production units that belong to this category are not part of the domestic territory'), 
           threeparttable = TRUE)
```

# Economics sectors and GDP

```{r out.width="90%"}
# Data
# Estadísticas por tema > Cuentas nacionales >
# Cuentas Nacionales Anuales > Agregados macroeconómicos - Retropolación base 2015 > Anexo Retropolación Base 2015
gdp_sector_annual_current_1975_2013 <- readxl::read_excel("000_data/002_anexo_retropolacion_base_2015.xlsx",
                                                          sheet = 2,
                                                          range = "B13:AP26",
                                                          col_names = c("id", "sector", as.character(1975:2013)),
                                                          trim_ws = TRUE)  %>% 
  slice(-13) %>% 
  mutate(id = replace(id, is.na(id), "Impuestos - Subsidios"),
         id = replace(id, id == "R +S + T", "R + S + T")) %>% 
  pivot_longer(cols = `1975`:`2013`, 
               names_to = "year",
               values_to = "value") %>% 
  mutate(year = as.integer(year))

gdp_sector_annual_current_2014_2021 <- read_excel("000_data/002_agregados_macroeconomicos_cuentas_nal_anuales_2005_2020p_2021pr.xlsx", 
                                                  sheet = 2, 
                                                  range = "B15:V114",
                                                  col_names = c("id", "id_25", "id_61", "sector", as.character(2005:2021))) %>% 
  mutate(id = replace(id, sector == "Impuestos menos subvenciones sobre los productos", "Impuestos - Subsidios")) %>% 
  filter(!is.na(id)) %>% 
  select(-c("id_25", "id_61")) %>% 
  pivot_longer(cols = `2005`:`2021`, 
               names_to = "year",
               values_to = "value") %>% 
  mutate(year = as.integer(year)) %>% 
  filter(year %in% 2014:2021)

gdp_sector_annual_current_1975_2021 <- gdp_sector_annual_current_1975_2013 %>%
  bind_rows(gdp_sector_annual_current_2014_2021) %>% 
  group_by(year) %>% 
  mutate(pib = sum(value)) %>% 
  ungroup() %>% 
  mutate(pct_value = value / pib,
         text_id_pct_value = str_glue("{id}: {scales::percent(pct_value, accuracy = 0.01)}"))

gdp_sector_annual_current_year <- gdp_sector_annual_current_1975_2021 %>% 
  # Specify the year
  filter(year %in% 1975)

gdp_sector_annual_current_year %>% 
  ggplot(aes(area = pct_value, fill = id, label = text_id_pct_value)) +
  geom_treemap(color = "black",
               show.legend = FALSE) + 
  geom_treemap_text(place = "centre") + 
  scale_fill_tq() + 
  labs(title = "Sectors and share in GDP for Colombia", 
       subtitle = str_glue("Period: {unique(gdp_sector_annual_current_year$year)}
                           Observation: Monetary values at current prices are used because of chain volume measures non-additivity"), 
       caption = "Source: DANE - Cuentas Nacionales Anuales - Retropolación Base 2015 - PIB enfoque de la producción y del gasto a precios corrientes
       Update: 2022-06-12") +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"))
```

# Economics sectors and GDP

```{r out.width="90%"}
gdp_sector_annual_current_year <- gdp_sector_annual_current_1975_2021 %>% 
  # Specify the year
  filter(year %in% 2000)

gdp_sector_annual_current_year %>% 
  ggplot(aes(area = pct_value, fill = id, label = text_id_pct_value)) +
  geom_treemap(color = "black",
               show.legend = FALSE) + 
  geom_treemap_text(place = "centre") + 
  scale_fill_tq() + 
  labs(title = "Sectors and share in GDP for Colombia", 
       subtitle = str_glue("Period: {unique(gdp_sector_annual_current_year$year)}
                           Observation: Monetary values at current prices are used because of chain volume measures non-additivity"), 
       caption = "Source: DANE - Cuentas Nacionales Anuales - Retropolación Base 2015 - PIB enfoque de la producción y del gasto a precios corrientes
       Update: 2022-06-14") +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"))
```

# Economics sectors and GDP

```{r out.width="90%"}
gdp_sector_annual_current_year <- gdp_sector_annual_current_1975_2021 %>% 
  # Specify the year
  filter(year %in% 2021)

gdp_sector_annual_current_year %>% 
  ggplot(aes(area = pct_value, fill = id, label = text_id_pct_value)) +
  geom_treemap(color = "black",
               show.legend = FALSE) + 
  geom_treemap_text(place = "centre") + 
  scale_fill_tq() + 
  labs(title = "Sectors and share in GDP for Colombia", 
       subtitle = str_glue("Period: {unique(gdp_sector_annual_current_year$year)}
                           Observation: Monetary values at current prices are used because of chain volume measures non-additivity"), 
       caption = "Source: DANE - Cuentas Nacionales Anuales - Retropolación Base 2015 - PIB enfoque de la producción y del gasto a precios corrientes
       Update: 2021-06-14") +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"))
```

# Macroeconomic identity

- Uses that firms, non-profit institutions, government bodies, households and the external sector give to production

$$\begin{split}
  \mathbf{Demand}_t  = & \text{Gasto de consumo final individual de los hogares y las ISFLH}_t + \\
  & \text{Gasto de consumo final del gobierno general}_t + \\
  & \text{Formación bruta de capital}_t + \\
  & \text{Exportaciones}_t \\
  = & C_t + G_t + I_t + X_t
  \end{split}$$

# Macroeconomic identity

- Aggregate value generated by the production units in a territory, plus taxes minus subsidies on products, plus products produced outside the territory offered to the territory

$$\begin{split}
  \mathbf{Supply}_t  = & \text{Valor agregado bruto}_t + \\
  & \text{Impuestos menos subvenciones sobre los productos}_t + \\
  & \text{Importaciones}_t \\
  = & \text{Producto interno bruto}_t + \\
  & \text{Importaciones}_t \\
  = & GDP_t + M_t
  \end{split}$$

# Macroeconomic identity

$$\begin{split}
  \mathbf{Supply}_t & = \mathbf{Demand}_t \\
  GDP_t + M_t & = C_t + G_t + I_t + X_t \\
  GDP_t & = (C_t + G_t + I_t) + (X_t - M_t)
  \end{split}$$

# Macroeconomic identity

```{r out.width="90%"}
gdp_agg_demand_pct <- wbstats::wb_data(country   = c('COL'), 
                                       indicator = c('NE.CON.PRVT.ZS',
                                                     'NE.CON.GOVT.ZS',
                                                     'NE.GDI.TOTL.ZS',
                                                     'NE.EXP.GNFS.ZS',
                                                     'NE.IMP.GNFS.ZS'), 
                                       return_wide = FALSE) %>% 
  mutate(indicator_short = case_when(
    indicator_id == 'NE.CON.PRVT.ZS' ~ 'C',
    indicator_id == 'NE.CON.GOVT.ZS' ~ 'G',
    indicator_id == 'NE.GDI.TOTL.ZS' ~ 'I',
    indicator_id == 'NE.EXP.GNFS.ZS' ~ 'X',
    indicator_id == 'NE.IMP.GNFS.ZS' ~ 'IM',
    TRUE ~ indicator))

gdp_agg_demand_pct %>% 
  ggplot(aes(x = date, y = value)) + 
  geom_point(aes(fill = indicator_short), 
             color = "black", 
             shape = 21, 
             show.legend = FALSE) +
  geom_line(aes(color = indicator_short, group = indicator_short)) +
  scale_color_tq() +
  scale_fill_tq() +
  scale_y_continuous(labels = scales::number_format(suffix = "%")) +
  labs(x        = "Year",
       y        = "Percent",
       title    = "Share of GDP components in Colombia", 
       subtitle = str_glue("C (% GDP) code WDI: NE.CON.PRVT.CN
                           G (% GDP) code WDI: NE.CON.GOVT.CN 
                           I (% GDP) code WDI: NE.GDI.TOTL.CN 
                           X (% GDP) code WDI: NE.EXP.GNFS.CN 
                           M (% GDP) code WDI: NE.IMP.GNFS.CN"),
       color    = 'Variables',
       caption  = str_glue("Source: World Development Indicators (WDI) - World Bank
                              Last update date: {unique(gdp_agg_demand_pct$last_updated)}")) +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        legend.position   = "bottom",
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"))
```

# Resource of interest

If you want a summary about the topic of Gross Domestic Product check out[^1]:

- [https://youtu.be/YXTjJSGWnsE](https://youtu.be/YXTjJSGWnsE)

[^1]: Banco de la República de Colombia - Cátedras de Macroeconomía y Banca Central

# Acknowledgments

- To my family that supports me

- To the taxpayers of Colombia and the __[UMNG students](https://www.umng.edu.co/estudiante)__ who pay my salary

- To the __[Business Science](https://www.business-science.io/)__ and __[R4DS Online Learning](https://www.rfordatasci.com/)__ communities where I learn __[R](https://www.r-project.org/about.html)__ 

- To the __[R Core Team](https://www.r-project.org/contributors.html)__, the creators of __[RStudio IDE](https://rstudio.com/products/rstudio/)__ and the authors and maintainers of the packages  __[tidyverse](https://CRAN.R-project.org/package=tidyverse)__, __[readxl](https://CRAN.R-project.org/package=readxl)__, __[knitr](https://CRAN.R-project.org/package=knitr)__, __[kableExtra](https://CRAN.R-project.org/package=kableExtra)__, __[janitor](https://CRAN.R-project.org/package=janitor)__, __[treemapify](https://CRAN.R-project.org/package=treemapify)__,
__[tidyquant](https://CRAN.R-project.org/package=tidyquant)__, __[wbstats](https://CRAN.R-project.org/package=wbstats)__ and __[tinytex](https://CRAN.R-project.org/package=tinytex)__ for allowing me to access these tools without paying for a license

- To the __[Linux kernel community](https://www.kernel.org/category/about.html)__ for allowing me the possibility to use some __[Linux distributions](https://static.lwn.net/Distributions/)__ as my main __[OS](https://en.wikipedia.org/wiki/Operating_system)__ without paying for a license

# References
