---
title: Economic Growth II
author: 
  - Luis Francisco Gómez López
institute: 
  - FAEDIS
date: last-modified
format: 
  beamer:
    colortheme: dolphin
    fonttheme: structurebold
    theme: AnnArbor
    link-citations: true
    colorlinks: true
    toc: true
    slide-level: 2
    section-titles: false
    include-in-header:
      file: 000_tex_files/preamble.tex
fig-cap-location: bottom
tbl-cap-location: top
knitr: 
  opts_chunk: 
    echo: false
    warning: false
    fig-align: center
    out-width: 85%
lang: en
bibliography: intro_econ.bib
---

```{r}
#| label: libraries

library(tidyverse)
library(knitr)
library(janitor)
library(kableExtra)
library(wbstats)
library(ggrepel)
```

# Please Read Me

## 

-   Check the message **Welcome greeting** published in the News Bulletin Board.

-   Dear student please edit your profile uploading a photo where your face is clearly visible.

-   The purpose of the virtual meetings is to answer questions and not to make a summary of the study material.

-   If you want to participate, please fill out the following survey: Primer corte 30% \> Learning Activities \> Tu opinión sobre la economía colombiana

-   This presentation is based on [@cardenas_introduccion_2020, Chapter 3]

# Purpose

## 

Analyze the determinants of economic growth

# Inputs to obtain GDP

## 

Gross domestic product is obtained by using the following inputs:

-   Labor

    -   Quantity of labor:
        -   Employment[^2]
        -   Hours worked
    -   Quality of Labor:
        -   Employment by educational attainment
        -   Compensation by educational attainment

[^2]: Includes individuals employed aged 15 years or over where this age range is necessary for international comparisons.

## 

Gross domestic product is obtained by using the following inputs[^3]:

-   Produced non-financial fixed assets[^4]

    -   Dwellings
    -   Other buildings and structures
        -   Buildings other than dwellings
        -   Other structures
        -   Land improvements
    -   Machinery and equipment
        -   Transport equipment
        -   Information and Computer Technology (ICT) equipment
        -   Other machinery and equipment
    -   Weapons systems
    -   Cultivated biological resources

[^3]: These categories are taken from [@united_nations_system_2009, p. 555]

[^4]: Other non-financial assets can be included but these are the ones that are usually measured. For more information check out [@oecd_measuring_2009] and [@oecd_measuring_2001]

## 

Other factors that affect the Gross Domestic Product different from labor and produced non-financial fixed assets

-   These factors are not directly observable but are used in growth accounting to calculate **Total Factor Productivity/Multifactor productivity** as an approximation to technological change

    -   Technological change is defined as changes in the Gross Domestic Product that are not due to changes in inputs

# Growth accounting

## 

![Labor productivity and total factor productivity [@de_vries_total_2022, page 21, fig 6]](_000_images/003_labor_productivity_total_factor_productivity.png){#fig-labor-factor-productivity-total-factor-productivity fig-alt="Labor productivity and total factor productivity" fig-align="center" width="400px" height="250px"}

## 

```{r}
#| label: tbl-growth-accounting-col
#| tbl-cap: Growth accounting applied to Colombia for selected years

# https://www.conference-board.org/us/ > 
# myTBC Member Sign In > Data >
# Data Central > TCB DATASETS > 
# Total Economy Database - Growth Accounting and Total Factor Productivity
# You need to have a free account
# Username: l*********m@gmail.com
# Password: U**********4
data_ted_col <- read_csv(file = '000_data/003_tidy_TED2.csv.gz') |>
  filter(unit_code == 'COL', 
         date %in% c(1990, 2010, 2020, 2024),
         variable_label %in% c('Growth in real GDP',
                               'Contribution of Labor Quality to real GDP growth',
                               'Contribution of Labor Quantity to real GDP growth',
                               'Contribution of Total Capital Services to real GDP growth',
                               'Growth of Total Factor Productivity')) |> 
  select(date, variable_label, values) |> 
  pivot_wider(id_cols = variable_label, 
              names_from = date, values_from = values) |> 
  rename(Measure = variable_label)

data_ted_col |>
  kable(booktabs = TRUE, 
        escape = FALSE, 
        digits = 2, 
        format = "latex") |>
  kable_styling(font_size = 9) |> 
  column_spec(column = 1, width = "2.5in") |>
  row_spec(row = 1, background = '#e31a1c') |> 
  row_spec(row = 2:3, background = '#18BC9C') |>
  row_spec(row = 4, background = '#CCBE93') |>
  row_spec(row = 5, background = '#FF7F00') |>
  footnote(general       = c('Source: Total Economy Database - Growth Accounting and Total Factor Productivity', 
                             'Last update: 2024'),
           general_title = "",
           threeparttable = TRUE)
```

# Case study: number of hours worked

##

```{r}
#| label: fig-gdp-emp-hw
#| fig-cap: Mean income vs hours worked by employees

year <- 2022

gdp_per_person_employed <- wbstats::wb_data(start_date = year,
                                            end_date   = year,
                                            indicator = c('SL.GDP.PCAP.EM.KD'), 
                                            return_wide = FALSE)

gdp_per_person_employed_clean <- gdp_per_person_employed |> 
  select(iso3c, country, date, value)

# This part needs to be update manually
average_annual_hours_actually_worked_per_worker <- read_csv(file = "000_data/003_average_annual_hours_actually_worked_per_worker.csv")

average_annual_hours_actually_worked_per_worker_clean <- average_annual_hours_actually_worked_per_worker |>  
  clean_names() |> 
  filter(time_period == {{year}}) |>
  # Worker status = Employees
  ## provides a more focused view of working hours 
  ## specifically for those in traditional employment 
  ## relationships.
  filter(worker_status_2 == "Employees") |> 
  select(ref_area, obs_value)

gdp_emp_hw <- average_annual_hours_actually_worked_per_worker_clean |> 
  inner_join(y = gdp_per_person_employed_clean, 
             by = join_by(ref_area == iso3c)) |> 
  relocate(obs_value, .before = value)

gdp_emp_hw |> 
  ggplot(aes(x = obs_value, y = value, 
             label = country)) + 
  geom_point() + 
  geom_text_repel(min.segment.length = 0, 
                  seed = 1234) +
  scale_y_continuous(labels = scales::label_number(big.mark = ",")) +
  labs(x = str_glue("{unique(average_annual_hours_actually_worked_per_worker$STRUCTURE_NAME)
}"),
       y = str_glue("{unique(gdp_per_person_employed$indicator) }"),
       subtitle = str_glue("Year: {year}
                           Number of countries: {nrow(gdp_emp_hw)}"),
                          # For last update source 1 check out in the page
                          # in overview
                          ## This needs to be update manually
       caption = str_glue("Source 1: OECD Data Explorer - OCDE
                          Source 2: World Development Indicators (WDI) − World Bank
                          Last update source 1: 2025-01-09
                          Last update source 1: {unique(gdp_per_person_employed$last_updated)}")) +
  theme(panel.border    = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        legend.position   = "bottom",
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"),
        axis.text.x       = element_text(angle = 90,
                                         vjust = 0.5,
                                         size = 12),
        axis.text.y       = element_text(size = 12))
```

# Proximate and fundamental determinants of economic growth

## 

-   Proximate determinants

    -   Increase in workers and worked hours
    -   Increase in produced non-financial fixed assets
    -   Increase in educational attainment, experience and skills (lifelong learning)
    -   Increase in overall efficiency of production (total factor productivity)

## 

-   Fundamental determinants

    -   Better institutions [@cardenas_introduccion_2020, Chapter 4]
    -   Integration into the global economy [@cardenas_introduccion_2020, Chapter 5]
    -   Geographical conditions

# Acknowledgments

## 

-   To my family that supports me

-   To the taxpayers of Colombia and the [**UMNG students**](https://www.umng.edu.co/estudiante) who pay my salary

-   To the [**Business Science**](https://www.business-science.io/) and [**R4DS Online Learning**](https://www.rfordatasci.com/) communities where I learn [**R**](https://www.r-project.org/about.html) and [**$\pi$-thon**](https://www.python.org/about/)

-   To the [**R Core Team**](https://www.r-project.org/contributors.html), the creators of [**RStudio IDE**](https://rstudio.com/products/rstudio/), [**Quarto**](https://quarto.org/) and the authors and maintainers of the packages [**tidyverse**](https://CRAN.R-project.org/package=tidyverse), [**knitr**](https://CRAN.R-project.org/package=knitr), [**janitor**](https://CRAN.R-project.org/package=janitor), [**kableExtra**](https://CRAN.R-project.org/package=kableExtra),
[**wbstats**](https://CRAN.R-project.org/package=wbstats),
[**ggrepel**](https://CRAN.R-project.org/package=ggrepel) and [**tinytex**](https://CRAN.R-project.org/package=tinytex) for allowing me to access these tools without paying for a license

-   To the [**Linux kernel community**](https://www.kernel.org/category/about.html) for allowing me the possibility to use some [**Linux distributions**](https://static.lwn.net/Distributions/) as my main [**OS**](https://en.wikipedia.org/wiki/Operating_system) without paying for a license

# References {.allowframebreaks}
