---
title: Economic Growth
subtitle: Session 1
author:
  - name: 
      given: Luis F.
      family: Gómez L.
    orcid: 0000-0002-2877-9015
    email: luis.gomezl@unimilitar.edu.co
    affiliations:
      - name: Distance Learning Faculty
toc: true
toc-depth: 1
slide-level: 2
format:
  presentation-umng-revealjs:
    logo-vertical-strip: 000_logos/logo-white-vertical.png 
    embed-resources: true
    reference-location: document
date: last-modified
date-format: "D MMMM, YYYY"
knitr: 
  opts_chunk: 
    echo: false
    warning: false
    fig.align: center
    out.width: 85%
lang: en
bibliography: intro_econ.bib
csl: apa-7.csl
editor: source
---

```{r}
#| label: libraries

library(tidyverse)
library(readxl)
library(scales)
```

```{r}
#| label: palette

umng_palette <- c(
  "#043074",
  "#fdc600",
  "#0d5e30",
  "#ee2a24",
  "#fc6700",
  "#00b3f0",
  "#6e3c1a",
  "#f8941c",
  "#8e44ad",
  "#2c3e50",
  "#16a085",
  "#c0392b",
  "#e91e63"
)
```

# Please Read Me

##

-   Check the message **Welcome greeting** published in the News Bulletin Board.

-   Dear student please edit your profile uploading a photo where your face is clearly visible.

-   If you want to participate, please fill out the following survey: **Primer corte 30% > Learning Activities > Tu opinión sobre la economía colombiana**

-   This presentation is based on [@cardenas_introduccion_2020, Chapter 3]

## 

- **Purpose**

    - Analyze the determinants of economic growth

# What is economic growth?

## {.smaller}

-   Economic growth can be define as an increase in the quantity and quality of products that a society produces and consumes [@roser_economic_2013]

-   The definition of economic growth is straightforward but this concept is extremely difficult to measure [@roser_economic_2013]

-   Economists often measure economic growth as an increase in Gross Domestic Product per capita by applying inflation adjustments. Furthermore, if international comparisons are necessary also purchase power parity (PPP) adjustments are applied [@roser_economic_2013]

-   From the long-term perspective of social history economic growth is a recent phenomena [@roser_economic_2013]

-   For data related to economic growth from a long-term perspective check out [@bolt_maddisonstyle_2024] and [@bolt_maddison_2024]

# Economic growth from a long-term perspective

## 

```{r}
# Data
# https://www.rug.nl/ggdc/ > 
# Historical Development > 
# Historical Development >
# Maddison Historical Statistics > 
# Releases > 
# 2023 > 
# Excel

# Or using the direct link
# https://www.rug.nl/ggdc/historicaldevelopment/maddison/releases/maddison-project-database-2023 >
# Excel

# Or using a doi or persistent link
# 10.34894/INZBF2
# https://dataverse.nl/dataset.xhtml?persistentId=doi:10.34894/INZBF2

# Import
data_madison_region_gdp_per_capita <- readxl::read_excel("000_data/003_mpd2023.xlsx",
                                                         sheet = 6,
                                                         range = "A2:J30",
                                                         col_names = TRUE,
                                                         trim_ws = TRUE)
# Tidy
data_madison_region_gdp_per_capita_clean <- data_madison_region_gdp_per_capita |> 
  rename(year = `...1`, World = `...10`) |> 
  mutate(year = as.integer(year)) |> 
  pivot_longer(cols = -c(year), 
               names_to = "region", 
               values_to = "gpd_per_capita") |>
  mutate(region = fct_reorder(.f = region,
                              .x = gpd_per_capita,
                              .fun = last,
                              .na_rm = FALSE,
                              .desc = TRUE))

# Helpers
## Check https://dataverse.nl/dataset.xhtml?persistentId=doi:10.34894/INZBF2
last_update_data_madison <- ymd("2024-04-26") 
```

```{r}
#| label: fig-gdp-pc-ppp-latin-america-world
#| fig-cap: GDP per−capita purchasing power parity, Latin America and the World

# Visualization
data_madison_region_gdp_per_capita_clean |> 
  ggplot(aes(x = year,
             y = gpd_per_capita, 
             group = region)) +
  geom_line(
    aes(color = region)
    ) +
  scale_x_continuous(breaks = c(seq(from = min(data_madison_region_gdp_per_capita_clean$year),
                                    to   = 2020,
                                    by   = 10))) +
  scale_y_continuous(labels = label_number(scale    = 1e-3,
                                           suffix   = "K",
                                           accuracy = 1)) +
  scale_color_manual(values = umng_palette[1:9]) +
  expand_limits(y = 0) +
  labs(x        = "Year",
       y        = "Thousands",
       color = NULL,
       subtitle = str_glue("Variable units: constant 2011 international USD
                           Period: {min(data_madison_region_gdp_per_capita_clean$year)}-{max(data_madison_region_gdp_per_capita_clean$year)}"),
       caption  = str_glue("Source: Maddison Project Database, version 2023 (Bolt and Van Zanden 2024a)
                           Last update date: {last_update_data_madison}")) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 90,
                               vjust = 0.5)
  )
```

# Economic growth paths

## 

```{r}
# Import
data_madison_country_gdp_per_capita <- read_excel("000_data/003_mpd2023.xlsx", 
                                                  sheet = 5, 
                                                  # It generates an error with
                                                  # "A1:F131145"
                                                  ## So cell_limits is used
                                                  range = cell_limits(c(1, 1), 
                                                                      c(131145, 6)),
                                                  trim_ws = TRUE)

# Tidy
data_madison_country_gdp_per_capita <- data_madison_country_gdp_per_capita |> 
  mutate(country = fct_reorder(.f = country,
                              .x = gdppc,
                              .fun = last,
                              .na_rm = FALSE,
                              .desc = TRUE))

data_madison_country_usa_col_kor_sgp <- data_madison_country_gdp_per_capita |>
  filter(countrycode %in% c("USA", "COL", "KOR", "SGP"),
         year %in% 1911:2022)
```

```{r}
#| label: fig-growth-paths-usa-col-kor
#| fig-cap: GDP per-capita purchasing power parity, Colombia, USA, South Korea and Singapore

# Visualization
data_madison_country_usa_col_kor_sgp |> 
  ggplot(aes(x = year,
             y = gdppc)) +
  geom_line(aes(color = country,
                group = country)) +
  scale_x_continuous(breaks = seq.int(from = min(data_madison_country_usa_col_kor_sgp$year), 
                                      to = max(data_madison_country_usa_col_kor_sgp$year), 
                                      by = 5)) + 
  scale_y_continuous(labels = label_number(scale    = 1e-3,
                                           suffix   = "K",
                                           accuracy = 1)) +
  scale_color_manual(values = umng_palette[1:4]) +
  expand_limits(y = 0) +
  labs(x        = "Year",
       y        = "Thousands",
       color = NULL,
       subtitle = str_glue("Variable units: constant 2011 international USD
                           Period: {min(data_madison_country_usa_col_kor_sgp$year)}-{max(data_madison_country_usa_col_kor_sgp$year)}
                           No information for South Korea in 1944-1945
                           No information for Singapore in 1940-1949"),
       caption  = str_glue("Source: Maddison Project Database, version 2023 (Bolt and Van Zanden 2024a)
                           Last update date: {last_update_data_madison}")) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 90,
                               vjust = 0.5)
  )
```

##

```{r}
# Tidy
data_madison_country_hti_dom <- data_madison_country_gdp_per_capita |>
  filter(countrycode %in% c("HTI", "DOM"),
         year %in% 1950:2022)
```

```{r}
#| label: fig-growth-paths-hti-dom
#| fig-cap: GDP per-capita purchasing power parity, Haiti and Dominican Republic

# Visualization
data_madison_country_hti_dom |> 
  ggplot(aes(x = year,
             y = gdppc)) +
  geom_line(aes(color = country,
                group = country)) +
  scale_x_continuous(breaks = seq.int(from = min(data_madison_country_hti_dom$year), 
                                      to = max(data_madison_country_hti_dom$year), 
                                      by = 10)) +
  scale_y_continuous(labels = label_number(scale    = 1e-3,
                                           suffix   = "K",
                                           accuracy = 1)) +
  scale_color_manual(values = umng_palette[1:2]) +
  expand_limits(y = 0) +
  labs(x        = "Year",
       y        = "Thousands",
       color = NULL,
       subtitle = str_glue("Variable units: constant 2011 international USD
                           Period: {min(data_madison_country_hti_dom$year)}-{max(data_madison_country_hti_dom$year)}"),
       caption  = str_glue("Source: Maddison Project Database, version 2023 (Bolt and Van Zanden 2024a)
                           Last update date: {last_update_data_madison}")) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 90,
                               vjust = 0.5)
  )
```

##

```{r}
# Tidy
data_madison_country_col_ven <- data_madison_country_gdp_per_capita |>
  filter(countrycode %in% c("COL", "VEN"),
         year %in% 1900:2022)
```

```{r}
#| label: fig-growth-paths-col-ven
#| fig-cap: GDP per-capita purchasing power parity, Colombia and Venezuela

# Visualization
data_madison_country_col_ven |> 
  ggplot(aes(x = year,
             y = gdppc)) +
  geom_line(aes(color = country,
                group = country)) +
  scale_x_continuous(breaks = seq.int(from = min(data_madison_country_col_ven$year), 
                                      to = max(data_madison_country_col_ven$year), 
                                      by = 10)) + 
  scale_y_continuous(labels = scales::label_number(scale    = 1e-3,
                                                   suffix   = "K",
                                                   accuracy = 1)) +
  scale_color_manual(values = umng_palette[1:2]) +
  expand_limits(y = 0) +
  labs(x        = "Year",
       y        = "Thousands",
       color = NULL,
       subtitle = str_glue("Variable units: constant 2011 international USD
                           Period: {min(data_madison_country_col_ven$year)}-{max(data_madison_country_col_ven$year)}"),
       caption  = str_glue("Source: Maddison Project Database, version 2023 (Bolt and Van Zanden 2024a)
                           Last update date: {last_update_data_madison}")) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 90,
                               vjust = 0.5)
  )
```

##

```{r}
# Tidy
data_madison_country_zwe_bwa <- data_madison_country_gdp_per_capita |>
  filter(countrycode %in% c("ZWE", "BWA"),
         year %in% 1950:2022)
```

```{r}
#| label: fig-growth-paths-zwe-bwa
#| fig-cap: GDP per-capita purchasing power parity, Zimbabwe and Botswana

# Visualization
data_madison_country_zwe_bwa |> 
  ggplot(aes(x = year,
             y = gdppc)) +
  geom_line(aes(color = country,
                group = country)) +
  scale_x_continuous(breaks = seq.int(from = min(data_madison_country_zwe_bwa$year), 
                                      to = max(data_madison_country_zwe_bwa$year), 
                                      by = 10)) + 
  scale_y_continuous(labels = scales::label_number(scale    = 1e-3,
                                                   suffix   = "K",
                                                   accuracy = 1)) +
  scale_color_manual(values = umng_palette[1:2]) +
  expand_limits(y = 0) +
  labs(x        = "Year",
       y        = "Thousands",
       color = NULL,
       subtitle = str_glue("Variable units: constant 2011 international USD
                           Period: {min(data_madison_country_zwe_bwa$year)}-{max(data_madison_country_zwe_bwa$year)}"),
       caption  = str_glue("Source: Maddison Project Database, version 2023 (Bolt and Van Zanden 2024a)
                           Last update date: {last_update_data_madison}")) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 90,
                               vjust = 0.5)
  )
```


# Acknowledgments

##



# References {.allowframebreaks}