---
title: "Economy and institutions"
author: 
  - Luis Francisco Gómez López
institute: 
  - FAEDIS
date: last-modified
format: 
  beamer:
    colortheme: dolphin
    fonttheme: structurebold
    theme: AnnArbor
    link-citations: true
    colorlinks: true
    toc: true
    slide-level: 2
    section-titles: false
    keep-tex: true
    include-in-header:
      file: 000_tex_files/preamble.tex
fig-cap-location: bottom
tbl-cap-location: top
knitr: 
  opts_chunk: 
    echo: false
    warning: false
    fig-align: center
    out-width: 85%
lang: en
bibliography: intro_econ.bib
---

```{r}
#| label: libraries

library(tidyverse)
library(wbstats)
library(tidyquant)
library(janitor)
library(ggrepel)
```

# Please Read Me

## 

-   Check the message **Welcome greeting** published in the News Bulletin Board.

-   Dear student please edit your profile uploading a photo where your face is clearly visible.

-   The purpose of the virtual meetings is to answer questions and not to make a summary of the study material.

-   This presentation is based on [@cardenas_introduccion_2020, Chapter 4]

# Purpose

## 

Analyze the role of institutions in the economy

# Definition of institutions

## 

-   "Institutions are the rules of the game in a society or, more formally, are the humanly devised constraints that shape human interaction" [@north_institutions_1990, p. 3]

-   Institutions can be informal or formal [@north_institutions_1991]

    -   **Informal**: sanctions, taboos, customs, traditions, and codes of conduct

    -   **Formal**: constitutions, laws, property right

# Measuring formal institutions

## 

-   Governance

    -   Set of traditions and institutions by which authority in a country is exercised where it includes [@kaufmann_worldwide_2011, p. 222]:

        -   The process by which governments are selected, monitored and replaced
        -   The capacity of the government to effectively formulate and implement sound policies
        -   The respect of citizens and the state for the institutions that govern economic and social interactions among them
        
## 

```{dot}
//| label: fig-governance-indicators
//| fig-cap: Governance dimensions [@kaufmann_worldwide_2011, p. 223]
//| fig-width: 5
//| fig-height: 2

digraph boxes_and_circles {

  graph [rankdir = LR]

  node [shape = circle, color = "#2C3E50"]

  "Governance";

  node [shape = oval, color = "#E31A1C"]

  "Process by which governments\nare selected, monitored, and\nreplaced";
  "Capacity of the government to\neffectively formulate and implement\nsound policies";
  "Respect of citizens and the state\nfor the institutions that govern economic and\nsocial interactions among them";

  node [shape = oval, color = "#18BC9C"]

  "Voice and\nAccountability";
  "Political Stability and\nAbsence of Violence/Terrorism";
  "Government\neffectiveness";
  "Regulatory\nquality";
  "Rule of law";
  "Control of\ncorruption";

  "Governance" -> "Process by which governments\nare selected, monitored, and\nreplaced";
  "Governance" -> "Capacity of the government to\neffectively formulate and implement\nsound policies";
  "Governance" -> "Respect of citizens and the state\nfor the institutions that govern economic and\nsocial interactions among them";

  "Process by which governments\nare selected, monitored, and\nreplaced" -> "Voice and\nAccountability";
  "Process by which governments\nare selected, monitored, and\nreplaced" -> "Political Stability and\nAbsence of Violence/Terrorism";
  "Capacity of the government to\neffectively formulate and implement\nsound policies" -> "Government\neffectiveness";
  "Capacity of the government to\neffectively formulate and implement\nsound policies" -> "Regulatory\nquality";
  "Respect of citizens and the state\nfor the institutions that govern economic and\nsocial interactions among them" -> "Rule of law";
  "Respect of citizens and the state\nfor the institutions that govern economic and\nsocial interactions among them" -> "Control of\ncorruption";
}
```

## 

-   Where can I find data?

    -   Worldwide Governance Indicators (WGI) - WorldBank

        -   [https://databank.worldbank.org/source/worldwide-governance-indicators](https://databank.worldbank.org/source/worldwide-governance-indicators)

-   What is the methodology?

    -   [https://info.worldbank.org/governance/wgi/](https://info.worldbank.org/governance/wgi/) \> Documentation

## 

```{r}
#| label: fig-governance-dimensions-col
#| fig-cap: Governance dimensions for Colombia
       
initial_year <- 1996
last_year <- 2022
middle_year <- ceiling((last_year - initial_year) / 2) + initial_year

# Percent Rank indicators
wgi_col <- wbstats::wb_data(indicator = c(
  # Voice and accountability
  "VA.PER.RNK",
  # Political Stability and Absence of Violence/Terrorism
  "PV.PER.RNK",
  # Government effectiveness
  "GE.PER.RNK",
  # Regulatory quality
  "RQ.PER.RNK",
  # Rule of law
  "RL.PER.RNK",
  # Control of Corruption
  "CC.PER.RNK"),
  country = "COL",
  start_date = initial_year,
  end_date = last_year,
  return_wide = FALSE) |>
  filter(date %in% c(initial_year, middle_year, last_year)) |>
  mutate(indicator = str_replace(string = indicator,
                                 pattern = ": Percentile Rank",
                                 replacement = ""),
         indicator = ifelse(test = indicator == "Political Stability and Absence of Violence/Terrorism",
                            yes = "Political Stability and Absence\nof Violence/Terrorism",
                            no = indicator),
         indicator = fct_relevel(indicator,
                                 "Voice and Accountability",
                                 "Political Stability and Absence\nof Violence/Terrorism",
                                 "Government Effectiveness",
                                 "Regulatory Quality",
                                 "Rule of Law",
                                 "Control of Corruption") |>
           fct_rev(),
         date = as.factor(date) |> fct_rev())

wgi_col |>
  ggplot(aes(y = indicator, x = value, fill = date)) +
  geom_col(position = "dodge", color = "black") +
  geom_text(aes(label = round(value, digits = 2)),
            position = position_dodge(width = 1),
            hjust = -0.2) +
  scale_fill_tq() +
  guides(fill = guide_legend(reverse = TRUE)) +
  expand_limits(x = c(0, 70)) +
  labs(x = NULL,
       y = NULL,
       fill = NULL,
       subtitle = str_glue("All indicators are expressed as percentile ranks
                           Percentile rank indicates the country's rank among all countries covered by the aggregate indicator,
                           with 0 corresponding to lowest rank, and 100 to highest rank"),
       caption = str_glue("Source: Worldwide Governance Indicators (WGI) - World Bank
                          Last update: {unique(wgi_col$last_updated)}")) +
  theme(panel.border    = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        legend.position   = "bottom",
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"))
```

# Crime

## 

```{r}
#| label: fig-intentional-homicides-col 
#| fig-cap: Intentional homicides (per 100,000 people)

initial_year <- 1990
last_year <- 2021

homicide <- wbstats::wb_data(# Intentional homicides (per 100,000 people)
  indicator = "VC.IHR.PSRC.P5", 
  start_date = initial_year, 
  end_date = last_year)

homicide_pivot_wider <- homicide |> 
  filter(!is.na(VC.IHR.PSRC.P5), date %in% c(initial_year, last_year)) |> 
  select(iso2c, iso3c, country, date, VC.IHR.PSRC.P5) |> 
  pivot_wider(names_from = date, values_from = VC.IHR.PSRC.P5) |> 
  drop_na()

names <- names(homicide_pivot_wider)

homicide_pivot_wider |> 
  # Check this notation in
  ## https://ggplot2.tidyverse.org/articles/ggplot2-in-packages.html
  ### In particular the case about: 
  #### You have the column name as a character vector 
  ggplot(aes(x = .data[[names[4]]], y = .data[[names[5]]]))  + 
  geom_abline(slope = 1, color = 'black') +
  geom_point(shape = 21, 
             size = 4,
             fill = "red") + 
  geom_label_repel(data = homicide_pivot_wider |> 
                     filter(country == "Colombia"),
                   aes(label = country), 
                   seed = 1234,                    
                   box.padding = 1) + 
  scale_fill_tq() +
  labs(fill = NULL, 
       title = str_glue("Period: {initial_year} vs {last_year}"),
       subtitle = str_glue("Number of countries: {nrow(homicide_pivot_wider)}"),
       caption = str_glue("Source: World Development Indicators - World Bank
                          Last update: {unique(homicide$last_updated)}")) + 
  theme(panel.border    = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        legend.position   = "bottom",
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"))
```

## 

```{r}
#| label: fig-intentional-homicides-female-col 
#| fig-cap: Intentional homicides, female (per 100,000 female)

initial_year <- 1990
last_year <- 2021

homicide_female <- wbstats::wb_data(
  # Intentional homicides, female (per 100,000 female)
  indicator = "VC.IHR.PSRC.FE.P5", 
  start_date = initial_year, 
  end_date = last_year)

homicide_female_pivot_wider <- homicide_female |> 
  filter(!is.na(VC.IHR.PSRC.FE.P5), date %in% c(initial_year, last_year)) |> 
  select(iso2c, iso3c, country, date, VC.IHR.PSRC.FE.P5) |> 
  pivot_wider(names_from = date, values_from = VC.IHR.PSRC.FE.P5) |> 
  drop_na()

names_female <- names(homicide_female_pivot_wider) 

homicide_female_pivot_wider |> 
  # Check this notation in
  ## https://ggplot2.tidyverse.org/articles/ggplot2-in-packages.html
  ### In particular the case about: 
  #### You have the column name as a character vector 
  ggplot(aes(x = .data[[names_female[4]]], y = .data[[names_female[5]]])) + 
  geom_abline(slope = 1, color = 'black') +
  geom_point(shape = 21, 
             size = 4,
             fill = "red") + 
  geom_label_repel(data = homicide_female_pivot_wider |> 
                     filter(country == "Colombia"),
                   aes(label = country), 
                   seed = 1234,                    
                   box.padding = 1) + 
  scale_fill_tq() +
  labs(fill = NULL, 
       title = str_glue("Period: {initial_year} vs {last_year}"),
       subtitle = str_glue("Number of countries: {nrow(homicide_female_pivot_wider)}"),
       caption = str_glue("Source: World Development Indicators - World Bank
                          Last update: {unique(homicide_female$last_updated)}")) + 
  theme(panel.border    = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        legend.position   = "bottom",
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"))
```

## 

```{r}
#| label: fig-intentional-homicides-male-col 
#| fig-cap: Intentional homicides, male (per 100,000 male)

initial_year <- 1990
last_year <- 2021

homicide_male <- wbstats::wb_data(
  # Intentional homicides, male (per 100,000 male)
  indicator = "VC.IHR.PSRC.MA.P5",
  start_date = initial_year, 
  end_date = last_year)

homicide_male_pivot_wider <- homicide_male |> 
  filter(!is.na(VC.IHR.PSRC.MA.P5), date %in% c(initial_year, last_year)) |> 
  select(iso2c, iso3c, country, date, VC.IHR.PSRC.MA.P5) |> 
  pivot_wider(names_from = date, values_from = VC.IHR.PSRC.MA.P5) |> 
  drop_na()

names_male <- names(homicide_male_pivot_wider) 

homicide_male_pivot_wider |> 
  # Check this notation in
  ## https://ggplot2.tidyverse.org/articles/ggplot2-in-packages.html
  ### In particular the case about: 
  #### You have the column name as a character vector 
  ggplot(aes(x = .data[[names_male[4]]], y = .data[[names_male[5]]])) + 
  geom_abline(slope = 1, color = 'black') +
  geom_point(shape = 21, 
             size = 4,
             fill = "red") + 
  geom_label_repel(data = homicide_male_pivot_wider |> 
                     filter(country == "Colombia"),
                   aes(label = country), 
                   seed = 1234,                    
                   box.padding = 1) + 
  scale_fill_tq() +
  labs(fill = NULL, 
       title = str_glue("Period: {initial_year} vs {last_year}"),
       subtitle = str_glue("Number of countries: {nrow(homicide_male_pivot_wider)}"),
       caption = str_glue("Source: World Development Indicators - World Bank
                          Last update: {unique(homicide_male$last_updated)}")) + 
  theme(panel.border    = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        legend.position   = "bottom",
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"))
```

##

```{r}
#| label: fig-map-homicides-col 
#| fig-cap: Intentional homicides (per 100,000 people) by department

crime_col_2023 <- read_rds(file = "000_data/004_crime_col_2023.rds")

crime_col_2023 |> 
  ggplot(aes(fill = homicidio_per)) + 
  geom_sf(aes(geometry = geometry),
          color = "black") + 
  scale_fill_gradient(low = "#1F78B4", 
                      high = "#E31A1C") +
  labs(fill = NULL,
       subtitle = str_glue("Year: {unique(crime_col_2023$year)}"),
       caption = str_glue("Source 1: Dirección de Investigación Criminal e INTERPOL - Policía Nacional
                          Last update 1: 2024-05-01
                          Source 2: Serie departamental de población por área, para el periodo 2020-2050 
                          con base en el CNPV 2018
                          Last update 2: 2023-03-04")) +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        legend.position   = "right",
        plot.subtitle     = element_text(face = "bold"),
        axis.text         = element_blank(),
        axis.ticks        = element_blank())
```

##

```{r}
#| label: fig-map-theft-from-person-col 
#| fig-cap: Theft from person[^1] (per 100,000 people) by department

crime_col_2023 |> 
  ggplot(aes(fill = hurto_personas_per)) + 
  geom_sf(aes(geometry = geometry),
          color = "black") + 
  scale_fill_gradient(low = "#1F78B4", 
                      high = "#E31A1C") +
  labs(fill = NULL,
       subtitle = str_glue("Year: {unique(crime_col_2023$year)}"),
       caption = str_glue("Source 1: Dirección de Investigación Criminal e INTERPOL - Policía Nacional
                          Last update 1: 2024-05-01
                          Source 2: Serie departamental de población por área, para el periodo 2020-2050 
                          con base en el CNPV 2018
                          Last update 2: 2023-03-04")) +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        legend.position   = "right",
        plot.subtitle     = element_text(face = "bold"),
        axis.text         = element_blank(),
        axis.ticks        = element_blank())
```

[^1]: *Hurto a personas* in spanish

##

```{r}
#| label: fig-map-theft-from-commercial-establishments-col 
#| fig-cap: Theft from commercial establishments[^2] (per 100,000 people) by department

crime_col_2023 |> 
  ggplot(aes(fill = hurto_comerciales_per)) + 
  geom_sf(aes(geometry = geometry),
          color = "black") + 
  scale_fill_gradient(low = "#1F78B4", 
                      high = "#E31A1C") +
  labs(fill = NULL,
       subtitle = str_glue("Year: {unique(crime_col_2023$year)}"),
       caption = str_glue("Source 1: Dirección de Investigación Criminal e INTERPOL - Policía Nacional
                          Last update 1: 2024-05-01
                          Source 2: Serie departamental de población por área, para el periodo 2020-2050 
                          con base en el CNPV 2018
                          Last update 2: 2023-03-04")) +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        legend.position   = "right",
        plot.subtitle     = element_text(face = "bold"),
        axis.text         = element_blank(),
        axis.ticks        = element_blank())
```

[^2]: *Hurto a comercio* in spanish

##

```{r}
#| label: fig-map-robbery-financial-institutions-col 
#| fig-cap: Robbery of Financial Institutions[^3] (per 100,000 people) by department

crime_col_2023 |> 
  ggplot(aes(fill = hurto_financieras_per)) + 
  geom_sf(aes(geometry = geometry),
          color = "black") + 
  scale_fill_gradient(low = "#1F78B4", 
                      high = "#E31A1C") +
  labs(fill = NULL,
       subtitle = str_glue("Year: {unique(crime_col_2023$year)}"),
       caption = str_glue("Source 1: Dirección de Investigación Criminal e INTERPOL - Policía Nacional
                          Last update 1: 2024-05-01
                          Source 2: Serie departamental de población por área, para el periodo 2020-2050 
                          con base en el CNPV 2018
                          Last update 2: 2023-03-04")) +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        legend.position   = "right",
        plot.subtitle     = element_text(face = "bold"),
        axis.text         = element_blank(),
        axis.ticks        = element_blank())
```

[^3]: *Hurto a entidades financieras* in spanish

##

```{r}
#| label: fig-map-kidnapping-col 
#| fig-cap: Kidnapping (per 100,000 people) by department

crime_col_2023 |> 
  ggplot(aes(fill = secuestro_per)) + 
  geom_sf(aes(geometry = geometry),
          color = "black") + 
  scale_fill_gradient(low = "#1F78B4", 
                      high = "#E31A1C") +
  labs(fill = NULL,
       subtitle = str_glue("Year: {unique(crime_col_2023$year)}"),
       caption = str_glue("Source 1: Dirección de Investigación Criminal e INTERPOL - Policía Nacional
                          Last update 1: 2024-05-01
                          Source 2: Serie departamental de población por área, para el periodo 2020-2050 
                          con base en el CNPV 2018
                          Last update 2: 2023-03-04")) +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        legend.position   = "right",
        plot.subtitle     = element_text(face = "bold"),
        axis.text         = element_blank(),
        axis.ticks        = element_blank())  
```

# Acknowledgments

## 

-   To my family that supports me

-   To the taxpayers of Colombia and the [**UMNG students**](https://www.umng.edu.co/estudiante) who pay my salary

-   To the [**Business Science**](https://www.business-science.io/) and [**R4DS Online Learning**](https://www.rfordatasci.com/) communities where I learn [**R**](https://www.r-project.org/about.html) and [**$\pi$-thon**](https://www.python.org/about/)

-   To the [**R Core Team**](https://www.r-project.org/contributors.html), the creators of [**RStudio IDE**](https://rstudio.com/products/rstudio/), [**Quarto**](https://quarto.org/) and the authors and maintainers of the packages [**tidyverse**](https://CRAN.R-project.org/package=tidyverse), [**wbstats**](https://CRAN.R-project.org/package=wbstats), [**tidyquant**](https://CRAN.R-project.org/package=tidyquant), [**janitor**](https://CRAN.R-project.org/package=janitor), [**ggrepel**](https://CRAN.R-project.org/package=ggrepel), [**sf**]( https://CRAN.R-project.org/package=sf) and [**tinytex**](https://CRAN.R-project.org/package=tinytex) for allowing me to access these tools without paying for a license

-   To the [**Linux kernel community**](https://www.kernel.org/category/about.html) for allowing me the possibility to use some [**Linux distributions**](https://static.lwn.net/Distributions/) as my main [**OS**](https://en.wikipedia.org/wiki/Operating_system) without paying for a license

# References {.allowframebreaks}
