---
title: Economy and institutions
subtitle: Session 1
author:
  - name: 
      given: Luis F.
      family: Gómez L.
    orcid: 0000-0002-2877-9015
    email: luis.gomezl@unimilitar.edu.co
    affiliations:
      - name: Distance Learning Faculty
toc: true
toc-depth: 1
slide-level: 2
format:
  presentation-umng-revealjs:
    logo-vertical-strip: 000_logos/logo-white-vertical.png 
    embed-resources: true
    mermaid-format: png
    reference-location: document
date: last-modified
date-format: "D MMMM, YYYY"
knitr: 
  opts_chunk: 
    echo: false
    warning: false
    fig.align: center
    out.width: 85%
lang: en
bibliography: intro_econ.bib
csl: apa-7.csl
editor: source
---

```{r}
#| label: libraries

library(tidyverse)
library(wbstats)
library(ggrepel)

# See 004_script_crime_col.R
# library(janitor)
# library(sf)
# library(readxl)
```

```{r}
#| label: palette

umng_palette <- c(
  "#043074",
  "#fdc600",
  "#0d5e30",
  "#ee2a24",
  "#fc6700",
  "#00b3f0",
  "#6e3c1a",
  "#f8941c",
  "#8e44ad",
  "#2c3e50",
  "#16a085",
  "#c0392b",
  "#e91e63"
)
```

# Please Read Me

##

-   Check the message **Welcome greeting** published in Announcements forum.

-   Dear student please edit your profile uploading a photo where your face is clearly visible.

-   If you want to participate, please fill out the following survey: **Primer corte 30% > Learning Activities > Tu opinión sobre la economía colombiana**

-   This presentation is based on [@cardenas_introduccion_2020, Chapter 4]

##

- **Purpose**

    - Analyze the role of institutions in the economy

# Definition of institutions

## 

-   **"Institutions are the rules of the game in a society or, more formally, are the humanly devised constraints that shape human interaction"** [@north_institutions_1990, p. 3]

-   Institutions can be informal or formal [@north_institutions_1991]

    -   **Informal**: sanctions, taboos, customs, traditions, and codes of conduct

    -   **Formal**: constitutions, laws, property right

# Measuring formal institutions

## {.smaller} 

-   Governance

    -   Set of traditions and institutions by which authority in a country is exercised where it includes [@kaufmann_worldwide_2011, p. 222]:

        -   The process by which governments are selected, monitored and replaced
        -   The capacity of the government to effectively formulate and implement sound policies
        -   The respect of citizens and the state for the institutions that govern economic and social interactions among them

## 

```{mermaid}
%%| label: fig-governance-indicators
%%| fig-cap: Governance dimensions [@kaufmann_worldwide_2011, p. 223]

flowchart TB
    A(Governance) --> A1("`Process by which 
    governments are 
    selected, monitored, 
    and replaced`")
    A --> A2("`Capacity of the 
    government to 
    effectively formulate 
    and implement 
    sound policies`")
    A --> A3("`Respect of citizens 
    and the state for 
    the institutions that 
    govern economic and 
    social interactions 
    among them`")
    
    A1 --> A11("`Voice and 
    accountability`")
    A1 --> A12("`Political Stability 
    and Absence of 
    Violence/Terrorism`")
    
    A2 --> A21("`Government 
    effectiveness`")
    A2 --> A22("`Regulatory 
    Quality`")

    A3 --> A31("`Rule of 
    law`")
    A3 --> A32("`Control of
    Corruption`")

    classDef UmngNode fill:#fdc600,stroke:#043074,stroke-width:2px;
    class A,A1,A11,A12,A2,A21,A22,A3,A31,A32 UmngNode;
```

## 

-   Where can I find data?

    -   Worldwide Governance Indicators (WGI) - WorldBank

        -   [https://databank.worldbank.org/source/worldwide-governance-indicators](https://databank.worldbank.org/source/worldwide-governance-indicators)

-   What is the methodology?

    -   [https://www.worldbank.org/en/publication/worldwide-governance-indicators](https://www.worldbank.org/en/publication/worldwide-governance-indicators) \> Documentation
    
##

```{r}
# Helpers
initial_year <- 1996
last_year <- 2023
middle_year <- ceiling((last_year - initial_year) / 2) + initial_year

# Import
## Percent Rank indicators
wgi_col <- wb_data(
  indicator = c(
    # Voice and accountability
    "VA.PER.RNK",
    # Political Stability and Absence of Violence/Terrorism
    "PV.PER.RNK",
    # Government effectiveness
    "GE.PER.RNK",
    # Regulatory quality
    "RQ.PER.RNK",
    # Rule of law
    "RL.PER.RNK",
    # Control of Corruption
    "CC.PER.RNK"
  ),
  country = "COL",
  start_date = initial_year,
  end_date = last_year,
  return_wide = FALSE
)

# Tidy
wgi_col_clean <- wgi_col |>
  filter(date %in% c(initial_year, middle_year, last_year)) |>
  mutate(
    indicator = str_replace(
      string = indicator,
      pattern = ": Percentile Rank",
      replacement = ""
    ),
    indicator = ifelse(
      test = indicator ==
        "Political Stability and Absence of Violence/Terrorism",
      yes = "Political Stability and Absence\nof Violence/Terrorism",
      no = indicator
    ),
    indicator = fct_relevel(
      indicator,
      "Voice and Accountability",
      "Political Stability and Absence\nof Violence/Terrorism",
      "Government Effectiveness",
      "Regulatory Quality",
      "Rule of Law",
      "Control of Corruption"
    ) |>
      fct_rev(),
    date = as.factor(date) |> fct_rev()
  )
```

```{r}
#| label: fig-governance-dimensions-col
#| fig-cap: Governance dimensions for Colombia

# Visualization
wgi_col_clean |>
  ggplot(aes(y = indicator, x = value, fill = date)) +
  geom_col(
    position = "dodge",
    color = "black"
  ) +
  geom_text(
    aes(label = round(value, digits = 2)),
    position = position_dodge(width = 1),
    hjust = -0.2
  ) +
  scale_fill_manual(values = umng_palette[1:3]) +
  guides(fill = guide_legend(reverse = TRUE)) +
  expand_limits(x = c(0, 70)) +
  labs(
    x = NULL,
    y = NULL,
    fill = NULL,
    subtitle = str_glue(
      "All indicators are expressed as percentile ranks
                           Percentile rank indicates the country's rank among all countries covered by the aggregate indicator,
                           with 0 corresponding to lowest rank, and 100 to highest rank"
    ),
    caption = str_glue(
      "Source: Worldwide Governance Indicators (WGI) - World Bank
                          Last update: {unique(wgi_col_clean$last_updated)}"
    )
  ) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10),
    legend.position = "bottom"
  )
```

# Crime

##

```{r}
# Helpers
initial_year <- 1990
last_year <- 2023

# Import
homicide <- wb_data(
  # Intentional homicides (per 100,000 people)
  indicator = "VC.IHR.PSRC.P5",
  start_date = initial_year,
  end_date = last_year
)

# Tidy
homicide_pivot_wider <- homicide |>
  filter(!is.na(VC.IHR.PSRC.P5), date %in% c(initial_year, last_year)) |>
  select(iso2c, iso3c, country, date, VC.IHR.PSRC.P5) |>
  pivot_wider(names_from = date, values_from = VC.IHR.PSRC.P5) |>
  drop_na()

names <- names(homicide_pivot_wider)
```

```{r}
#| label: fig-intentional-homicides-col
#| fig-cap: Intentional homicides (per 100,000 people)

# Visualization
homicide_pivot_wider |>
  # Check this notation in
  ## https://ggplot2.tidyverse.org/articles/ggplot2-in-packages.html
  ### In particular the case about:
  #### You have the column name as a character vector
  ggplot(aes(x = .data[[names[5]]], y = .data[[names[4]]])) +
  geom_abline(slope = 1, color = umng_palette[1]) +
  geom_point(
    shape = 21,
    size = 4,
    fill = umng_palette[2]
  ) +
  geom_label_repel(
    data = homicide_pivot_wider |>
      filter(country == "Colombia"),
    aes(label = country),
    seed = 1234,
    box.padding = 1
  ) +
  labs(
    x = str_glue(
      "Intentional homicides (per 100,000 people) in {initial_year}"
    ),
    y = str_glue("Intentional homicides (per 100,000 people) in {last_year}"),
    fill = NULL,
    title = str_glue("Period: {initial_year} vs {last_year}"),
    subtitle = str_glue("Number of countries: {nrow(homicide_pivot_wider)}"),
    caption = str_glue(
      "Source: World Development Indicators - World Bank
                          Last update: {unique(homicide$last_updated)}"
    )
  ) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10)
  )
```

##

```{r}
# Helpers
initial_year <- 1990
last_year <- 2023

# Import
homicide_female <- wb_data(
  # Intentional homicides, female (per 100,000 female)
  indicator = "VC.IHR.PSRC.FE.P5",
  start_date = initial_year,
  end_date = last_year
)

# Tidy
homicide_female_pivot_wider <- homicide_female |>
  filter(!is.na(VC.IHR.PSRC.FE.P5), date %in% c(initial_year, last_year)) |>
  select(iso2c, iso3c, country, date, VC.IHR.PSRC.FE.P5) |>
  pivot_wider(names_from = date, values_from = VC.IHR.PSRC.FE.P5) |>
  drop_na()

names_female <- names(homicide_female_pivot_wider)
```

```{r}
#| label: fig-intentional-homicides-female-col
#| fig-cap: Intentional homicides, female (per 100,000 female)

# Visualization
homicide_female_pivot_wider |>
  # Check this notation in
  ## https://ggplot2.tidyverse.org/articles/ggplot2-in-packages.html
  ### In particular the case about:
  #### You have the column name as a character vector
  ggplot(aes(x = .data[[names_female[5]]], y = .data[[names_female[4]]])) +
  geom_abline(slope = 1, color = umng_palette[1]) +
  geom_point(shape = 21, size = 4, fill = umng_palette[2]) +
  geom_label_repel(
    data = homicide_female_pivot_wider |>
      filter(country == "Colombia"),
    aes(label = country),
    seed = 1234,
    box.padding = 1
  ) +
  labs(
    x = str_glue(
      "Intentional homicides female (per 100,000 people) in {initial_year}"
    ),
    y = str_glue(
      "Intentional homicides female (per 100,000 people) in {last_year}"
    ),
    fill = NULL,
    title = str_glue("Period: {initial_year} vs {last_year}"),
    subtitle = str_glue(
      "Number of countries: {nrow(homicide_female_pivot_wider)}"
    ),
    caption = str_glue(
      "Source: World Development Indicators - World Bank
                          Last update: {unique(homicide_female$last_updated)}"
    )
  ) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10)
  )
```

##

```{r}
# Helpers
initial_year <- 1990
last_year <- 2023

# Import
homicide_male <- wb_data(
  # Intentional homicides, male (per 100,000 male)
  indicator = "VC.IHR.PSRC.MA.P5",
  start_date = initial_year,
  end_date = last_year
)

# Tidy
homicide_male_pivot_wider <- homicide_male |>
  filter(!is.na(VC.IHR.PSRC.MA.P5), date %in% c(initial_year, last_year)) |>
  select(iso2c, iso3c, country, date, VC.IHR.PSRC.MA.P5) |>
  pivot_wider(names_from = date, values_from = VC.IHR.PSRC.MA.P5) |>
  drop_na()

names_male <- names(homicide_male_pivot_wider)
```

```{r}
#| label: fig-intentional-homicides-male-col
#| fig-cap: Intentional homicides, male (per 100,000 male)

# Visualization
homicide_male_pivot_wider |>
  # Check this notation in
  ## https://ggplot2.tidyverse.org/articles/ggplot2-in-packages.html
  ### In particular the case about:
  #### You have the column name as a character vector
  ggplot(aes(x = .data[[names_male[5]]], y = .data[[names_male[4]]])) +
  geom_abline(slope = 1, color = umng_palette[1]) +
  geom_point(shape = 21, size = 4, fill = umng_palette[2]) +
  geom_label_repel(
    data = homicide_male_pivot_wider |>
      filter(country == "Colombia"),
    aes(label = country),
    seed = 1234,
    box.padding = 1
  ) +
  labs(
    x = str_glue(
      "Intentional homicides male (per 100,000 people) in {initial_year}"
    ),
    y = str_glue(
      "Intentional homicides male (per 100,000 people) in {last_year}"
    ),
    fill = NULL,
    title = str_glue("Period: {initial_year} vs {last_year}"),
    subtitle = str_glue(
      "Number of countries: {nrow(homicide_male_pivot_wider)}"
    ),
    caption = str_glue(
      "Source: World Development Indicators - World Bank
                          Last update: {unique(homicide_male$last_updated)}"
    )
  ) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10)
  )
```

##

- In Colombia, the **Observatorio de Derechos Humanos y Defensa Nacional** is responsible for producing, consolidating, and analyzing statistical data related to defense and security.  

    - For detailed information on definitions and indicators regarding this topic, please refer to [@ministerio_de_defensa_nacional_metodologipara_2024].
    
<!-- See https://www.mindefensa.gov.co/ > 
         Defensa y Seguridad > 
         Datos y cifras >
         Documentos Metodológicos -->
         
## {.nostretch}

```{r}
# Check 000_scripts/004_script_crime_col2.R

# Import
crime_col <- read_rds(file = "000_data/004_crime_col_2025.rds")

# Helpers
last_update_pol_nal <- ymd("2026-01-21")
last_update_pob <- ymd("2025-08-08")
crime_year <- 2025
```

```{r}
#| label: fig-map-homicides-col
#| fig-cap: Intentional homicides (per 100,000 people) by department
#| fig-height: 6

# Visualization
crime_col |>
  ggplot(aes(fill = homicidio_per)) +
  geom_sf(aes(geometry = geometry), color = "black") +
  scale_fill_gradient(low = umng_palette[2], high = umng_palette[1]) +
  labs(
    fill = NULL,
    subtitle = str_glue("Year: {crime_year}"),
    caption = str_glue(
      "Source 1: Ministerio de Defensa y Dirección de Investigación Criminal e INTERPOL - Policía Nacional
                          Last update 1: {last_update_pol_nal}
                          Source 2: Serie departamental de población por área, para el periodo 2018-2050 
                          con base en el CNPV 2018
                          Last update 2: {last_update_pob}"
    )
  ) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )
```

## {.nostretch}

```{r}
#| label: fig-map-theft-from-person-col
#| fig-cap: Theft from person[^1] (per 100,000 people) by department
#| fig-height: 6

# Visualization
crime_col |>
  ggplot(aes(fill = hurto_personas_per)) +
  geom_sf(aes(geometry = geometry), color = "black") +
  scale_fill_gradient(low = umng_palette[2], high = umng_palette[1]) +
  labs(
    fill = NULL,
    subtitle = str_glue("Year: {crime_year}"),
    caption = str_glue(
      "Source 1: Ministerio de Defensa y Dirección de Investigación Criminal e INTERPOL - Policía Nacional
                          Last update 1: {last_update_pol_nal}
                          Source 2: Serie departamental de población por área, para el periodo 2018-2050 
                          con base en el CNPV 2018
                          Last update 2: {last_update_pob}"
    )
  ) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )
```

[^1]: *Hurto a personas* in spanish

## {.nostretch}

```{r}
#| label: fig-map-theft-from-commercial-establishments-col
#| fig-cap: Theft from commercial establishments[^2] (per 100,000 people) by department
#| fig-height: 6

# Visualization
crime_col |>
  ggplot(aes(fill = hurto_comerciales_per)) +
  geom_sf(aes(geometry = geometry), color = "black") +
  scale_fill_gradient(low = umng_palette[2], high = umng_palette[1]) +
  labs(
    fill = NULL,
    subtitle = str_glue("Year: {crime_year}"),
    caption = str_glue(
      "Source 1: Ministerio de Defensa y Dirección de Investigación Criminal e INTERPOL - Policía Nacional
                          Last update 1: {last_update_pol_nal}
                          Source 2: Serie departamental de población por área, para el periodo 2018-2050 
                          con base en el CNPV 2018
                          Last update 2: {last_update_pob}"
    )
  ) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )
```

[^2]: *Hurto a comercio* in spanish

## {.nostretch}

```{r}
#| label: fig-map-robbery-financial-institutions-col
#| fig-cap: Robbery of Financial Institutions[^3] (per 100,000 people) by department
#| fig-height: 6

# Visualization
crime_col |>
  ggplot(aes(fill = hurto_financieras_per)) +
  geom_sf(aes(geometry = geometry), color = "black") +
  scale_fill_gradient(low = umng_palette[2], high = umng_palette[1]) +
  labs(
    fill = NULL,
    subtitle = str_glue("Year: {crime_year}"),
    caption = str_glue(
      "Source 1: Ministerio de Defensa y Dirección de Investigación Criminal e INTERPOL - Policía Nacional
                          Last update 1: {last_update_pol_nal}
                          Source 2: Serie departamental de población por área, para el periodo 2018-2050 
                          con base en el CNPV 2018
                          Last update 2: {last_update_pob}"
    )
  ) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )
```

[^3]: *Hurto a entidades financieras* in spanish

## {.nostretch}

```{r}
#| label: fig-map-kidnapping-col
#| fig-cap: Kidnapping (per 100,000 people) by department
#| fig-height: 6

# Visualization
crime_col |>
  ggplot(aes(fill = secuestro_per)) +
  geom_sf(aes(geometry = geometry), color = "black") +
  scale_fill_gradient(low = umng_palette[2], high = umng_palette[1]) +
  labs(
    fill = NULL,
    subtitle = str_glue("Year: {crime_year}"),
    caption = str_glue(
      "Source 1: Ministerio de Defensa y Dirección de Investigación Criminal e INTERPOL - Policía Nacional
                          Last update 1: {last_update_pol_nal}
                          Source 2: Serie departamental de población por área, para el periodo 2018-2050 
                          con base en el CNPV 2018
                          Last update 2: {last_update_pob}"
    )
  ) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )
```

# Acknowledgments

##

-   To my family that supports me

-   To the taxpayers of Colombia and the [**UMNG students**](https://www.umng.edu.co/estudiante) who pay my salary

-   To the [**Business Science**](https://www.business-science.io/) and [**R4DS Online Learning**](https://www.rfordatasci.com/) communities where I learn [**R**](https://www.r-project.org/about.html) and [**$\pi$-thon**](https://www.python.org/about/)

-   To the [**R Core Team**](https://www.r-project.org/contributors.html), the creators of [**RStudio IDE**](https://posit.co/download/rstudio-desktop/), [**Positron**](https://positron.posit.co/), [**Quarto**](https://quarto.org/) and the authors and maintainers of the packages [**tidyverse**](https://CRAN.R-project.org/package=tidyverse), [**wbstats**](https://CRAN.R-project.org/package=wbstats),
[**ggrepel**](https://CRAN.R-project.org/package=ggrepel),
[**janitor**](https://CRAN.R-project.org/package=janitor), 
[**sf**](https://CRAN.R-project.org/package=sf),
[**readxl**](https://CRAN.R-project.org/package=readxl) and [**tinytex**](https://CRAN.R-project.org/package=tinytex) for allowing me to access these tools without paying for a license
    
# References {.allowframebreaks}
