---
title: Economy and institutions II
author: 
  - Luis Francisco Gómez López
institute: 
  - FAEDIS
date: last-modified
format: 
  beamer:
    colortheme: dolphin
    fonttheme: structurebold
    theme: AnnArbor
    link-citations: true
    colorlinks: true
    toc: true
    slide-level: 2
    section-titles: false
    include-in-header:
      - text: |
          \usepackage{bigdelim}
      - file: 000_tex_files/preamble.tex
fig-cap-location: bottom
tbl-cap-location: top
knitr: 
  opts_chunk: 
    echo: false
    warning: false
    fig-align: center
    out-width: 85%
lang: en
bibliography: intro_econ.bib
---

```{r}
#| label: libraries

library(tidyverse)
library(tidyquant)
library(janitor)
```

# Please Read Me

## 

-   Check the message **Welcome greeting** published in the News Bulletin Board.

-   Dear student please edit your profile uploading a photo where your face is clearly visible.

-   The purpose of the virtual meetings is to answer questions and not to make a summary of the study material.

-   If you want to participate, please fill out the following survey: Primer corte 30% \> Learning Activities \> Tu opinión sobre la economía colombiana

-   This presentation is based on [@cardenas_introduccion_2020, Chapter 4]

# Purpose

## 

Analyze the role of institutions in the economy

# Some informal institutions

## 

```{r}
#| label: fig-not-vote-for
#| fig-cap: Percentage of people who would never vote for

# Data
# https://www.dane.gov.co/ > Datos Abiertos Microdatos y Metadatos >
# Sociedad > Gobierno >
# Encuesta de Cultura Política - ECP - 2023  > 
## Tabla de viviendas 
## Elecciones y partidos 

viviendas <- read_csv(file = "000_data/004_tabla_de_viviendas_2023.csv")
# Personas de 18 años y más
elecciones_partidos <- read_csv(file = "000_data/004_elecciones_y_partidos_2023.csv")

viviendas_clean <- viviendas |> 
  janitor::clean_names() |> 
  select(
    # Llave vivienda
    directorio, 
    # Región a la que corresponde 
    # la vivienda
    region, 
    # Total de hogares en la 
    # vivienda?
    p70,
    # Factor de Expansión
    fex_p)

voto_por <- elecciones_partidos |> 
  janitor::clean_names() |> 
  select(
    # Llave vivienda 
    directorio,
    # Tabla en que se cargo la informacion
    ## 252. Tabla de Personas
    nro_encuesta,
    # Llave hogar
    hogar_numero,
    # Llave persona
    persona_numero,
    # 15. Votaría alguna vez por:
    ## 1. Si
    ## 2. No
    ## 3. No sabe/No responde
    ### a. Una mujer
    p2009s1,
    ### b. Un(a) indígena
    p2009s2,
    ### c. Una persona en condición de discapacidad
    p2009s3,
    ### d. Un negro(a), mulato(a), afrocolombiano(a) o afrodescendiente
    p2009s4,
    ### e. Un(a) campesino(a)
    p2009s8,
    ### f. Una persona LGTBI o con otras opciones sexuales diversas
    p2009s9,
    ### g. Una persona joven 
    p2009s10)

voto_por_merge <- voto_por  |> 
  left_join(y = viviendas_clean, 
            by = join_by(directorio))

total <- sum(voto_por_merge$fex_p)

voto_por_clean <- voto_por_merge |> 
  mutate(across(.cols = starts_with(match = "p2"), 
                .fns = ~ case_when(
                  .x == 1  ~ "Si",
                  .x == 2  ~ "No",
                  .x == 99 ~ "No sabe/No informa"
                  ), 
                .names = "{.col}" 
                ), 
         .keep = "unused" 
         ) |> 
  pivot_longer(cols = starts_with(match = "p2"), 
               names_to  = "group",
               values_to = "response") |> 
  filter(response == "No") |> 
  group_by(group) |> 
  summarize(n = sum(fex_p)) |> 
  ungroup() |> 
  mutate(pct_n = n/{{total}}) |> 
  mutate(group_label = case_when(
        group == "p2009s1"  ~ "Una mujer",
        group == "p2009s2"  ~ "Un(a) indígena",
        group == "p2009s3"  ~ str_glue("Una persona en condición de 
                                       discapacidad"),
        group == "p2009s4"  ~ str_glue("Un negro(a), mulato(a), 
                                       afrocolombiano(a) o afrodescendiente"),
        group == "p2009s8"  ~ "Un(a) campesino(a)",
        group == "p2009s9"  ~ str_glue("Una persona LGTBI o con 
                                       otras opciones sexuales diversas"),
        group == "p2009s10" ~ "Una persona joven"
    )
  ) |> 
  mutate(pct_n_label = scales::number(x = pct_n, 
                                      accuracy = 0.01,  
                                      scale = 100, 
                                      suffix = "%")) |> 
  mutate(group_label = fct_reorder(.f = group_label, 
                                   .x = pct_n))

voto_por_clean |>  
  ggplot(aes(pct_n, 
             group_label)) +
  geom_point(shape = 21, 
             fill = "#E31A1C",
             size = 3) +
  geom_segment(aes(xend = 0, yend = group_label), 
               color = "#2C3E50") + 
  geom_text(aes(label = pct_n_label),
            hjust = -0.5, size = 5) + 
  scale_x_continuous(limits = c(0, 0.7)) +
  labs(x = NULL,
       y = NULL,
       subtitle = str_glue("Year: 2023
                           Data only apply for people 18 years of age and older 
                           Total target population: {scales::number(total, accuracy = 1, big.mark = ',')}"),
       caption = str_glue('Source: DANE - Encuesta de Cultura Política - ECP - 2023
                          Last update date: 2024-04-15')) + 
  theme(legend.position = "bottom",
        panel.border      = element_rect(fill = NA, 
                                         color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y       = element_text(size = 12))
```

## 

```{r}
#| label: fig-not-neighbor
#| fig-cap: Percentage of people who would never like to have a certain neighbor

# Data
# https://www.dane.gov.co/ > Datos Abiertos Microdatos y Metadatos >
# Sociedad > Gobierno >
# Encuesta de Cultura Política - ECP - 2023  > 
## Tabla de viviendas
### Already upload in another code chunk
## Elecciones y partidos 

# Personas de 18 años y más
capital_social <- read_csv(file = "000_data/004_componente_capital_social.csv")

tener_vecino <- capital_social |> 
  janitor::clean_names() |> 
  select(
    # Llave vivienda 
    directorio,
    # Tabla en que se cargo la informacion
    ## 252. Tabla de Personas
    nro_encuesta,
    # Llave hogar
    hogar_numero,
    # Llave persona
    persona_numero,
    # 43. De las siguientes personas, a quién 
    # no quisiera tener de vecino(a):
    ### a. Personas con discapacidad 
    p2041s1,
    ### b. Personas LGTBI
    p2041s2,
    ### c. Personas de una raza o etnia diferente a la suya
    p2041s3,
    ### d. Personas en situación de pobreza 
    p2041s4,
    ### e.Víctimas de conflicto armado
    p2041s5,
    ### f. Exguerrilleros(as) 
    p2041s6,
    ### g. Exparamilitares
    p2041s7,
    ### h. Consumidores(as) de sustancias psicoactivas
    p2041s8,
    ### i. Personas en situación de prostitución
    p2041s9,
    ### j. Personas con VIH/SIDA
    p2041s10,
    ### k. Inmigrantes o trabajadores extranjeros
    p2041s11,
    ### l. Personas que profesan una religión diferente
    p2041s12,
    ### m. Ninguna de las anteriores
    #### En este caso la persona no escoge 
    #### ninguna de la a. a la l.
    p2041s13)

tener_vecino_merge <- tener_vecino |> 
  left_join(y = viviendas_clean, 
            by = join_by(directorio))

tener_vecino_clean <- tener_vecino_merge |> 
  mutate(across(.cols = starts_with(match = "p2"), 
                .fns = ~ case_when(
                  .x == 1  ~ "No",
                  ), 
                .names = "{.col}" 
                ), 
         .keep = "unused" 
         ) |> 
  pivot_longer(cols = starts_with(match = "p2"), 
               names_to  = "group",
               values_to = "response") |> 
  filter(response == "No") |> 
  group_by(group) |> 
  summarize(n = sum(fex_p)) |> 
  ungroup() |> 
  mutate(pct_n = n/{{total}}) |> 
  mutate(group_label = case_when(
        group == "p2041s1"  ~ "Personas con discapacidad",
        group == "p2041s2"  ~ "Personas LGTBI",
        group == "p2041s3"  ~ str_glue("Personas de una 
                                       raza o etnia diferente a la suya"),
        group == "p2041s4"  ~ str_glue("Personas en situación de 
                                       pobreza"),
        group == "p2041s5"  ~ "Víctimas de conflicto armado",
        group == "p2041s6"  ~ "Exguerrilleros(as)",
        group == "p2041s7"  ~ "Exparamilitares",
        group == "p2041s8"  ~ str_glue("Consumidores(as) de sustancias 
                                       psicoactivas"),
        group == "p2041s9"  ~ str_glue("Personas en situación de 
                                       prostitución"),
        group == "p2041s10"  ~ "Personas con VIH/SIDA",
        group == "p2041s11" ~ str_glue("Inmigrantes o trabajadores 
                                       extranjeros"),
        group == "p2041s12" ~ str_glue("Personas que profesan una 
                                       religión diferente")
    )
  ) |> 
  mutate(pct_n_label = scales::number(x = pct_n, 
                                      accuracy = 0.01,  
                                      scale = 100, 
                                      suffix = "%")) |> 
  mutate(group_label = fct_reorder(.f = group_label, 
                                   .x = pct_n))

tener_vecino_clean |>  
  ggplot(aes(pct_n, 
             group_label)) +
  geom_point(shape = 21, 
             fill = "#E31A1C",
             size = 3) +
  geom_segment(aes(xend = 0, yend = group_label), 
               color = "#2C3E50") + 
  geom_text(aes(label = pct_n_label),
            hjust = -0.5, size = 5) + 
  scale_x_continuous(limits = c(0, 0.7)) +
  labs(x = NULL,
       y = NULL,
       subtitle = str_glue("Year: 2023
                           Data only apply for people 18 years of age and older 
                           Total target population: {scales::number(total, accuracy = 1, big.mark = ',')}"),
       caption = str_glue('Source: DANE - Encuesta de Cultura Política - ECP - 2023
                          Last update date: 2024-04-15')) + 
  theme(legend.position = "bottom",
        panel.border      = element_rect(fill = NA, 
                                         color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y       = element_text(size = 12))
```

# Institutions and economic growth

##

- The big question: **Why are some countries much poorer than others?**

    - According to [@acemoglu_chapter_2005, pp. 389-393] the key variables are the **political institutions** and the **distribution of resources**
    
- The **political institutions** are the rules that govern incentives in politics. They determine [@acemoglu_why_2013, pp. 79-80]: 

    - How the government is chosen
    - Which part of the government has the right to do what
    - Who has power in society
    - To what ends that power can be used

##

- The **political institutions** determine the **economic institutions**

    - The **economic institutions** are [@acemoglu_why_2013, pp. 73-74]:
    
        - The rules influencing how the economy works
        - The incentives that motivate people
        
## 

- The argument by [@acemoglu_chapter_2005, pp. 389-393] in relation to how **political institutions** and the **distribution of resources** determine economic performance is as follows:

    - Economic institutions shape the incentives of key economic actors in society where they influence investments in physical and human capital and technology, and the organization of production [@acemoglu_chapter_2005, pp. 389-390].
    
$$\text{economic institutions}_t \Longrightarrow
  \begin{cases}
   \text{economic performance}_t \\
   \text{distribution of resources}_{t+1}
  \end{cases}$$

##

- The argument by [@acemoglu_chapter_2005, pp. 389-393] in relation to how **political institutions** and the **distribution of resources** determine economic performance is as follows:

    - Economic institutions are determined as collective choices of the society. However, there is no guarantee that all individuals will prefer the same set of economic institutions because they lead to different distributions of resources. So, there will typically be a conflict of interest among various individuals over the choice of economic institutions [@acemoglu_chapter_2005, p. 390].
    
    - So how are economic institutions determined? Whichever individuals have more political power are likely to secure the set of economic institutions that they prefer.
    
$$\text{political power}_t \Longrightarrow
   \text{economic institutions}_t$$

##

- There are 2 components of political power [@acemoglu_chapter_2005, pp. 390-391]:

    - **De jure (institutional)**: refers to power that originates from the political institutions in society
    
    - **De facto**: the power that derives from actual influence, irrespective of the formal institutional framework.  It can be based on factors such as wealth, control of resources, or the ability to mobilize people, among others.
    
##

- In the case of the **de jure (institutional)** power according to [@acemoglu_chapter_2005, p. 391] this aspect implies that:

$$\text{political institutions}_t \Longrightarrow
   \text{de jure political power}_t$$

##
   
- In the case of the **de facto** power according to [@acemoglu_chapter_2005, p. 391] this aspect depends on 2 sources:

    - The ability of the group of individuals to act together and avoid the incentives of some individuals to  benefit without contributing.
    
        - In the economic literature this is known as **solving the collective action problem** and the **free-rider problem**
        
    - The economic resources of the group which determine the ability to use existing political institutions and force against other individuals
    
- Because [@acemoglu_chapter_2005, p. 391] don't have a satisfactory theory about how individuals solve the **collective action problem** they focus on:

$$\text{distributions of resources}_t \Longrightarrow
   \text{de facto political power}_t$$

##

- Also [@acemoglu_chapter_2005, pp. 391-392] point out that political institutions and the distribution of resources tend to change relatively slowly but determine economic institutions and economic performance both directly and indirectly.

$$\text{political power}_t \Longrightarrow
   \text{political institutions}_{t+1}$$

##

- Putting all these pieces together [@acemoglu_chapter_2005, p. 392] describe in a simplistic schematic representation this framework in the following way:

\footnotesize

\begin{tabular}{l l l l l l l l l}
political & & de jure & \rdelim\}{9}{*} & & economic & & \ldelim\{{5}{*} & economic \\
institutions$_{\scriptscriptstyle t}$ & $\Rightarrow$ & political & & & institutions$_{\scriptscriptstyle t}$ & $\Rightarrow$ & & performance$_{\scriptscriptstyle t}$ \\
& & power$_{\scriptscriptstyle t}$ & & & & & & \\
& & & & & & & & distribution \\
& & & & $\Rightarrow$ & & & & of resources$_{\scriptscriptstyle t+1}$\\
& & & & & & & & \\
distribution & & de facto & & & political & & & \\ 
of resources$_{\scriptscriptstyle t}$ & $\Rightarrow$ & political & & & institutions$_{\scriptscriptstyle t + 1}$ & & & \\
& & power$_{\scriptscriptstyle t}$ & & & & & &
\end{tabular}

# Acknowledgments

## 

-   To my family that supports me

-   To the taxpayers of Colombia and the [**UMNG students**](https://www.umng.edu.co/estudiante) who pay my salary

-   To the [**Business Science**](https://www.business-science.io/) and [**R4DS Online Learning**](https://www.rfordatasci.com/) communities where I learn [**R**](https://www.r-project.org/about.html) and [**$\pi$-thon**](https://www.python.org/about/)

-   To the [**R Core Team**](https://www.r-project.org/contributors.html), the creators of [**RStudio IDE**](https://rstudio.com/products/rstudio/), [**Quarto**](https://quarto.org/) and the authors and maintainers of the packages  and [**tinytex**](https://CRAN.R-project.org/package=tinytex) for allowing me to access these tools without paying for a license

-   To the [**Linux kernel community**](https://www.kernel.org/category/about.html) for allowing me the possibility to use some [**Linux distributions**](https://static.lwn.net/Distributions/) as my main [**OS**](https://en.wikipedia.org/wiki/Operating_system) without paying for a license

# References {.allowframebreaks}
