---
title: "Constructing a bilateral real exchange rate"
author: "Anonymous"
date: '2021-09-19'
output:
  pdf_document: default
  html_document:
    number_sections: yes
    code_folding: hide
bibliography: intro_econ.bib
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE}
library(tidyverse)
library(tidyquant)
library(timetk)
```

# Introduction

This document expands the information that is pointed out in [@neely_constructing_2020]

# Bilateral real exchange rate

$$BRER_{ij} = \frac{NER_{ij} * P_{j}}{P_i}$$

- Territory $i$
- Territory $j$
- $BRER_{ij}$: Bilateral real exchange rate between territory $i$ and territory $j$ from the perspective of territory $i$
- $NER_{ij}$: Nominal exchange rate between territory $i$ and territory $j$ from the perspective of territory $i$

    + $NER_{ij} = \frac{Currency_i}{Currency_j}$
    
For example if $i = US$ and $j = EUA$ then the idea is to calculate the bilateral real exchange rate between the United States of America and the Euro Area

# Data

## Case study: Bilateral real exchange rate between the US and the EUA 

- Source: https://fred.stlouisfed.org/
- Variables: 

    + $NER_{US,EUA}$: U.S. / Euro Foreign Exchange Rate
        + Code: DEXUSEU
        
    + $P_{US}$: Consumer Price Index for All Urban Consumers: All Items in U.S. City Average 
        + Code: CPIAUCSL
    
    + $P_{EUA}$: Harmonized Index of Consumer Prices: All Items for Euro area (19 countries) 
        + Code: CP0000EZ19M086NEST 
        
## Import and tidy

```{r}
# Import
nex_us_eua <- tq_get(c("DEXUSEU"),
       get = "economic.data", 
       complete_cases = TRUE,
       from = "1999-01-01")

pi_us_eua <- tq_get(c("CPIAUCSL", "CP0000EZ19M086NEST"),
                     get = "economic.data", 
                     complete_cases = TRUE,
                     from = "1999-01-01")

# Tidy
nex_us_eua <- nex_us_eua %>% 
  summarize_by_time(.date_var = date, 
                    .by = "month", 
                    DEXUSEU = mean(price, na.rm = TRUE))

pi_us_eua <- pi_us_eua %>% 
  pivot_wider(id_cols = date, 
              names_from = symbol, 
              values_from = price)

data_clean <- nex_us_eua %>% 
  inner_join(y = pi_us_eua, by = "date") %>% 
  set_names(nm = c("date", "nex_us_eua", "p_us", "p_eua"))
```

## Transform

### Bilateral real exchange rate

```{r}
data_transform <- data_clean %>% 
  mutate(brer_us_eua = nex_us_eua * (p_eua / p_us))
```

### Rescale bilateral real exchange rate

```{r}
brer_us_eua_100 <- data_transform$brer_us_eua[1]

data_transform <- data_transform %>% 
  mutate(rescale_brer_us_eua = (100 * brer_us_eua) / brer_us_eua_100)
```

# Visualize

```{r}
data_transform %>% 
  ggplot(aes(x = date, y = rescale_brer_us_eua)) + 
  geom_line()
```

# References