---
title: "External sector II"
author: 
  - Luis Francisco Gómez López
institute: 
  - FAEDIS
date: last-modified
format: 
  beamer:
    colortheme: dolphin
    fonttheme: structurebold
    theme: AnnArbor
    link-citations: true
    colorlinks: true
    toc: true
    slide-level: 2
    section-titles: false
    keep-tex: true
    include-in-header:
      file: 000_tex_files/preamble.tex
fig-cap-location: bottom
tbl-cap-location: top
knitr: 
  opts_chunk: 
    echo: false
    warning: false
    fig-align: center
    out-width: 85%
lang: en
bibliography: intro_econ.bib
---

```{r}
library(tidyverse)
library(wbstats)
library(tidyquant)
library(knitr)
```

# Please Read Me

##

-   Check the message **Welcome greeting** published in the News Bulletin Board.

-   Dear student please edit your profile uploading a photo where your face is clearly visible.

-   The purpose of the virtual meetings is to answer questions and not to make a summary of the study material.

-   This presentation is based on [@cardenas_introduccion_2020, Chapter 5]

# Purpose

##

Explain the composition and determinants of Colombian foreign trade

# The concept of comparative advantage

##

-   **Absolute advantage**: the ability of a party (individual, company, country) to produce a product more efficiently than any other party

    -   A party should concentrate in producing a good in which it has an absolute advantage

    -   ¿What happend when you don't have an absolute advantage?

-   **Comparative advantage**: the ability of a party (individual, company, country) to produce a product at a lower relative opportunity cost

    -   Under certain conditions, even if a party doesn't have an **absolute advantage**, trade with other agent can be benefical for both of them

##

-   Example of gains of trade using the concept of **comparative advantage** [@ridley_when_2010]

    -   **Agent 1** produce 1 spear in 4 hours and 1 axe in 3 hours.

        -   To produce both products **agent 1** will need to allocate 7 hours

    -   **Agent 2** produce 1 spear in 1 hour and 1 axe in 2 hours.

        -   **Agent 2** has the absolute advantage in the prodution of spears and axes.
        -   To produce both products **agent 2** will need to allocate 3 hours

    -   ¿It will be benefical for agent 1 and 2 to trade? Yes!!! **Agent 1** can produce 2 axes in 6 hours and **agent 2** 2 spears in 2 hours. Then they can trade 1 spear for 1 axe.

        -   **Agent 1** has now 1 spear and 1 axe only working 6 hours.
        -   Furthermore **agent 2** has now 1 spear and 1 axe only working 2 hours.
        -   Each of them saves 1 hour of work in contrast with a situation in which they don't trade and produce themselves both products.

# Volume of international trade

##

```{r}
#| label: fig-imports-exports-col
#| fig-cap: Colombia imports and exports

# Data
data <- wbstats::wb_data(country   = "COL",
                         indicator = c("NE.IMP.GNFS.KN", "NE.EXP.GNFS.KN"),
                         return_wide = FALSE) |>
  filter(!is.na(value))

data |>
  select(country, date, value, indicator_id) |>
  mutate(indicator_id = case_when(
    indicator_id == "NE.IMP.GNFS.KN" ~ "Imports",
    indicator_id == "NE.EXP.GNFS.KN" ~ "Exports",
    TRUE                             ~ indicator_id)) |>

  #Plot
  ggplot() +
  geom_point(aes(x = date, y = value, fill = indicator_id),
             shape = 21,
             color = "black",
             show.legend = FALSE) +
  geom_line(aes(x     = date,
                y     = value,
                group = indicator_id,
                color = indicator_id)) +
  scale_y_continuous(labels = scales::number_format(scale    = 1e-9,
                                                    suffix   = "B",
                                                    accuracy = 1)) +
  scale_color_tq() +
  scale_fill_tq() +
  labs(x        = "Year",
       y        = "Billions (Short scale)",
       color    = "",
       title    = str_glue("Period: {min(data$date)}-{max(data$date)}"),
       subtitle = str_glue("Variables Units: constant Local Currency Units Base Year 2015
                              Imports WDI code : NE.IMP.GNFS.KN
                              Exports WDI: NE.EXP.GNFS.KN"),
       caption  = str_glue("Source: World Development Indicators - World Bank,
                              Last update: {unique(data$last_updated)}")) +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"))
```

# Harmonized System (HS)

##

-   The **Harmonized System (HS)** is a standardized numerical method of classifying traded goods that is internationally accepted and maintained by the **World Customs Organization (WCO)**

-   The HS code consists of 6-digits that are the same independent of the country:

    -   First 2 digits designate the **HS** chapter
    -   Second 2 digits designate the **HS** heading
    -   Third 2 digits designate the **HS** subheading

-   Example of **HS** code: 090111 (Coffee, not roasted, not decaffeinated)

    -   Chapter **09**: Cofee, tea, mate and spices
    -   Heading **01**: Coffee, whether or not roasted or decaffeinated; coffee husks and skins; coffee substitutes containing coffee in any proportion
    -   Subheading **11**: Coffee, not roasted, not decaffeinated

# Patterns of international trade: goods

##

![Colombian good exports **HS 4 digits (HS92) edition** year 1995 [@oec_observatory_2022]](000_data/005_col_exports_1995_hs4_92.png){#fig-colombian-goods-exports-1995 fig-alt="Colombian goods exports 1995" width="450px" height="450px"}

##

![Colombian good exports **HS 4 digits (HS92) edition** year 2022 [@oec_observatory_2022]](000_data/005_col_exports_2022_hs4_92.png){#fig-colombian-goods-exports-2022 fig-alt="Colombian goods exports 2022" width="450px" height="450px"}

##

![Colombian goods imports **HS 4 digits (HS92) edition** year 1995 [@oec_observatory_2022]](000_data/005_col_imports_1995_hs4_92.png){#fig-colombian-goods-imports-1995 fig-alt="Colombian goods imports 1995" width="450px" height="450px"}

##

![Colombian goods imports **HS 4 digits (HS92) edition** year 2022 [@oec_observatory_2022]](000_data/005_col_imports_2022_hs4_92.png){#fig-colombian-goods-imports-2022 fig-alt="Colombian goods imports 2022" width="450px" height="450px"}

# Extended Balance of Payments Services classification (EBOPS)

##

-   The EBOPS classification provides a breakdown of the Balance of Payments Trade in Services by types of services.

-   The classification thereby meets a number of user requirements, including the provision of more detailed information on Trade in services

    -   **EBOPS 2002**
    -   **EBOPS 2010**

-   For more information check out

    -   [@vereinte_nationen_manual_2002]
    -   [@united_nations_manual_2012]

# Patterns of international trade: services

##

![Colombian services exports **(EBOPS Version 2002)** year 2000 [@oec_observatory_2022]](000_data/005_col_export_ebops_version2002_2000.png){#fig-colombian-services-exports-2000 fig-alt="Colombian services exports 2000" width="280px" height="280px"}

##

![Colombian services exports **(EBOPS Version 2002)** year 2020 [@oec_observatory_2022]](000_data/005_col_export_ebops_version2002_2020.png){#fig-colombian-services-exports-2020 fig-alt="Colombian services exports 2020" width="250px" height="250px"}

##

![Colombian services imports **(EBOPS Version 2002)** year 2000 [@oec_observatory_2022]](000_data/005_col_import_ebops_version2002_2000.png){#fig-colombian-services-imports-2000 fig-alt="Colombian services imports 2000" width="280px" height="280px"}

##

![Colombian services imports **(EBOPS Version 2002)** year 2020 [@oec_observatory_2022]](000_data/005_col_import_ebops_version2002_2020.png){#fig-colombian-services-imports-2020 fig-alt="Colombian services imports 2020" width="250px" height="250px"}

# Observatory of Economic Complexity (OEC)

##

-   *"The Observatory of Economic Complexity (OEC) is an online data visualization and distribution platform focused on the geography and dynamics of economic activities"* [@oec_observatory_2022]

-   *"The OEC is currently designed and developed by Datawheel, but it began as a research project at MIT's Collective Learning group (former Macro Connections Group)"* [@oec_observatory_2022]

##

-   Tree map

    -   [https://oec.world/](https://oec.world/) \> TOOLS \> Tree map

        -   Country
        -   Product
        -   Bilateral

-   Profiles

    -   [https://oec.world/](https://oec.world/) \> PROFILES \> Countries \> Colombia

# Acknowledgments

##

-   To my family that supports me

-   To the taxpayers of Colombia and the [**UMNG students**](https://www.umng.edu.co/estudiante) who pay my salary

-   To the [**Business Science**](https://www.business-science.io/) and [**R4DS Online Learning**](https://www.rfordatasci.com/) communities where I learn [**R**](https://www.r-project.org/about.html) and [**$\pi$-thon**](https://www.python.org/about/)

-   To the [**R Core Team**](https://www.r-project.org/contributors.html), the creators of [**RStudio IDE**](https://rstudio.com/products/rstudio/), [**Quarto**](https://quarto.org/) and the authors and maintainers of the packages [**tidyverse**](https://CRAN.R-project.org/package=tidyverse), [**wbstats**](https://CRAN.R-project.org/package=wbstats), [**tidyquant**](https://CRAN.R-project.org/package=tidyquant), [**knitr**](https://CRAN.R-project.org/package=knitr), and [**tinytex**](https://CRAN.R-project.org/package=tinytex) for allowing me to access these tools without paying for a license

-   To the [**Linux kernel community**](https://www.kernel.org/category/about.html) for allowing me the possibility to use some [**Linux distributions**](https://static.lwn.net/Distributions/) as my main [**OS**](https://en.wikipedia.org/wiki/Operating_system) without paying for a license

# References {.allowframebreaks}
