---
title: External sector
subtitle: Session 2
author:
  - name: 
      given: Luis F.
      family: Gómez L.
    orcid: 0000-0002-2877-9015
    email: luis.gomezl@unimilitar.edu.co
    affiliations:
      - name: Distance Learning Faculty
toc: true
toc-depth: 1
slide-level: 2
format:
  presentation-umng-revealjs:
    logo-vertical-strip: 000_logos/logo-white-vertical.png 
    embed-resources: true
    reference-location: document
date: last-modified
date-format: "D MMMM, YYYY"
knitr: 
  opts_chunk: 
    echo: false
    warning: false
    fig.align: center
    out.width: 85%
lang: en
bibliography: intro_econ.bib
csl: apa-7.csl
editor: source
---

```{r}
#| label: libraries

library(tidyverse)
library(wbstats)
library(scales)
```

```{r}
#| label: palette

umng_palette <- c(
  "#043074",
  "#fdc600",
  "#0d5e30",
  "#ee2a24",
  "#fc6700",
  "#00b3f0",
  "#6e3c1a",
  "#f8941c",
  "#8e44ad",
  "#2c3e50",
  "#16a085",
  "#c0392b",
  "#e91e63"
)
```

# Please Read Me

##

-   Check the message **Welcome greeting** published in Announcements forum.

-   Dear student please edit your profile uploading a photo where your face is clearly visible.

-   If you want to participate, please fill out the following survey: **Primer corte 30% > Learning Activities > Tu opinión sobre la economía colombiana**

-   This presentation is based on [@cardenas_introduccion_2020, Chapter 5]

##

- **Purpose**

    - Explain the composition and determinants of Colombian foreign trade

# The concept of comparative advantage

## {.smaller}

-   **Absolute advantage**: the ability of a party (individual, company, country) to produce a product more efficiently than any other party

    -   A party should concentrate in producing a good in which it has an absolute advantage

    -   ¿What happend when you don't have an absolute advantage?

-   **Comparative advantage**: the ability of a party (individual, company, country) to produce a product at a lower relative opportunity cost

    -   Under certain conditions, even if a party doesn't have an **absolute advantage**, trade with other agent can be benefical for both of them

## {.smaller}

-   Example of gains of trade using the concept of **comparative advantage** [@ridley_when_2010]

    -   **Agent 1** produce 1 spear in 4 hours and 1 axe in 3 hours.

        -   To produce both products **agent 1** will need to allocate 7 hours

    -   **Agent 2** produce 1 spear in 1 hour and 1 axe in 2 hours.

        -   **Agent 2** has the absolute advantage in the prodution of spears and axes.
        -   To produce both products **agent 2** will need to allocate 3 hours

    -   ¿It will be benefical for agent 1 and 2 to trade? Yes!!! **Agent 1** can produce 2 axes in 6 hours and **agent 2** 2 spears in 2 hours. Then they can trade 1 spear for 1 axe.

        -   **Agent 1** has now 1 spear and 1 axe only working 6 hours.
        -   Furthermore **agent 2** has now 1 spear and 1 axe only working 2 hours.
        -   Each of them saves 1 hour of work in contrast with a situation in which they don't trade and produce themselves both products.

# International trade: goods

##

```{r}
# Import
trade_col <- wb_data(
  country = "COL",
  indicator = c("NE.IMP.GNFS.KN", "NE.EXP.GNFS.KN"),
  return_wide = FALSE
)

# Tidy
trade_col_clean <- trade_col |>
  filter(!is.na(value)) |>
  select(country, date, value, indicator_id) |>
  mutate(indicator_id = case_when(
    indicator_id == "NE.IMP.GNFS.KN" ~ "Imports",
    indicator_id == "NE.EXP.GNFS.KN" ~ "Exports",
    TRUE ~ indicator_id
  ))
```

```{r}
#| label: fig-imports-exports-col
#| fig-cap: Colombia imports and exports

# Visualization
trade_col_clean |>
  ggplot() +
  geom_point(aes(x = date, y = value, fill = indicator_id),
    shape = 21,
    color = "black",
    show.legend = FALSE
  ) +
  geom_line(aes(
    x = date,
    y = value,
    group = indicator_id,
    color = indicator_id
  )) +
  scale_y_continuous(labels = number_format(
    scale = 1e-9,
    suffix = "B",
    accuracy = 1
  )) +
  scale_color_manual(values = umng_palette[1:2]) +
  scale_fill_manual(values = umng_palette[1:2]) +
  labs(
    x = "Year",
    y = "Billions (Short scale)",
    color = "",
    title = str_glue("Period: {min(trade_col_clean$date)}-{max(trade_col_clean$date)}"),
    subtitle = str_glue("Variables Units: constant Local Currency Units Base Year 2015
                              Imports WDI code : NE.IMP.GNFS.KN
                              Exports WDI: NE.EXP.GNFS.KN"),
    caption = str_glue("Source: World Development Indicators - World Bank,
                              Last update: {unique(trade_col$last_updated)}")
  ) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10)
  )
```

## {.smaller}

-   The **Harmonized System (HS)** is a standardized numerical method of classifying traded goods that is internationally accepted and maintained by the **World Customs Organization (WCO)**

-   The HS code consists of 6-digits that are the same independent of the country:

    -   First 2 digits designate the **HS** chapter
    -   Second 2 digits designate the **HS** heading
    -   Third 2 digits designate the **HS** subheading

-   Example of **HS** code: 090111 (Coffee, not roasted, not decaffeinated)

    -   Chapter **09**: Cofee, tea, mate and spices
    -   Heading **01**: Coffee, whether or not roasted or decaffeinated; coffee husks and skins; coffee substitutes containing coffee in any proportion
    -   Subheading **11**: Coffee, not roasted, not decaffeinated

##

![Colombian good exports **HS 4 digits (HS92) edition** year 1995 [@simoes_economic_2011]](000_data/005_col_exports_1995_hs4_92.png){.nostretch #fig-colombian-goods-imports-1995 width=100% fig-align="center"}

##

![Colombian good exports **HS 4 digits (HS92) edition** year 2023 [@simoes_economic_2011]](000_data/005_col_exports_2023_hs4_92.png){.nostretch #fig-colombian-goods-exports-2023 width=100% fig-align="center"}

##

![Colombian goods imports **HS 4 digits (HS92) edition** year 1995 [@simoes_economic_2011]](000_data/005_col_imports_1995_hs4_92.png){.nostretch #fig-colombian-goods-imports-1995 width=100% fig-align="center"}

##

![Colombian goods imports **HS 4 digits (HS92) edition** year 2023 [@simoes_economic_2011]](000_data/005_col_imports_2023_hs4_92.png){.nostretch #fig-colombian-goods-imports-2023 width=100% fig-align="center"}

# International trade: services

## {.smaller}

-   The Extended Balance of Payments Services Classification (EBOPS) provides a breakdown of the Balance of Payments Trade in Services by types of services.

-   The classification thereby meets a number of user requirements, including the provision of more detailed information on Trade in services

    -   **EBOPS 2002**
    -   **EBOPS 2010**

-   For more information check out

    -   [@vereinte_nationen_manual_2002]
    -   [@united_nations_manual_2012]
    
##

```{r}
# Data
## https://comtradeplus.un.org/ >
## Data >
## Trade >
### Type of Product: Services
### Frequency: Annual
### Classifications: 2002
### EBOPS2002 service codes: 200, 205,
###                          236, 245,
###                          249, 253,
###                          260, 262,
###                          266, 268,
###                          287, 291
### Reporters: Colombia
### Trade Flows: Exports, Imports
### Periods:
### Partners: World

# Import
## There is a problem in the data where
## a comma is included and the end for
## rows greater or equal than 2
trade_services <- read_csv(file = "000_data/005_trade_data_services_un_comtrade.csv")

# Helpers
last_update_trade_services <- ymd("2025-07-28")

# Tidy
trade_services_clean <- trade_services |> 
  select(refYear,
         reporterDesc,
         flowDesc,
         cmdCode,
         cmdDesc,
         cifvalue)

# Wrangle
trade_services_clean_top_8 <- trade_services_clean |>
  filter(cmdCode != 200) |> 
  mutate(cmdDesc = str_replace_all(string = cmdDesc, 
                                   pattern = "[0-9.]", 
                                   replacement = "")) |>
  mutate(cmdDesc = str_trim(string = cmdDesc)) |> 
  group_by(refYear,
           flowDesc) |> 
  slice_max(order_by = cifvalue, 
            n = 8) |> 
  mutate(cmdDesc = fct_reorder(.f = cmdDesc, 
                               .x = cifvalue)) |> 
  ungroup() |> 
  mutate(cifvalue_label = number(x = cifvalue, 
                                 scale = 1e-9, 
                                 suffix = "B", 
                                 accuracy = 0.01))

## Exports first year
trade_services_clean_top_8_export_first <- trade_services_clean_top_8 |> 
  filter(refYear == min(refYear), flowDesc == "Export")

## Exports last year
trade_services_clean_top_8_export_last <- trade_services_clean_top_8 |> 
  filter(refYear == max(refYear), flowDesc == "Export")

## Imports first year
trade_services_clean_top_8_import_first <- trade_services_clean_top_8 |>
  filter(refYear == min(refYear), flowDesc == "Import")

## Imports last year
trade_services_clean_top_8_import_last <- trade_services_clean_top_8 |> 
  filter(refYear == max(refYear), flowDesc == "Import")
```

```{r}
#| label: fig-exports-col-services-first-year
#| fig-cap: Colombia service exports

# Visualization
trade_services_clean_top_8_export_first |>
  mutate(cmdDesc = fct_reorder(.f = cmdDesc, 
                               .x = cifvalue)) |>
  ggplot(aes(
    cifvalue,
    cmdDesc
  )) +
  geom_segment(aes(xend = 0, yend = cmdDesc),
               color = umng_palette[1]
  ) +
  geom_point(
    shape = 21,
    fill = umng_palette[2],
    size = 3
  ) +
  geom_text(aes(label = cifvalue_label),
            hjust = -0.5, size = 5
  ) +
  scale_x_continuous(limits = c(
    0,
    max(trade_services_clean_top_8_export_first$cifvalue) + 0.6e9
  )) +
  labs(
    x = NULL,
    y = NULL,
    subtitle = str_glue("Year: {min(trade_services_clean$refYear)}
                        Values in Billions (Short Scale: 10^9)"),
    caption = str_glue("Source: UN Comtrade Database - Trade Services
                          Last update date: {last_update_trade_services}")
  ) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )
```

##

```{r}
#| label: fig-exports-col-services-last-year
#| fig-cap: Colombia service exports

# Visualization
trade_services_clean_top_8_export_last |>
  mutate(cmdDesc = fct_reorder(.f = cmdDesc, 
                               .x = cifvalue)) |>
  ggplot(aes(
    cifvalue,
    cmdDesc
  )) +
  geom_segment(aes(xend = 0, yend = cmdDesc),
               color = umng_palette[1]
  ) +
  geom_point(
    shape = 21,
    fill = umng_palette[2],
    size = 3
  ) +
  geom_text(aes(label = cifvalue_label),
            hjust = -0.5, size = 5
  ) +
  scale_x_continuous(limits = c(
    0,
    max(trade_services_clean_top_8_export_last$cifvalue) + 0.6e9
  )) +
  labs(
    x = NULL,
    y = NULL,
    subtitle = str_glue("Year: {max(trade_services_clean$refYear)}
                        Values in Billions (Short Scale: 10^9)"),
    caption = str_glue("Source: UN Comtrade Database - Trade Services
                          Last update date: {last_update_trade_services}")
  ) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )
```

##

```{r}
#| label: fig-imports-col-services-first-year
#| fig-cap: Colombia service imports

# Visualization
trade_services_clean_top_8_import_first |>
  mutate(cmdDesc = fct_reorder(.f = cmdDesc, 
                               .x = cifvalue)) |>
  ggplot(aes(
    cifvalue,
    cmdDesc
  )) +
  geom_segment(aes(xend = 0, yend = cmdDesc),
               color = umng_palette[1]
  ) +
  geom_point(
    shape = 21,
    fill = umng_palette[2],
    size = 3
  ) +
  geom_text(aes(label = cifvalue_label),
            hjust = -0.5, size = 5
  ) +
  scale_x_continuous(limits = c(
    0,
    max(trade_services_clean_top_8_import_first$cifvalue) + 0.6e9
  )) +
  labs(
    x = NULL,
    y = NULL,
    subtitle = str_glue("Year: {min(trade_services_clean$refYear)}
                        Values in Billions (Short Scale: 10^9)"),
    caption = str_glue("Source: UN Comtrade Database - Trade Services
                          Last update date: {last_update_trade_services}")
  ) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )
```

##

```{r}
#| label: fig-imports-col-services-last-year
#| fig-cap: Colombia service imports

# Visualization
trade_services_clean_top_8_import_last |>
  mutate(cmdDesc = fct_reorder(.f = cmdDesc, 
                               .x = cifvalue)) |>
  ggplot(aes(
    cifvalue,
    cmdDesc
  )) +
  geom_segment(aes(xend = 0, yend = cmdDesc),
               color = umng_palette[1]
  ) +
  geom_point(
    shape = 21,
    fill = umng_palette[2],
    size = 3
  ) +
  geom_text(aes(label = cifvalue_label),
            hjust = -0.5, size = 5
  ) +
  scale_x_continuous(limits = c(
    0,
    max(trade_services_clean_top_8_import_last$cifvalue) + 0.6e9
  )) +
  labs(
    x = NULL,
    y = NULL,
    subtitle = str_glue("Year: {max(trade_services_clean$refYear)}
                        Values in Billions (Short Scale: 10^9)"),
    caption = str_glue("Source: UN Comtrade Database - Trade Services
                          Last update date: {last_update_trade_services}")
  ) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )
```

# Observatory of Economic Complexity (OEC)

##

-   *"The Observatory of Economic Complexity (OEC) is an online data visualization and distribution platform focused on the geography and dynamics of economic activities"*

-   *"The OEC is currently designed and developed by Datawheel, but it began as a research project at MIT's Collective Learning group (former Macro Connections Group)"*

##

-   Tree map

    -   [https://oec.world/](https://oec.world/) \> TOOLS \> Tree map

        -   Country
        -   Product
        -   Bilateral

-   Profiles

    -   [https://oec.world/](https://oec.world/) \> PROFILES \> Countries \> Colombia

# Acknowledgments

##

-   To my family that supports me

-   To the taxpayers of Colombia and the [**UMNG students**](https://www.umng.edu.co/estudiante) who pay my salary

-   To the [**Business Science**](https://www.business-science.io/) and [**R4DS Online Learning**](https://www.rfordatasci.com/) communities where I learn [**R**](https://www.r-project.org/about.html) and [**$\pi$-thon**](https://www.python.org/about/)

-   To the [**R Core Team**](https://www.r-project.org/contributors.html), the creators of [**RStudio IDE**](https://posit.co/download/rstudio-desktop/), [**Positron**](https://positron.posit.co/), [**Quarto**](https://quarto.org/) and the authors and maintainers of the packages [**tidyverse**](https://CRAN.R-project.org/package=tidyverse),
[**wbstats**](https://CRAN.R-project.org/package=wbstats),
[**scales**](https://CRAN.R-project.org/package=scales), and [**tinytex**](https://CRAN.R-project.org/package=tinytex) for allowing me to access these tools without paying for a license

# References {.allowframebreaks}