---
title: "External sector II"
author: Luis Francisco Gomez Lopez
date: 2024-01-28
execute: 
  echo: false
  warning: false
format:
  beamer:
    colortheme: dolphin
    fonttheme: structurebold
    theme: AnnArbor
    link-citations: true
    linkcolor: blue
    fig-align: center
    include-in-header:
      - text: |
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{multirow}
          \usepackage{wrapfig}
          \usepackage{float}
          \usepackage{colortbl}
          \usepackage{pdflscape}
          \usepackage{tabu}
          \usepackage{threeparttable}
          \usepackage{threeparttablex}
          \usepackage[normalem]{ulem}
          \usepackage{makecell}
          \usepackage{xcolor}
          \usepackage{fontawesome5}
bibliography: intro_econ.bib
---

```{r}
library(tidyverse)
library(wbstats)
library(tidyquant)
library(knitr)
```

# Contents

- Please Read Me
- Purpose
- The concept of comparative advantage
- Volume of international trade
- Harmonized System (HS)
- Patterns of international trade: goods
- Extended Balance of Payments Services classification (EBOPS)
- Patterns of international trade: services
- Observatory of Economic Complexity (OEC)
- Acknowledgments
- References

# Please Read Me

- Check the message __Welcome greeting__ published in the News Bulletin Board.

- Dear student please edit your profile uploading a photo where your face is clearly visible.

- The purpose of the virtual meetings is to answer questions and not to make a summary of the study material.

- This presentation is based on [@cardenas_introduccion_2020, Chapter 5]

# Purpose

Explain the composition and determinants of Colombian foreign trade

# The concept of comparative advantage

- __Absolute advantage__: the ability of a party (individual, company, country) to produce a product more efficiently than any other party
    
    + A party should concentrate in producing a good in which it has an absolute advantage
        
    + ¿What happend when you don't have an absolute advantage?
    
- __Comparative advantage__: the ability of a party (individual, company, country) to produce a product at a lower relative opportunity cost
    
    + Under certain conditions, even if a party doesn't have an __absolute advantage__, trade with other agent can be benefical for both of them
    
# The concept of comparative advantage

- Example of gains of trade using the concept of __comparative advantage__ [@ridley_when_2010]
    
    + __Agent 1__ produce 1 spear in 4 hours and 1 axe in 3 hours. 
    
        + To produce both products __agent 1__ will need to allocate 7 hours 
        
    + __Agent 2__ produce 1 spear in 1 hour and 1 axe in 2 hours. 
    
        + __Agent 2__ has the absolute advantage in the prodution of spears and axes. 
        + To produce both products __agent 2__ will need to allocate 3 hours 
            
    + ¿It will be benefical for agent 1 and 2 to trade? Yes!!! __Agent 1__ can produce 2 axes in 6 hours and __agent 2__ 2 spears in 2 hours. Then they can trade 1 spear for 1 axe.
        
        + __Agent 1__ has now 1 spear and 1 axe only working 6 hours. 
        + Furthermore __agent 2__ has now 1 spear and 1 axe only working 2 hours. 
        + Each of them saves 1 hour of work in contrast with a situation in which they don't trade and produce themselves both products.
        
# Volume of international trade

```{r }
#| out-width: 90%
#| fig-align: center

# Data
data <- wbstats::wb_data(country   = "COL",
                         indicator = c("NE.IMP.GNFS.KN", "NE.EXP.GNFS.KN"),
                         return_wide = FALSE) |>
  filter(!is.na(value))

data |>
  select(country, date, value, indicator_id) |>
  mutate(indicator_id = case_when(
    indicator_id == "NE.IMP.GNFS.KN" ~ "Imports",
    indicator_id == "NE.EXP.GNFS.KN" ~ "Exports",
    TRUE                             ~ indicator_id)) |>

  #Plot
  ggplot() +
  geom_point(aes(x = date, y = value, fill = indicator_id),
             shape = 21,
             color = "black",
             show.legend = FALSE) +
  geom_line(aes(x     = date,
                y     = value,
                group = indicator_id,
                color = indicator_id)) +
  scale_y_continuous(labels = scales::number_format(scale    = 1e-9,
                                                    suffix   = "B",
                                                    accuracy = 1)) +
  scale_color_tq() +
  scale_fill_tq() +
  labs(x        = "Year",
       y        = "Billions (Short scale)",
       color    = "",
       title    = str_glue("Colombia imports and exports: {min(data$date)}-{max(data$date)}"),
       subtitle = str_glue("Variables Units: constant Local Currency Units Base Year 2015
                              Imports WDI code : NE.IMP.GNFS.KN
                              Exports WDI: NE.EXP.GNFS.KN"),
       caption  = str_glue("Source: World Development Indicators - World Bank,
                              Last update: {unique(data$last_updated)}")) +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"))
```


# Harmonized System (HS)

- The __Harmonized System (HS)__ is a standardized numerical method of classifying traded goods that is internationally accepted and maintained by the __World Customs Organization (WCO)__

- The HS code consists of 6-digits that are the same independent of the country:

    + First 2 digits designate the __HS__ chapter
    + Second 2 digits designate the __HS__ heading
    + Third 2 digits designate the __HS__ subheading
    
- Example of __HS__ code: 090111 (Coffee, not roasted, not decaffeinated)

    + Chapter __09__: Cofee, tea, mate and spices
    + Heading __01__: Coffee, whether or not roasted or decaffeinated; coffee husks and skins; coffee substitutes containing coffee in any proportion
    + Subheading __11__: Coffee, not roasted, not decaffeinated
    
# Patterns of international trade: goods

![Colombian good exports __HS 4 digits (HS92) edition__ year 1995 [@oec_observatory_2022]](000_data/005_col_exports_1995_hs4_92.png){#fig-colombian-goods-exports-1995 fig-alt="Colombian goods exports 1995" width="450px" height="450px"}

# Patterns of international trade: goods

![Colombian good exports __HS 4 digits (HS92) edition__ year 2021 [@oec_observatory_2022]](000_data/005_col_exports_2021_hs4_92.png){#fig-colombian-goods-exports-2021 fig-alt="Colombian goods exports 2021" width="450px" height="450px"}

# Patterns of international trade: goods

![Colombian goods imports __HS 4 digits (HS92) edition__ year 1995 [@oec_observatory_2022]](000_data/005_col_imports_1995_hs4_92.png){#fig-colombian-goods-imports-1995 fig-alt="Colombian goods imports 1995" width="450px" height="450px"}

# Patterns of international trade: goods

![Colombian goods imports __HS 4 digits (HS92) edition__ year 2021 [@oec_observatory_2022]](000_data/005_col_imports_2021_hs4_92.png){#fig-colombian-goods-imports-2021 fig-alt="Colombian goods imports 2021" width="450px" height="450px"}

# Extended Balance of Payments Services classification (EBOPS)

-  The EBOPS classification provides a breakdown of the Balance of Payments Trade in Services by types of services. 

- The classification thereby meets a number of user requirements, including the provision of more detailed information on Trade in services

    + **EBOPS 2002**
    + **EBOPS 2010**
    
- For more information check out

    + [@vereinte_nationen_manual_2002]
    + [@united_nations_manual_2012]
    
# Patterns of international trade: services

![Colombian services exports __(EBOPS Version 2002)__ year 2000 [@oec_observatory_2022]](000_data/005_col_export_ebops_version2002_2000.png){#fig-colombian-services-exports-2000 fig-alt="Colombian services exports 2000" width="280px" height="280px"}

# Patterns of international trade: services

![Colombian services exports __(EBOPS Version 2002)__ year 2020 [@oec_observatory_2022]](000_data/005_col_export_ebops_version2002_2020.png){#fig-colombian-services-exports-2020 fig-alt="Colombian services exports 2020" width="250px" height="250px"}

# Patterns of international trade: services

![Colombian services imports __(EBOPS Version 2002)__ year 2000 [@oec_observatory_2022]](000_data/005_col_import_ebops_version2002_2000.png){#fig-colombian-services-imports-2000 fig-alt="Colombian services imports 2000" width="280px" height="280px"}

# Patterns of international trade: services

![Colombian services imports __(EBOPS Version 2002)__ year 2020 [@oec_observatory_2022]](000_data/005_col_import_ebops_version2002_2020.png){#fig-colombian-services-imports-2020 fig-alt="Colombian services imports 2020" width="250px" height="250px"}

# Observatory of Economic Complexity (OEC)

- "The Observatory of Economic Complexity (OEC) is an online data visualization and distribution platform focused on the geography and dynamics of economic activities" [@oec_observatory_2022]

- "The OEC is currently designed and developed by Datawheel, but it began as a research project at MIT's Collective Learning group (former Macro Connections Group)" [@oec_observatory_2022]

# Observatory of Economic Complexity (OEC)

- Tree map

    + [https://oec.world/](https://oec.world/) > TOOLS > Tree map
    
        + Country 
        + Product
        + Bilateral
        
- Profiles

    + [https://oec.world/](https://oec.world/) > PROFILES > Countries > Colombia

# Acknowledgments

- To my family that supports me

- To the taxpayers of Colombia and the __[UMNG students](https://www.umng.edu.co/estudiante)__ who pay my salary

- To the __[Business Science](https://www.business-science.io/)__ and __[R4DS Online Learning](https://www.rfordatasci.com/)__ communities where I learn __[R](https://www.r-project.org/about.html)__ 

- To the __[R Core Team](https://www.r-project.org/contributors.html)__, the creators of __[RStudio IDE](https://rstudio.com/products/rstudio/)__, __[Quarto](https://quarto.org/)__ and the authors and maintainers of the packages  __[tidyverse](https://CRAN.R-project.org/package=tidyverse)__, __[wbstats](https://CRAN.R-project.org/package=wbstats)__, __[tidyquant](https://CRAN.R-project.org/package=tidyquant)__, __[knitr](https://CRAN.R-project.org/package=knitr)__, and __[tinytex](https://CRAN.R-project.org/package=tinytex)__ for allowing me to access these tools without paying for a license

- To the __[Linux kernel community](https://www.kernel.org/category/about.html)__ for allowing me the possibility to use some __[Linux distributions](https://static.lwn.net/Distributions/)__ as my main __[OS](https://en.wikipedia.org/wiki/Operating_system)__ without paying for a license

# References {.allowframebreaks}
