---
title: Public Finances
subtitle: Session 2
author:
  - name: 
      given: Luis F.
      family: Gómez L.
    orcid: 0000-0002-2877-9015
    email: luis.gomezl@unimilitar.edu.co
    affiliations:
      - name: Distance Learning Faculty
toc: true
toc-depth: 1
slide-level: 2
format:
  presentation-umng-revealjs:
    logo-vertical-strip: 000_logos/logo-white-vertical.png 
    embed-resources: true
    reference-location: document
    mermaid-format: png
date: last-modified
date-format: "D MMMM, YYYY"
knitr: 
  opts_chunk: 
    echo: false
    warning: false
    fig.align: center
    out.width: 85%
lang: en
bibliography: intro_econ.bib
csl: apa-7.csl
editor: source
---

```{r}
#| label: libraries

library(tidyverse)
library(janitor)
library(gt)
library(scales)
library(treemapify)
```

```{r}
#| label: palette

umng_palette <- c(
  "#043074",
  "#fdc600",
  "#0d5e30",
  "#ee2a24",
  "#fc6700",
  "#00b3f0",
  "#6e3c1a",
  "#f8941c",
  "#8e44ad",
  "#2c3e50",
  "#16a085",
  "#c0392b",
  "#e91e63"
)
```

# Please Read Me

##

-   Check the message **Welcome greeting** published in Announcements forum.

-   Dear student please edit your profile uploading a photo where your face is clearly visible.

-   If you want to participate, please fill out the following survey: **Primer corte 30% > Learning Activities > Tu opinión sobre la economía colombiana**

-   This presentation is based on [@cardenas_introduccion_2020, Chapter 6]

##

- **Purpose**

    - Explain the role of the State in the economy
    
# Expenditures

## 

-   Useful resources

    -   **Portal de transparencia económica**

    -   **Guía ciudadana a la tributación y el gasto del Estado colombiano** [@observatorio_fiscal_de_la_universidad_javeriana_guiciudadana_2018]

        -   [**https://www.ofiscal.org/guias**](https://www.ofiscal.org/guias) \> Tributación y gasto del Estado colombiano \> Ver guía

    -   **Presupuesto en Perspectiva Económica (PePE)**

        -   [**https://www.ofiscal.org/interactua**](https://www.ofiscal.org/interactua) \> PePE Presupuesto en Perspectiva Económica
        
## 

-   **Presupuesto General de la Nación (PGN)**

    -   It is the most important financial management instrument of fiscal policy

    -   This instrument presents the government's proposed revenues and spending for a specific period

    -   Every year the "Presupuesto General de la Nación (PGN)" is define for the next year

    -   The "Presupuesto General de la Nación (PGN)" must be approved by the congress and sanctioned as a law by the president
    
##

```{r}
# Option 1
## https://www.pte.gov.co  >
## PGN Presupuesto General de la Nación >
## Cuadro de Ejecución PGN >
## PGN >
## DESCARGAR DATOS

# Option 2
## https://www.datos.gov.co/ >
## Buscar: Información de Gastos del Presupuesto General de la Nación >
## Exportar >
## Exportar formato CSV
### Hacienda y Crédito Público
### Este conjunto de datos presenta la información del Presupuesto
### General de la Nación relacionadas con las entidades del Nivel
### Nacional

# Helpers
last_update_tax_revenue <- ymd("2026-01-22")

# Import
pgn <- read_csv(
  file = "000_data/006_informacion_de_gastos_del_presupuesto_general_de_la_nacion_2026-01-22.zip"
)

# Tidy
png_ano_mes_sector <- pgn |>
  clean_names() |>
  group_by(ano, mes, sector) |>
  summarise(apropiacion_vigente = sum(apropiacion_vigente)) |>
  ungroup() |>
  mutate(sector = str_to_title(sector))

## Specify the years to show
range_year <- unique(png_ano_mes_sector$ano)
initial_year <- min(range_year)
last_year <- max(range_year)
middle_year <- ceiling((last_year - initial_year) / 2) + initial_year

# Helpers
sector <- png_ano_mes_sector |>
  distinct(sector)

rows_sector <- nrow(sector)
first_rows <- ceiling(x = (rows_sector / 2))
last_rows <- rows_sector - first_rows
```

```{r}
#| label: tbl-classification-by-sector-png-1
#| tbl-cap: Classification by sector - PNG

# Table
sector |>
  slice_head(n = first_rows) |>
  gt() |>
  cols_label(
    sector = "Sector"
  ) |>
  tab_options(
    row.striping.include_table_body = TRUE,
    table.border.top.width = px(3),
    table.border.top.color = "black",
    column_labels.border.bottom.width = px(2),
    column_labels.border.bottom.color = "black",
    table.border.bottom.width = px(3),
    table.border.bottom.color = "black",
    table.font.size = px(14)
  )
```

##

```{r}
#| label: tbl-classification-by-sector-png-2
#| tbl-cap: Classification by sector - PNG

# Table
sector |>
  slice_tail(n = last_rows) |>
  gt() |>
  cols_label(
    sector = "Sector"
  ) |>
  tab_options(
    row.striping.include_table_body = TRUE,
    table.border.top.width = px(3),
    table.border.top.color = "black",
    column_labels.border.bottom.width = px(2),
    column_labels.border.bottom.color = "black",
    table.border.bottom.width = px(3),
    table.border.bottom.color = "black",
    table.font.size = px(14)
  )
```

##

```{r}
# Tidy
## Initial year
png_initial_year <- png_ano_mes_sector |>
  filter(ano == {{ initial_year }}) |>
  filter(mes == "Diciembre") |>
  arrange(desc(apropiacion_vigente))

png_spent_initial_year <- png_initial_year |>
  mutate(
    sector = case_when(
      sector %in% png_initial_year$sector[1:11] ~ sector,
      .default = 'Otros Sectores'
    )
  ) |>
  group_by(sector) |>
  summarise(apropiacion_vigente = sum(apropiacion_vigente)) |>
  ungroup() |>
  mutate(
    apropiacion_vigente_pct = apropiacion_vigente / sum(apropiacion_vigente),
    label = str_glue(
      "{number(apropiacion_vigente_pct, scale = 1e2, accuracy = 0.01, suffix = '%')}
                          {number(apropiacion_vigente, scale = 1e-12, accuracy = 0.01, suffix = 'B')}"
    ),
    nombre_sector = fct_reorder(.f = sector, .x = apropiacion_vigente) |>
      fct_rev()
  )
```

```{r}
#| label: fig-pgn-col-initial-year
#| fig-cap: !expr 'paste0("Who spent the nation''s general budget in ", initial_year, "?")'

# Visualization
png_spent_initial_year |>
  ggplot(aes(area = apropiacion_vigente, fill = sector, label = label)) +
  geom_treemap(color = "black") +
  geom_treemap_text(place = "centre") +
  scale_fill_manual(values = umng_palette[1:12]) +
  labs(
    fill = NULL,
    title = str_glue("Year: {initial_year}"),
    subtitle = str_glue(
      "Total national budget: {number(sum(png_spent_initial_year$apropiacion_vigente), big.mark = ',')} COP
                           Observation: B means billions in long scale (10^12)"
    ),
    caption = str_glue(
      "Portal de transparencia económica - https://www.pte.gov.co/
                          Last update date: {last_update_tax_revenue}"
    )
  ) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10),
    legend.position = "bottom"
  )
```

##

```{r}
# Tidy
## Middle year
png_middle_year <- png_ano_mes_sector |>
  filter(ano == {{ middle_year }}) |>
  filter(mes == "Diciembre") |>
  arrange(desc(apropiacion_vigente))

png_spent_middle_year <- png_middle_year |>
  mutate(
    sector = case_when(
      sector %in% png_middle_year$sector[1:11] ~ sector,
      .default = 'Otros Sectores'
    )
  ) |>
  group_by(sector) |>
  summarise(apropiacion_vigente = sum(apropiacion_vigente)) |>
  ungroup() |>
  mutate(
    apropiacion_vigente_pct = apropiacion_vigente / sum(apropiacion_vigente),
    label = str_glue(
      "{number(apropiacion_vigente_pct, scale = 1e2, accuracy = 0.01, suffix = '%')}
                          {number(apropiacion_vigente, scale = 1e-12, accuracy = 0.01, suffix = 'B')}"
    ),
    nombre_sector = fct_reorder(.f = sector, .x = apropiacion_vigente) |>
      fct_rev()
  )
```

```{r}
#| label: fig-pgn-col-middle-year
#| fig-cap: !expr 'paste0("Who spent the nation''s general budget in ", middle_year, "?")'

# Visualization
png_spent_middle_year |>
  ggplot(aes(area = apropiacion_vigente, fill = sector, label = label)) +
  geom_treemap(color = "black") +
  geom_treemap_text(place = "centre") +
  scale_fill_manual(values = umng_palette[1:12]) +
  labs(
    fill = NULL,
    title = str_glue("Year: {middle_year}"),
    subtitle = str_glue(
      "Total national budget: {number(sum(png_spent_middle_year$apropiacion_vigente), big.mark = ',')} COP
                           Observation: B means billions in long scale (10^12)"
    ),
    caption = str_glue(
      "Portal de transparencia económica - https://www.pte.gov.co/
                          Last update date: {last_update_tax_revenue}"
    )
  ) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10),
    legend.position = "bottom"
  )
```

##

```{r}
# Tidy
# Last year
## You need to change this
## depending on the information
## of the last month
month_last_year <- "Diciembre"

png_last_year <- png_ano_mes_sector |>
  filter(ano == {{ last_year }}) |>
  filter(mes == {{ month_last_year }}) |>
  arrange(desc(apropiacion_vigente))

png_spent_last_year <- png_last_year |>
  mutate(
    sector = case_when(
      sector %in% png_middle_year$sector[1:11] ~ sector,
      .default = 'Otros Sectores'
    )
  ) |>
  group_by(sector) |>
  summarise(apropiacion_vigente = sum(apropiacion_vigente)) |>
  ungroup() |>
  mutate(
    apropiacion_vigente_pct = apropiacion_vigente / sum(apropiacion_vigente),
    label = str_glue(
      "{number(apropiacion_vigente_pct, scale = 1e2, accuracy = 0.01, suffix = '%')}
                          {number(apropiacion_vigente, scale = 1e-12, accuracy = 0.01, suffix = 'B')}"
    ),
    nombre_sector = fct_reorder(.f = sector, .x = apropiacion_vigente) |>
      fct_rev()
  )
```

```{r}
#| label: fig-pgn-col-last-year
#| fig-cap: !expr 'paste0("Who spent the nation''s general budget in ", last_year, "?")'

# Visualization
png_spent_last_year |>
  ggplot(aes(area = apropiacion_vigente, fill = sector, label = label)) +
  geom_treemap(color = "black") +
  geom_treemap_text(place = "centre") +
  scale_fill_manual(values = umng_palette[1:12]) +
  labs(
    fill = NULL,
    title = str_glue("Year: {last_year}"),
    subtitle = str_glue(
      "Total national budget: {number(sum(png_spent_last_year$apropiacion_vigente), big.mark = ',')} COP
                           Observation: B means billions in long scale (10^12)"
    ),
    caption = str_glue(
      "Portal de transparencia económica - https://www.pte.gov.co/
                          Last update date: {last_update_tax_revenue}"
    )
  ) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10),
    legend.position = "bottom"
  )
```

# Case study: Education sector

##

```{r}
# Tidy
## Education entities
pgn_education <- pgn |>
  clean_names() |>
  mutate(sector = str_to_title(sector)) |>
  mutate(nombre_entidad = str_to_title(nombre_entidad)) |>
  # Check this in case you don't have all the
  # data from all the year
  # filter(ano == {{ last_year }} - 1) |>
  filter(ano == {{ last_year }}) |>
  filter(mes == "Diciembre") |>
  filter(sector == "Educacion") |>
  select(nombre_entidad, apropiacion_vigente) |>
  group_by(nombre_entidad) |>
  summarise(apropiacion_vigente = sum(apropiacion_vigente)) |>
  ungroup()

## Wrangle
pgn_education_clean <- pgn_education |>
  mutate(
    apropiacion_vigente_pct = apropiacion_vigente / sum(apropiacion_vigente)
  ) |>
  mutate(
    apropiacion_vigente_pct_label = number(
      x = apropiacion_vigente_pct,
      scale = 100,
      accuracy = 0.01,
      suffix = " %"
    )
  ) |>
  arrange(desc(apropiacion_vigente_pct_label)) |>
  select(nombre_entidad, apropiacion_vigente_pct_label)
```

<!-- Check this if you need to change  r last_year - 1 or r last_year depending if you have all the data for this year -->

- Resources assigned to the Education sector in the year `r last_year` according to the **Presupuesto General de la Nación**

    - `r sum(pgn_education$apropiacion_vigente) |> scales::number(big.mark = ",")` COP

<!-- Check this if you need to change  r last_year - 1 or r last_year depending if you have all the data for this year -->

- Number of entities in the Education sector in the year `r last_year` according to the **Presupuesto General de la Nación**

    - `r nrow(pgn_education_clean)`

##

```{r}
# Helpers
n_entities <- nrow(pgn_education_clean)
first_n_entities <- ceiling(x = n_entities / 2)
last_n_entities <- n_entities - first_n_entities
```

<!-- Check this if you need to change  r last_year - 1 or r last_year depending if you have all the data for this year -->

```{r}
#| label: tbl-participation-entities-education-sector-1
#| tbl-cap: !expr 'paste0("Participation in the resources assigned for each entity in the education sector (Year ", last_year, ")")'

# Table
pgn_education_clean |>
  slice_head(n = first_n_entities) |>
  gt() |>
  cols_label(
    nombre_entidad = "Entity",
    apropiacion_vigente_pct_label = "Percentage participation in the resources assigned"
  ) |>
  tab_options(
    row.striping.include_table_body = TRUE,
    table.border.top.width = px(3),
    table.border.top.color = "black",
    column_labels.border.bottom.width = px(2),
    column_labels.border.bottom.color = "black",
    table.border.bottom.width = px(3),
    table.border.bottom.color = "black",
    table.font.size = px(14),
    data_row.padding = px(3)
  )
```

##

<!-- Check this if you need to change  r last_year - 1 or r last_year depending if you have all the data for this year -->

```{r}
#| label: tbl-participation-entities-education-sector-2
#| tbl-cap: !expr 'paste0("Participation in the resources assigned for each entity in the education sector (Year ", last_year, ")")'

# Table
pgn_education_clean |>
  slice_tail(n = last_n_entities) |>
  gt() |>
  cols_label(
    nombre_entidad = "Entity",
    apropiacion_vigente_pct_label = "Percentage participation in the resources assigned"
  ) |>
  tab_options(
    row.striping.include_table_body = TRUE,
    table.border.top.width = px(3),
    table.border.top.color = "black",
    column_labels.border.bottom.width = px(2),
    column_labels.border.bottom.color = "black",
    table.border.bottom.width = px(3),
    table.border.bottom.color = "black",
    table.font.size = px(14),
    data_row.padding = px(3)
  )
```

# Acknowledgments

##

-   To my family that supports me

-   To the taxpayers of Colombia and the [**UMNG students**](https://www.umng.edu.co/estudiante) who pay my salary

-   To the [**Business Science**](https://www.business-science.io/) and [**R4DS Online Learning**](https://www.rfordatasci.com/) communities where I learn [**R**](https://www.r-project.org/about.html) and [**$\pi$-thon**](https://www.python.org/about/)

-   To the [**R Core Team**](https://www.r-project.org/contributors.html), the creators of [**RStudio IDE**](https://posit.co/download/rstudio-desktop/), [**Positron**](https://positron.posit.co/), [**Quarto**](https://quarto.org/) and the authors and maintainers of the packages [**tidyverse**](https://CRAN.R-project.org/package=tidyverse),
[**janitor**](https://CRAN.R-project.org/package=janitor),
[**gt**](https://CRAN.R-project.org/package=gt),
[**scales**](https://CRAN.R-project.org/package=scales),
[**treemapify**](https://CRAN.R-project.org/package=treemapify) and [**tinytex**](https://CRAN.R-project.org/package=tinytex) for allowing me to access these tools without paying for a license 

# References {.allowframebreaks}