---
title: "Public Finances II"
author: 
  - Luis Francisco Gómez López
institute: 
  - FAEDIS
date: last-modified
format: 
  beamer:
    colortheme: dolphin
    fonttheme: structurebold
    theme: AnnArbor
    link-citations: true
    colorlinks: true
    toc: true
    slide-level: 2
    section-titles: false
    include-in-header:
      file: 000_tex_files/preamble.tex
fig-cap-location: bottom
tbl-cap-location: top
knitr: 
  opts_chunk: 
    echo: false
    warning: false
    fig-align: center
    out-width: 85%
lang: en
bibliography: intro_econ.bib
---

```{r}
#| label: libraries

library(tidyverse)
library(readxl)
library(knitr)
library(kableExtra)
library(tidyquant)
library(treemapify)
library(janitor)
library(latex2exp)
```

# Please Read Me

## 

-   Check the message **Welcome greeting** published in the News Bulletin Board.

-   Dear student please edit your profile uploading a photo where your face is clearly visible.

-   The purpose of the virtual meetings is to answer questions and not to make a summary of the study material.

-   If you want to participate, please fill out the following survey: Primer corte 30% \> Learning Activities \> Tu opinión sobre la economía colombiana

-   This presentation is based on [@cardenas_introduccion_2020, Chapter 6]

# Purpose

## 

Explain the role of the State in the economy

# Expenditures

## 

-   Useful resources

    -   **Portal de transparencia económica**

        -   [**https://www.pte.gov.co**](https://www.pte.gov.co) \> PGN Presupuesto General de la Nación \> Cuadro de Ejecución PGN \> PGN \> DESCARGAR DATOS

    -   **Guía ciudadana a la tributación y el gasto del Estado colombiano** [@observatorio_fiscal_de_la_universidad_javeriana_guiciudadana_2018]

        -   [**https://www.ofiscal.org/**](https://www.ofiscal.org/) \> Guias \> Tributación y gasto del Estado colombiano \> ver y descargar ↓

    -   **Presupuesto en Perspectiva Económica (PePE)**

        -   [**https://www.ofiscal.org/**](https://www.ofiscal.org/) \> Interactúa \> PePE

## 

-   **Presupuesto General de la Nación (PGN)**

    -   It is the most important financial management instrument of fiscal policy

    -   This instrument presents the government's proposed revenues and spending for a specific period

    -   Every year the "Presupuesto General de la Nación (PGN)" is define for the next year

    -   The "Presupuesto General de la Nación (PGN)" must be approved by the congress and sanctioned as a law by the president

    -   For more information check out **Guía ciudadana al Presupuesto General de la Nación** [@observatorio_fiscal_de_la_universidad_javeriana_guiciudadana_2022]

        -   [**https://www.ofiscal.org/**](https://www.ofiscal.org/) \> Guias \> Presupuesto General de la Nación \> ver y descargar ↓

## 

```{r}
#| include: false

# Option 1
## https://www.pte.gov.co  > 
## PGN Presupuesto General de la Nación >
## Cuadro de Ejecución PGN >
## PGN >
## DESCARGAR DATOS

# Option 2
## https://www.datos.gov.co/ >
## Buscar: Información de Gastos del Presupuesto General de la Nación >
## Exportar >
## Exportar formato CSV
### Hacienda y Crédito Público
### Este conjunto de datos presenta la información del Presupuesto 
### General de la Nación relacionadas con las entidades del Nivel 
### Nacional
pgn <- read_csv(file = "000_data/006_informacion_de_gastos_del_presupuesto_general_de_la_nacion_2025-02-05.zip")

png_ano_mes_sector <- pgn |> 
  clean_names() |> 
  group_by(ano, mes, sector) |> 
  summarise(apropiacion_vigente = sum(apropiacion_vigente)) |> 
  ungroup() |> 
  mutate(sector = str_to_title(sector))

# Specify the years to show
range_year <- unique(png_ano_mes_sector$ano)
initial_year <- min(range_year)
last_year <- max(range_year)
middle_year <- ceiling((last_year - initial_year) / 2) + initial_year
```

```{r}
#| label: tbl-classification-by-sector-png-1
#| tbl-cap: Classification by sector - PNG

sector <- png_ano_mes_sector |> 
  distinct(sector)

rows_sector <- nrow(sector)
first_rows  <- ceiling(x = (rows_sector / 2))
last_rows   <- rows_sector - first_rows

sector |>
  slice_head(n = first_rows) |> 
  kable(format = "latex", 
        booktabs = TRUE, 
        col.names = c("Sector")) |>
  kable_styling(font_size = 8,
                latex_options = "striped") |>
  row_spec(row = 0, bold = TRUE)
```

##

```{r}
#| label: tbl-classification-by-sector-png-2
#| tbl-cap: Classification by sector - PNG

sector |>
  slice_tail(n = first_rows) |> 
  kable(format = "latex", 
        booktabs = TRUE, 
        col.names = c("Sector")) |>
  kable_styling(font_size = 8,
                latex_options = "striped") |>
  row_spec(row = 0, bold = TRUE)
```

##

```{r}
#| label: fig-pgn-col-initial-year
#| fig-cap: !expr 'paste0("Who spent the nation''s general budget in ", initial_year, "?")'

# Initial year
png_initial_year <- png_ano_mes_sector |> 
  filter(ano == {{initial_year}}) |> 
  filter(mes == "Diciembre") |> 
  arrange(desc(apropiacion_vigente))

png_spent_initial_year <- png_initial_year |> 
  mutate(sector = case_when(
    sector %in% png_initial_year$sector[1:11] ~ sector,
    .default = 'Otros Sectores')) |> 
  group_by(sector) |> 
  summarise(apropiacion_vigente = sum(apropiacion_vigente)) |> 
  ungroup() |> 
  mutate(apropiacion_vigente_pct = apropiacion_vigente / sum(apropiacion_vigente),
         label = str_glue("{scales::number(apropiacion_vigente_pct, scale = 1e2, accuracy = 0.01, suffix = '%')}
                          {scales::number(apropiacion_vigente, scale = 1e-12, accuracy = 0.01, suffix = 'B')}"),
         nombre_sector = fct_reorder(.f = sector, .x = apropiacion_vigente) |> 
           fct_rev())

png_spent_initial_year |> 
  ggplot(aes(area = apropiacion_vigente,
             fill = sector,
             label = label)) +
  geom_treemap(color = "black") +
  geom_treemap_text(place = "centre") +
  scale_fill_tq() +
  labs(fill = NULL,
       title = str_glue("Year: {initial_year}"),
       subtitle = str_glue("Total national budget: {scales::number(sum(png_spent_initial_year$apropiacion_vigente), big.mark = ',')} COP
                           Observation: B means billions in long scale (10¹²)"),
       caption = str_glue("Portal de transparencia económica - https://www.pte.gov.co/
                          Last update date: 2025-02-05")) +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        legend.position   = "bottom",
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"))
```

##

```{r}
#| label: fig-pgn-col-middle-year
#| fig-cap: !expr 'paste0("Who spent the nation''s general budget in ", middle_year, "?")'

# Middle year  
png_middle_year <- png_ano_mes_sector |> 
  filter(ano == {{middle_year}}) |> 
  filter(mes == "Diciembre") |> 
  arrange(desc(apropiacion_vigente))

png_spent_middle_year <- png_middle_year |> 
  mutate(sector = case_when(
    sector %in% png_middle_year$sector[1:11] ~ sector,
    .default = 'Otros Sectores')) |> 
  group_by(sector) |> 
  summarise(apropiacion_vigente = sum(apropiacion_vigente)) |> 
  ungroup() |> 
  mutate(apropiacion_vigente_pct = apropiacion_vigente / sum(apropiacion_vigente),
         label = str_glue("{scales::number(apropiacion_vigente_pct, scale = 1e2, accuracy = 0.01, suffix = '%')}
                          {scales::number(apropiacion_vigente, scale = 1e-12, accuracy = 0.01, suffix = 'B')}"),
         nombre_sector = fct_reorder(.f = sector, .x = apropiacion_vigente) |> 
           fct_rev())

png_spent_middle_year |> 
  ggplot(aes(area = apropiacion_vigente,
             fill = sector,
             label = label)) +
  geom_treemap(color = "black") +
  geom_treemap_text(place = "centre") +
  scale_fill_tq() +
  labs(fill = NULL,
       title = str_glue("Year: {middle_year}"),
       subtitle = str_glue("Total national budget: {scales::number(sum(png_spent_middle_year$apropiacion_vigente), big.mark = ',')} COP
                           Observation: B means billions in long scale (10¹²)"),
       caption = str_glue("Portal de transparencia económica - https://www.pte.gov.co/
                          Last update date: 2025-02-05")) +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        legend.position   = "bottom",
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"))
```

##

```{r}
#| label: fig-pgn-col-last-year
#| fig-cap: !expr 'paste0("Who spent the nation''s general budget in ", last_year, "?")'

# Last year
## You need to change this
## depending on the information 
## of the last month
month_last_year <- "Enero" 

png_last_year <- png_ano_mes_sector |> 
  filter(ano == {{last_year}}) |> 
  filter(mes == {{month_last_year}}) |> 
  arrange(desc(apropiacion_vigente))

png_spent_last_year <- png_last_year |> 
  mutate(sector = case_when(
    sector %in% png_middle_year$sector[1:11] ~ sector,
    .default = 'Otros Sectores')) |> 
  group_by(sector) |> 
  summarise(apropiacion_vigente = sum(apropiacion_vigente)) |> 
  ungroup() |> 
  mutate(apropiacion_vigente_pct = apropiacion_vigente / sum(apropiacion_vigente),
         label = str_glue("{scales::number(apropiacion_vigente_pct, scale = 1e2, accuracy = 0.01, suffix = '%')}
                          {scales::number(apropiacion_vigente, scale = 1e-12, accuracy = 0.01, suffix = 'B')}"),
         nombre_sector = fct_reorder(.f = sector, .x = apropiacion_vigente) |> 
           fct_rev())

png_spent_last_year |> 
  ggplot(aes(area = apropiacion_vigente,
             fill = sector,
             label = label)) +
  geom_treemap(color = "black") +
  geom_treemap_text(place = "centre") +
  scale_fill_tq() +
  labs(fill = NULL,
       title = str_glue("Year: {last_year}"),
       subtitle = str_glue("Total national budget: {scales::number(sum(png_spent_last_year$apropiacion_vigente), big.mark = ',')} COP
                           Observation: B means billions in long scale (10¹²)"),
       caption = str_glue("Portal de transparencia económica - https://www.pte.gov.co/
                          Last update date: 2025-02-05")) +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        legend.position   = "bottom",
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"))
```

# Case study: Education sector

##

```{r}
#| include: false

# Education entities
pgn_education <- pgn |>
  clean_names() |> 
  mutate(sector = str_to_title(sector)) |> 
  mutate(nombre_entidad = str_to_title(nombre_entidad)) |> 
  filter(ano == {{last_year}} - 1) |> 
  filter(mes == "Diciembre") |>
  filter(sector == "Educacion") |> 
  select(nombre_entidad,
         apropiacion_vigente) |> 
  group_by(nombre_entidad) |> 
  summarise(apropiacion_vigente = sum(apropiacion_vigente)) |> 
  ungroup()

# Clean data
pgn_education_clean <- pgn_education |> 
  mutate(apropiacion_vigente_pct = apropiacion_vigente / sum(apropiacion_vigente)) |> 
  mutate(apropiacion_vigente_pct_label = scales::number(x = apropiacion_vigente_pct, 
                                                        scale = 100, 
                                                        accuracy = 0.01, 
                                                        suffix = " %")) |> 
  arrange(desc(apropiacion_vigente_pct_label)) |> 
  select(nombre_entidad, apropiacion_vigente_pct_label)
```

- Resources assigned to the Education sector in the year `r last_year - 1` according to the **Presupuesto General de la Nación**

    - `r sum(pgn_education$apropiacion_vigente) |> scales::number(big.mark = ",")` COP

- Number of entities in the Education sector in the year `r last_year - 1` according to the **Presupuesto General de la Nación**

    - `r nrow(pgn_education_clean)`

##

```{r}
#| include: false

n_entities <- nrow(pgn_education_clean)
first_n_entities <- ceiling(x = n_entities / 2)  
last_n_entities  <- n_entities - first_n_entities
```

```{r}
#| label: tbl-participation-entities-education-sector-1
#| tbl-cap: !expr 'paste0("Participation in the resources assigned for each entity in the education sector (Year ", last_year - 1, ")")'

pgn_education_clean |>
  slice_head(n = first_n_entities) |> 
  kable(format = "latex", 
        booktabs = TRUE, 
        col.names = c("Entity", "Percentage participation in the resources assigned")) |>
  kable_styling(font_size = 5,
                latex_options = "striped") |>
  row_spec(row = 0, bold = TRUE)
```

##

```{r}
#| label: tbl-participation-entities-education-sector-2
#| tbl-cap: !expr 'paste0("Participation in the resources assigned for each entity in the education sector (Year ", last_year - 1, ")")'

pgn_education_clean |>
  slice_tail(n = last_n_entities) |> 
  kable(format = "latex", 
        booktabs = TRUE, 
        col.names = c("Entity", "Percentage participation in the resources assigned")) |>
  kable_styling(font_size = 5,
                latex_options = "striped") |>
  row_spec(row = 0, bold = TRUE)
```

# Acknowledgments

## 

-   To my family that supports me

-   To the taxpayers of Colombia and the [**UMNG students**](https://www.umng.edu.co/estudiante) who pay my salary

-   To the [**Business Science**](https://www.business-science.io/) and [**R4DS Online Learning**](https://www.rfordatasci.com/) communities where I learn [**R**](https://www.r-project.org/about.html) and [**$\pi$-thon**](https://www.python.org/about/)

-   To the [**R Core Team**](https://www.r-project.org/contributors.html), the creators of [**RStudio IDE**](https://rstudio.com/products/rstudio/), [**Quarto**](https://quarto.org/) and the authors and maintainers of the packages [**tidyverse**](https://CRAN.R-project.org/package=tidyverse), [**readxl**](https://CRAN.R-project.org/package=readxl), [**knitr**](https://CRAN.R-project.org/package=knitr), [**kableExtra**](https://CRAN.R-project.org/package=kableExtra), [**tidyquant**](https://CRAN.R-project.org/package=tidyquant), [**janitor**](https://CRAN.R-project.org/package=janitor), [**treemapify**](https://CRAN.R-project.org/package=treemapify), and [**tinytex**](https://CRAN.R-project.org/package=tinytex) for allowing me to access these tools without paying for a license

-   To the [**Linux kernel community**](https://www.kernel.org/category/about.html) for allowing me the possibility to use some [**Linux distributions**](https://static.lwn.net/Distributions/) as my main [**OS**](https://en.wikipedia.org/wiki/Operating_system) without paying for a license

# References {.allowframebreaks}
