---
title: "Public Finances II"
author: 
  - Luis Francisco Gómez López
institute: 
  - FAEDIS
date: last-modified
format: 
  beamer:
    colortheme: dolphin
    fonttheme: structurebold
    theme: AnnArbor
    link-citations: true
    colorlinks: true
    toc: true
    slide-level: 2
    section-titles: false
    include-in-header:
      file: 000_tex_files/preamble.tex
fig-cap-location: bottom
tbl-cap-location: top
knitr: 
  opts_chunk: 
    echo: false
    warning: false
    fig-align: center
    out-width: 85%
lang: en
bibliography: intro_econ.bib
---

```{r}
#| label: libraries

library(tidyverse)
library(readxl)
library(knitr)
library(kableExtra)
library(tidyquant)
library(treemapify)
library(janitor)
library(latex2exp)
```

# Please Read Me

## 

-   Check the message **Welcome greeting** published in the News Bulletin Board.

-   Dear student please edit your profile uploading a photo where your face is clearly visible.

-   The purpose of the virtual meetings is to answer questions and not to make a summary of the study material.

-   If you want to participate, please fill out the following survey: Primer corte 30% \> Learning Activities \> Tu opinión sobre la economía colombiana

-   This presentation is based on [@cardenas_introduccion_2020, Chapter 6]

# Purpose

## 

Explain the role of the State in the economy

# Expenditures

## 

-   Useful resources

    -   **Portal de transparencia económica**

        -   [**https://www.pte.gov.co**](https://www.pte.gov.co) \> PRESUPUESTO GENERAL DE LA NACIÓN \> Ejecución

    -   **Guía ciudadana a la tributación y el gasto del Estado colombiano** [@observatorio_fiscal_de_la_universidad_javeriana_guiciudadana_2018]

        -   [**https://www.ofiscal.org/publicaciones**](https://www.ofiscal.org/publicaciones) \> guías ciudadanas \> Tributación y gasto del Estado colombiano \> ver y descargar ↓

    -   **Presupuesto en Perspectiva Económica (PePE)**

        -   [**https://www.ofiscal.org/**](https://www.ofiscal.org/) \> Interactúa \> PePE

## 

-   **Presupuesto General de la Nación (PGN)**

    -   It is the most important financial management instrument of fiscal policy

    -   This instrument presents the government's proposed revenues and spending for a specific period

    -   Every year the "Presupuesto General de la Nación (PGN)" is define for the next year

    -   The "Presupuesto General de la Nación (PGN)" must be approved by the congress and sanctioned as a law by the president

    -   For more information check out **Guía ciudadana al Presupuesto General de la Nación** [@observatorio_fiscal_de_la_universidad_javeriana_guiciudadana_2022]

        -   [**https://www.ofiscal.org/**](https://www.ofiscal.org/) \> Guias \> Presupuesto General de la Nación \> ver y descargar ↓

## 

```{r}
#| include: false

# https://www.pte.gov.co  > 
# PRESUPUESTO GENERAL DE LA NACIÓN >
# Ejecución >
# Ejecución por Sector para la Vigencia Actual
pgn_what_spent <- read_csv(file = "000_data/006_ejecucion_presupuestal_por_sector_pgn.csv",
                           locale = locale(encoding = "latin1",
                                           decimal_mark = '.')) |>
  arrange(desc(AsignacionVigente)) |> 
  clean_names() |> 
  mutate(nombre_sector = str_to_title(nombre_sector))
```

```{r}
#| label: fig-pgn-col
#| fig-cap: Who spends nation's general budget spent on?

pgn_what_spent_clean <- pgn_what_spent |> 
  select(nombre_sector, asignacion_vigente) |> 
  mutate(nombre_sector = case_when(
    nombre_sector %in% pgn_what_spent$nombre_sector[1:11] ~ nombre_sector,
    .default = 'Otros Sectores')) |> 
  group_by(nombre_sector) |> 
  summarise(asignacion_vigente = sum(asignacion_vigente)) |> 
  ungroup() |> 
  mutate(asignacion_vigente_pct = asignacion_vigente / sum(asignacion_vigente),
         label = str_glue("{scales::number(asignacion_vigente_pct, scale = 1e2, accuracy = 0.01, suffix = '%')}
                          {scales::number(asignacion_vigente, scale = 1e-12, accuracy = 0.01, suffix = 'B')}"),
         nombre_sector = fct_reorder(.f = nombre_sector, .x = asignacion_vigente) |> 
           fct_rev())

pgn_what_spent_clean |> 
  ggplot(aes(area = asignacion_vigente,
             fill = nombre_sector,
             label = label)) +
  geom_treemap(color = "black") +
  geom_treemap_text(place = "centre") +
  scale_fill_tq() +
  labs(fill = NULL,
       title = str_glue("Year: {unique(pgn_what_spent$anio)}"),
       subtitle = str_glue("Total national budget: {scales::number(sum(pgn_what_spent$asignacion_vigente), big.mark = ',')} COP
                           Observation: B means billions in long scale (10¹²)"),
       caption = str_glue("Portal de transparencia económica - https://www.pte.gov.co/
                          Last update date: 2024-07-16")) +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        legend.position   = "bottom",
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"))
```

# Acknowledgments

## 

-   To my family that supports me

-   To the taxpayers of Colombia and the [**UMNG students**](https://www.umng.edu.co/estudiante) who pay my salary

-   To the [**Business Science**](https://www.business-science.io/) and [**R4DS Online Learning**](https://www.rfordatasci.com/) communities where I learn [**R**](https://www.r-project.org/about.html) and [**$\pi$-thon**](https://www.python.org/about/)

-   To the [**R Core Team**](https://www.r-project.org/contributors.html), the creators of [**RStudio IDE**](https://rstudio.com/products/rstudio/), [**Quarto**](https://quarto.org/) and the authors and maintainers of the packages [**tidyverse**](https://CRAN.R-project.org/package=tidyverse), [**readxl**](https://CRAN.R-project.org/package=readxl), [**knitr**](https://CRAN.R-project.org/package=knitr), [**kableExtra**](https://CRAN.R-project.org/package=kableExtra), [**tidyquant**](https://CRAN.R-project.org/package=tidyquant), [**janitor**](https://CRAN.R-project.org/package=janitor), [**treemapify**](https://CRAN.R-project.org/package=treemapify), and [**tinytex**](https://CRAN.R-project.org/package=tinytex) for allowing me to access these tools without paying for a license

-   To the [**Linux kernel community**](https://www.kernel.org/category/about.html) for allowing me the possibility to use some [**Linux distributions**](https://static.lwn.net/Distributions/) as my main [**OS**](https://en.wikipedia.org/wiki/Operating_system) without paying for a license

# References {.allowframebreaks}
