---
title: "Money, prices and the exchange rate II"
author: 
  - Luis Francisco Gómez López
institute: 
  - FAEDIS
date: last-modified
format: 
  beamer:
    colortheme: dolphin
    fonttheme: structurebold
    theme: AnnArbor
    link-citations: true
    colorlinks: true
    toc: true
    slide-level: 2
    section-titles: false
    include-in-header:
      file: 000_tex_files/preamble.tex
fig-cap-location: bottom
tbl-cap-location: top
knitr: 
  opts_chunk: 
    echo: false
    warning: false
    fig-align: center
    out-width: 85%
lang: en
bibliography: intro_econ.bib
---

```{r}
#| label: libraries

library(tidyverse)
library(knitr)
library(kableExtra)
library(wbstats)
library(tidyquant)
```

# Please Read Me

##

-   Check the message **Welcome greeting** published in the News Bulletin Board.

-   Dear student please edit your profile uploading a photo where your face is clearly visible.

-   The purpose of the virtual meetings is to answer questions and not to make a summary of the study material.

-   If you want to participate, please fill out the following survey: Primer corte 30% \> Learning Activities \> Tu opinión sobre la economía colombiana

-   This presentation is based on [@cardenas_introduccion_2020, Chapter 7]

# Purpose

##

Analyze the money market and introduce the concepts of inflation, nominal exchange rate and interest rate

# Monetary policy intervention rate

##

```{r}

# Political interest rate 2024-07-02 (Colombia)
i_t <- 0.1125
date <- lubridate::ymd("2024-07-02")
```

-   In the link [**https://www.banrep.gov.co/es/estadisticas/tasas-interes-politica-monetaria**](https://www.banrep.gov.co/es/estadisticas/tasas-interes-politica-monetaria) on **`r date`** it was pointed out that:

    -   *"La tasa de intervención de política monetaria o tasa de referencia es la tasa de interés mínima que el Banco de la República (BanRep) cobra a las entidades financieras por la liquidez que les suministra mediante las operaciones de mercado abierto (OMA). Esta tasa es el principal instrumento de intervención de política monetaria utilizado por el BanRep para afectar la cantidad de dinero que circula en la economía."*

    -   Also the *"tasa de intervención de política monetaria"* on **`r date`** was **`r scales::number(i_t, scale = 1e2, accuracy = 0.01, suffix = "%")`**

##

```{r}
#| label: fig-daily-pir-col
#| fig-cap: Monetary policy intervention rate of the Bank of the Republic (Colombia)

# https://www.banrep.gov.co/es > 
# Estadísticas económicas >
# Gráficas interactivas > 
# CATÁLOGO DE SERIES >  
# Tasas de interés y sector financiero > 
# Tasas de interés >
# Tasa de política monetaria > 
# Diaria > 
# DESCARGAR >
# Descargar datos en CSV
pir_daily <- read_csv(file = '000_data/007_tip_serie_historica_diaria.csv') |> 
  set_names(nm = c('date', 'pir'))

# Plot
static_plot1 <- pir_daily |> 
  ggplot(aes(x = date, y = pir)) +
  geom_line(color = palette_light()[[1]]) +
  labs(x        = '', 
       y        = 'Percent',
       caption  = str_glue('Source: Banco de la República - Colombia
                           Last update: {pir_daily$date[length(pir_daily$date)]}'),
       subtitle = str_glue('Interest rate expressed as annual effective overdue
                             Frequency: daily')) +
  scale_y_continuous(labels = scales::label_number(accuracy = 0.1, suffix = "%"),
                     breaks = seq(from = 0, to = max(pir_daily$pir), by = 2.5)) + 
  expand_limits(y = 0) +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        legend.position   = "none",
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"))

static_plot1
```

##

```{r}
#| label: fig-daily-pir-tib-ibr-col
#| fig-cap: Monetary policy intervention rate, TIB and IBR

# https://www.banrep.gov.co/es > 
# Estadísticas económicas >
# Gráficas interactivas > 
# CATÁLOGO DE SERIES >  
# Tasas de interés y sector financiero > 
# Tasas de interés
# Tasa interbancaria (TIB) > 
# Diaria > 
# DESCARGAR >
# Descargar datos en CSV
tib_daily <- read_csv(file  = '000_data/007_tib_serie_historica_diaria.csv') |> 
  set_names(nm = c('date', 'tib'))

# https://www.banrep.gov.co/es > 
# Estadísticas económicas >
# Gráficas interactivas > 
# CATÁLOGO DE SERIES >  
# Tasas de interés y sector financiero > 
# Tasas de interés
# Indicador Bancario de Referencia (IBR) overnight, efectiva > 
# Diaria > 
# DESCARGAR >
# Descargar datos en CSV
ibr_daily <- read_csv(file = "000_data/007_ibr_plazo_overnight_efectiva_diaria.csv") |> 
  set_names(nm = c('date', 'ibr'))

pir_tib_tbr_daily <- pir_daily |> 
  left_join(tib_daily, by = "date") |> 
  left_join(ibr_daily, by = "date") |>
  drop_na() |> 
  pivot_longer(cols = pir:ibr, names_to = "variable", values_to = "value") |> 
  mutate(variable = case_when(
    variable == "pir" ~ "Monetary policy intervention rate",
    variable == "tib" ~ "Tasa interbancaria (TIB)",
    variable == "ibr" ~ "Indicador Bancario de Referencia (IBR)",
    TRUE ~ variable))

# Plot
static_plot2 <- pir_tib_tbr_daily |> 
  ggplot(aes(x = date, 
             y = value, 
             group = variable,
             color = variable)) +
  geom_line(linewidth = 0.5, alpha = 0.5) +
  labs(x        = '', 
       y        = 'Percent',
       color    = "",
       caption  = str_glue('Source: Banco de la República - Colombia
                           Last update: {pir_tib_tbr_daily$date[length(pir_tib_tbr_daily$date)]}'),
       subtitle = str_glue('Interest rates expressed as annual effective overdue
                             Frequency: daily')) +
  scale_y_continuous(labels = scales::label_number(accuracy = 0.1, suffix = "%"),
                     breaks = seq(from = 0, to = 25, by = 2.5)) + 
  scale_color_tq() +
  expand_limits(y = 0) +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        legend.position   = "bottom",
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"))

static_plot2
```

##

```{r}
#| label: fig-daily-pir-cdt-90-180-360-col
#| fig-cap: Monetary policy intervention rate and CDT interest rates

# https://www.banrep.gov.co/es > 
# Estadísticas económicas >
# Gráficas interactivas > 
# CATÁLOGO DE SERIES >  
# Tasas de interés y sector financiero > 
# Tasas de interés >
# Tasa de Certificado de Depósito a Término (CDT) a 90 días > 
# Diaria > 
# DESCARGAR >
# Descargar datos en CSV
cdt_90 <- read_csv(file  = '000_data/007_cdt_90_diaria.csv') |> 
  set_names(nm = c('date', 'cdt_90'))

# https://www.banrep.gov.co/es > 
# Estadísticas económicas >
# Gráficas interactivas > 
# CATÁLOGO DE SERIES >  
# Tasas de interés y sector financiero > 
# Tasas de interés
# Tasa de Certificado de Depósito a Término (CDT) a 180 días > 
# Diaria > 
# DESCARGAR >
# Descargar datos en CSV
cdt_180 <- read_csv(file = "000_data/007_cdt_180_diaria.csv") |> 
  set_names(nm = c('date', 'cdt_180'))

# https://www.banrep.gov.co/es > 
# Estadísticas económicas >
# Gráficas interactivas > 
# CATÁLOGO DE SERIES >  
# Tasas de interés y sector financiero > 
# Tasas de interés
# Tasa de Certificado de Depósito a Término (CDT) a 360 días > 
# Diaria > 
# DESCARGAR >
# Descargar datos en CSV
cdt_360 <- read_csv(file = "000_data/007_cdt_360_diaria.csv") |> 
  set_names(nm = c('date', 'cdt_360'))

pir_cdt_90_180_360_daily <- pir_daily |>  
  left_join(cdt_90, by = "date") |> 
  left_join(cdt_180, by = "date") |>
  left_join(cdt_360, by = "date") |>
  drop_na() |> 
  pivot_longer(cols = pir:cdt_360, names_to = "variable", values_to = "value") |> 
  mutate(variable = case_when(
    variable == "pir" ~ "Monetary policy intervention rate",
    variable == "cdt_90" ~ "CDT 90",
    variable == "cdt_180" ~ "CDT 180",
    variable == "cdt_360" ~ "CDT 360",
    .default = variable),
    variable = variable |> 
      fct_relevel(c("Monetary policy intervention rate",
                    "CDT 90", "CDT 180", "CDT 360")))

# Plot
static_plot3 <- pir_cdt_90_180_360_daily |> 
  ggplot(aes(x = date, 
             y = value, 
             group = variable, 
             color = variable)) +
  geom_line(linewidth = 0.5, alpha = 0.5) +
  labs(x        = '', 
       y        = 'Percent',
       color    = "",
       caption  = str_glue('Source: Banco de la República - Colombia
                           Last update: {pir_cdt_90_180_360_daily$date[length(pir_cdt_90_180_360_daily$date)]}'),
       subtitle = str_glue('Interest rates expressed as annual effective overdue
                             Frequency: daily')) +
  scale_y_continuous(labels = scales::label_number(accuracy = 0.1, suffix = "%"),
                     breaks = seq(from = 0, to = 25, by = 2.5)) + 
  scale_color_tq() +
  expand_limits(y = 0) +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        legend.position   = "bottom",
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"))

static_plot3 
```

# Monetary policy transmission channels

##

![**Interest rate and credit** [@banrep_econo-cimientos_2022, fig. Canal de tasa de interés y de crédito: ejemplo gráfico]](_000_images/007_econo_canales_transmision_1_tasa_interes_credito.jpg){#fig-canal-tasa-interes-credito fig-alt="Interest rate and credit" width="400px" height="250px"}

##

![**Exchange rate** [@banrep_econo-cimientos_2022, fig. Canal de la casa de cambio: ejemplo gráfico]](_000_images/007_econo_canales_transmision_2_tasa_cambio.jpg){#fig-canal-tasa-de-cambio fig-alt="Tasa de Cambio" width="400px" height="150px"}

##

![**Expectations** [@banrep_econo-cimientos_2022, fig. Canal de las expectativas: ejemplo gráfico]](_000_images/007_econo_canales_transmision_3_expectativas.jpg){#fig-canal-expectativas fig-alt="Expectativas" width="400px" height="250px"}

# Acknowledgments

##

-   To my family that supports me

-   To the taxpayers of Colombia and the [**UMNG students**](https://www.umng.edu.co/estudiante) who pay my salary

-   To the [**Business Science**](https://www.business-science.io/) and [**R4DS Online Learning**](https://www.rfordatasci.com/) communities where I learn [**R**](https://www.r-project.org/about.html) and [**$\pi$-thon**](https://www.python.org/about/)

-   To the [**R Core Team**](https://www.r-project.org/contributors.html), the creators of [**RStudio IDE**](https://rstudio.com/products/rstudio/), [**Quarto**](https://quarto.org/) and the authors and maintainers of the packages [**tidyverse**](https://CRAN.R-project.org/package=tidyverse), [**knitr**](https://CRAN.R-project.org/package=knitr), [**kableExtra**](https://CRAN.R-project.org/package=kableExtra), [**wbstats**](https://CRAN.R-project.org/package=wbstats), [**tidyquant**](https://CRAN.R-project.org/package=tidyquant), and [**tinytex**](https://CRAN.R-project.org/package=tinytex) for allowing me to access these tools without paying for a license

-   To the [**Linux kernel community**](https://www.kernel.org/category/about.html) for allowing me the possibility to use some [**Linux distributions**](https://static.lwn.net/Distributions/) as my main [**OS**](https://en.wikipedia.org/wiki/Operating_system) without paying for a license

# References {.allowframebreaks}