---
title: Money, prices and the exchange rate
subtitle: Session 1
author:
  - name: 
      given: Luis F.
      family: Gómez L.
    orcid: 0000-0002-2877-9015
    email: luis.gomezl@unimilitar.edu.co
    affiliations:
      - name: Distance Learning Faculty
toc: true
toc-depth: 1
slide-level: 2
format:
  presentation-umng-revealjs:
    logo-vertical-strip: 000_logos/logo-white-vertical.png 
    embed-resources: true
    reference-location: document
date: last-modified
date-format: "D MMMM, YYYY"
knitr: 
  opts_chunk: 
    echo: false
    warning: false
    fig.align: center
    out.width: 85%
lang: en
bibliography: intro_econ.bib
csl: apa-7.csl
editor: source
---

```{r}
#| label: libraries

library(tidyverse)
library(gt)
```

```{r}
#| label: palette

umng_palette <- c(
  "#043074",
  "#fdc600",
  "#0d5e30",
  "#ee2a24",
  "#fc6700",
  "#00b3f0",
  "#6e3c1a",
  "#f8941c",
  "#8e44ad",
  "#2c3e50",
  "#16a085",
  "#c0392b",
  "#e91e63"
)
```

# Please Read Me

##

-   Check the message **Welcome greeting** published in Announcements forum.

-   Dear student please edit your profile uploading a photo where your face is clearly visible.

-   If you want to participate, please fill out the following survey: **Primer corte 30% > Learning Activities > Tu opinión sobre la economía colombiana**

-   This presentation is based on [@cardenas_introduccion_2020, Chapter 7]

##

- **Purpose**

    - Analyze the money market and introduce the concepts of inflation, nominal exchange rate and interest rate
    
# Money and Central Banks

##

-   Money is an **asset**[^1] that fulfills in general 3 functions

    -   Medium of exchange
    -   Unit of account
    -   Store of value

-   *"A central bank[^2] is the institution that issues and administrates legal currency and exercises the function of banker of banks"* [@banco_de_la_republica_central_2021]

[^1]: This part is important because for example a credit card can fulfills the 3 functions but it is not an asset as defined by [@ifrs_conceptual_2018, chap. 4]
    
[^2]: For a list of central banks around the world see [**https://www.bis.org/cbanks.htm**](https://www.bis.org/cbanks.htm)

## {.smaller}

-   In the case of Colombia, some of the functions of the central bank are [@banrep_funciones_2025]:

    - Maintain a low and stable inflation rate in order to preserve the purchasing power of money (**Monetary Policy - Política monetaria**)

    - Intervene in the foreign exchange market in such a way that the interventions are consistent with the inflation targeting strategy (**Foreign exchange intervention policy - Política de intervención cambiaria**).
    
    - Regulate the activity of credit institutions, control the liquidity of the financial system and ensure the proper functioning of payments in the economy (**Credit policity - Política crediticia**).

## {.smaller}

-   In the case of Colombia, some of the functions of the central bank are [@banrep_funciones_2025]:

    - Manage the international reserves[^3] (**Reservar internacionales**)
    
    - Lender of last resort for credit institutions (**Prestamista de última instancia de los establecimientos de crédito**)
    
    - Support the government in certain economic operations for the management of its liquidity, its payments, and its financial portfolio (**Government fiscal agent - Agente fiscal del gobierno**)
    
[^3]: For more information about the international reserves you can check out [@banrep_reservas_2025]

##

```{r}
#| label: tbl-ingresos-tributarios-no-tributarios-col
#| tbl-cap: Establishment date of central banking institutions [@herger_understanding_2019, p. 15]

# Table
tibble(Bank = c("Sveriges Riksbank",
                "Bank of England",
                "Banque de France", 
                "Bank of Finland",
                "Nederlandsche Bank",
                "Austrian National Bank",
                "Norges Bank",
                "Danmarks Nationalbank", 
                "Banco de Portugal",
                "Belgian National Bank", 
                "Banco de España", 
                "German Reichsbank", 
                "Bank of Japan",
                "Banca d’Italia",
                "Banco de la Republica (Colombia)"),
       Established = c(1668, 1694, 1800, 1800, 1814, 1816,
                       1816, 1818, 1846, 1850, 1874, 1876,
                       1882, 1893, 1923)) |> 
  gt() |>
  tab_style(
    style = cell_fill(color = umng_palette[6]), 
    locations = cells_body(
      rows = 15
    )  
  ) |> 
  tab_options(
    row.striping.include_table_body = TRUE,
    table.border.top.width = px(3),
    table.border.top.color = "black",
    column_labels.border.bottom.width = px(2),
    column_labels.border.bottom.color = "black",
    table.border.bottom.width = px(3),
    table.border.bottom.color = "black",
    table.font.size = px(14)
  )
```

# Inflation and Consumer Price Index

##

-   **Inflation** is a persistent increase in the price level:

    -   Inflation refers to a general increase in the price level

    -   This increase must be persistent

    -   A measure that represents the behavior of the price level is required where it is necessary to specify a basket of products

-   Review the videos found in:

    -   **Tercer corte 40% \> Learning Resources \> Links of interest**

##

-   The **price level** is an **index number** [@ralph_practical_2015]

    -   An **index number** is a quantity that by varying shows the changes of a magnitude over time or space

-   The variation in the **price level** from one period to another is used to measure inflation

-   The value that the index number takes to measure the **price level** will depend on the products that are taken into account in the basket that is used

##

-   Usually the **price level** used to measure inflation in Colombia is called **Indice de Precios al Consumidor (IPC)** or **Consumer Price Index (CPI)**

    -   In Colombia, this index measures the average variation of the prices of a basket of products representative of household consumption [@dane_metodologigeneral_2019]

    -   You can check out the historical basket of products included in the **(IPC)** or **(CPI)** in: [@dane_metodologigeneral_2019] \> Información adicional \> Estructura histórica de ponderaciones y canasta de seguimiento del IPC

##

-   The **CPI** is build using information from **Encuesta Nacional de Presupuesto de los Hogares –ENPH- (2016-2017)** where it is conducted every 10 years [@dane_metodologigeneral_2019, section 2.1.5].

-   Additionally, usually monthly, but in some cases bimonthly, quadrimestral and biannual, price information is collected for approximately 443[^4] items in 38 cities where divisions included in the **CPI** are based on **Classification of Individual Consumption According to Purpose (COICOP)** [@united_nations_classifications_2000] [^5] and adapted for Colombia.

[^4]: This can change sometimes every year
[^5]: An update version [@united_nations_classification_2023] is not used according to [@dane_metodologigeneral_2019, section 2.1.5, g.] 

##

```{r}
#| label: tbl-coicop-cpi
#| tbl-cap: COICOP divisions (International version) 1999

## COICOP Revision
### https://unstats.un.org/unsd/classifications/econ/
tibble(Division = c("01", "02", "03",
                    "04", "05", "06",
                    "07", "08", "09",
                    "10", "11", "12"),
       Name     = c("Food and non-alcoholic beverages",
                    "Alcoholic beverages, tobacco and narcotics",
                    "Clothing and footwear",
                    "Housing, water, electricity, gas and other fuels",
                    "Furnishings, household equipment and routine household maintenance",
                    "Health",
                    "Transport",
                    "Communication",
                    "Recreation, sport and culture",
                    "Education",
                    "Restaurants and hotels",
                    "Miscellaneous goods and services")
       ) |> 
  gt() |> 
  tab_options(
    row.striping.include_table_body = TRUE,
    table.border.top.width = px(3),
    table.border.top.color = "black",
    column_labels.border.bottom.width = px(2),
    column_labels.border.bottom.color = "black",
    table.border.bottom.width = px(3),
    table.border.bottom.color = "black",
    table.font.size = px(18)
  )
```

# Acknowledgments

##



# References {.allowframebreaks}