---
title: Financial Market
subtitle: Session 2
author:
  - name: 
      given: Luis F.
      family: Gómez L.
    orcid: 0000-0002-2877-9015
    email: luis.gomezl@unimilitar.edu.co
    affiliations:
      - name: Distance Learning Faculty
toc: true
toc-depth: 1
slide-level: 2
format:
  presentation-umng-revealjs:
    logo-vertical-strip: 000_logos/logo-white-vertical.png 
    embed-resources: true
    reference-location: document
date: last-modified
date-format: "D MMMM, YYYY"
knitr: 
  opts_chunk: 
    echo: false
    warning: false
    fig.align: center
    out.width: 85%
lang: en
bibliography: intro_econ.bib
csl: apa-7.csl
editor: source
---

```{r}
#| label: libraries

library(tidyverse)
library(scales)
library(readxl)
library(janitor)
library(gt)
library(tidyquant)
library(wbstats)
```

```{r}
#| label: palette

umng_palette <- c(
  "#043074",
  "#fdc600",
  "#0d5e30",
  "#ee2a24",
  "#fc6700",
  "#00b3f0",
  "#6e3c1a",
  "#f8941c",
  "#8e44ad",
  "#2c3e50",
  "#16a085",
  "#c0392b",
  "#e91e63"
)
```

# Please Read Me

##

-   Check the message **Welcome greeting** published in Announcements forum.

-   Dear student please edit your profile uploading a photo where your face is clearly visible.

-   If you want to participate, please fill out the following survey: **Primer corte 30% > Learning Activities > Tu opinión sobre la economía colombiana**

-   This presentation is based on [@cardenas_introduccion_2020, Chapter 8]

##

- **Purpose**

    - Analyze the functioning of the financial market, identifying the types of intermediaries and instruments that are part of it
    
# Principal instruments *"Mercado no intermediado"*

##

-   **Fixed income (Renta fija)**: provides returns in the form of regular interest payments and repayments of the principal

    -   Títulos de tesorería (TES)

        -   Debt securities issued by the national government and administered by the Banco de la República.

        -   The national government use this instrument to finance its activities

    -   Certificados de Depósito a Término (CDT)

##

```{r}
# Data
# https://www.banrep.gov.co/es > 
# Estadísticas económicas >
# ¡Nuevo portal de estadísticas económicas! >
# Series de estadísticas económicas >
# Catálogo >
# Tasas de interés y sector financiero >
# Tasas de interés >
# Tasas Cero Cupón TES
## Use Tasa de interés Cero Cupón, Títulos de Tesorería (TES), pesos

# Import ----
tes_daily_t_1_5_10 <- read_csv(file = "000_data/008_tasa_cero_cupon_titulos_de_tesoreria.csv") |> 
  # check column names
  select(`Periodo(MMM DD, AAAA)`,
         `Tasa de interés Cero Cupón, Títulos de Tesorería (TES), pesos - 1 año`,
         `Tasa de interés Cero Cupón, Títulos de Tesorería (TES), pesos - 5 años`,
         `Tasa de interés Cero Cupón, Títulos de Tesorería (TES), pesos - 10 años`)

# Tidy ----
tes_daily_t_1_5_10_clean <- tes_daily_t_1_5_10 |> 
  set_names(nm = c("date", 
                   "tes_pesos_1", "tes_pesos_5", "tes_pesos_10")) |> 
  pivot_longer(cols = starts_with(match = "tes"), 
               names_to = "tes_type", 
               values_to = "interest_rate") |> 
  mutate(tes_type = case_when(
    tes_type == "tes_pesos_1"  ~ "TES, pesos 1 year",
    tes_type == "tes_pesos_5"  ~ "TES, pesos 5 years",
    tes_type == "tes_pesos_10" ~ "TES, pesos 10 years"))
```

```{r}
#| label: fig-cero-cupon-tes-1-5-10-anos-col
#| fig-cap: Interest rates TES zero coupon (COP, 1, 5 and 10 years)

# Visualization ----
tes_daily_t_1_5_10_clean |> 
  ggplot(aes(x = date,
             y = interest_rate, 
             group = tes_type)) +
  geom_line(aes(color = tes_type)) +
  scale_y_continuous(labels = label_number(suffix = "%", 
                                           accuracy = 1)) +
  scale_color_manual(values = umng_palette[1:3]) +
  labs(x        = NULL,
       y        = "Percent",
       color = NULL,
       subtitle = str_glue("Periodicity: Daily
                           Period: {first(tes_daily_t_1_5_10_clean$date)} - {last(tes_daily_t_1_5_10_clean$date)}"),
       caption  = str_glue("Source: Banco de la Republica
                             Last update date: {last(tes_daily_t_1_5_10_clean$date)}")) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10),
    legend.position = "bottom"
  )
```

## {.smaller} 

- TES zero coupon

    - Actual bond price: $P$
    - Nominal value: $NV$ where $P < NV$
    - Term: $n$
    - Interest rate: $i$

$$\begin{split}
   P & = \frac{NV}{(1 + i)^n} \\
   (1 + i)^n & = \frac{NV}{P} \\
   i & = \left( \frac{NV}{P} \right)^{\frac{1}{n}} - 1 \\
  \end{split}$$

## {.smaller} 

- An example: TES zero coupon

    - Actual bond price: $P = 600000$
    - Nominal value: $NV = 1000000$
    - Term: $n = 10$
    
$$\begin{split}
  i & = \left( \frac{NV}{P} \right)^{\frac{1}{n}} - 1 \\
    & = \left( \frac{1000000}{600000} \right)^{\frac{1}{10}} - 1 \\
    & \approx 0.0524 \text{ (5.24 \%)}  
  \end{split}$$

## {.smaller} 

-   **Equity (Renta variable)**:

    -   Shares/Stocks

        -   These instruments are issued by companies to raise funds from the general public

        -   They represent a fractional ownership in the company that issue them
        
        -   If the company does well, its value may increase, and the value of your shares may also rise. You can then sell your shares for a profit. If the company performs poorly, the value of your shares may decrease, and you could lose money.
        
        -   Some companies share their profits with shareholders by paying dividends. 
        
        -   The decision to pay dividends or reinvest the profist rests with the company's board of directors. Whether a company pays dividends or reinvests its profits depends on its specific circumstances and strategic goals.

#

```{r}
# Import ----
msci_colcap <- read_csv(file = '000_data/008_imbc_colcap_iqy.csv')

# Tidy ----
msci_colcap_clean <- msci_colcap |>
  set_names(nm = c("date", "colcap"))
```

```{r}
#| label: fig-msci-colcap-index-col
#| fig-cap: MSCI COLCAP Index

# Visualization ----
msci_colcap_clean |>
  ggplot(aes(x = date,
             y = colcap)) +
  geom_hline(yintercept = 1000,
             color = umng_palette[3]) +
  geom_line(color = umng_palette[1],
            show.legend = FALSE) +
  labs(x        = NULL,
       y        = NULL,
       color = NULL,
       subtitle = str_glue("Periodicity: Daily
                           Period: {msci_colcap_clean$date[1]} - {msci_colcap_clean$date[nrow(msci_colcap_clean)]}"),
       caption  = str_glue("Source: Bolsa de Valores de Colombia (BVC)
                             Last update date: {last(msci_colcap_clean$date)}")) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10),
    legend.position = "bottom"
  )
```

##

```{r}
# Data
# https://www.blackrock.com/co >
# Productos >
# iShares ETFs >
# Fondo Bursátil iShares MSCI COLCAP >

## Direct link
## https://www.blackrock.com/co/productos/251708/ishares-fondo-burstil-ishares-colcap-fund

# Import
current_index_basket_msci_colcap <- read_excel(path = "000_data/008_current_index_basket_msci_colcap_index_col.xlsx",
                                                     sheet = "Canasta_ICAP", 
                                                     range = "B6:F29")

# Tidy
current_index_basket_msci_colcap_clean <- current_index_basket_msci_colcap |> 
  clean_names() |> 
  select(nemotecnico_ticker,  
         razon_social_emisor_issuer,
         percent_de_part) |> 
  set_names(nm = c("ticker", "legal_name", "percent")) |> 
  mutate(percent = percent * 100) |> 
  arrange(desc(percent))

# Helpers
n_rows <- nrow(current_index_basket_msci_colcap_clean) 
first_rows <- ceiling(n_rows / 2)
last_rows <- n_rows - first_rows

last_update_current_index_basket_msci_colcap <- ymd("2025-08-13")
```

```{r}
#| label: tbl-current-index-basket-msci-colcap-index-col-1
#| tbl-cap: Current MSCI COLCAP Index basket

# Table  
table_current_index_basket_msci_colcap_clean_1 <- current_index_basket_msci_colcap_clean |> 
  slice_head(n = first_rows) |> 
  gt() |>
  fmt_number(
    columns = percent,
    decimals = 2
  ) |>
  cols_label(
    ticker = "Ticker",
    legal_name = "Legal Name", 
    percent = "Percent"
  ) |>
  tab_source_note(
    source_note = "Source: BVC - Bolsa de Valores de Colombia"
  ) |>
  tab_source_note(
    source_note = str_glue("Last update: {last_update_current_index_basket_msci_colcap}")
  ) |> 
  tab_options(
    row.striping.include_table_body = TRUE,
    table.border.top.width = px(3),
    table.border.top.color = "black",
    column_labels.border.bottom.width = px(2),
    column_labels.border.bottom.color = "black",
    table.border.bottom.width = px(3),
    table.border.bottom.color = "black",
    table.font.size = px(18)
  )

table_current_index_basket_msci_colcap_clean_1
```

##

```{r}
#| label: tbl-current-index-basket-msci-colcap-index-col-2
#| tbl-cap: Current MSCI COLCAP Index basket

# Table  
table_current_index_basket_msci_colcap_clean_2 <- current_index_basket_msci_colcap_clean |> 
  slice_tail(n = last_rows) |> 
  gt() |>
  fmt_number(
    columns = percent,
    decimals = 2
  ) |>
  cols_label(
    ticker = "Ticker",
    legal_name = "Legal Name", 
    percent = "Percent"
  ) |>
  tab_source_note(
    source_note = "Source: BVC - Bolsa de Valores de Colombia"
  ) |>
  tab_source_note(
    source_note = str_glue("Last update: {last_update_current_index_basket_msci_colcap}")
  ) |> 
  tab_options(
    row.striping.include_table_body = TRUE,
    table.border.top.width = px(3),
    table.border.top.color = "black",
    column_labels.border.bottom.width = px(2),
    column_labels.border.bottom.color = "black",
    table.border.bottom.width = px(3),
    table.border.bottom.color = "black",
    table.font.size = px(18)
  )

table_current_index_basket_msci_colcap_clean_2
```

##

```{r}
# Import
icolcap_cl <- tq_get("ICOLCAP.CL",
       get = "stock.prices",
       # Only avaible from this date
       from = "2013-05-02")

# Tidy
icolcap_cl_clean <- icolcap_cl |> 
  select(date, close) |> 
  set_names(nm = c("date", "icolcap_cl"))
```

```{r}
#| label: fig-ishares-msci-colcap-etf
#| fig-cap: iShares MSCI COLCAP Exchange-Traded Fund (ETF)

# Visualization
icolcap_cl_clean |>
  ggplot(aes(x = date,
             y = icolcap_cl)) +
  geom_line(color = umng_palette[1],
            show.legend = FALSE) +
  labs(x        = NULL,
       y        = "Colombian persos (COP)",
       color = NULL,
       subtitle = str_glue("Periodicity: Daily
                           Period: {icolcap_cl_clean$date[1]} - {icolcap_cl_clean$date[nrow(icolcap_cl_clean)]}"),
       caption  = str_glue("Source: Yahoo Finance
                             Last update date: {last(icolcap_cl_clean $date)}")) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10),
    legend.position = "bottom"
  )
```

# Banking market or *"Mercado Intermediado"*

## {.smaller} 

-   **What is a bank?**

    -   *"A bank is an institution **whose current operations** consist in granting loans and receiving deposits from the public"* [@freixas_microeconomics_2008, p. 1]

        -   Therefore the core activities of banks are related to deposits and loans

-   **What functions banks perform?** [@freixas_microeconomics_2008, p. 2]

    -   Offering liquidity and payment services
    -   Transforming assets
    -   Managing risks
    -   Processing information and monitoring borrowers

## {.smaller}

-   **Offering liquidity and payment services** [@freixas_microeconomics_2008, p. 2-4]

    -   Banks offer short term credits to companies and individuals and have created networks that facilitate the transfer of funds between the bank accounts of economic agents

-   **Transforming assets** [@freixas_microeconomics_2008, p. 5-6]

    -   **Convenience of denomination**: banks collect small deposits to offer large loans
    -   **Quality transformation**: bank deposits offer better risk-return characteristics than direct investments
    -   **Maturity transformation**: banks transform securities with short maturities, offered to depositors, into securities with long maturities, which borrowers desire

## {.smaller}

-   **Managing risks**:

    -   **Credit risk**: it is related to the the probability that a loan is no repaid
    -   **Interest rate risk**: it is related to the difference between deposit rates, which change more, and lending rates, which are more stable
    -   **Liquidity risk**: it is related to the difficulty a bank has in selling a loan compared to the ease with which a depositor withdraws his savings

-   In the case of the Colombian context the framework to manage risks is known as **Sistema Integral de Administración de Riesgos (SIAR)** [@superintendencia_financiera_de_colombia_circular_2021, chapter 31]

##

```{r}
# Import ----
bank <- read_excel("000_data/008_entidades_vigiladas_superintendencia_financiera_colombia.xls",
                         sheet = 1, 
                         range = "B15:P462") 

# Tidy
bank_clean <- bank |> 
  slice(-1) |> 
  select(1:2,5:6) |> 
  set_names(nm = c("Type", "Code", "Abbreviate Name", "NIT")) |> 
  filter(Type == 1)

# Helpers
n_row_bank_clean <- nrow(bank_clean)
cut <- round(n_row_bank_clean / 2)
last_update_bank <- ymd("2025-08-13")
```

```{r}
#| label: tbl-supervised-banking-establishments-1-col
#| tbl-cap: Supervised banking establishments

# Table
table_bank_clean_1 <- bank_clean |> 
  slice(1:cut) |> 
  gt() |> 
  tab_source_note(
    source_note = "Source: Superintendencia Financiera de Colombia"
  ) |>
  tab_source_note(
    source_note = str_glue("Last update: {last_update_bank}")
  ) |> 
  tab_options(
    row.striping.include_table_body = TRUE,
    table.border.top.width = px(3),
    table.border.top.color = "black",
    column_labels.border.bottom.width = px(2),
    column_labels.border.bottom.color = "black",
    table.border.bottom.width = px(3),
    table.border.bottom.color = "black",
    table.font.size = px(16), 
    data_row.padding = px(5)
  )

table_bank_clean_1
```

##

```{r}
#| label: tbl-supervised-banking-establishments-2-col
#| tbl-cap: Supervised banking establishments

# Table
table_bank_clean_2 <- bank_clean |> 
  slice((cut+1):n_row_bank_clean) |> 
  gt() |> 
  tab_source_note(
    source_note = "Source: Superintendencia Financiera de Colombia"
  ) |>
  tab_source_note(
    source_note = str_glue("Last update: {last_update_bank}")
  ) |> 
  tab_options(
    row.striping.include_table_body = TRUE,
    table.border.top.width = px(3),
    table.border.top.color = "black",
    column_labels.border.bottom.width = px(2),
    column_labels.border.bottom.color = "black",
    table.border.bottom.width = px(3),
    table.border.bottom.color = "black",
    table.font.size = px(16), 
    data_row.padding = px(5)
  )

table_bank_clean_2
```

##

```{r}
# Import ----
i_rate_deposit_lending <- wbstats::wb_data(country = "COL",
                                           # Deposit interest rate (%)
                                           ## FR.INR.DPST
                                           # Lending interest rate (%)
                                           ## FR.INR.LEND
                                           indicator    = c("FR.INR.DPST", 
                                                            "FR.INR.LEND"),
                                           return_wide  = FALSE) 

# Tidy 
i_rate_deposit_lending_clean <- i_rate_deposit_lending |>
  filter(!is.na(value))
```

```{r}
#| label: fig-deposit-lending-interest-rate-col
#| fig-cap: Deposit and lending interest rate in Colombia

# Visualization
i_rate_deposit_lending_clean |> 
  ggplot(aes(x = date,
             y = value)) +
  geom_line(aes(color = indicator)) +
  scale_y_continuous(labels = label_number(suffix = "%")) +
  scale_color_manual(values = umng_palette[1:2]) +
  labs(x = NULL,
       y = NULL,
       color = NULL,
       subtitle = str_glue("Period: {min(i_rate_deposit_lending$date)}-{max(i_rate_deposit_lending$date)}"),
       caption = str_glue("Source: World Development Indicators (WDI) - World Bank
                          Last update: {unique(i_rate_deposit_lending$last_updated)}")) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10),
    legend.position = "bottom"
  )
```

# Acknowledgments

##

-   To my family that supports me

-   To the taxpayers of Colombia and the [**UMNG students**](https://www.umng.edu.co/estudiante) who pay my salary

-   To the [**Business Science**](https://www.business-science.io/) and [**R4DS Online Learning**](https://www.rfordatasci.com/) communities where I learn [**R**](https://www.r-project.org/about.html) and [**$\pi$-thon**](https://www.python.org/about/)

-   To the [**R Core Team**](https://www.r-project.org/contributors.html), the creators of [**RStudio IDE**](https://posit.co/download/rstudio-desktop/), [**Positron**](https://positron.posit.co/), [**Quarto**](https://quarto.org/) and the authors and maintainers of the packages [**tidyverse**](https://CRAN.R-project.org/package=tidyverse), [**scales**](https://CRAN.R-project.org/package=scales), [**readxl**](https://CRAN.R-project.org/package=readxl),
[**janitor**](https://CRAN.R-project.org/package=janitor),
[**gt**](https://CRAN.R-project.org/package=gt),
[**tidyquant**](https://CRAN.R-project.org/package=tidyquant),
[**wbstats**](https://CRAN.R-project.org/package=wbstats) and [**tinytex**](https://CRAN.R-project.org/package=tinytex) for allowing me to access these tools without paying for a license

# References {.allowframebreaks}