---
title: "Financial Market II"
author: 
  - Luis Francisco Gómez López
institute: 
  - FAEDIS
date: last-modified
format: 
  beamer:
    colortheme: dolphin
    fonttheme: structurebold
    theme: AnnArbor
    link-citations: true
    colorlinks: true
    toc: true
    slide-level: 2
    section-titles: false
    include-in-header:
      file: 000_tex_files/preamble.tex
fig-cap-location: bottom
tbl-cap-location: top
knitr: 
  opts_chunk: 
    echo: false
    warning: false
    fig-align: center
    out-width: 85%
lang: en
bibliography: intro_econ.bib
---

```{r}
#| label: libraries

library(DiagrammeR)
library(tidyverse)
library(wbstats)
library(tidyquant)
library(ggrepel)
library(lubridate)
library(knitr)
library(kableExtra)
library(readxl)
```

# Please Read Me

## 

-   Check the message **Welcome greeting** published in the News Bulletin Board.

-   Dear student please edit your profile uploading a photo where your face is clearly visible.

-   The purpose of the virtual meetings is to answer questions and not to make a summary of the study material.

-   If you want to participate, please fill out the following survey: Primer corte 30% \> Learning Activities \> Tu opinión sobre la economía colombiana

-   This presentation is based on [@cardenas_introduccion_2020, Chapter 8]

# Purpose

## 

Analyze the functioning of the financial market, identifying the types of intermediaries and instruments that are part of it

# Principal instruments *"Mercado no intermediado"*

## 

-   **Fixed income (Renta fija)**: provides returns in the form of regular interest payments and repayments of the principal

    -   Títulos de tesorería (TES)

        -   Debt securities issued by the national government and administered by the Banco de la República.

        -   The national government use this instrument to finance its activities

    -   Certificados de Depósito a Término (CDT)

## 

```{r}
#| label: fig-cero-cupon-tes-1-5-10-anos-col
#| fig-cap: Interest rates TES zero coupon (COP, 1, 5 and 10 years)


# https://www.banrep.gov.co/es > 
# Estadísticas económicas >
# ¡Nuevo portal de estadísticas económicas! >
# Series de estadísticas económicas >
# Catálogo >
# Tasas de interés y sector financiero >
# Tasas de interés >
# Tasas Cero Cupón TES
## Use Tasa de interés Cero Cupón, Títulos de Tesorería (TES), pesos 

tes_daily_t_1_5_10 <- read_csv(file = "000_data/008_tasa_cero_cupon_titulos_de_tesoreria.csv") |> 
  # check column names
  select(`Periodo(MMM DD, AAAA)`,
         `Tasa de interés Cero Cupón, Títulos de Tesorería (TES), pesos - 1 año`,
         `Tasa de interés Cero Cupón, Títulos de Tesorería (TES), pesos - 5 años`,
         `Tasa de interés Cero Cupón, Títulos de Tesorería (TES), pesos - 10 años`) |> 
  set_names(nm = c("date", "tes_pesos_1", "tes_pesos_5", "tes_pesos_10"))

tes_daily_t_1_5_10 |>
  pivot_longer(cols = starts_with(match = "tes"), 
               names_to = "tes_type", 
               values_to = "interest_rate") |> 
  mutate(tes_type = case_when(
    tes_type == "tes_pesos_1"  ~ "TES, pesos 1 year",
    tes_type == "tes_pesos_5"  ~ "TES, pesos 5 years",
    tes_type == "tes_pesos_10" ~ "TES, pesos 10 years")) |> 
  ggplot(aes(x = date,
             y = interest_rate, 
             group = tes_type)) +
  geom_line(aes(color = tes_type)) +
  scale_y_continuous(labels = scales::label_number(suffix = "%", 
                                                   accuracy = 1)) +
  scale_color_tq() +
  labs(x        = NULL,
       y        = "Percent",
       color = NULL,
       subtitle = str_glue("Periodicity: Daily
                           Period: {first(tes_daily_t_1_5_10$date)} - {last(tes_daily_t_1_5_10$date)}"),
       caption  = str_glue("Source: Banco de la Republica
                             Last update date: {last(tes_daily_t_1_5_10$date)}")) +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        legend.position   = "bottom",
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"))
```

## 

- TES zero coupon

    - Actual bond price: $P$
    - Nominal value: $NV$ where $P < NV$
    - Term: $n$
    - Interest rate: $i$

$$\begin{split}
   P & = \frac{NV}{(1 + i)^n} \\
   (1 + i)^n & = \frac{NV}{P} \\
   i & = \left( \frac{NV}{P} \right)^{\frac{1}{n}} - 1 \\
  \end{split}$$

## 

- An example: TES zero coupon

    - Actual bond price: $P = 600000$
    - Nominal value: $NV = 1000000$
    - Term: $n = 10$
    
$$\begin{split}
  i & = \left( \frac{NV}{P} \right)^{\frac{1}{n}} - 1 \\
    & = \left( \frac{1000000}{600000} \right)^{\frac{1}{10}} - 1 \\
    & \approx 0.0524 \text{ (5.24 \%)}  
  \end{split}$$

## 

-   **Equity (Renta variable)**:

    -   Shares/Stocks

        -   These instruments are issued by companies to raise funds from the general public

        -   They represent a fractional ownership in the company that issue them
        
        -   If the company does well, its value may increase, and the value of your shares may also rise. You can then sell your shares for a profit. If the company performs poorly, the value of your shares may decrease, and you could lose money.
        
        -   Some companies share their profits with shareholders by paying dividends. 
        
        -   The decision to pay dividends or reinvest the profist rests with the company's board of directors. Whether a company pays dividends or reinvests its profits depends on its specific circumstances and strategic goals.


## 

```{r}
#| label: fig-msci-colcap-index-col
#| fig-cap: MSCI COLCAP Index

# https://www.banrep.gov.co/es > 
# Estadísticas económicas >
# Gráficas interactivas > 
# CATÁLOGO DE SERIES >
# Tasas de interés y sector financiero > 
# Mercado accionario > 
# Índice COLCAP > 
# Diaria > 
# DESCARGAR >
# Descargar datos en CSV
msci_colcap <- read_csv(file = '000_data/008_imbc_colcap_iqy.csv') |>
  set_names(nm = c("date", "colcap"))

msci_colcap |>
  ggplot(aes(x = date,
             y = colcap)) +
  geom_line(color = unname(palette_light()[1]),
            show.legend = FALSE) +
  labs(x        = NULL,
       y        = NULL,
       color = NULL,
       subtitle = str_glue("Periodicity: Daily"),
       caption  = str_glue("Source: Bolsa de Valores de Colombia (BVC)
                             Last update date: {last(msci_colcap$date)}")) +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        legend.position   = "bottom",
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"))
```

# Banking market or *"Mercado Intermediado"*

## 

-   **What is a bank?**

    -   *"A bank is an institution **whose current operations** consist in granting loans and receiving deposits from the public"* [@freixas_microeconomics_2008, p 1]

        -   Therefore the core activities of banks are related to deposits and loans

-   **What functions banks perform?** [@freixas_microeconomics_2008, p 2]

    -   Offering liquidity and payment services
    -   Transforming assets
    -   Managing risks
    -   Processing information and monitoring borrowers

##

-   **Offering liquidity and payment services** [@freixas_microeconomics_2008, p 2-4]

    -   Banks offer short term credits to companies and individuals and have created networks that facilitate the transfer of funds between the bank accounts of economic agents

-   **Transforming assets** [@freixas_microeconomics_2008, p 5-6]

    -   **Convenience of denomination**: banks collect small deposits to offer large loans
    -   **Quality transformation**: bank deposits offer better risk-return characteristics than direct investments
    -   **Maturity transformation**: banks transform securities with short maturities, offered to depositors, into securities with long maturities, which borrowers desire

##

-   **Managing risks**:

    -   **Credit risk**: it is related to the the probability that a loan is no repaid
    -   **Interest rate risk**: it is related to the difference between deposit rates, which change more, and lending rates, which are more stable
    -   **Liquidity risk**: it is related to the difficulty a bank has in selling a loan compared to the ease with which a depositor withdraws his savings

-   In the case of the Colombian context the framework to manage risks is known as **Sistema Integral de Administración de Riesgos (SIAR)** [@superintendencia_financiera_de_colombia_circular_2021, chapter 31]

##

```{r}
#| label: tbl-supervised-banking-establishments-1-col
#| tbl-cap: Supervised banking establishments

# https://www.superfinanciera.gov.co > 
# Entidades vigiladas y controladas > 
# Entidades vigilidas por la Superfinanciera >
# Entidades vigiladas por la Superintendencia Financiera de Colombia >
# Lista general de entidades vigiladas por la Superintendencia Financiera de Colombia > 
# Descargue el archivo en formato Excel
bank_table <- read_excel("000_data/008_entidades_vigiladas_superintendencia_financiera_colombia.xls",
                         sheet = 1, 
                         range = "B15:P445") |> 
  slice(-1) |> 
  select(1:2,5:6) |> 
  set_names(nm = c("Type", "Code", "Abbreviate Name", "NIT")) |> 
  filter(Type == 1)

cut <- round(nrow(bank_table) / 2)

bank_table |>
  slice(1:cut) |> 
  kable(format = "latex", booktabs = TRUE, escape = FALSE, 
        digits = 2, linesep = "") |>
  kable_styling(font_size = 7) |>
  row_spec(row = 0, bold = TRUE) |> 
  # Manually update the last update
  footnote(general       = c('Source: Superintendencia Financiera de Colombia', 
                             'Last update: 2024-07-17'),
           general_title = "",
           threeparttable = TRUE)
```

##

```{r}
#| label: tbl-supervised-banking-establishments-2-col
#| tbl-cap: Supervised banking establishments

bank_table |>
  slice((cut+1):nrow(bank_table)) |> 
  kable(format = "latex", booktabs = TRUE, escape = FALSE, 
        digits = 2, linesep = "") |>
  kable_styling(font_size = 7) |>  
  row_spec(row = 0, bold = TRUE) |> 
  # Manually update the last update
  footnote(general       = c('Source: Superintendencia Financiera de Colombia', 
                             'Last update: 2024-07-17'),
           general_title = "",
           threeparttable = TRUE)
```

##

```{r}
#| label: fig-deposit-lending-interest-rate-col
#| fig-cap: Deposit and lending interest rate in Colombia

i_rate_deposit_lending <- wbstats::wb_data(country = "COL",
                                   # Deposit interest rate (%)
                                   ## FR.INR.DPST
                                   # Lending interest rate (%)
                                   ## FR.INR.LEND
                                   indicator    = c("FR.INR.DPST", "FR.INR.LEND"),
                                   return_wide  = FALSE) |>
  filter(!is.na(value))

i_rate_deposit_lending |> 
  ggplot(aes(x = date,
             y = value)) +
  geom_line(aes(color = indicator)) +
  scale_y_continuous(labels = scales::label_number(suffix = "%")) +
  scale_color_tq() +
  labs(x = NULL,
       y = NULL,
       color = NULL,
       subtitle = str_glue("Period: {min(i_rate_deposit_lending$date)}-{max(i_rate_deposit_lending$date)}"),
       caption = str_glue("Source: World Development Indicators (WDI) - World Bank
                          Last update: {unique(i_rate_deposit_lending$last_updated)}")) +
  theme(panel.border    = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        legend.position   = "bottom",
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"))
```

# Acknowledgments

-   To my family that supports me

-   To the taxpayers of Colombia and the [**UMNG students**](https://www.umng.edu.co/estudiante) who pay my salary

-   To the [**Business Science**](https://www.business-science.io/) and [**R4DS Online Learning**](https://www.rfordatasci.com/) communities where I learn [**R**](https://www.r-project.org/about.html) and [**$\pi$-thon**](https://www.python.org/about/)

-   To the [**R Core Team**](https://www.r-project.org/contributors.html), the creators of [**RStudio IDE**](https://rstudio.com/products/rstudio/), [**Quarto**](https://quarto.org/) and the authors and maintainers of the packages [**tidyverse**](https://CRAN.R-project.org/package=tidyverse), [**wbstats**](https://CRAN.R-project.org/package=wbstats), [**tidyquant**](https://CRAN.R-project.org/package=tidyquant), [**ggrepel**](https://CRAN.R-project.org/package=ggrepel), [**lubridate**](https://CRAN.R-project.org/package=lubridate), [**knitr**](https://CRAN.R-project.org/package=knitr), [**kableExtra**](https://CRAN.R-project.org/package=kableExtra), [**readxl**](https://CRAN.R-project.org/package=readxl), and [**tinytex**](https://CRAN.R-project.org/package=tinytex) for allowing me to access these tools without paying for a license

-   To the [**Linux kernel community**](https://www.kernel.org/category/about.html) for allowing me the possibility to use some [**Linux distributions**](https://static.lwn.net/Distributions/) as my main [**OS**](https://en.wikipedia.org/wiki/Operating_system) without paying for a license

# References {.allowframebreaks}