---
title: "Financial Market I"
author: 
  - Luis Francisco Gómez López
institute: 
  - FAEDIS
date: last-modified
format: 
  beamer:
    colortheme: dolphin
    fonttheme: structurebold
    theme: AnnArbor
    link-citations: true
    colorlinks: true
    toc: true
    slide-level: 2
    section-titles: false
    include-in-header:
      file: 000_tex_files/preamble.tex
fig-cap-location: bottom
tbl-cap-location: top
knitr: 
  opts_chunk: 
    echo: false
    warning: false
    fig-align: center
    out-width: 85%
lang: en
bibliography: intro_econ.bib
---

```{r}
#| label: libraries

library(DiagrammeR)
library(tidyverse)
library(wbstats)
library(tidyquant)
library(ggrepel)
library(lubridate)
library(knitr)
library(kableExtra)
library(readxl)
```

# Please Read Me

## 

-   Check the message **Welcome greeting** published in the News Bulletin Board.

-   Dear student please edit your profile uploading a photo where your face is clearly visible.

-   The purpose of the virtual meetings is to answer questions and not to make a summary of the study material.

-   If you want to participate, please fill out the following survey: Primer corte 30% \> Learning Activities \> Tu opinión sobre la economía colombiana

-   This presentation is based on [@cardenas_introduccion_2020, Chapter 8]

# Purpose

## 

Analyze the functioning of the financial market, identifying the types of intermediaries and instruments that are part of it

# Structure of the Colombian financial system

## 

-   The financial system is the set of entities whose main function is to channel the money of savers to those who wish to make investments

    -   We are only going to focus on supervised financial entities. Therefore entities outside the law or unregulated are not going to be analyzed

    -   Also ponzi schemes, pyramids or unregulated investment scheme are not analyzed.

        -   However if you are interested in this topic check out [@carvajal_ponzi_2009] and [@hofstetter_ponzi_2018]

## 

```{dot}
//| label: fig-financial-system-structure-supervised-col
//| fig-cap: Financial system structure by supervised entities [@congreso_de_colombia_decreto_1993]
//| fig-width: 4.5
//| fig-height: 2.7

digraph boxes_and_circles {

  graph [rankdir = LR]

  node [shape = circle, color = "#2C3E50"]
 
  "Sector Financiero";

  node [shape = oval, color = "#E31A1C"]

  "Establecimientos\nde crédito";
  "Sociedades de\nservicios financieros";
  "Sociedades de\nCapitalización";
  "Entidades Aseguradores\ne Intermediarios";
  
  node [shape = oval, color = "#18BC9C"]
  
  "Establecimientos bancarios";
  "Corporaciones Financieras";
  "Compañías de Financiamiento";
  "Cooperativas Financieras";
  "Sociedades fiduciarias";
  "Almacenes generales de depósito";
  "Sociedades administradoras de\nFondos de Pensiones y de Cesantías";
  "Entidades aseguradoras";
  "Intermediarios de seguros";
  "Intermediarios de reaseguros";
  
  "Sector Financiero" -> "Establecimientos\nde crédito";
  "Sector Financiero" -> "Sociedades de\nservicios financieros";
  "Sector Financiero" -> "Sociedades de\nCapitalización";
  "Sector Financiero" -> "Entidades Aseguradores\ne Intermediarios";
  
  "Establecimientos\nde crédito" -> "Establecimientos bancarios";
  "Establecimientos\nde crédito" -> "Corporaciones Financieras";
  "Establecimientos\nde crédito" -> "Compañías de Financiamiento";
  "Establecimientos\nde crédito" -> "Cooperativas Financieras";
  "Sociedades de\nservicios financieros" -> "Sociedades fiduciarias";
  "Sociedades de\nservicios financieros" -> "Almacenes generales de depósito";
  "Sociedades de\nservicios financieros" -> "Sociedades administradoras de\nFondos de Pensiones y de Cesantías";
  "Entidades Aseguradores\ne Intermediarios" -> "Entidades aseguradoras";
  "Entidades Aseguradores\ne Intermediarios" -> "Intermediarios de seguros";
  "Entidades Aseguradores\ne Intermediarios" -> "Intermediarios de reaseguros";
}
```

## 

```{dot}
//| label: fig-financial-system-structure-markets-col
//| fig-cap: Financial system structure by markets entities [@cardenas_introduccion_2020, Chapter 8, p. 264]
//| fig-width: 4.5
//| fig-height: 2.7

digraph boxes_and_circles {

  graph [rankdir = LR]

  node [shape = circle, color = "#2C3E50"]
 
  "Sector Financiero";

  node [shape = oval, color = "#E31A1C"]

  "Mercado monetario";
  "Mercado de capitales";
  "Mercado de divisas";
  "Otros mercados financieros";
  
  node [shape = oval, color = "#18BC9C"]
  
  "Bancario";
  "No bancario";
  
  node [shape = oval, color = "#CCBE93"]
  
  "Bancos";
  "Otros intermediarios";
  "Acciones";
  "Renta fija";
  "Derivados";
  "Otros mercados";
  
  "Sector Financiero" -> "Mercado monetario";
  "Sector Financiero" -> "Otros mercados financieros";
  "Sector Financiero" -> "Mercado de divisas";
  "Sector Financiero" -> "Mercado de capitales";
  
  "Mercado de capitales" -> "Bancario";
  "Mercado de capitales" -> "No bancario";
  
  "Bancario" -> "Bancos";
  "Bancario" -> "Otros intermediarios";
  "No bancario" -> "Acciones";
  "No bancario" -> "Renta fija";
  "No bancario" -> "Derivados";
  "No bancario" -> "Otros mercados";
}
```

# Financial depth

## 

-   *"Financial depth captures the financial sector relative to the economy. It is the size of banks, other financial institutions, and financial markets in a country, taken together and compared to a measure of economic output"* [@world_bank_financial_2016]

-   How it is measure using quantity indicators?[^1]

    -   **Domestic credit to private sector (% of GDP)**

    -   **Market capitalization of listed domestic companies (% of GDP)**

        -   Share price times the number of shares outstanding (including their several classes) for listed domestic companies

[^1]: These indicators doesn't measure the quality of financial depth

## 

-   According to the literature the *"evidence suggests that both financial intermediaries and markets matter for growth and that reverse causality alone is not driving this relationship"* [@levine_chapter_2005, p. 866]

##

- The financial system provides 5 broad categories of financial functions that can affect resource allocation and economic growth [@levine_chapter_2005, p. 870]

    - **Producing information and allocating capital** [@levine_chapter_2005, pp. 870-872]
    
        - Financial systems enhance the collection and processing of information about potential investments. 
        
        - This allows for better assessment of projects and more efficient allocation of resources that foster economic growth. 

##

- The financial system provides 5 broad categories of financial functions that can affect resource allocation and economic growth [@levine_chapter_2005, p. 870]

    - **Monitoring firms and exerting corporate governance** [@levine_chapter_2005, pp. 872-875]
    
        - Financial institutions, particularly banks and investors, play a role in monitoring the activities of firms they finance. 
        
        - This helps to ensure that managers act in the best interests of shareholders and that capital is used efficiently. 

##

- The financial system provides 5 broad categories of financial functions that can affect resource allocation and economic growth [@levine_chapter_2005, p. 870]

    - **Risk amelioration** [@levine_chapter_2005, pp. 875-879]
    
        - Financial systems reduce the negative impact of risk by enabling diversification, risk transfer, and liquidity. 
        
        - This encourages investment and innovation, fostering economic growth by making risky but productive ventures more viable.

##

- The financial system provides 5 broad categories of financial functions that can affect resource allocation and economic growth [@levine_chapter_2005, p. 870]

    - **Pooling of savings** [@levine_chapter_2005, pp. 879-880]
    
        - The financial system gathers dispersed funds into large capital pools, enabling substantial investments that individuals alone couldn't make. 
        
        - This process reduces transaction costs, facilitates large-scale projects, and efficiently allocates capital, driving economic growth by channeling savings into productive ventures.

##

- The financial system provides 5 broad categories of financial functions that can affect resource allocation and economic growth [@levine_chapter_2005, p. 870]

    - **Easing exchange** [@levine_chapter_2005, pp. 880-881]
    
        - The financial system reduce transaction costs and facilite trade allowing for smoother economic interactions.
        
        - This increases efficiency, promotes specialization, and expands markets, contributing to economic growth.

## 

```{r}
#| label: fig-domestic-credit-private-sector-gdp-pc-col
#| fig-cap: Financial depth vs Gross Domestic Product per-capita

start_date <- 2023
end_date <- 2023

credit_private_long <- wbstats::wb_data(country = "countries_only",
                                       # Domestic credit to private sector (% of GDP)
                                       ## FS.AST.PRVT.GD.ZS
                                       #  GDP per capita, PPP (constant 2021 international $)
                                       ## NY.GDP.PCAP.PP.KD
                                       indicator    = c("FS.AST.PRVT.GD.ZS", "NY.GDP.PCAP.PP.KD"),
                                       start_date = start_date,
                                       end_date = end_date,
                                       return_wide = FALSE)

credit_private_wide <- credit_private_long |> 
  select(iso2c, iso3c, country, date, indicator_id, value) |> 
  pivot_wider(id_cols = iso2c:date, 
              names_from = indicator_id,
              values_from = value) |> 
  filter(!is.na(FS.AST.PRVT.GD.ZS), !is.na(NY.GDP.PCAP.PP.KD)) |> 
  select(iso3c, country, date,
         FS.AST.PRVT.GD.ZS, NY.GDP.PCAP.PP.KD)

credit_private_wide_min_max_col <- credit_private_wide |> 
  filter(FS.AST.PRVT.GD.ZS == min(FS.AST.PRVT.GD.ZS) |
         FS.AST.PRVT.GD.ZS == max(FS.AST.PRVT.GD.ZS) |
         iso3c == "COL")

credit_private_wide |> 
  ggplot(aes(x = FS.AST.PRVT.GD.ZS,
             y = NY.GDP.PCAP.PP.KD)) +
  geom_point(color = "black",
             fill = unname(palette_light()[2]),
             shape = 21,
             show.legend = FALSE) +
  geom_label_repel(data = filter(credit_private_wide, iso3c %in% credit_private_wide_min_max_col$iso3c),
                   aes(label = country),
                   seed = 1234,
                   size = 3,
                   min.segment.length = 0,
                   box.padding = 1,
                   force = 2,
                   max.overlaps = Inf) +
  scale_x_log10(labels = scales::label_number()) +
  scale_y_log10(labels = scales::label_number()) +
  labs(x        = "Domestic credit to private sector (% of GDP)",
       y        = str_glue("GDP per capita, PPP
                           (constant 2021 international $)"),
       color = NULL,
       subtitle = str_glue("Period: {start_date}
                           Axis scales: Logarithmic Base 10
                           Number of territories: {nrow(credit_private_wide)}"),
       caption  = str_glue("Source: World Development Indicators (WDI) - World Bank
                             Last update date: {unique(credit_private_long$last_updated)}")) +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        legend.position   = "bottom",
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"))
```

## 

```{r}
#| label: fig-market-capitalization-gdp-pc-col
#| fig-cap: Financial depth vs Gross Domestic Product per-capita

start_date <- 2022
end_date <- 2022

market_cap_long <- wbstats::wb_data(country = "countries_only",
                               # Market capitalization of listed domestic companies (% of GDP)
                               ## CM.MKT.LCAP.GD.ZS
                               #  GDP per capita, PPP (constant 2021 international $)
                               ## NY.GDP.PCAP.PP.KD
                               indicator    = c("CM.MKT.LCAP.GD.ZS", "NY.GDP.PCAP.PP.KD"),
                               start_date = start_date,
                               end_date = end_date,
                               return_wide  = FALSE) 

market_cap_wide <- market_cap_long |>
  select(iso2c, iso3c, country, date, indicator_id, value) |>
  pivot_wider(id_cols = iso2c:date, 
              names_from = indicator_id,
              values_from = value) |> 
  filter(!is.na(CM.MKT.LCAP.GD.ZS), !is.na(NY.GDP.PCAP.PP.KD)) |>
  select(iso3c, country, date,
         CM.MKT.LCAP.GD.ZS, NY.GDP.PCAP.PP.KD)

market_cap_wide_min_max_col <- market_cap_wide |> 
  filter(CM.MKT.LCAP.GD.ZS == min(CM.MKT.LCAP.GD.ZS) |
         CM.MKT.LCAP.GD.ZS == max(CM.MKT.LCAP.GD.ZS) |
         iso3c == "COL")

market_cap_wide |>
  ggplot(aes(x = CM.MKT.LCAP.GD.ZS,
             y = NY.GDP.PCAP.PP.KD)) +
  geom_point(color = "black",
             fill = unname(palette_light()[2]),
             shape = 21,
             show.legend = FALSE) +
  geom_label_repel(data = filter(market_cap_wide, iso3c %in% market_cap_wide_min_max_col$iso3c),
                   aes(label = country),
                   seed = 1234,
                   size = 3,
                   min.segment.length = 0,
                   box.padding = 1,
                   force = 2,
                   max.overlaps = Inf) +
  scale_x_log10() +
  scale_y_log10(labels = scales::label_number()) +
  labs(x        = "Market capitalization of listed domestic companies (% of GDP)",
       y        = str_glue("GDP per capita, PPP
                           (constant 2021 international $)"),
       color = NULL,
       subtitle = str_glue("Period: {start_date}
                           Axis scales: Logarithmic Base 10
                           Number of territories: {nrow(market_cap_wide)}"),
       caption  = str_glue("Source: World Development Indicators (WDI) - World Bank
                             Last update date: {unique(market_cap_long$last_updated)}")) +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        legend.position   = "bottom",
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"))
```

# Uncertainty in financial markets

## 

-   Financing is essentially the exchange of a sum of money today for a promise to return more money in the future. Therefore it is not surprising that such exchange can be problematic

    -   **Information asymmetry**: in an exchange one party has more or better information than the other

        -   **Adverse selection**
        -   **Moral hazard**

## 

-   **Adverse selection** occurs when it is not possible to identify the quality of a product for a party that participates in a transaction. Therefore bad products are sold with good products where the consequence is that bad products take off good products from the market [@durlauf_adverse_1987]

    -   In the context of financial markets **adverse selection** occurs when an increase in interest rates induces good debtors to stop requesting loans, so that only those individuals with a higher probability of not paying the loan end up requesting loans

## 

-   **Moral hazard** is *"any situation in which one person makes the decision about how much risk to take, while someone else bears the cost if things go badly"* [@krugman_return_2009, p. 63]

    -   In the context of financial markets **moral hazard** occurs when debtors take riskier actions that increase the probability of default

# Acknowledgments

-   To my family that supports me

-   To the taxpayers of Colombia and the [**UMNG students**](https://www.umng.edu.co/estudiante) who pay my salary

-   To the [**Business Science**](https://www.business-science.io/) and [**R4DS Online Learning**](https://www.rfordatasci.com/) communities where I learn [**R**](https://www.r-project.org/about.html) and [**$\pi$-thon**](https://www.python.org/about/)

-   To the [**R Core Team**](https://www.r-project.org/contributors.html), the creators of [**RStudio IDE**](https://rstudio.com/products/rstudio/), [**Quarto**](https://quarto.org/) and the authors and maintainers of the packages [**tidyverse**](https://CRAN.R-project.org/package=tidyverse), [**wbstats**](https://CRAN.R-project.org/package=wbstats), [**tidyquant**](https://CRAN.R-project.org/package=tidyquant), [**ggrepel**](https://CRAN.R-project.org/package=ggrepel), [**lubridate**](https://CRAN.R-project.org/package=lubridate), [**knitr**](https://CRAN.R-project.org/package=knitr), [**kableExtra**](https://CRAN.R-project.org/package=kableExtra), [**readxl**](https://CRAN.R-project.org/package=readxl), and [**tinytex**](https://CRAN.R-project.org/package=tinytex) for allowing me to access these tools without paying for a license

-   To the [**Linux kernel community**](https://www.kernel.org/category/about.html) for allowing me the possibility to use some [**Linux distributions**](https://static.lwn.net/Distributions/) as my main [**OS**](https://en.wikipedia.org/wiki/Operating_system) without paying for a license

# References {.allowframebreaks}
