---
title: Financial Market
subtitle: Session 1
author:
  - name: 
      given: Luis F.
      family: Gómez L.
    orcid: 0000-0002-2877-9015
    email: luis.gomezl@unimilitar.edu.co
    affiliations:
      - name: Distance Learning Faculty
toc: true
toc-depth: 1
slide-level: 2
format:
  presentation-umng-revealjs:
    logo-vertical-strip: 000_logos/logo-white-vertical.png 
    embed-resources: true
    mermaid-format: png
    reference-location: document
date: last-modified
date-format: "D MMMM, YYYY"
knitr: 
  opts_chunk: 
    echo: false
    warning: false
    fig.align: center
    out.width: 85%
lang: en
bibliography: intro_econ.bib
csl: apa-7.csl
editor: source
---

```{r}
#| label: libraries

library(tidyverse)
library(wbstats)
library(scales)
library(ggrepel)
```

```{r}
#| label: palette

umng_palette <- c(
  "#043074",
  "#fdc600",
  "#0d5e30",
  "#ee2a24",
  "#fc6700",
  "#00b3f0",
  "#6e3c1a",
  "#f8941c",
  "#8e44ad",
  "#2c3e50",
  "#16a085",
  "#c0392b",
  "#e91e63"
)
```

# Please Read Me

##

-   Check the message **Welcome greeting** published in Announcements forum.

-   Dear student please edit your profile uploading a photo where your face is clearly visible.

-   If you want to participate, please fill out the following survey: **Primer corte 30% > Learning Activities > Tu opinión sobre la economía colombiana**

-   This presentation is based on [@cardenas_introduccion_2020, Chapter 8]

##

- **Purpose**

    - Analyze the functioning of the financial market, identifying the types of intermediaries and instruments that are part of it

# Structure of the Colombian financial system

##

-   The financial system is the set of entities whose main function is to channel the money of savers to those who wish to make investments

    -   We are only going to focus on supervised financial entities. Therefore entities outside the law or unregulated are not going to be analyzed

    -   Also ponzi schemes, pyramids or unregulated investment scheme are not analyzed.

        -   However if you are interested in this topic check out [@carvajal_ponzi_2009] and [@hofstetter_ponzi_2018]

## {.center} 

```{mermaid}
%%| label: fig-financial-system-structure-supervised-col
%%| fig-cap: Financial system structure by supervised entities [@congreso_de_colombia_decreto_1993]
%%| fig-height: 3.5

flowchart TB
    A("`Sector
    financiero`")  --> A1("`Establecimientos
    de crédito`")
    A1 --> A11("`Establecimientos
    bancarios`")
    A1 --> A12("`Corporaciones
    financieras`")
    A1 --> A13("`Compañias
    de
    financiamiento`")
    A1 --> A14("`Cooperativas
    financieras`")
    A --> A2("`Sociedades de
    servicios
    financieros`")
    A2 --> A21("`Sociedades
    fiduciarias`")
    A2 --> A22("`Almacenes
    generales
    de depósito`")
    A2 --> A23("`Sociedades
    administrativas
    de fondos
    de pensiones y 
    de cesantías`")
    A --> A3("`Sociedades de
    capitalización`")
    A --> A4("`Entidades
    aseguradoras e
    intermediarios`")
    A4 --> A41("`Entidades
    aseguradoras`")
    A4 --> A42("`Intermediarios
    de seguros`")
    A4 --> A43("`Intermediarios
    de reaseguros`")

    classDef UmngNode fill:#fdc600,stroke:#043074,stroke-width:2px;
    class A,A1,A11,A12,A13,A14,A2,A21,A22,A23,A3,A4,A41,A42,A43 UmngNode;
```

##

```{mermaid}
%%| label: fig-financial-system-structure-markets-col
%%| fig-cap: Financial system structure by markets entities [@cardenas_introduccion_2020, Chapter 8, p. 264]

flowchart TB
    A("`Sector
    financiero`")  --> A1("`Mercado
    monetario`")
    A --> A2("`Mercado
    de capitales`")
    A2 --> A21(Bancario)
    A21 --> A211(Bancos)
    A21 --> A212("`Otros
    intermediarios`")
    A2 --> A22(No bancario)
    A22 --> A221(Acciones)
    A22 --> A222("`Renta
    fija`")
    A22 --> A223(Derivados)
    A22 --> A224("`Otros
    mercados`")
    A --> A3("`Mercado
    de divisas`")
    A --> A4("`Otros mercados
    financieros`")

    classDef UmngNode fill:#fdc600,stroke:#043074,stroke-width:2px;
    class A,A1,A2,A21,A22,A211,A212,A221,A222,A223,A224,A3,A4 UmngNode;
```

# Financial depth

## {.smaller} 

-   *"Financial depth captures the financial sector relative to the economy. It is the size of banks, other financial institutions, and financial markets in a country, taken together and compared to a measure of economic output"* [@world_bank_financial_2016]

-   How it is measure using quantity indicators?[^1]

    -   **Domestic credit to private sector (% of GDP)**

    -   **Market capitalization of listed domestic companies (% of GDP)**

        -   Share price times the number of shares outstanding (including their several classes) for listed domestic companies

-   According to the literature the *"evidence suggests that both financial intermediaries and markets matter for growth and that reverse causality alone is not driving this relationship"* [@levine_chapter_2005, p. 866]

[^1]: These indicators doesn't measure the quality of financial depth

##

- The financial system provides 5 broad categories of financial functions that can affect resource allocation and economic growth [@levine_chapter_2005, p. 870]

    - **Producing information and allocating capital** [@levine_chapter_2005, pp. 870-872]
    
        - Financial systems enhance the collection and processing of information about potential investments. 
        
        - This allows for better assessment of projects and more efficient allocation of resources that foster economic growth. 

##

- The financial system provides 5 broad categories of financial functions that can affect resource allocation and economic growth [@levine_chapter_2005, p. 870]

    - **Monitoring firms and exerting corporate governance** [@levine_chapter_2005, pp. 872-875]
    
        - Financial institutions, particularly banks and investors, play a role in monitoring the activities of firms they finance. 
        
        - This helps to ensure that managers act in the best interests of shareholders and that capital is used efficiently. 

##

- The financial system provides 5 broad categories of financial functions that can affect resource allocation and economic growth [@levine_chapter_2005, p. 870]

    - **Risk amelioration** [@levine_chapter_2005, pp. 875-879]
    
        - Financial systems reduce the negative impact of risk by enabling diversification, risk transfer, and liquidity. 
        
        - This encourages investment and innovation, fostering economic growth by making risky but productive ventures more viable.

##

- The financial system provides 5 broad categories of financial functions that can affect resource allocation and economic growth [@levine_chapter_2005, p. 870]

    - **Pooling of savings** [@levine_chapter_2005, pp. 879-880]
    
        - The financial system gathers dispersed funds into large capital pools, enabling substantial investments that individuals alone couldn't make. 
        
        - This process reduces transaction costs, facilitates large-scale projects, and efficiently allocates capital, driving economic growth by channeling savings into productive ventures.

##

- The financial system provides 5 broad categories of financial functions that can affect resource allocation and economic growth [@levine_chapter_2005, p. 870]

    - **Easing exchange** [@levine_chapter_2005, pp. 880-881]
    
        - The financial system reduce transaction costs and facilite trade allowing for smoother economic interactions.
        
        - This increases efficiency, promotes specialization, and expands markets, contributing to economic growth.

##

```{r}
# Helpers
start_date <- 2024
end_date <- 2024

# Import
credit_private_long <- wb_data(
  country = "countries_only",
  # Domestic credit to private sector (% of GDP)
  ## FS.AST.PRVT.GD.ZS
  #  GDP per capita, PPP (constant 2021 international $)
  ## NY.GDP.PCAP.PP.KD
  indicator = c(
    "FS.AST.PRVT.GD.ZS",
    "NY.GDP.PCAP.PP.KD"
  ),
  start_date = start_date,
  end_date = end_date,
  return_wide = FALSE
)

# Tidy
credit_private_wide <- credit_private_long |>
  select(
    iso2c, iso3c, country,
    date,
    indicator_id,
    value
  ) |>
  pivot_wider(
    id_cols = iso2c:date,
    names_from = indicator_id,
    values_from = value
  ) |>
  filter(!is.na(FS.AST.PRVT.GD.ZS), !is.na(NY.GDP.PCAP.PP.KD)) |>
  select(
    iso3c, country, date,
    FS.AST.PRVT.GD.ZS, NY.GDP.PCAP.PP.KD
  )

## Min, max and Colombia
credit_private_wide_min_max_col <- credit_private_wide |>
  filter(FS.AST.PRVT.GD.ZS == min(FS.AST.PRVT.GD.ZS) |
    FS.AST.PRVT.GD.ZS == max(FS.AST.PRVT.GD.ZS) |
    iso3c == "COL")
```

```{r}
#| label: fig-domestic-credit-private-sector-gdp-pc-col
#| fig-cap: Financial depth vs Gross Domestic Product per-capita

# Visualization
credit_private_wide |>
  ggplot(aes(
    x = FS.AST.PRVT.GD.ZS,
    y = NY.GDP.PCAP.PP.KD
  )) +
  geom_point(
    color = "black",
    fill = umng_palette[2],
    shape = 21,
    show.legend = FALSE
  ) +
  geom_label_repel(
    data = filter(
      credit_private_wide,
      iso3c %in% credit_private_wide_min_max_col$iso3c
    ),
    aes(label = country),
    seed = 1234,
    size = 3,
    min.segment.length = 0,
    box.padding = 1,
    force = 2,
    max.overlaps = Inf
  ) +
  scale_x_log10(labels = label_number()) +
  scale_y_log10(labels = label_number()) +
  labs(
    x = "Domestic credit to private sector (% of GDP)",
    y = str_glue("GDP per capita, PPP
                           (constant 2021 international $)"),
    color = NULL,
    subtitle = str_glue("Period: {start_date}
                           Axis scales: Logarithmic Base 10
                           Number of territories: {nrow(credit_private_wide)}"),
    caption = str_glue("Source: World Development Indicators (WDI) - World Bank
                             Last update date: {unique(credit_private_long$last_updated)}")
  ) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10)
  )
```

##

```{r}
# Helpers
start_date <- 2024
end_date <- 2024

# Import
market_cap_long <- wb_data(country = "countries_only",
                           # Market capitalization of listed domestic companies (% of GDP)
                           ## CM.MKT.LCAP.GD.ZS
                           #  GDP per capita, PPP (constant 2021 international $)
                           ## NY.GDP.PCAP.PP.KD
                           indicator    = c("CM.MKT.LCAP.GD.ZS", 
                                            "NY.GDP.PCAP.PP.KD"),
                           start_date = start_date,
                           end_date = end_date,
                           return_wide  = FALSE) 

# Tidy
market_cap_wide <- market_cap_long |>
  select(iso2c, iso3c, country, date, indicator_id, value) |>
  pivot_wider(id_cols = iso2c:date, 
              names_from = indicator_id,
              values_from = value) |> 
  filter(!is.na(CM.MKT.LCAP.GD.ZS), !is.na(NY.GDP.PCAP.PP.KD)) |>
  select(iso3c, country, date,
         CM.MKT.LCAP.GD.ZS, NY.GDP.PCAP.PP.KD)

## Min, max and Colombia
market_cap_wide_min_max_col <- market_cap_wide |> 
  filter(CM.MKT.LCAP.GD.ZS == min(CM.MKT.LCAP.GD.ZS) |
           CM.MKT.LCAP.GD.ZS == max(CM.MKT.LCAP.GD.ZS) |
           iso3c == "COL")
```

```{r}
#| label: fig-market-capitalization-gdp-pc-col
#| fig-cap: Financial depth vs Gross Domestic Product per-capita

# Visualization
market_cap_wide |>
  ggplot(aes(x = CM.MKT.LCAP.GD.ZS,
             y = NY.GDP.PCAP.PP.KD)) +
  geom_point(color = "black",
             fill = umng_palette[2],
             shape = 21,
             show.legend = FALSE) +
  geom_label_repel(data = filter(market_cap_wide, iso3c %in% market_cap_wide_min_max_col$iso3c),
                   aes(label = country),
                   seed = 1234,
                   size = 3,
                   min.segment.length = 0,
                   box.padding = 1,
                   force = 2,
                   max.overlaps = Inf) +
  scale_x_log10(labels = label_number()) +
  scale_y_log10(labels = label_number()) +
  labs(x        = "Market capitalization of listed domestic companies (% of GDP)",
       y        = str_glue("GDP per capita, PPP
                           (constant 2021 international $)"),
       color = NULL,
       subtitle = str_glue("Period: {start_date}
                           Axis scales: Logarithmic Base 10
                           Number of territories: {nrow(market_cap_wide)}"),
       caption  = str_glue("Source: World Development Indicators (WDI) - World Bank
                             Last update date: {unique(market_cap_long$last_updated)}")) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10)
  )
```

# Uncertainty in financial markets

## 

-   Financing is essentially the exchange of a sum of money today for a promise to return more money in the future. Therefore it is not surprising that such exchange can be problematic

    -   **Information asymmetry**: in an exchange one party has more or better information than the other

        -   **Adverse selection**
        -   **Moral hazard**

## 

-   **Adverse selection** occurs when it is not possible to identify the quality of a product for a party that participates in a transaction. Therefore bad products are sold with good products where the consequence is that bad products take off good products from the market [@durlauf_adverse_1987]

    -   In the context of financial markets **adverse selection** occurs when an increase in interest rates induces good debtors to stop requesting loans, so that only those individuals with a higher probability of not paying the loan end up requesting loans

## 

-   **Moral hazard** is *"any situation in which one person makes the decision about how much risk to take, while someone else bears the cost if things go badly"* [@krugman_return_2009, p. 63]

    -   In the context of financial markets **moral hazard** occurs when debtors take riskier actions that increase the probability of default

# Acknowledgments

##

-   To my family that supports me

-   To the taxpayers of Colombia and the [**UMNG students**](https://www.umng.edu.co/estudiante) who pay my salary

-   To the [**Business Science**](https://www.business-science.io/) and [**R4DS Online Learning**](https://www.rfordatasci.com/) communities where I learn [**R**](https://www.r-project.org/about.html) and [**$\pi$-thon**](https://www.python.org/about/)

-   To the [**R Core Team**](https://www.r-project.org/contributors.html), the creators of [**RStudio IDE**](https://posit.co/download/rstudio-desktop/), [**Positron**](https://positron.posit.co/), [**Quarto**](https://quarto.org/) and the authors and maintainers of the packages [**tidyverse**](https://CRAN.R-project.org/package=tidyverse), [**wbstats**](https://CRAN.R-project.org/package=wbstats), [**scales**](https://CRAN.R-project.org/package=scales), [**ggrepel**](https://CRAN.R-project.org/package=ggrepel) and [**tinytex**](https://CRAN.R-project.org/package=tinytex) for allowing me to access these tools without paying for a license

# References {.allowframebreaks}