---
title: "Labor Market"
author: "Luis Francisco Gomez Lopez"
date: 2024-02-06
execute: 
  echo: false
  warning: false
format:
  beamer:
    colortheme: dolphin
    fonttheme: structurebold
    theme: AnnArbor
    link-citations: true
    linkcolor: blue
    include-in-header:
      - text: |
          \usepackage{booktabs}
          \usepackage{longtable}
          \usepackage{array}
          \usepackage{multirow}
          \usepackage{wrapfig}
          \usepackage{float}
          \usepackage{colortbl}
          \usepackage{pdflscape}
          \usepackage{tabu}
          \usepackage{threeparttable}
          \usepackage{threeparttablex}
          \usepackage[normalem]{ulem}
          \usepackage{makecell}
          \usepackage{xcolor}
          \usepackage{fontawesome5}
bibliography: intro_econ.bib
---

```{r}
library(tidyverse)
library(tidyquant)
library(readxl)
```

# Contents

- Please Read Me
- Purpose
- Labor classification of the population
- Labor market indicators
- Labor demand
- Wages
- Labor regulation
- Economic environment and the Company
- Acknowledgments
- References

# Please Read Me

- Check the message __Welcome greeting__ published in the News Bulletin Board.

- Dear student please edit your profile uploading a photo where your face is clearly visible.

- The purpose of the virtual meetings is to answer questions and not to make a summary of the study material.

- This presentation is based on [@cardenas_introduccion_2020, Chapter 9]

# Purpose

Examine the evolution and characteristics of the supply and demand of labor in Colombia

# Labor classification of the population

- To understand the indicators of the labor market from an economic perspective, it is necessary to classify the population from a labor point of view. Initially, the total population of a territory could be included but in practice the __Total Population (TP)__ that is included __doesn't__ cover [@hussmanns_surveys_1990, chapter 2]:

    + Members of the armed forces because this information is considered secret or because it is not easy to obtain information when its members are in barracks and military areas.
    
    + Residents in institutions such as people who are not part of the military forces but reside in military facilities, residents of penal or correctional centers and hospital residents.

- In that sense, the **Total Population (TP)** covers the non-institutional civilian population residing in households.

# Labor classification of the population

```{r }
#| out-width: 90%
#| fig-align: center

#Plot
## Labor Market Colombia
## Data
## https://www.dane.gov.co > Estadísticas por tema > 
## Mercado Laboral > Empleo y desempleo > Anexos 
## Total Nacional

labor_classification_concept <- read_excel(path = "000_data/009_anexo_empleo_dic_23.xlsx",
                                   sheet = "Total nacional", 
                                   range = "A23:A30", 
                                   col_names = "concept")

# This is the part that needs to be updated
labor_classification_last_data <- read_excel(path = "000_data/009_anexo_empleo_dic_23.xlsx",
                                           sheet = "Total nacional", 
                                           range = "JQ23:JQ30",
                                           col_names = "value") |> 
  mutate(value = value*1e3)

labor_classification <- bind_cols(labor_classification_concept,
                                  labor_classification_last_data)

poblacion_total <- labor_classification |> pull(2) |> _[1]
poblacion_en_edad_de_trabajar <- labor_classification |> pull(2) |> _[2]
poblacion_menor_a_15_anos <- poblacion_total - poblacion_en_edad_de_trabajar
fuerza_de_trabajo <- labor_classification |> pull(2) |> _[3]
poblacion_fuera_de_la_fuerza_de_trabajo <- poblacion_en_edad_de_trabajar - fuerza_de_trabajo
poblacion_ocupada <- labor_classification |> pull(2) |> _[4]
poblacion_desocupada <- fuerza_de_trabajo - poblacion_ocupada
subocupados <- labor_classification |> pull(2) |> _[7]
completamente_empleados <- poblacion_ocupada - subocupados

tibble(x = c( 0, 
              -1, 
              1,
              0,
              2,
              -2,
              2,
              -3,
              -1),
       y = c( 0, 
              -1, 
              -1,
              -2,
              -2,
              -3,
              -3,
              -4,
              -4),
       indicator = c('Total Population (TP)',
                     'Population below\n15 years',
                     'Working Age\nPopulation\n(WAP)',
                     'Labor\nForce\n(LF)',
                     'Not in\nLabor\nForce\n(NLF)',
                     'Employed (E)',
                     'Unemployed (U)',
                     'Full Employed\n(FE)',
                     'Underemployed\n(UE)'
       ),
       level     = c('0', 
                     '1', '1', 
                     '2', '2', 
                     '3', '3',
                     '4', '4'),
       # Población total
       value     = c(poblacion_total, 
                     # Población menor a 15 años
                     poblacion_menor_a_15_anos,    
                     # Población en edad de trabajar
                     poblacion_en_edad_de_trabajar, 
                     # Fuerza de trabajo
                     fuerza_de_trabajo, 
                     # Población fuera de la fuerza de trabajo
                     poblacion_fuera_de_la_fuerza_de_trabajo, 
                     # Población Ocupada
                     poblacion_ocupada, 
                     # Población Desocupada
                     poblacion_desocupada, 
                     # Completamente empleados
                     completamente_empleados, 
                     # Subocupados
                     subocupados))  |> 
  mutate(label_text = str_glue('{indicator}
                                 {value |> 
                                    scales::number(big.mark = ",")}')) |> 
  ggplot(aes(x = x, y = y, label = label_text)) +
  geom_segment(x = 0, y = 0, xend = 0, yend = -1) +
  geom_segment(x = -1, y = -1, xend = 1, yend = -1) +
  geom_segment(x = 1, y = -1, xend = 1, yend = -2) +
  geom_segment(x = 0, y = -2, xend = 2, yend = -2) +
  geom_segment(x = 0, y = -2, xend = 0, yend = -3) +
  geom_segment(x = -2, y = -3, xend = 2, yend = -3) +
  geom_segment(x = -2, y = -3, xend = -2, yend = -4) +
  geom_segment(x = -3, y = -4, xend = -1, yend = -4) +
  geom_label(aes(fill = level), 
             color       = "white", 
             fontface    = "bold",
             size        = 3.5) +
  scale_fill_tq() +
  coord_cartesian(xlim = c(-5, 5), 
                  ylim = c(-4.2, 0.2)) +
  labs(title    = 'Labor classification of the population in Colombia',
                  # This part needs to be update manually
       subtitle = str_glue('Period: December 2023'),
                  # This part needs to be update manually in relation
                  # to the item "Last update"
       caption  = str_glue('Sources: DANE, Gran Encuesta Integrada de Hogares - GEIH
                              Last update: 2024-01-31'),
       fill     = "Level"
  ) + 
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        panel.grid        = element_blank(),
        legend.background = element_rect(fill = "#f3fcfc"),
        plot.title        = element_text(face = "bold"),
        axis.title        = element_blank(),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_blank(), 
        axis.ticks        = element_blank())
```

# Labor market indicators

- __Percentage of the working age population__ ("Porcentaje de población en edad de trabajar") 

    + $\%WAP = \frac{WAP}{PT} \times 100$
    
- __Labor participation rate (LPR)__ ("Tasa global de participación") 

    + $LPR = \frac{LF}{WAP} \times 100$
    
# Labor market indicators

- __Employment rate (ER)__ ("Tasa de ocupación")

    + $UR = \frac{E}{WAP} \times 100$ 

- __Unemployment rate (UR)__ ("Tasa de desempleo")

    + $UR = \frac{U}{LF} \times 100$

- __Underemployment (UER)__ ("Tasa de subempleo")

    + $UER = \frac{UE}{LF} \times 100$

# Labor demand

- The labor demand is the process by which the economy generates job vacancies

- How companies decide how many workers to contract?

    + Demand of the products that offer by the company
    + Wages (the price of labor)
    + Labor regulation
    + The costs of other inputs different form labor and used in production
    + The technology used by companies
    
# Labor market indicators

```{r }
#| out-width: 90%
#| fig-align: center

data <- wbstats::wb_data(country     = c('COL'),
                         indicator   = c('SL.UEM.TOTL.ZS', 'SL.EMP.TOTL.SP.ZS'), 
                         return_wide = FALSE)

data_clean <- data |> 
  drop_na(value) |> 
  select(country, date, value, indicator) |>
  mutate(indicator = case_when(
         indicator == 'Unemployment, total (% of total labor force) (modeled ILO estimate)' ~ 'Unemployment rate (UR) (ILO estimate))',
         indicator == 'Employment to population ratio, 15+, total (%) (modeled ILO estimate)' ~ 'Employment rate (ER) (ILO estimate))',
         .default = indicator))

data_clean |> 
  ggplot(aes(x = date, 
               y = value)) + 
  geom_point(shape = 21, fill = palette_light()[[2]], color = "black") +
  geom_line() +
  geom_vline(xintercept = c(1999, 2020),
             color      = palette_light()[[1]]) +
  expand_limits(y = 0) +
  facet_wrap(facet    = vars(indicator), 
               ncol     = 1,
               nrow     = 2, 
               scales   = "free_y") +
  scale_x_continuous(breaks = c(1991, 1999, 2010, max(data_clean$date))) +
  scale_y_continuous(labels = scales::number_format(accuracy = 1,
                                                      suffix = "%")) +
  coord_cartesian(xlim = c(min(data_clean$date), max(data_clean$date))) +
  labs(x        = 'Year',
       y        = 'Percent',
       title    = 'Annual Unemployment rate (UR) and Employment rate (ER) in Colombia', 
       subtitle = str_glue('UR code WDI: SL.UEM.TOTL.ZS
                             UR units: percent
                             ER code WDI: SL.EMP.TOTL.SP.ZS
                             ER units: percent
                             Observation: ILO estimations where ages 15 and older are considered the WAP'),
       caption  = str_glue('Source: World Development Indicators (WDI) - World Bank
                             Last update date: {unique(data$last_updated)}')) +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"),
        strip.background.x = element_rect(color = "black", fill = "#fefff0"),
        strip.text.x = element_text(margin = margin(t = 5,
                                                    r = 5,
                                                    b = 5,
                                                    l = 5, 
                                                    unit = "pt"), face = "bold"))
```

# Labor demand

- In times of low economic growth, companies reduce production and therefore hire fewer workers. 

```{r }
#| out-width: 75%
#| fig-align: center

                                             # GDP per capita growth (annual %)
econ_growth_unemployment_long <- wbstats::wb_data(indicator = c("NY.GDP.PCAP.KD.ZG",
                                                           # Unemployment, total (% of total labor force)
                                                           # It is calculated not but the ILO but by the Statistical Office of each country
                                                           "SL.UEM.TOTL.NE.ZS"), 
                                                  country = "COL", 
                                                  return_wide = FALSE)

econ_growth_unemployment_wide <- econ_growth_unemployment_long |> 
  select(iso2c, iso3c, country, date, indicator_id, value) |> 
  pivot_wider(id_cols =  iso2c:date, 
              names_from = indicator_id,
              values_from = value) |> 
  drop_na(NY.GDP.PCAP.KD.ZG, SL.UEM.TOTL.NE.ZS)

econ_growth_unemployment_wide |> 
  ggplot(aes(NY.GDP.PCAP.KD.ZG, SL.UEM.TOTL.NE.ZS)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_continuous(labels = scales::number_format(suffix = "%")) +
  scale_y_continuous(labels = scales::number_format(suffix = "%")) +
  labs(x        = "GDP per capita growth (annual %)",
       y        = str_glue("Annual total unemployment (% of total labor force)"),
       color = NULL,
       title    = "Economic growth vs unemployment rate",
       subtitle = str_glue("Period: {min(econ_growth_unemployment_long$date)} - {max(econ_growth_unemployment_long$date)}"),
       # Update last update date manually
       caption  = str_glue("Source: World Development Indicators (WDI) - World Bank
                             Last update date: {unique(econ_growth_unemployment_long$last_updated)}")) +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        legend.position   = "bottom",
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"))
```

# Wages

```{r }
#| out-width: 90%
#| fig-align: center

read_excel(path = "000_data/009_anexos_GEIH_fuerza_laboral_educacion_2020.xls", 
           sheet = 7, 
           range = "A19:L23", 
           col_names = FALSE) |> 
  select(1, 12) |> 
  set_names(nm = "wage_range", "value") |> 
  mutate(category = "Ninguno") |> 
  rbind(
    read_excel(path = "000_data/009_anexos_GEIH_fuerza_laboral_educacion_2020.xls", 
               sheet = 7, 
               range = "A26:L30", 
               col_names = FALSE) |> 
      select(1, 12) |> 
      set_names(nm = "wage_range", "value") |> 
      mutate(category = "Bachiller") 
  ) |> 
  rbind(
    read_excel(path = "000_data/009_anexos_GEIH_fuerza_laboral_educacion_2020.xls", 
               sheet = 7, 
               range = "A33:L37", 
               col_names = FALSE) |> 
      select(1, 12) |> 
      set_names(nm = "wage_range", "value") |> 
      mutate(category = "Técnico o tecnológico") 
  ) |> 
  rbind(
    read_excel(path = "000_data/009_anexos_GEIH_fuerza_laboral_educacion_2020.xls", 
               sheet = 7, 
               range = "A40:L44", 
               col_names = FALSE) |> 
      select(1, 12) |> 
      set_names(nm = "wage_range", "value") |> 
      mutate(category = "Universitario y/o postgrado")
  ) |> 
  mutate(wage_range = as_factor(wage_range),
         category = as_factor(category),
         value = value*100) |> 
  ggplot(aes(y = wage_range, value)) +
  geom_col() +
  scale_x_continuous(labels = scales::number_format(scale = 1e-3, suffix = "K")) +
  facet_wrap(facets = vars(category)) + 
  labs(x = "Número de trabajadores en miles (K)",
       y = NULL,
       title = "Población ocupada según último título educativo obtenido y rangos de ingresos",
       subtitle = str_glue("Periodo: 2020
                           En la categoría 'No informa' se incluye los trabajadores sin renumeración")) +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"),
        strip.background.x = element_rect(color = "black", fill = "#fefff0"),
        strip.text.x = element_text(margin = margin(t = 5,
                                                    r = 5,
                                                    b = 5,
                                                    l = 5, 
                                                    unit = "pt"), face = "bold"))
```

# Wages

- Mincer equation

  + $log_e(w_{it}) = log_e(w_o) + \beta_1Ed_{it} +  \beta_2Ex_{it} + \beta_2Ex_{it}^2 + \epsilon_{it}$

      + $log_e$: natural logarithm
      + $w_{it}$: wage of individual $i$ in period $t$
      + $w_o$: wage without any level of education an experience
      + $\epsilon_{it}$: years of education of individual $i$ in period $t$
      + $Ex_{it}$: potential experience of individual $i$ in period $t$ which is equal to the age minus the years of education minus six
      + $\epsilon_{it}$: are other components of individual $i$ in period $t$ that might affect the wage
      
# Labor regulation 

- The labor regulation refers to:

    + The laws governing labor contracts 
    + Laws governing labor relations that empower unions to represent workers collectively
    + Social security laws governing the response to social needs such as unemployment, maternity, old age, disability, death and sickness

# Labor regulation

- Resources

    + **Calculadora Laboral - Mintrabajo**
    
        + [https://www.mintrabajo.gov.co/atencion-al-ciudadano/tramites-y-servicios/mi-calculadora](https://www.mintrabajo.gov.co/atencion-al-ciudadano/tramites-y-servicios/mi-calculadora)
        
# Economic environment and the Company

- Set of economic factors and forces that influence the development of and organization.

```{r econ_env_macro}
#| out-width: 75%
#| fig-align: center

#Plot
ggplot(data = tibble(x = 0, y = 0)) + 
    ## Company
    annotate(geom = "text", 
             x        = 0, 
             y        = 0,
             label    = "Company",
             size     = 3,
             fontface = "bold") +
    geom_rect(aes(xmin = -0.4,
                  xmax =  0.4,
                  ymin = -0.4,
                  ymax =  0.4),
              alpha = 0.3,
              fill  = "red",
              color = "red") +
    ## Micro environment
    geom_rect(aes(xmin = -2,
                  xmax =  2,
                  ymin = -2,
                  ymax =  2),
              alpha = 0.25,
              color = "blue") +
    annotate(geom = "text", 
             x        = 0, 
             y        = 1,
             label    = str_glue("Micro environment"),
             size     = 6,
             fontface = "bold") +
    annotate(geom = "text", 
             x     = 0, 
             y     = -1.2,
             label = str_glue("Clients - Suppliers - Intermediaries - Competitors - General Public"),
             size  = 3.5) +
    ## Macro environment
    geom_rect(aes(xmin = -3,
                  xmax =  3,
                  ymin = -3,
                  ymax =  3),
              alpha = 0.1,
              color = "purple") +
    annotate(geom     = "text", 
             x        = 0, 
             y        = 2.5,
             label    = str_glue("Macro environment"),
             size     = 8,
             fontface = "bold") +
    annotate(geom = "text", 
             x     =  0, 
             y     = -2.5,
             label = str_glue("Economic - Political - Social - Technological - Ecological - Legal"),
             size  = 5) +
    theme_void() +
    labs(caption = "Based on https://foda-dafo.com/ ")
```

# Acknowledgments

- To my family that supports me

- To the taxpayers of Colombia and the __[UMNG students](https://www.umng.edu.co/estudiante)__ who pay my salary

- To the __[Business Science](https://www.business-science.io/)__ and __[R4DS Online Learning](https://www.rfordatasci.com/)__ communities where I learn __[R](https://www.r-project.org/about.html)__ 

- To the __[R Core Team](https://www.r-project.org/contributors.html)__, the creators of __[RStudio IDE](https://rstudio.com/products/rstudio/)__, __[Quarto](https://quarto.org/)__ and the authors and maintainers of the packages  __[tidyverse](https://CRAN.R-project.org/package=tidyverse)__,
__[tidyquant](https://CRAN.R-project.org/package=tidyquant)__, 
__[readxl](https://CRAN.R-project.org/package=readxl)__, and __[tinytex](https://CRAN.R-project.org/package=tinytex)__ for allowing me to access these tools without paying for a license

- To the __[Linux kernel community](https://www.kernel.org/category/about.html)__ for allowing me the possibility to use some __[Linux distributions](https://static.lwn.net/Distributions/)__ as my main __[OS](https://en.wikipedia.org/wiki/Operating_system)__ without paying for a license

# References {.allowframebreaks}
