---
title: Labor Market
subtitle: Session 2
author:
  - name: 
      given: Luis F.
      family: Gómez L.
    orcid: 0000-0002-2877-9015
    email: luis.gomezl@unimilitar.edu.co
    affiliations:
      - name: Distance Learning Faculty
toc: true
toc-depth: 1
slide-level: 2
format:
  presentation-umng-revealjs:
    logo-vertical-strip: 000_logos/logo-white-vertical.png 
    embed-resources: true
    reference-location: document
    mermaid-format: png
date: last-modified
date-format: "D MMMM, YYYY"
knitr: 
  opts_chunk: 
    echo: false
    warning: false
    fig.align: center
    out.width: 85%
lang: en
bibliography: intro_econ.bib
csl: apa-7.csl
editor: source
---

```{r}
#| label: libraries

library(tidyverse)
library(readxl)
library(gt)
library(janitor)
library(Rilostat)
library(ggrepel)
```

```{r}
#| label: palette

umng_palette <- c(
  "#043074",
  "#fdc600",
  "#0d5e30",
  "#ee2a24",
  "#fc6700",
  "#00b3f0",
  "#6e3c1a",
  "#f8941c",
  "#8e44ad",
  "#2c3e50",
  "#16a085",
  "#c0392b",
  "#e91e63"
)
```

# Please Read Me

##

-   Check the message **Welcome greeting** published in Announcements forum.

-   Dear student please edit your profile uploading a photo where your face is clearly visible.

-   If you want to participate, please fill out the following survey: **Primer corte 30% > Learning Activities > Tu opinión sobre la economía colombiana**

-   This presentation is based on [@cardenas_introduccion_2020, Chapter 9]

##

- **Purpose**

    - Examine the evolution and characteristics of the supply and demand of labor in Colombia
    
# Unemployment

##

```{r}
# https://www.dane.gov.co >
# Estadísticas por tema >
# Mercado laboral >
# Empleo y desempleo >
# Documento >
# Anexos >
## Descargar

# Import ----
## Helper variables
dates <- seq.Date(
  from = ymd("2001-01-01"),
  # Change the final date
  to = ymd("2025-12-01"),
  by = "month"
) |>
  as.character()
col_names <- c("concept", dates)

## Data
indicators <- read_excel(
  path = "000_data/009_anexo_empleo_dic_2025.xlsx",
  sheet = "Total nacional",
  # Change the range of the data
  range = "A14:KO21",
  col_names = col_names
)

# Tidy
indicators_longer <- indicators |>
  pivot_longer(cols = -c(concept), names_to = "date", values_to = "value") |>
  mutate(date = ymd(date))

## Filter the last 10 years
years <- year(indicators_longer$date) |>
  unique() |>
  tail(n = 10)

indicators_longer_td <- indicators_longer |>
  filter(concept == "Tasa de Desocupación (TD)") |>
  mutate(year = year(date)) |>
  mutate(year = as_factor(year)) |>
  mutate(month = month(date, label = TRUE)) |>
  filter(year %in% years)
```

```{r}
#| label: fig-unemployment-rate-first-years
#| fig-cap: Behavior of unemployment rate in Colombia for selected years

# Visualization
## Helpers
year_pandemic <- 2020
### Update this part using the imported file
last_update_date <- ymd("2026-01-30")

### First years
first_years <- indicators_longer_td |>
  filter(year %in% head(years, n = 5)) |>
  # Delete the year about the pandemic
  ## It is an atypical year
  filter(year != year_pandemic)

## Plot
first_years |>
  ggplot(aes(x = month, y = value)) +
  geom_line(aes(group = year, color = year)) +
  scale_color_manual(values = umng_palette[1:4]) +
  labs(
    x = NULL,
    y = "Percent (%)",
    color = NULL,
    subtitle = str_glue(
      "Period: {head(unique(first_years$year), n = 1)} - {tail(unique(first_years$year), n = 1)}
                           Observation: {year_pandemic} is not included because is an atypical year taking into account the COVID-19 pandemic"
    ),
    caption = str_glue(
      "Source: DANE - Gran Encuesta Integrada de Hogares (GEIH)
                          Last update date: {last_update_date}"
    )
  ) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10)
  )
```

##

```{r}
#| label: fig-unemployment-rate-last-years
#| fig-cap: Behavior of unemployment rate in Colombia for selected years

# Visualization
## Last years
last_years <- indicators_longer_td |>
  filter(year %in% tail(years, n = 5)) |>
  # Delete the year about the pandemic
  ## It is an atypical year
  filter(year != year_pandemic)

last_years |>
  ggplot(aes(x = month, y = value)) +
  geom_line(aes(group = year, color = year)) +
  scale_color_manual(values = umng_palette[1:5]) +
  labs(
    x = NULL,
    y = "Percent (%)",
    color = NULL,
    subtitle = str_glue(
      "Period: {head(unique(last_years$year), n = 1)} - {tail(unique(last_years$year), n = 1)}"
    ),
    caption = str_glue(
      "Source: DANE - Gran Encuesta Integrada de Hogares (GEIH)
                          Last update date: {last_update_date}"
    )
  ) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10)
  )
```

##

```{r}
# https://www.dane.gov.co >
# Estadísticas por tema >
# Mercado laboral >
# Empleo y desempleo >
# Documento >
# Anexos >
## Descargar

# Import data ----
## Helpers
### Different from "Total nacional"
dates <- seq.Date(
  from = ymd("2010-01-01"),
  # Change the final date
  to = ymd("2025-12-01"),
  by = "month"
) |>
  as.character()
col_names <- c("concept", dates)

## Data
indicators_male <- read_excel(
  path = "000_data/009_anexo_empleo_dic_2025.xlsx",
  sheet = "Total_nacional_IML_Sexo",
  # Change the range of the data
  range = "A14:GK18",
  col_names = col_names
) |>
  mutate(sex = "Male")

indicators_female <- read_excel(
  path = "000_data/009_anexo_empleo_dic_2025.xlsx",
  sheet = "Total_nacional_IML_Sexo",
  # Change the range of the data
  range = "A33:GK37",
  col_names = col_names
) |>
  mutate(sex = "Female")

# Tidy
indicators_male_longer <- indicators_male |>
  pivot_longer(
    cols = -c(concept, sex),
    names_to = "date",
    values_to = "value"
  ) |>
  mutate(date = ymd(date))

indicators_female_longer <- indicators_female |>
  pivot_longer(
    cols = -c(concept, sex),
    names_to = "date",
    values_to = "value"
  ) |>
  mutate(date = ymd(date))

## Merge
indicators_sex_longer <- indicators_male_longer |>
  bind_rows(indicators_female_longer)

### Filter the last 10 years
years <- year(indicators_longer$date) |>
  unique() |>
  tail(n = 10)

indicators_sex_longer_td <- indicators_sex_longer |>
  filter(concept == "Tasa de Desocupación (TD)") |>
  mutate(year = year(date)) |>
  mutate(year = as_factor(year)) |>
  mutate(month = month(date, label = TRUE)) |>
  filter(year %in% years)
```

```{r}
#| label: fig-unemployment-rate-male-female
#| fig-cap: Unemployment rate in Colombia for men and women

# Visualization
indicators_sex_longer_td |>
  ggplot(aes(x = date, y = value)) +
  geom_line(aes(group = sex, color = sex)) +
  scale_color_manual(values = umng_palette[1:2]) +
  labs(
    x = NULL,
    y = "Percent (%)",
    color = NULL,
    subtitle = str_glue(
      "Period: {head(unique(indicators_sex_longer_td$date), n = 1)} - {tail(unique(indicators_sex_longer_td$date), n = 1)}"
    ),
    caption = str_glue(
      "Source: DANE - Gran Encuesta Integrada de Hogares (GEIH)
                          Last update date: {last_update_date}"
    )
  ) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10)
  )
```

# Employed

##

```{r}
# Import ----
## Helpers
date <- ymd("2025-12-01")
date_label <- date |> format("%Y - %B")
# Different from "Total nacional"
dates <- seq.Date(
  from = ymd("2015-01-01"),
  # Change the final date
  to = date,
  by = "month"
) |>
  as.character()
col_names <- c("concept", dates)

## Data
indicators_sector <- read_excel(
  path = "000_data/009_anexo_empleo_dic_2025.xlsx",
  sheet = "Ocupados TN_T13_rama",
  # Change the range of the data
  range = "A16:EC30",
  col_names = col_names
)

# Tidy
indicators_sector_clean <- indicators_sector |>
  select(1, ncol(indicators_sector)) |>
  set_names(nm = c("concept", "last_period")) |>
  mutate(last_period = (last_period * 1e3) |> round()) |>
  arrange(desc(last_period))
```

```{r}
#| label: tbl-employed-sector-last-period
#| tbl-cap: Employed by sector in Colombia

# Table
indicators_sector_clean |>
  gt() |>
  cols_label(
    concept = "Concept",
    last_period = date_label
  ) |>
  tab_source_note(
    source_note = "Source: DANE - Gran Encuesta Integrada de Hogares (GEIH)"
  ) |>
  tab_source_note(
    source_note = str_glue("Last update date: {last_update_date}")
  ) |>
  tab_options(
    row.striping.include_table_body = TRUE,
    table.border.top.width = px(3),
    table.border.top.color = "black",
    column_labels.border.bottom.width = px(2),
    column_labels.border.bottom.color = "black",
    table.border.bottom.width = px(3),
    table.border.bottom.color = "black",
    table.font.size = px(16),
    data_row.padding = px(5)
  )
```

##

```{r}
# Import ----
## Helpers
date <- ymd("2025-12-01")
date_label <- date |> format("%Y - %B")
# Different from "Total nacional"
dates <- seq.Date(
  from = ymd("2010-01-01"),
  # Change the final date
  to = date,
  by = "month"
) |>
  as.character()
col_names <- c("concept", dates)

## Data
indicators_occupational_position <- read_excel(
  path = "000_data/009_anexo_empleo_dic_2025.xlsx",
  sheet = "Ocupados TN_posición",
  # Change the range of the data
  range = "A16:GK24",
  col_names = col_names
)

# Tidy
indicators_occupational_position_clean <- indicators_occupational_position |>
  select(1, ncol(indicators_occupational_position)) |>
  set_names(nm = c("concept", "last_period")) |>
  mutate(last_period = (last_period * 1e3) |> round()) |>
  arrange(desc(last_period))
```

```{r}
#| label: tbl-employed-occupational-position-last-period
#| tbl-cap: Employed by occupational position in Colombia

# Table
indicators_occupational_position_clean |>
  gt() |>
  cols_label(
    concept = "Concept",
    last_period = date_label
  ) |>
  tab_source_note(
    source_note = "Source: DANE - Gran Encuesta Integrada de Hogares (GEIH)"
  ) |>
  tab_source_note(
    source_note = str_glue("Last update date: {last_update_date}")
  ) |>
  tab_options(
    row.striping.include_table_body = TRUE,
    table.border.top.width = px(3),
    table.border.top.color = "black",
    column_labels.border.bottom.width = px(2),
    column_labels.border.bottom.color = "black",
    table.border.bottom.width = px(3),
    table.border.bottom.color = "black",
    table.font.size = px(20),
    data_row.padding = px(5)
  )
```

# Informality

## {.smaller}

- What is informality?

    - The classification used in Colombia is based initially on the occupational position. For more information about definitions of the occupational position check out [@dane_manual_2024, pp. 88-92]
    
        -  **Obrero, empleado del gobierno** is not considered informal in any case
        -  **Obrero o empleado de empresa particular**, **Empleado doméstico** and **Trabajador sin remuneración en empresas o negocios de otros hogares** who do not have health or pension contributions made due to their employment relationship with the employer who hired them are considered informal
        - **Jornalero o Peón** and **Trabajador familiar sin remuneración** are considered informal
        
## {.smaller}

- What is informality?

    - The classification used in Colombia is based initially on the occupational position. For more information about definitions of the occupational position check out [@dane_manual_2024, pp. 88-92]
        
        - **Patrón o empleador** and **Trabajador por cuenta propia** classified as informal[^1]
        - Those employed in the sectors with ISIC codes 84 and 99 with the exception of **Jornalero o Peón** and **Trabajador familiar sin remuneración** are not considered informal  
        
[^1]: The rules are not simple in this case. Check [https://www.dane.gov.co/index.php/estadisticas-por-tema/mercado-laboral/empleo-informal-y-seguridad-social](https://www.dane.gov.co/index.php/estadisticas-por-tema/mercado-laboral/empleo-informal-y-seguridad-social) > Anexos in relation to the sheets **Código_SAS** or **Código_STATA**          

##

- Informality proportion (PI)

    - Employed (E)
    - Informal Employed (I) 

$$
PI = \frac{I}{E} \times 100
$$

- Formality proportion (PF)

$$
PF = 1- PI = 1 - \frac{I}{E} \times 100
$$

##

```{r}
# https://www.dane.gov.co >
# Estadísticas por tema >
# Mercado laboral >
# Empleo informal y seguridad social  >
# Anexos >
## Descargar

# Import ----
## Helpers
dates <- seq.Date(
  from = ymd("2021-01-01"),
  # Change the final date
  to = ymd("2025-11-01"),
  by = "month"
) |>
  as.character()
col_names <- c("concept", dates)

## Data
informality_col <- read_excel(
  path = "000_data/009_anexo_empleo_informal_seguridad_social_nov_2025.xlsx",
  # Check white spaces
  sheet = "Grandes dominios ",
  # Change the range of the data
  range = "A14:BH16",
  col_names = col_names
)

# Tidy
informality_col_wider <- informality_col |>
  pivot_longer(cols = -c(concept), names_to = "date", values_to = "value") |>
  pivot_wider(id_cols = date, names_from = concept, values_from = value) |>
  clean_names() |>
  mutate(prop_informalidad = (informal / poblacion_ocupada) * 100) |>
  mutate(prop_formalidad = 100 - prop_informalidad) |>
  mutate(date = ymd(date))

informality_col_longer <- informality_col_wider |>
  select(date, prop_informalidad, prop_formalidad) |>
  pivot_longer(cols = -c(date), names_to = "variable", values_to = "value")
```

```{r}
#| label: fig-informality-formality-proportion
#| fig-cap: Formality and informality proportion in Colombia

# Visualization
informality_col_longer |>
  ggplot(aes(x = date, y = value)) +
  geom_line(aes(group = variable, color = variable)) +
  scale_color_manual(
    labels = c("Formality proportion (PF)", "Informality proportion (PI)"),
    values = umng_palette[1:2]
  ) +
  labs(color = NULL) +
  labs(
    x = NULL,
    y = "Percent (%)",
    color = NULL,
    subtitle = str_glue(
      "Period: {head(unique(informality_col_longer$date), n = 1)} - {tail(unique(informality_col_longer$date), n = 1)}"
    ),
    caption = str_glue(
      "Source: DANE - Gran Encuesta Integrada de Hogares (GEIH)
                          Last update date: {last_update_date}"
    )
  ) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10)
  )
```

# Informality proportion vs unemployment rate

##

```{r}
# Data
## https://ilostat.ilo.org/data/

# Helpers
toc <- get_ilostat_toc()
last_update <- toc |>
  filter(str_detect(string = id, pattern = "SDG_0852_SEX_AGE_RT_A")) |>
  pull(last.update) |>
  str_sub(start = 1, end = 10) |>
  dmy()

## Countries
lac_iso <- c(
  "ARG", # Argentina
  "ATG", # Antigua and Barbuda
  "BHS", # Bahamas
  "BLZ", # Belize
  "BOL", # Bolivia
  "BRA", # Brazil
  "BRB", # Barbados
  "CHL", # Chile
  "COL", # Colombia
  "CRI", # Costa Rica
  "CUB", # Cuba
  "DMA", # Dominica
  "DOM", # Dominican Republic
  "ECU", # Ecuador
  "GRD", # Grenada
  "GTM", # Guatemala
  "GUY", # Guyana
  "HND", # Honduras
  "HTI", # Haiti
  "JAM", # Jamaica
  "LCA", # Saint Lucia
  "MEX", # Mexico
  "NIC", # Nicaragua
  "PAN", # Panama
  "PER", # Peru
  "PRY", # Paraguay
  "SLV", # El Salvador
  "SUR", # Suriname
  "TTO", # Trinidad and Tobago
  "URY", # Uruguay
  "VEN" # Venezuela
)

# id: SDG_0852_SEX_AGE_RT_A
# indicator: SDG_0852_SEX_AGE_RT
# indicator.label: SDG indicator 8.5.2 - Unemployment rate (%)
## Change this part
year <- 2024
unemployment_rate_lac <- get_ilostat(
  id = "SDG_0852_SEX_AGE_RT_A",
  segment = "indicator",
  filters = list(
    time = year,
    classif1 = "AGE_YTHADULT_YGE15",
    sex = "SEX_T",
    ref_area = lac_iso
  )
)

# Tidy
unemployment_rate_lac_clean <- unemployment_rate_lac |>
  select(ref_area, time, obs_value) |>
  rename(unemployment_rate = obs_value) |>
  mutate(time = as.double(time))

# id: SDG_0831_SEX_ECO_RT_A
# indicator: SDG_0831_SEX_ECO_RT
# indicator.SDG indicator 8.3.1 - Proportion of informal employment in total employment by sex and sector (%)
informality_rate_lac <- get_ilostat(
  id = "SDG_0831_SEX_ECO_RT_A",
  segment = "indicator",
  time_format = "date",
  filters = list(
    time = year,
    classif1 = "ECO_AGGREGATE_TOTAL",
    sex = "SEX_T",
    ref_area = lac_iso
  )
)

# Tidy
informality_rate_lac_clean <- informality_rate_lac |>
  select(ref_area, time, obs_value) |>
  rename(informality_rate = obs_value) |>
  mutate(time = year(time)) |>
  mutate(time = as.double(time))

# Merge
unemployment_informality_rate <- unemployment_rate_lac_clean |>
  inner_join(y = informality_rate_lac_clean, by = join_by(ref_area, time)) |>
  label_ilostat()
```

```{r}
#| label: fig-informality-vs-unemployment
#| fig-cap: Informality proportion vs unemployment rate for some countries in Latin America

# Visualization
unemployment_informality_rate |>
  ggplot(aes(x = informality_rate, y = unemployment_rate)) +
  geom_vline(
    xintercept = mean(unemployment_informality_rate$informality_rate),
    colour = umng_palette[3]
  ) +
  geom_hline(
    yintercept = mean(unemployment_informality_rate$unemployment_rate),
    colour = umng_palette[4]
  ) +
  geom_point(shape = 21, color = "black", fill = umng_palette[2]) +
  geom_label_repel(
    aes(label = ref_area.label),
    seed = 1234,
    size = 3,
    min.segment.length = 0,
    box.padding = 1,
    force = 2,
    max.overlaps = Inf,
    color = umng_palette[1]
  ) +
  labs(
    x = "Informality proportion",
    y = "Unemployment rate",
    subtitle = str_glue(
      "Period: {year}
                           Observation: Horizontal and vertical lines correspond to the mean of the variables in each axis"
    ),
    caption = str_glue(
      "Source: International Labour Organization - ILOSTAT Data Explorer
                             Last update date: {last_update}"
    )
  ) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10)
  )
```

# The big picture and connecting ideas

##

![Macro, micro and organizational level business [@leal_filho_business_2020]](_000_images/001_business_environment.png){.nostretch #fig-business-environment width=85% fig-align="center"}

## 

![A model of the economy: flows of resources [@the_core_econ_team_economy_2023, chap. 1, sec. 1.13]](_000_images/001_model_of_the_economy.png){.nostretch #fig-model-of-the-economy width=75% fig-align="center"}

##

```{mermaid}
%%| label: fig-management-functions
%%| fig-cap: POLC framework [@kinicki_management_2022, pp. 8-10]

flowchart TB
    A("`Management
    functions`")  --> A1("`**P**lanning`") --> A11("`Setting goals 
    and deciding 
    how to achieve 
    them`")
    A --> A2("`**O**rganizing`") --> A21("`Arranging tasks, 
    people and 
    other resources 
    to accomplish 
    the work`")
    A --> A3("`**L**eading`") --> A31("`Motivating, 
    directing and 
    influencing 
    people to 
    achieve the 
    organization’s
    goals`")
    A --> A4("`**C**ontrolling`") --> A41("`Monitoring 
    performance, 
    comparing it 
    with goals and 
    taking corrective 
    action as needed`")

    classDef UmngNode fill:#fdc600,stroke:#043074,stroke-width:2px;
    class A,A1,A11,A2,A21,A3,A31,A4,A41 UmngNode;
```

##

```{mermaid}
%%| label: fig-manager-roles
%%| fig-cap: The roles managers play [@openstax_featured_principles_2019, chap. 2, sec. 1.2]

flowchart TB
    A("`Manager
    roles`") --> A1(Informational)
    A1 --> A11("`Monitor
    Disseminator
    Spokesperson`")
    A --> A2(Interpersonal)
    A2 --> A21("`Figurehead
    Leader
    Liason`")
    A --> A3(Decisional)
    A3 --> A31("`Entrepreneur
    Disturbance handler
    Resource allocator
    Negotiator`")

    classDef UmngNode fill:#fdc600,stroke:#043074,stroke-width:2px;
    class A,A1,A11,A2,A21,A3,A31 UmngNode;
```

# Acknowledgments

##

-   To my family that supports me

-   To the taxpayers of Colombia and the [**UMNG students**](https://www.umng.edu.co/estudiante) who pay my salary

-   To the [**Business Science**](https://www.business-science.io/) and [**R4DS Online Learning**](https://www.rfordatasci.com/) communities where I learn [**R**](https://www.r-project.org/about.html) and [**$\pi$-thon**](https://www.python.org/about/)

-   To the [**R Core Team**](https://www.r-project.org/contributors.html), the creators of [**RStudio IDE**](https://posit.co/download/rstudio-desktop/), [**Positron**](https://positron.posit.co/), [**Quarto**](https://quarto.org/) and the authors and maintainers of the packages [**tidyverse**](https://CRAN.R-project.org/package=tidyverse), [**readxl**](https://CRAN.R-project.org/package=readxl), 
[**gt**](https://CRAN.R-project.org/package=gt),
[**janitor**](https://CRAN.R-project.org/package=janitor), [**Rilostat**](https://CRAN.R-project.org/package=Rilostat), [**ggrepel**](https://CRAN.R-project.org/package=ggrepel) and [**tinytex**](https://CRAN.R-project.org/package=tinytex) for allowing me to access these tools without paying for a license

# References {.allowframebreaks}