---
title: "Labor Market II"
author: 
  - Luis Francisco Gómez López
institute: 
  - FAEDIS
date: last-modified
format: 
  beamer:
    colortheme: dolphin
    fonttheme: structurebold
    theme: AnnArbor
    link-citations: true
    colorlinks: true
    toc: true
    slide-level: 2
    section-titles: false
    include-in-header:
      file: 000_tex_files/preamble.tex
fig-cap-location: bottom
tbl-cap-location: top
knitr: 
  opts_chunk: 
    echo: false
    warning: false
    fig-align: center
    out-width: 85%
lang: en
bibliography: intro_econ.bib
---

```{r}
#| label: libraries

library(tidyverse)
library(tidyquant)
library(readxl)
library(kableExtra)
library(janitor)
library(Rilostat)
library(ggrepel)
```

# Please Read Me

##

-   Check the message **Welcome greeting** published in the News Bulletin Board.

-   Dear student please edit your profile uploading a photo where your face is clearly visible.

-   The purpose of the virtual meetings is to answer questions and not to make a summary of the study material.

-   If you want to participate, please fill out the following survey: Primer corte 30% \> Learning Activities \> Tu opinión sobre la economía colombiana

-   This presentation is based on [@cardenas_introduccion_2020, Chapter 9]

# Purpose

##

Examine the evolution and characteristics of the supply and demand of labor in Colombia

# Unemployment

##

```{r}
#| include: false

# https://www.dane.gov.co >
# Estadísticas por tema >
# Mercado laboral >
# Empleo y desempleo >
# Documento >
# Anexos >
## Descargar

# Import data ----
## Auxiliary variable
dates <- seq.Date(from = ymd("2001-01-01"),
                  # Change the final date
                  to = ymd("2025-01-01"), 
                  by = "month") |> 
  as.character()
col_names <- c("concept", dates)

indicators <- read_excel(path = "000_data/009_anexo_empleo_feb_28.xlsx", 
                         sheet = "Total nacional",
                         # Change the range of the data 
                         range = "A14:KD21",
                         col_names = col_names)

indicators_longer <- indicators |> 
  pivot_longer(cols = -c(concept), 
               names_to = "date", 
               values_to = "value") |> 
  mutate(date = ymd(date))

# Filter the last 10 years
years <- year(indicators_longer$date) |> 
  unique() |> 
  tail(n = 10)

indicators_longer_td <- indicators_longer |> 
  filter(concept == "Tasa de Desocupación (TD)") |> 
  mutate(year = year(date)) |> 
  mutate(year = as_factor(year)) |> 
  mutate(month = month(date, label = TRUE)) |> 
  filter(year %in% years)
```

```{r}
#| label: fig-unemployment-rate-first-years
#| fig-cap: Behavior of unemployment rate in Colombia for selected years

# Visualize
year_pandemic <- 2020
## Update this part using the imported file
last_update_date <- ymd("2025-02-28")

## First years
first_years <- indicators_longer_td |> 
  filter(year %in% head(years, n = 5)) |> 
  # Delete the year about the pandemic
  ## It is an atypical year
  filter(year != year_pandemic)

g1 <- first_years |> 
  ggplot(aes(x = month, y = value)) + 
  geom_line(aes(group = year, color = year)) +
  scale_color_tq() +
  labs(x = NULL,
       y = "Percent (%)",
       color = NULL,
       subtitle = str_glue("Period: {head(unique(first_years$year), n = 1)} - {tail(unique(first_years$year), n = 1)}
                           Observation: {year_pandemic} is not included because is an atypical year taking into account the COVID-19 pandemic"),
       caption = str_glue("Source: DANE - Gran Encuesta Integrada de Hogares (GEIH)
                          Last update date: {last_update_date}")) + 
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        legend.position   = "bottom",
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"))

g1
```

##

```{r}
#| label: fig-unemployment-rate-last-years
#| fig-cap: Behavior of unemployment rate in Colombia for selected years

## Last years
last_years <- indicators_longer_td |> 
  filter(year %in% tail(years, n = 5)) |> 
  # Delete the year about the pandemic
  ## It is an atypical year
  filter(year != year_pandemic)

g2 <- last_years |> 
  ggplot(aes(x = month, y = value)) + 
  geom_line(aes(group = year, color = year)) +
  scale_color_tq() +
  labs(x = NULL,
       y = "Percent (%)",
       color = NULL,
       subtitle = str_glue("Period: {head(unique(last_years$year), n = 1)} - {tail(unique(last_years$year), n = 1)}"),
       caption = str_glue("Source: DANE - Gran Encuesta Integrada de Hogares (GEIH)
                          Last update date: {last_update_date}")) + 
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        legend.position   = "bottom",
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"))

g2
```

##

```{r}
#| include: false

# https://www.dane.gov.co >
# Estadísticas por tema >
# Mercado laboral >
# Empleo y desempleo >
# Documento >
# Anexos >
## Descargar

# Import data ----
## Auxiliary variable
                  # Different from "Total nacional"
dates <- seq.Date(from = ymd("2010-01-01"),
                  # Change the final date
                  to = ymd("2025-01-01"), 
                  by = "month") |> 
  as.character()
col_names <- c("concept", dates)

indicators_male <- read_excel(path = "000_data/009_anexo_empleo_feb_28.xlsx", 
                              sheet = "Total_nacional_IML_Sexo",
                              # Change the range of the data 
                              range = "A14:FZ18",
                              col_names = col_names) |> 
  mutate(sex = "Male")

indicators_female <- read_excel(path = "000_data/009_anexo_empleo_feb_28.xlsx", 
                                sheet = "Total_nacional_IML_Sexo",
                                # Change the range of the data 
                                range = "A33:FZ37",
                                col_names = col_names) |> 
  mutate(sex = "Female")

indicators_male_longer <- indicators_male |> 
  pivot_longer(cols = -c(concept, sex), 
               names_to = "date", 
               values_to = "value") |> 
  mutate(date = ymd(date))

indicators_female_longer <- indicators_female |> 
  pivot_longer(cols = -c(concept, sex), 
               names_to = "date", 
               values_to = "value") |> 
  mutate(date = ymd(date))

indicators_sex_longer <- indicators_male_longer |> 
  bind_rows(indicators_female_longer)

# Filter the last 10 years
years <- year(indicators_longer$date) |> 
  unique() |> 
  tail(n = 10)

indicators_sex_longer_td <- indicators_sex_longer |> 
  filter(concept == "Tasa de Desocupación (TD)") |> 
  mutate(year = year(date)) |> 
  mutate(year = as_factor(year)) |> 
  mutate(month = month(date, label = TRUE)) |> 
  filter(year %in% years)
```

```{r}
#| label: fig-unemployment-rate-male-female
#| fig-cap: Unemployment rate in Colombia for men and women

# Visualize
g3 <- indicators_sex_longer_td |> 
  ggplot(aes(x = date, y = value)) + 
  geom_line(aes(group = sex, color = sex)) +
  scale_color_tq() +
  labs(x = NULL,
       y = "Percent (%)",
       color = NULL,
       subtitle = str_glue("Period: {head(unique(indicators_sex_longer_td$date), n = 1)} - {tail(unique(indicators_sex_longer_td$date), n = 1)}"),
       caption = str_glue("Source: DANE - Gran Encuesta Integrada de Hogares (GEIH)
                          Last update date: {last_update_date}")) + 
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        legend.position   = "bottom",
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"))

g3
```

# Employed

##

```{r}
#| label: tbl-employed-sector-last-period
#| tbl-cap: Employed by sector in Colombia

# Import data ----
## Auxiliary variable
date <- ymd("2025-01-01")
date_label <- date |> format("%Y - %B")
                  # Different from "Total nacional"
dates <- seq.Date(from = ymd("2015-01-01"),
                  # Change the final date
                  to = date, 
                  by = "month") |> 
  as.character()
col_names <- c("concept", dates)

indicators_sector <- read_excel(path = "000_data/009_anexo_empleo_feb_28.xlsx", 
                                sheet = "Ocupados TN_T13_rama",
                                # Change the range of the data 
                                range = "A16:DR30",
                                col_names = col_names)

indicators_sector_clean <- indicators_sector |> 
  select(1, ncol(indicators_sector)) |> 
  set_names(nm = c("concept", "last_period")) |> 
  mutate(last_period = (last_period*1e3) |> round()) |> 
  arrange(desc(last_period))

indicators_sector_table <- indicators_sector_clean  |> 
  kable(booktabs = TRUE,
        format = "latex", 
        col.names = c("Concept", date_label)) |>
  kable_styling(font_size = 7) |> 
  row_spec(row = 0, bold = TRUE) |> 
  footnote(general = c('Source: DANE - Gran Encuesta Integrada de Hogares (GEIH)',
                       # Change this
                       'Last update: 2025-02-28'))

indicators_sector_table
```

##

```{r}
#| label: tbl-employed-occupational-position-last-period
#| tbl-cap: Employed by occupational position in Colombia

# Import data ----
## Auxiliary variable
date <- ymd("2025-01-01")
date_label <- date |> format("%Y - %B")
                  # Different from "Total nacional"
dates <- seq.Date(from = ymd("2010-01-01"),
                  # Change the final date
                  to = date, 
                  by = "month") |> 
  as.character()
col_names <- c("concept", dates)

indicators_occupational_position <- read_excel(path = "000_data/009_anexo_empleo_feb_28.xlsx", 
                                               sheet = "Ocupados TN_posición",
                                               # Change the range of the data 
                                               range = "A16:FZ24",
                                               col_names = col_names)

indicators_occupational_position_clean <- indicators_occupational_position |> 
  select(1, ncol(indicators_occupational_position)) |> 
  set_names(nm = c("concept", "last_period")) |> 
  mutate(last_period = (last_period*1e3) |> round()) |> 
  arrange(desc(last_period))

indicators_occupational_position_table <- indicators_occupational_position_clean |> 
  kable(booktabs = TRUE, escape = FALSE,
        format = "latex", 
        col.names = c("Concept", format(date, "%Y - %B"))) |>
  kable_styling(font_size = 7) |> 
  row_spec(row = 0, bold = TRUE) |> 
  footnote(general = c('Source: DANE - Gran Encuesta Integrada de Hogares (GEIH)',
                       # Change this
                       last_update_date))

indicators_occupational_position_table
```

# Informality

##

- What is informality?

    - The classification used in Colombia is based initially on the occupational position. For more information about definitions of the occupational position check out [@dane_manual_2024, pp. 88-92]
    
        -  **Obrero, empleado del gobierno** is not considered informal in any case
        -  **Obrero o empleado de empresa particular**, **Empleado doméstico** and **Trabajador sin remuneración en empresas o negocios de otros hogares** who do not have health or pension contributions made due to their employment relationship with the employer who hired them are considered informal
        - **Jornalero o Peón** and **Trabajador familiar sin remuneración** are considered informal
        
##

- What is informality?

    - The classification used in Colombia is based initially on the occupational position. For more information about definitions of the occupational position check out [@dane_manual_2024, pp. 88-92]
        
        - **Patrón o empleador** and **Trabajador por cuenta propia** classified as informal[^1]
        - Those employed in the sectors with ISIC codes 84 and 99 with the exception of **Jornalero o Peón** and **Trabajador familiar sin remuneración** are not considered informal  
        
[^1]: The rules are not simple in this case. Check [https://www.dane.gov.co/index.php/estadisticas-por-tema/mercado-laboral/empleo-informal-y-seguridad-social](https://www.dane.gov.co/index.php/estadisticas-por-tema/mercado-laboral/empleo-informal-y-seguridad-social) > Anexos in relation to the sheets **Código_SAS** or **Código_STATA**          

##

- Informality proportion (PI)

    - Employed (E)
    - Informal Employed (I) 

$$PI = \frac{I}{E} \times 100$$

- Formality proportion (PF)

$$PF = 1- PI = 1 - \frac{I}{E} \times 100$$

##

```{r}
#| label: fig-informality-formality-proportion
#| fig-cap: Formality and informality proportion in Colombia

# Import data ----
## Auxiliary variable
dates <- seq.Date(from = ymd("2021-01-01"),
                  # Change the final date
                  to = ymd("2025-01-01"), 
                  by = "month") |> 
  as.character()
col_names <- c("concept", dates)

informality_col <- read_excel(path = "000_data/009_anexo_empleo_informal_seguridad_social_feb_28.xlsx",
                              # Check white spaces
                              sheet = "Grandes dominios ",
                              # Change the range of the data 
                              range = "A14:AX16",
                              col_names = col_names)

informality_col_wider <- informality_col |> 
  pivot_longer(cols = -c(concept), 
               names_to = "date", 
               values_to = "value") |> 
  pivot_wider(id_cols = date, 
              names_from = concept, 
              values_from = value) |> 
  clean_names() |> 
  mutate(prop_informalidad = (informal / poblacion_ocupada)*100) |> 
  mutate(prop_formalidad = 100 - prop_informalidad) |> 
  mutate(date = ymd(date))

informality_col_longer <- informality_col_wider |> 
  select(date, prop_informalidad, prop_formalidad) |> 
  pivot_longer(cols = -c(date), 
               names_to = "variable", 
               values_to = "value")

g4 <- informality_col_longer |>
  ggplot(aes(x = date, y = value)) + 
  geom_line(aes(group = variable, color = variable)) +
  scale_color_tq(labels = c("Formality proportion (PF)", "Informality proportion (PI)")) +
  labs(x = NULL,
       y = "Percent (%)",
       color = NULL,
       subtitle = str_glue("Period: {head(unique(informality_col_longer$date), n = 1)} - {tail(unique(informality_col_longer$date), n = 1)}"),
       caption = str_glue("Source: DANE - Gran Encuesta Integrada de Hogares (GEIH)
                          Last update date: {last_update_date}")) + 
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        legend.position   = "bottom",
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"))

g4
```

# Informality proportion vs unemployment rate

##

```{r}
#| include: false

# Data
## https://ilostat.ilo.org/data/

toc <- get_ilostat_toc()
last_update <- toc |> filter(str_detect(string = id, 
                                        pattern = "SDG_0852_SEX_AGE_RT_A")) |> 
  pull(last.update) |> 
  str_sub(start = 1, end = 10) |> 
  dmy()

lac_iso <- c(
  "ARG", # Argentina
  "BOL", # Bolivia
  "BRA", # Brazil
  "CHL", # Chile
  "COL", # Colombia
  "CRI", # Costa Rica
  "CUB", # Cuba
  "DOM", # Dominican Republic
  "ECU", # Ecuador
  "SLV", # El Salvador
  "GTM", # Guatemala
  "HND", # Honduras
  "MEX", # Mexico
  "NIC", # Nicaragua
  "PAN", # Panama
  "PRY", # Paraguay
  "PER", # Peru
  "URY", # Uruguay
  "VEN"  # Venezuela
)

# id: SDG_0852_SEX_AGE_RT_A
# indicator: SDG_0852_SEX_AGE_RT
# indicator.label: SDG indicator 8.5.2 - Unemployment rate (%)
## Change this part
year <- 2023
unemployment_rate_lac <- get_ilostat(id = "SDG_0852_SEX_AGE_RT_A", 
                                     segment = "indicator", 
                                     filters = list(time = year,
                                                    classif1 = "AGE_YTHADULT_YGE15",
                                                    sex      = "SEX_T",
                                                    ref_area = lac_iso))

unemployment_rate_lac_clean <- unemployment_rate_lac |> 
  select(ref_area, time, obs_value) |> 
  rename(unemployment_rate = obs_value) |> 
  mutate(time = as.double(time))

# id: SDG_0831_SEX_ECO_RT_A
# indicator: SDG_0831_SEX_ECO_RT
# indicator.SDG indicator 8.3.1 - Proportion of informal employment in total employment by sex and sector (%)
informality_rate_lac <- get_ilostat(id = "SDG_0831_SEX_ECO_RT_A", 
                                    segment = "indicator", 
                                    time_format = "date",
                                    filters = list(time = year,
                                                   classif1 = "ECO_AGGREGATE_TOTAL",
                                                   sex      = "SEX_T",
                                                   ref_area = lac_iso))

informality_rate_lac_clean <- informality_rate_lac |> 
  select(ref_area, time, obs_value) |> 
  rename(informality_rate = obs_value) |> 
  mutate(time = year(time)) |> 
  mutate(time = as.double(time))

## Merge
unemployment_informality_rate <- unemployment_rate_lac_clean |> 
  inner_join(y = informality_rate_lac_clean,
             by = join_by(ref_area, time)) |> 
  label_ilostat()
```

```{r}
#| label: fig-informality-vs-unemployment
#| fig-cap: Informality proportion vs unemployment rate for some countries in Latin America

# Visualization
g5 <- unemployment_informality_rate |>
  ggplot(aes(x = informality_rate, y = unemployment_rate)) + 
  geom_vline(xintercept = mean(unemployment_informality_rate$informality_rate), 
             colour = "#E31A1C") +
  geom_hline(yintercept = mean(unemployment_informality_rate$unemployment_rate), 
             colour = "#18BC9C") +
  geom_point() + 
  geom_label_repel(aes(label = ref_area.label),
                   seed = 1234,
                   size = 3,
                   min.segment.length = 0,
                   box.padding = 1,
                   force = 2,
                   max.overlaps = Inf) +
  labs(x        = "Informality proportion",
       y        = "Unemployment rate",
       subtitle = str_glue("Period: {year}
                           Observation: Horizontal and vertical lines correspond to the mean of the variables in each axis"),
       caption  = str_glue("Source: International Labour Organization - ILOSTAT Data Explorer
                             Last update date: {last_update}")) +
  theme(panel.border      = element_rect(fill = NA, color = "black"),
        plot.background   = element_rect(fill = "#f3fcfc"),
        panel.background  = element_rect(fill = "#f3f7fc"),
        legend.background = element_rect(fill = "#f3fcfc"),
        legend.position   = "bottom",
        plot.title        = element_text(face = "bold"),
        axis.title        = element_text(face = "bold"),
        legend.title      = element_text(face = "bold"),
        axis.text         = element_text(face = "bold"))

g5
```

# The economy as a system

##

![A model of the economy: flows of resources [@the_core_econ_team_economy_2023, fig. 1.21]](_000_images/001_model_economy.png){#fig-model-of-the-economy fig-alt="A model of the economy: flows of resources" width="400px" height="250px"}

# Economic environment and the Company

##

```{r}
#| label: fig-econ-env-macro
#| fig-cap: Set of economic factors and forces that influence the development of an organization

#Plot
ggplot(data = tibble(x = 0, y = 0)) + 
    ## Company
    annotate(geom = "text", 
             x        = 0, 
             y        = 0,
             label    = "Company",
             size     = 3,
             fontface = "bold") +
    geom_rect(aes(xmin = -0.4,
                  xmax =  0.4,
                  ymin = -0.4,
                  ymax =  0.4),
              alpha = 0.3,
              fill  = "#E31A1C",
              color = "#E31A1C") +
    ## Micro environment
    geom_rect(aes(xmin = -2,
                  xmax =  2,
                  ymin = -2,
                  ymax =  2),
              alpha = 0.25,
              color = "#2C3E50") +
    annotate(geom = "text", 
             x        = 0, 
             y        = 1,
             label    = str_glue("Micro environment"),
             size     = 6,
             fontface = "bold") +
    annotate(geom = "text", 
             x     = 0, 
             y     = -1.2,
             label = str_glue("Clients - Suppliers - Intermediaries - Competitors - General Public"),
             size  = 3.5) +
    ## Macro environment
    geom_rect(aes(xmin = -3,
                  xmax =  3,
                  ymin = -3,
                  ymax =  3),
              alpha = 0.1,
              color = "#6A3D9A") +
    annotate(geom     = "text", 
             x        = 0, 
             y        = 2.5,
             label    = str_glue("Macro environment"),
             size     = 8,
             fontface = "bold") +
    annotate(geom = "text", 
             x     =  0, 
             y     = -2.5,
             label = str_glue("Economic - Political - Social - Technological - Ecological - Legal"),
             size  = 5) +
    theme_void() +
    labs(caption = "Based on https://foda-dafo.com/ ")
```

# Acknowledgments

-   To my family that supports me

-   To the taxpayers of Colombia and the [**UMNG students**](https://www.umng.edu.co/estudiante) who pay my salary

-   To the [**Business Science**](https://www.business-science.io/) and [**R4DS Online Learning**](https://www.rfordatasci.com/) communities where I learn [**R**](https://www.r-project.org/about.html) and [**$\pi$-thon**](https://www.python.org/about/)

-   To the [**R Core Team**](https://www.r-project.org/contributors.html), the creators of [**RStudio IDE**](https://rstudio.com/products/rstudio/), [**Quarto**](https://quarto.org/) and the authors and maintainers of the packages [**tidyverse**](https://CRAN.R-project.org/package=tidyverse), [**tidyquant**](https://CRAN.R-project.org/package=tidyquant), [**readxl**](https://CRAN.R-project.org/package=readxl), and [**tinytex**](https://CRAN.R-project.org/package=tinytex) for allowing me to access these tools without paying for a license

-   To the [**Linux kernel community**](https://www.kernel.org/category/about.html) for allowing me the possibility to use some [**Linux distributions**](https://static.lwn.net/Distributions/) as my main [**OS**](https://en.wikipedia.org/wiki/Operating_system) without paying for a license

# References {.allowframebreaks}
