---
title: Labor Market
subtitle: Session 1
author:
  - name: 
      given: Luis F.
      family: Gómez L.
    orcid: 0000-0002-2877-9015
    email: luis.gomezl@unimilitar.edu.co
    affiliations:
      - name: Distance Learning Faculty
toc: true
toc-depth: 1
slide-level: 2
format:
  presentation-umng-revealjs:
    logo-vertical-strip: 000_logos/logo-white-vertical.png 
    embed-resources: true
    reference-location: document
date: last-modified
date-format: "D MMMM, YYYY"
knitr: 
  opts_chunk: 
    echo: false
    warning: false
    fig.align: center
    out.width: 85%
lang: en
bibliography: intro_econ.bib
csl: apa-7.csl
editor: source
---

```{r}
#| label: libraries

library(tidyverse)
library(scales)
library(readxl)
```

```{r}
#| label: palette

umng_palette <- c(
  "#043074",
  "#fdc600",
  "#0d5e30",
  "#ee2a24",
  "#fc6700",
  "#00b3f0",
  "#6e3c1a",
  "#f8941c",
  "#8e44ad",
  "#2c3e50",
  "#16a085",
  "#c0392b",
  "#e91e63"
)
```

# Please Read Me

##

-   Check the message **Welcome greeting** published in Announcements forum.

-   Dear student please edit your profile uploading a photo where your face is clearly visible.

-   If you want to participate, please fill out the following survey: **Primer corte 30% > Learning Activities > Tu opinión sobre la economía colombiana**

-   This presentation is based on [@cardenas_introduccion_2020, Chapter 9]

##

- **Purpose**

    - Examine the evolution and characteristics of the supply and demand of labor in Colombia
    
# Labor classification of the population

## {.smaller}

-   To understand the indicators of the labor market from an economic perspective, it is necessary to classify the population from a labor point of view. Initially, the total population of a territory could be included but in practice the **Total Population (TP)** that is included **doesn't** cover [@hussmanns_surveys_1990, chapter 2]:

    -   Members of the armed forces because this information is considered secret or because it is not easy to obtain information when its members are in barracks and military areas.

    -   Residents in institutions such as people who are not part of the military forces but reside in military facilities, residents of penal or correctional centers and hospital residents.

-   In that sense, the **Total Population (TP)** covers the non-institutional civilian population residing in households.

##

```{r}
# Data
# https://www.dane.gov.co > 
# Estadísticas por tema > 
# Mercado Laboral > 
# Empleo y desempleo > 
# Documento >
# Anexos

# Import ----
labor_classification_concept <- read_excel(path = "000_data/009_anexo_empleo_jul_31.xlsx",
                                           sheet = "Total nacional", 
                                           range = "A23:A30", 
                                           col_names = "concept")

## This is the part that needs to be updated
labor_classification_last_data <- read_excel(path = "000_data/009_anexo_empleo_jul_31.xlsx",
                                             sheet = "Total nacional", 
                                             range = "KI23:KI30",
                                             col_names = "value") |> 
  mutate(value = value*1e3)

labor_classification <- bind_cols(labor_classification_concept,
                                  labor_classification_last_data)

# Helpers
poblacion_total <- labor_classification |> pull(2) |> _[1]
poblacion_en_edad_de_trabajar <- labor_classification |> pull(2) |> _[2]
poblacion_menor_a_15_anos <- poblacion_total - poblacion_en_edad_de_trabajar
fuerza_de_trabajo <- labor_classification |> pull(2) |> _[3]
poblacion_fuera_de_la_fuerza_de_trabajo <- poblacion_en_edad_de_trabajar - fuerza_de_trabajo
poblacion_ocupada <- labor_classification |> pull(2) |> _[4]
poblacion_desocupada <- fuerza_de_trabajo - poblacion_ocupada
subocupados <- labor_classification |> pull(2) |> _[7]
completamente_empleados <- poblacion_ocupada - subocupados

last_update_labor <- ymd("2025-07-31")
```

```{r}
#| label: fig-labor-classification-col
#| fig-cap: Labor classification of the population in Colombia

# visualization
## Data
labor <- tibble(x = c( 0, 
              -1, 
              1,
              0,
              2,
              -2,
              2,
              -3,
              -1),
       y = c( 0, 
              -1, 
              -1,
              -2,
              -2,
              -3,
              -3,
              -4,
              -4),
       indicator = c('Total Population (TP)',
                     'Population below\n15 years',
                     'Working Age\nPopulation\n(WAP)',
                     'Labor\nForce\n(LF)',
                     'Not in\nLabor\nForce\n(NLF)',
                     'Employed (E)',
                     'Unemployed (U)',
                     'Full Employed\n(FE)',
                     'Underemployed\n(UE)'
       ),
       level     = c('0', 
                     '1', '1', 
                     '2', '2', 
                     '3', '3',
                     '4', '4'),
       # Población total
       value     = c(poblacion_total, 
                     # Población menor a 15 años
                     poblacion_menor_a_15_anos,    
                     # Población en edad de trabajar
                     poblacion_en_edad_de_trabajar, 
                     # Fuerza de trabajo
                     fuerza_de_trabajo, 
                     # Población fuera de la fuerza de trabajo
                     poblacion_fuera_de_la_fuerza_de_trabajo, 
                     # Población Ocupada
                     poblacion_ocupada, 
                     # Población Desocupada
                     poblacion_desocupada, 
                     # Completamente empleados
                     completamente_empleados, 
                     # Subocupados
                     subocupados))  |> 
  mutate(label_text = str_glue('{indicator}
                                 {value |> 
                                    number(big.mark = ",")}'))

# Plot
labor |> 
  ggplot(aes(x = x, y = y, label = label_text)) +
  geom_segment(x = 0, y = 0, xend = 0, yend = -1) +
  geom_segment(x = -1, y = -1, xend = 1, yend = -1) +
  geom_segment(x = 1, y = -1, xend = 1, yend = -2) +
  geom_segment(x = 0, y = -2, xend = 2, yend = -2) +
  geom_segment(x = 0, y = -2, xend = 0, yend = -3) +
  geom_segment(x = -2, y = -3, xend = 2, yend = -3) +
  geom_segment(x = -2, y = -3, xend = -2, yend = -4) +
  geom_segment(x = -3, y = -4, xend = -1, yend = -4) +
  geom_label(aes(fill = level), 
             color       = "white", 
             fontface    = "bold",
             size        = 3.5) +
  scale_fill_manual(values = umng_palette[1:5]) +
  coord_cartesian(xlim = c(-5, 5), 
                  ylim = c(-4.2, 0.2)) +
  labs(x = NULL,
       y = NULL,
       # This part needs to be update manually
       subtitle = str_glue('Period: June 2025'),
       caption  = str_glue('Sources: DANE, Gran Encuesta Integrada de Hogares - GEIH
                              Last update: {last_update_labor}'),
    fill     = "Level"
  ) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text         = element_blank(), 
    axis.ticks        = element_blank()
  )
```

# Acknowledgments

##



# References {.allowframebreaks}