---
title: Labor Market
subtitle: Session 1
author:
  - name: 
      given: Luis F.
      family: Gómez L.
    orcid: 0000-0002-2877-9015
    email: luis.gomezl@unimilitar.edu.co
    affiliations:
      - name: Distance Learning Faculty
toc: true
toc-depth: 1
slide-level: 2
format:
  presentation-umng-revealjs:
    logo-vertical-strip: 000_logos/logo-white-vertical.png 
    embed-resources: true
    reference-location: document
date: last-modified
date-format: "D MMMM, YYYY"
knitr: 
  opts_chunk: 
    echo: false
    warning: false
    fig.align: center
    out.width: 85%
lang: en
bibliography: intro_econ.bib
csl: apa-7.csl
editor: source
---

```{r}
#| label: libraries

library(tidyverse)
library(scales)
library(readxl)
library(wbstats)
```

```{r}
#| label: palette

umng_palette <- c(
  "#043074",
  "#fdc600",
  "#0d5e30",
  "#ee2a24",
  "#fc6700",
  "#00b3f0",
  "#6e3c1a",
  "#f8941c",
  "#8e44ad",
  "#2c3e50",
  "#16a085",
  "#c0392b",
  "#e91e63"
)
```

# Please Read Me

##

-   Check the message **Welcome greeting** published in Announcements forum.

-   Dear student please edit your profile uploading a photo where your face is clearly visible.

-   If you want to participate, please fill out the following survey: **Primer corte 30% > Learning Activities > Tu opinión sobre la economía colombiana**

-   This presentation is based on [@cardenas_introduccion_2020, Chapter 9]

##

- **Purpose**

    - Examine the evolution and characteristics of the supply and demand of labor in Colombia
    
# Labor classification of the population

## {.smaller}

-   To understand the indicators of the labor market from an economic perspective, it is necessary to classify the population from a labor point of view. Initially, the total population of a territory could be included but in practice the **Total Population (TP)** that is included **doesn't** cover [@hussmanns_surveys_1990, chapter 2]:

    -   Members of the armed forces because this information is considered secret or because it is not easy to obtain information when its members are in barracks and military areas.

    -   Residents in institutions such as people who are not part of the military forces but reside in military facilities, residents of penal or correctional centers and hospital residents.

-   In that sense, the **Total Population (TP)** covers the non-institutional civilian population residing in households.

##

```{r}
# Data
# https://www.dane.gov.co >
# Estadísticas por tema >
# Mercado Laboral >
# Empleo y desempleo >
# Documento >
# Anexos

# Path
## Update this
path_labor <- "000_data/009_anexo_empleo_dic_2025.xlsx"

# Import ----
labor_classification_concept <- read_excel(
  path = path_labor,
  sheet = "Total nacional",
  range = "A23:A30",
  col_names = "concept"
)

## This is the part that needs to be updated
labor_classification_last_data <- read_excel(
  path = path_labor,
  sheet = "Total nacional",
  range = "KO23:KO30",
  col_names = "value"
) |>
  mutate(value = value * 1e3)

labor_classification <- bind_cols(
  labor_classification_concept,
  labor_classification_last_data
)

# Helpers
poblacion_total <- labor_classification |> pull(2) |> _[1]
poblacion_en_edad_de_trabajar <- labor_classification |> pull(2) |> _[2]
poblacion_menor_a_15_anos <- poblacion_total - poblacion_en_edad_de_trabajar
fuerza_de_trabajo <- labor_classification |> pull(2) |> _[3]
poblacion_fuera_de_la_fuerza_de_trabajo <- poblacion_en_edad_de_trabajar -
  fuerza_de_trabajo
poblacion_ocupada <- labor_classification |> pull(2) |> _[4]
poblacion_desocupada <- fuerza_de_trabajo - poblacion_ocupada
subocupados <- labor_classification |> pull(2) |> _[7]
completamente_empleados <- poblacion_ocupada - subocupados

# Helpers
## Update this
last_update_labor <- ymd("2026-01-30")
## Check the period related to
## the information
period_labor <- "December 2025"
```

```{r}
#| label: fig-labor-classification-col
#| fig-cap: Labor classification of the population in Colombia

# visualization
## Data
labor <- tibble(
  x = c(0, -1, 1, 0, 2, -2, 2, -3, -1),
  y = c(0, -1, -1, -2, -2, -3, -3, -4, -4),
  indicator = c(
    'Total Population (TP)',
    'Population below\n15 years',
    'Working Age\nPopulation\n(WAP)',
    'Labor\nForce\n(LF)',
    'Not in\nLabor Force\n(NLF)',
    'Employed (E)',
    'Unemployed (U)',
    'Full Employed\n(FE)',
    'Underemployed\n(UE)'
  ),
  level = c('0', '1', '1', '2', '2', '3', '3', '4', '4'),
  # Población total
  value = c(
    poblacion_total,
    # Población menor a 15 años
    poblacion_menor_a_15_anos,
    # Población en edad de trabajar
    poblacion_en_edad_de_trabajar,
    # Fuerza de trabajo
    fuerza_de_trabajo,
    # Población fuera de la fuerza de trabajo
    poblacion_fuera_de_la_fuerza_de_trabajo,
    # Población Ocupada
    poblacion_ocupada,
    # Población Desocupada
    poblacion_desocupada,
    # Completamente empleados
    completamente_empleados,
    # Subocupados
    subocupados
  )
) |>
  mutate(
    label_text = str_glue(
      '{indicator}
                               {value |> number(big.mark = ",")}'
    )
  )

# Plot
labor |>
  ggplot(aes(x = x, y = y, label = label_text)) +
  geom_segment(x = 0, y = 0, xend = 0, yend = -1) +
  geom_segment(x = -1, y = -1, xend = 1, yend = -1) +
  geom_segment(x = 1, y = -1, xend = 1, yend = -2) +
  geom_segment(x = 0, y = -2, xend = 2, yend = -2) +
  geom_segment(x = 0, y = -2, xend = 0, yend = -3) +
  geom_segment(x = -2, y = -3, xend = 2, yend = -3) +
  geom_segment(x = -2, y = -3, xend = -2, yend = -4) +
  geom_segment(x = -3, y = -4, xend = -1, yend = -4) +
  geom_label(
    aes(fill = level),
    color = "white",
    fontface = "bold",
    size = 3.5,
    label.size = 0.5,
    label.r = unit(0.30, "lines")
  ) +
  scale_fill_manual(values = umng_palette[1:5]) +
  coord_cartesian(xlim = c(-5, 5), ylim = c(-4.2, 0.2)) +
  labs(
    x = NULL,
    y = NULL,
    # This part needs to be update manually
    subtitle = str_glue('Period: {period_labor}'),
    caption = str_glue(
      'Sources: DANE, Gran Encuesta Integrada de Hogares - GEIH
                              Last update: {last_update_labor}'
    ),
    fill = "Level"
  ) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )
```

# Labor market indicators

##

-   **Percentage of the working age population** ("Porcentaje de población en edad de trabajar")

    -   $\%WAP = \frac{WAP}{PT} \times 100$

-   **Labor participation rate (LPR)** ("Tasa global de participación")

    -   $LPR = \frac{LF}{WAP} \times 100$

##

-   **Employment rate (ER)** ("Tasa de ocupación")

    -   $ER = \frac{E}{WAP} \times 100$

-   **Unemployment rate (UR)** ("Tasa de desempleo")

    -   $UR = \frac{U}{LF} \times 100$

-   **Underemployment (UER)** ("Tasa de subempleo")

    -   $UER = \frac{UE}{LF} \times 100$

##

```{r}
# Import ----
employment_unemployment <- wb_data(
  country = c('COL'),
  indicator = c('SL.UEM.TOTL.ZS', 'SL.EMP.TOTL.SP.ZS'),
  return_wide = FALSE
)

# Tidy
employment_unemployment_clean <- employment_unemployment |>
  drop_na(value) |>
  select(country, date, value, indicator) |>
  mutate(
    indicator = case_when(
      indicator ==
        'Unemployment, total (% of total labor force) (modeled ILO estimate)' ~ 'Unemployment rate (UR) (ILO estimate))',
      indicator ==
        'Employment to population ratio, 15+, total (%) (modeled ILO estimate)' ~ 'Employment rate (ER) (ILO estimate))',
      .default = indicator
    )
  )
```

```{r}
#| label: fig-ur-er-col
#| fig-cap: Annual Unemployment rate (UR) and Employment rate (ER) in Colombia

# Visualization
employment_unemployment_clean |>
  ggplot(aes(x = date, y = value)) +
  geom_point(shape = 21, fill = umng_palette[2], color = "black") +
  geom_line(color = umng_palette[1]) +
  geom_vline(xintercept = c(1999, 2020), color = umng_palette[3]) +
  expand_limits(y = 0) +
  facet_wrap(facet = vars(indicator), ncol = 1, nrow = 2, scales = "free_y") +
  scale_x_continuous(
    breaks = c(1991, 1999, 2010, max(employment_unemployment_clean$date))
  ) +
  scale_y_continuous(labels = label_number(accuracy = 1, suffix = "%")) +
  coord_cartesian(
    xlim = c(
      min(employment_unemployment_clean$date),
      max(employment_unemployment_clean$date)
    )
  ) +
  labs(
    x = 'Year',
    y = 'Percent',
    subtitle = str_glue(
      'UR code WDI: SL.UEM.TOTL.ZS
                             UR units: percent
                             ER code WDI: SL.EMP.TOTL.SP.ZS
                             ER units: percent
                             Observation: ILO estimations where ages 15 and older are considered the WAP'
    ),
    caption = str_glue(
      'Source: World Development Indicators (WDI) - World Bank
                             Last update date: {unique(employment_unemployment$last_updated)}'
    )
  ) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10),
    strip.background.x = element_rect(color = "black", fill = umng_palette[2]),
    strip.text.x = element_text(
      margin = margin(t = 5, r = 5, b = 5, l = 5, unit = "pt"),
      face = "bold"
    )
  )
```

# Labor demand

##

-   The labor demand is the process by which the economy generates job vacancies

-   How companies decide how many workers to contract?

    -   Demand of the products that offer by the company
    -   Wages (the price of labor)
    -   Labor regulation
    -   The costs of other inputs different form labor and used in production
    -   The technology used by companies

##

```{r}
# Import ----
## GDP per capita growth (annual %)
econ_growth_unemployment_long <- wb_data(
  indicator = c(
    "NY.GDP.PCAP.KD.ZG",
    # Unemployment, total (% of total labor force)
    # It is calculated not but the ILO but by the Statistical Office of each country
    "SL.UEM.TOTL.NE.ZS"
  ),
  country = "COL",
  return_wide = FALSE
)

# Tidy
econ_growth_unemployment_wide <- econ_growth_unemployment_long |>
  select(iso2c, iso3c, country, date, indicator_id, value) |>
  pivot_wider(
    id_cols = iso2c:date,
    names_from = indicator_id,
    values_from = value
  ) |>
  drop_na(NY.GDP.PCAP.KD.ZG, SL.UEM.TOTL.NE.ZS)
```

```{r}
#| label: fig-economic-growth-ur-col
#| fig-cap: Economic growth vs unemployment rate

# Visualization
econ_growth_unemployment_wide |>
  ggplot(aes(NY.GDP.PCAP.KD.ZG, SL.UEM.TOTL.NE.ZS)) +
  geom_smooth(method = "lm", color = umng_palette[1], se = FALSE) +
  geom_point(shape = 21, color = "black", fill = umng_palette[2]) +
  scale_x_continuous(labels = label_number(suffix = "%")) +
  scale_y_continuous(labels = label_number(suffix = "%")) +
  labs(
    x = "GDP per capita growth (annual %)",
    y = str_glue("Annual total unemployment (% of total labor force)"),
    color = NULL,
    title = "In times of low economic growth, companies reduce production and therefore hire fewer workers",
    subtitle = str_glue(
      "Period: {min(econ_growth_unemployment_long$date)} - {max(econ_growth_unemployment_long$date)}"
    ),
    # Update last update date manually
    caption = str_glue(
      "Source: World Development Indicators (WDI) - World Bank
                             Last update date: {unique(econ_growth_unemployment_long$last_updated)}"
    )
  ) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10),
    legend.position = "bottom"
  )
```

# Wages

##

```{r}
# Check
# 000_scripts/009_script_GEIH.R

# Import ----
group_ingreso_nivel_educativo_categories <- read_rds(
  file = "000_data/009_income_educational_level_geih.rds"
)
levels_geih <- read_rds(file = "000_data/009_levels.rds")

# Helpers
## Generating variables
### Remember to change this
year <- 2025
month <- "11"

level_1 <- levels_geih$educativo[1:4]
level_2 <- levels_geih$educativo[5:6]
level_3 <- levels_geih$educativo[7:10]
level_4 <- levels_geih$educativo[11:13]

nivel_ingreso_levels <- levels_geih$ingreso

ncol <- 2
```

```{r}
#| label: fig-income-education-level-1
#| fig-cap: Employed population according to last educational degree obtained and income bracket

# Visualization
group_ingreso_nivel_educativo_categories |>
  # Filter data
  filter(nivel_educativo %in% level_1) |>
  ggplot(aes(x = FEX_C18, y = nivel_ingreso)) +
  geom_point() +
  geom_segment(aes(xend = 0, yend = nivel_ingreso)) +
  scale_x_continuous(
    labels = label_number(scale = 1e-3, big.mark = ",", suffix = "K")
  ) +
  scale_y_discrete(breaks = nivel_ingreso_levels) +
  facet_wrap(facets = vars(nivel_educativo), ncol = ncol) +
  labs(
    x = "Number of employees in thousands (K)",
    y = str_glue("Current legal monthly minimum wages"),
    subtitle = str_glue("Period: {year}-{month}"),
    caption = str_glue(
      "Gran Encuesta Integrada de Hogares - GEIH - {year}-{month}"
    )
  ) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10),
    strip.background.x = element_rect(color = "black", fill = umng_palette[2]),
    strip.text.x = element_text(
      margin = margin(t = 5, r = 5, b = 5, l = 5, unit = "pt"),
      face = "bold"
    )
  )
```

##

```{r}
#| label: fig-income-education-level-2
#| fig-cap: Employed population according to last educational degree obtained and income bracket

# Visualization
group_ingreso_nivel_educativo_categories |>
  # Filter data
  filter(nivel_educativo %in% level_2) |>
  ggplot(aes(x = FEX_C18, y = nivel_ingreso)) +
  geom_point() +
  geom_segment(aes(xend = 0, yend = nivel_ingreso)) +
  scale_x_continuous(
    labels = label_number(scale = 1e-3, big.mark = ",", suffix = "K")
  ) +
  scale_y_discrete(breaks = nivel_ingreso_levels) +
  facet_wrap(facets = vars(nivel_educativo), ncol = ncol) +
  labs(
    x = "Number of employees in thousands (K)",
    y = str_glue("Current legal monthly minimum wages"),
    subtitle = str_glue("Period: {year}-{month}"),
    caption = str_glue(
      "Gran Encuesta Integrada de Hogares - GEIH - {year}-{month}"
    )
  ) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10),
    strip.background.x = element_rect(color = "black", fill = umng_palette[2]),
    strip.text.x = element_text(
      margin = margin(t = 5, r = 5, b = 5, l = 5, unit = "pt"),
      face = "bold"
    )
  )
```

##

```{r}
#| label: fig-income-education-level-3
#| fig-cap: Employed population according to last educational degree obtained and income bracket

# Visualization
group_ingreso_nivel_educativo_categories |>
  # Filter data
  filter(nivel_educativo %in% level_3) |>
  ggplot(aes(x = FEX_C18, y = nivel_ingreso)) +
  geom_point() +
  geom_segment(aes(xend = 0, yend = nivel_ingreso)) +
  scale_x_continuous(
    labels = label_number(scale = 1e-3, big.mark = ",", suffix = "K")
  ) +
  scale_y_discrete(breaks = nivel_ingreso_levels) +
  facet_wrap(facets = vars(nivel_educativo), ncol = ncol) +
  labs(
    x = "Number of employees in thousands (K)",
    y = str_glue("Current legal monthly minimum wages"),
    subtitle = str_glue("Period: {year}-{month}"),
    caption = str_glue(
      "Gran Encuesta Integrada de Hogares - GEIH - {year}-{month}"
    )
  ) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10),
    strip.background.x = element_rect(color = "black", fill = umng_palette[2]),
    strip.text.x = element_text(
      margin = margin(t = 5, r = 5, b = 5, l = 5, unit = "pt"),
      face = "bold"
    )
  )
```

##

```{r}
#| label: fig-income-education-level-4
#| fig-cap: Employed population according to last educational degree obtained and income bracket

# Visualization
group_ingreso_nivel_educativo_categories |>
  # Filter data
  filter(nivel_educativo %in% level_4) |>
  ggplot(aes(x = FEX_C18, y = nivel_ingreso)) +
  geom_point() +
  geom_segment(aes(xend = 0, yend = nivel_ingreso)) +
  scale_x_continuous(
    labels = label_number(scale = 1e-3, big.mark = ",", suffix = "K")
  ) +
  scale_y_discrete(breaks = nivel_ingreso_levels) +
  facet_wrap(facets = vars(nivel_educativo), ncol = ncol) +
  labs(
    x = "Number of employees in thousands (K)",
    y = str_glue("Current legal monthly minimum wages"),
    subtitle = str_glue("Period: {year}-{month}"),
    caption = str_glue(
      "Gran Encuesta Integrada de Hogares - GEIH - {year}-{month}"
    )
  ) +
  theme(
    panel.border = element_rect(
      fill = NA,
      color = "black"
    ),
    plot.background = element_rect(fill = "white"),
    panel.background = element_rect(fill = "white"),
    legend.background = element_rect(fill = "white"),
    axis.text = element_text(size = 10),
    strip.background.x = element_rect(color = "black", fill = umng_palette[2]),
    strip.text.x = element_text(
      margin = margin(t = 5, r = 5, b = 5, l = 5, unit = "pt"),
      face = "bold"
    )
  )
```

## {.smaller}

-   Mincer equation

    -   $log_e(w_{it}) = log_e(w_o) + \beta_1Ed_{it} +  \beta_2Ex_{it} + \beta_2Ex_{it}^2 + \epsilon_{it}$

        -   $log_e$: natural logarithm
        -   $w_{it}$: wage of individual $i$ in period $t$
        -   $w_o$: wage without any level of education an experience
        -   $\epsilon_{it}$: years of education of individual $i$ in period $t$
        -   $Ex_{it}$: potential experience of individual $i$ in period $t$ which is equal to the age minus the years of education minus six
        -   $\epsilon_{it}$: are other components of individual $i$ in period $t$ that might affect the wage

# Labor regulation

##

-   The labor regulation refers to:

    -   The laws governing labor contracts
    -   Laws governing labor relations that empower unions to represent workers collectively
    -   Social security laws governing the response to social needs such as unemployment, maternity, old age, disability, death and sickness

##

-   Resources

    -   **Calculadora Laboral - Mintrabajo**

        <!-- comment Check always the link to see if it works -->
        
        <!-- comment See  https://github.com/quarto-dev/quarto-cli/issues/5100 about possible change in the quarto documentatio -->
    
        -   [https://www.mintrabajo.gov.co/web/guest/inicio](https://www.mintrabajo.gov.co/web/guest/inicio) \> Atención al Ciudadano \> Trámites y Servicios \> Mi Calculadora

# Acknowledgments

##

-   To my family that supports me

-   To the taxpayers of Colombia and the [**UMNG students**](https://www.umng.edu.co/estudiante) who pay my salary

-   To the [**Business Science**](https://www.business-science.io/) and [**R4DS Online Learning**](https://www.rfordatasci.com/) communities where I learn [**R**](https://www.r-project.org/about.html) and [**$\pi$-thon**](https://www.python.org/about/)

-   To the [**R Core Team**](https://www.r-project.org/contributors.html), the creators of [**RStudio IDE**](https://posit.co/download/rstudio-desktop/), [**Positron**](https://positron.posit.co/), [**Quarto**](https://quarto.org/) and the authors and maintainers of the packages [**tidyverse**](https://CRAN.R-project.org/package=tidyverse), [**scales**](https://CRAN.R-project.org/package=scales), [**readxl**](https://CRAN.R-project.org/package=readxl),
[**wbstats**](https://CRAN.R-project.org/package=wbstats) and [**tinytex**](https://CRAN.R-project.org/package=tinytex) for allowing me to access these tools without paying for a license

# References {.allowframebreaks}